### 2010-02-10 bmc_article.tex
Template for CDK-3 paper

%% BioMed_Central_Tex_Template_v1.05
%%                                      %
%  bmc_article.tex            ver: 1.05 %
%                                       %

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%  LaTeX template for BioMed Central  %%
%%     journal article submissions     %%
%%                                     %%
%%         <27 January 2006>           %%
%%                                     %%
%%                                     %%
%% Uses:                               %%
%% cite.sty, url.sty, bmc_article.cls  %%
%% ifthen.sty. multicol.sty		       %%
%%									   %%
%%                                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%	
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.pdf and the instructions for    %%
%% authors page on the biomed central website                      %%
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\NeedsTeXFormat{LaTeX2e}[1995/12/01]
\documentclass[10pt]{bmc_article}

% Load packages
\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\urlstyle{rm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%	
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%   
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %% 
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %% 
%%  submitted article.                         %% 
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\def\includegraphic{}
\def\includegraphics{}

\setlength{\topmargin}{0.0cm}
\setlength{\textheight}{21.5cm}
\setlength{\oddsidemargin}{0cm} 
\setlength{\textwidth}{16.5cm}
\setlength{\columnsep}{0.6cm}

\newboolean{publ}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                              %%
%% You may change the following style settings  %%
%% Should you wish to format your article       %%
%% in a publication style for printing out and  %%
%% sharing with colleagues, but ensure that     %%
%% before submitting to BMC that the style is   %%
%% returned to the Review style setting.        %%
%%                                              %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Review style settings
\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}

%Publication style settings
%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}

% Begin ...
\begin{document}
\begin{bmcformat}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{A sample article title}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Ensure \and is entered between all but   %%
%% the last two authors. This will be       %%
%% replaced by a comma in the final article %%
%%                                          %%
%% Ensure there are no trailing spaces at   %% 
%% the ends of the lines                    %%     	
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author{Charles A Darwin\correspondingauthor$^{1,2}$%
       \email{Charles A Darwin\correspondingauthor - charles@londonzoo.co.uk}%
      \and
         Jane E Doe\correspondingauthor$^2$%
         \email{Jane E Doe\correspondingauthor - jane.e.doe@cambridge.co.uk}
       and 
         John RS Smith$^3$%
         \email{John RS Smith - john.RS.Smith@cambridge.co.uk}%
      }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address{%
    \iid(1)Life Sciences Department, Kings College London, Cornwall House,%
        Waterloo Road, London, UK\\
    \iid(2)Department of Zoology, Cambridge, Waterloo Road, London, UK\\
    \iid(3)Marine Ecology Department, Institute of Marine Sciences Kiel, %
        D\"{u}sternbrooker Weg 20, 24105 Kiel, Germany
}%

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% The Section headings here are those for  %%
%% a Research article submitted to a        %%
%% BMC-Series journal.                      %%  
%%                                          %%
%% If your article is not of this type,     %%
%% then refer to the Instructions for       %%
%% authors on http://www.biomedcentral.com  %%
%% and change the section headings          %%
%% accordingly.                             %%   
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
        \paragraph*{Background:} Text for this section of the abstract.

        \paragraph*{Results:} Text for this section of the abstract \ldots

        \paragraph*{Conclusions:} Text for this section of the abstract \ldots
\end{abstract}

\ifthenelse{\boolean{publ}}{\begin{multicols}{2}}{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% The Section headings here are those for  %%
%% a Research article submitted to a        %%
%% BMC-Series journal.                      %%  
%%                                          %%
%% If your article is not of this type,     %%
%% then refer to the instructions for       %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and change the section headings          %%
%% accordingly.                             %% 
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
 Text for this section.\cite{koon,oreg,khar,zvai,xjon,schn,pond,smith,marg,hunn,advi,koha,mouse}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Results and Discussion %%
%%
\section*{Results and Discussion}
  \subsection*{Results sub-heading}
    \subsubsection*{This is a sub-sub-heading}
      Sub-sub-sub-headings are made with the \textsl{\\subsubsection} command. \pb
      pb at end of lines ensures correct paragraph spacing.\pb
	  Text for this sub-sub-section \ldots
    \subsubsection*{Another sub-sub-sub-heading}
      Text for this sub-sub-section \ldots

  \subsection*{Another results sub-heading}
    Text for this sub-section \ldots

  \subsection*{Yet another results sub-heading}
    Text for this sub-section.  More results \ldots

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions}
  Text for this section \ldots

%%%%%%%%%%%%%%%%%%
\section*{Methods}
  \subsection*{Methods sub-heading for this section}
    Text for this sub-section \ldots

  \subsection*{Another methods sub-heading for this section}
    Text for this sub-section \ldots

  \subsection*{Yet another sub-heading for this section}
    Text for this sub-section \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Authors contributions}
    Text for this section \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
  \ifthenelse{\boolean{publ}}{\small}{}
  Text for this section \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%              
%%  Bmc_article.bst  will be used to                       %%
%%  create a .BBL file for submission, which includes      %%
%%  XML structured for BMC.                                %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %% 
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %% 
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
 \bibliographystyle{bmc_article}  % Style BST file
  \bibliography{bmc_article} }     % Bibliography file (usually '*.bib' )

%%%%%%%%%%%

\ifthenelse{\boolean{publ}}{\end{multicols}}{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

\section*{Figures}
  \subsection*{Figure 1 - Sample figure title}
      A short description of the figure content
      should go here.

  \subsection*{Figure 2 - Sample figure title}
      Figure legend text.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - Sample table title}
    Here is an example of a \emph{small} table in \LaTeX\ using  
    \verb|\tabular{...}|. This is where the description of the table 
    should go. \par \mbox{}
    \par
    \mbox{
      \begin{tabular}{|c|c|c|}
        \hline \multicolumn{3}{|c|}{My Table}\\ \hline
        A1 & B2  & C3 \\ \hline
        A2 & ... & .. \\ \hline
        A3 & ..  & .  \\ \hline
      \end{tabular}
      }
  \subsection*{Table 2 - Sample table title}
    Large tables are attached as separate files but should
    still be described here.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Additional Files}
  \subsection*{Additional file 1 --- Sample additional file title}
    Additional file descriptions text (including details of how to
    view the file, if it is in a non-standard format or the file extension).  This might
    refer to a multi-page table or a figure.

  \subsection*{Additional file 2 --- Sample additional file title}
    Additional file descriptions text.

\end{bmcformat}
\end{document}

### 2010-02-10 bmc_article.tex
Replaced demo material of template, with very rough outline of paper

@@
 
-\title{A sample article title}
+\title{The Chemistry Development Kit (CDK). 3. Atom typing, ...}
 
@@
 
-\author{Charles A Darwin\correspondingauthor$^{1,2}$%
-       \email{Charles A Darwin\correspondingauthor - charles@londonzoo.co.uk}%
+\author{
+  Egon L Willighagen\correspondingauthor$^{1}$%
+  \email{Egon L Willighagen\correspondingauthor - egon.willighagen@gmail.com}
 \and
-         Jane E Doe\correspondingauthor$^2$%
-         \email{Jane E Doe\correspondingauthor - jane.e.doe@cambridge.co.uk}
-       and 
-         John RS Smith$^3$%
-         \email{John RS Smith - john.RS.Smith@cambridge.co.uk}%
+  Christoph Steinbeck\correspondingauthor$^2$%
+  \email{Christoph Steinbeck\correspondingauthor - steinbeck@ebi.ac.uk}
 }
@@
 
-\address{%
-    \iid(1)Life Sciences Department, Kings College London, Cornwall House,%
+\address{
+    \iid(1)Life Sciences Department, Kings College London, Cornwall House,
         Waterloo Road, London, UK\\
-    \iid(2)Department of Zoology, Cambridge, Waterloo Road, London, UK\\
-    \iid(3)Marine Ecology Department, Institute of Marine Sciences Kiel, %
-        D\"{u}sternbrooker Weg 20, 24105 Kiel, Germany
-}%
+    \iid(2)Department of Zoology, Cambridge, Waterloo Road, London, UK
+}
 
@@
 \section*{Background}
- Text for this section.\cite{koon,oreg,khar,zvai,xjon,schn,pond,smith,marg,hunn,advi,koha,mouse}
+
+Open Source cheminformatics state \ldots (EGON)
+
+Adoption of the CDK ... Cinfony~\cite{OBoyle2008}, and other bindings \ldots ...
+
+Recent prominent use cases of the CDK~\cite{Steinbeck2003,Steinbeck2006} \ldots (EGON)
 
@@
 \section*{Results and Discussion}
-  \subsection*{Results sub-heading}
-    \subsubsection*{This is a sub-sub-heading}
-      Sub-sub-sub-headings are made with the \textsl{\\subsubsection} command. \pb
-      pb at end of lines ensures correct paragraph spacing.\pb
-	  Text for this sub-sub-section \ldots
-    \subsubsection*{Another sub-sub-sub-heading}
-      Text for this sub-sub-section \ldots
 
-  \subsection*{Another results sub-heading}
-    Text for this sub-section \ldots
+\subsection*{New Functionality}
 
-  \subsection*{Yet another results sub-heading}
-    Text for this sub-section.  More results \ldots
+Everyone is invited to contribute and write up the functionality they contributed;
+only requirement: the code must have been incorporated into the main CDK library by the
+time of submission of this paper.
+
+  \subsubsection*{Atom Typing}
+
+  The CDK has adopted a new atom typing algorithm \ldots (EGON)
+
+\subsection*{Improved Coding Standards}
 
@@
 \section*{Methods}
-  \subsection*{Methods sub-heading for this section}
-    Text for this sub-section \ldots
-
-  \subsection*{Another methods sub-heading for this section}
-    Text for this sub-section \ldots
-
-  \subsection*{Yet another sub-heading for this section}
-    Text for this sub-section \ldots
+  \subsection*{Peer Review: Automated}
+    \subsubsection*{PMD}
+    \subsubsection*{JavaDoc Checks}
+  \subsection*{Peer Review: Manual}
+  \subsection*{Unit Testing}
 
@@
 
-\section*{Figures}
-  \subsection*{Figure 1 - Sample figure title}
-      A short description of the figure content
-      should go here.
-
-  \subsection*{Figure 2 - Sample figure title}
-      Figure legend text.
+%\section*{Figures}
+%  \subsection*{Figure 1 - Sample figure title}
+%      A short description of the figure content
+%      should go here.
+%
+%  \subsection*{Figure 2 - Sample figure title}
+%      Figure legend text.
 
@@
 %%
-\section*{Tables}
-  \subsection*{Table 1 - Sample table title}
-    Here is an example of a \emph{small} table in \LaTeX\ using  
-    \verb|\tabular{...}|. This is where the description of the table 
-    should go. \par \mbox{}
-    \par
-    \mbox{
-      \begin{tabular}{|c|c|c|}
-        \hline \multicolumn{3}{|c|}{My Table}\\ \hline
-        A1 & B2  & C3 \\ \hline
-        A2 & ... & .. \\ \hline
-        A3 & ..  & .  \\ \hline
-      \end{tabular}
-      }
-  \subsection*{Table 2 - Sample table title}
-    Large tables are attached as separate files but should
-    still be described here.
+%\section*{Tables}
+%  \subsection*{Table 1 - Sample table title}
+%    Here is an example of a \emph{small} table in \LaTeX\ using  
+%    \verb|\tabular{...}|. This is where the description of the table 
+%    should go. \par \mbox{}
+%    \par
+%    \mbox{
+%      \begin{tabular}{|c|c|c|}
+%        \hline \multicolumn{3}{|c|}{My Table}\\ \hline
+%        A1 & B2  & C3 \\ \hline
+%        A2 & ... & .. \\ \hline
+%        A3 & ..  & .  \\ \hline
+%      \end{tabular}
+%      }
+%  \subsection*{Table 2 - Sample table title}
+%    Large tables are attached as separate files but should
+%    still be described here.
 
@@
 
-\section*{Additional Files}
-  \subsection*{Additional file 1 --- Sample additional file title}
-    Additional file descriptions text (including details of how to
-    view the file, if it is in a non-standard format or the file extension).  This might
-    refer to a multi-page table or a figure.
-
-  \subsection*{Additional file 2 --- Sample additional file title}
-    Additional file descriptions text.
+%\section*{Additional Files}
+%  \subsection*{Additional file 1 --- Sample additional file title}
+%    Additional file descriptions text (including details of how to
+%    view the file, if it is in a non-standard format or the file extension).  This might
+%    refer to a multi-page table or a figure.
+%
+%  \subsection*{Additional file 2 --- Sample additional file title}
+%    Additional file descriptions text.
 

### 2011-12-19 bmc_article.tex
Use biblatex
Written a draft abstract

@@
 % Load packages
-\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
 \usepackage{url}  % Formatting web addresses  
@@
 \urlstyle{rm}
+\usepackage[citestyle=numeric-comp]{biblatex}                                                                     
+\setlength{\bibitemsep}{0cm}                                                                                                                
+\renewbibmacro{in:}{%                                                                                                                       
+  \ifentrytype{article}{}{%                                                                                                                 
+  \printtext{\bibstring{in}\intitlepunct}}}                                                                                                 
+\addbibresource{bmc_article.bib}
 
@@
 
-\title{The Chemistry Development Kit (CDK). 3. Atom typing, ...}
+\title{The Chemistry Development Kit (CDK). 3. Atom typing, Rendering, Molecular Formula, ...}
 
@@
 \author{
-  Egon L Willighagen\correspondingauthor$^{1}$%
-  \email{Egon L Willighagen\correspondingauthor - egon.willighagen@gmail.com}
+  Egon L Willighagen$^{1}$% // FIXED FIRST AUTHOR
+  \email{Egon L Willighagen - egon.willighagen@ki.se}
 \and
-  Christoph Steinbeck\correspondingauthor$^2$%
-  \email{Christoph Steinbeck\correspondingauthor - steinbeck@ebi.ac.uk}
+  Miguel Rojas Cherto$^2$%
+  \email{Miguel Rojas Cherto - steinbeck@ebi.ac.uk}
+\and
+  You C B Here$^3$%
+  \email{You C B Here - you.here@example.org}
+\and
+  C D K Project\correspondingauthor$^98$% // FIXED SECOND BUT LAST AUTHOR
+  \email{C D K Project - cdk-devel@lists.sourceforge.net}
+\and
+  Rajarshi Guha$^99$% // FIXED ONE BUT LAST AUTHOR
+  \email{Rajarshi Guha - steinbeck@ebi.ac.uk}
+\and
+  Christoph Steinbeck$^100$% // FIXED LAST AUTHOR
+  \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
 }
@@
 \address{
-    \iid(1)Life Sciences Department, Kings College London, Cornwall House,
-        Waterloo Road, London, UK\\
-    \iid(2)Department of Zoology, Cambridge, Waterloo Road, London, UK
+    \iid(1)Division of Molecular Toxicology, Institute of Environmental Medicine, Karolinska Institutet, SE-17177 Stockholm, Sweden\\
+    \iid(2)Division of Analytical Biosciences, Leiden/Amsterdam Center for Drug Research, Leiden, The Netherlands \\
+    \iid(3)Example Ltd., Glasgow, UK \\
+    \iid(98)\url{http://cdk.sourceforge.net/} \\
+    \iid(99)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA \\
+    \iid(100)Chemoinformatics and Metabolism team, European Bioinformatics Institute, Hinxton, UK 
 }
@@
 % Do not use inserted blank lines (ie \\) until main body of text.
-        \paragraph*{Background:} Text for this section of the abstract.
+\paragraph*{Background:}
+Cheminformatics is a well-established field with many applications in chemistry,
+drug discovery, and the life sciences. Particularly in the latter field, however,
+molecular properties are not well-understood yet, and the interaction between
+molecules and larger systems remains a challenge to be modeled. Therefore, the
+development of new data mining and pattern recognition continues to be important
+and the need for access to a cheminformatics library that one can change remains.
 
-        \paragraph*{Results:} Text for this section of the abstract \ldots
+\paragraph*{Results:}
+We here report the 1.4 series of the Chemistry Development Kit and outline how
+it evolved since the previous report. We demonstrate the new APIs we introduced
+in this version and discuss the extensive quality control mechanism we have
+adopted. New APIs have been introduced for substructure searching, rendering of
+molecules, handling of molecular formulas, atom type perception, and InChI
+support. At a quality control side we have introduced automated building,
+extensive unit and use case testing, and adopted peer review.
 
-        \paragraph*{Conclusions:} Text for this section of the abstract \ldots
+\paragraph*{Conclusions:}
+With this paper we have shown the continued effort to provide a free, Open Source
+cheminformatics library, and show that such colleborative projects can survive for
+a long period. We have taken advantage from the community support, and show
+that an open source cheminformatics project can act as a peer reviewed
+publishing platform for scientific computating software.
 \end{abstract}
@@
 
+  \subsubsection*{Rendering API}
+
+  Initiated by ...
+
+  \subsubsection*{Molecular Formula}
+
+  ....
+
+  \subsubsection*{SMARTS parser}
+
+  ....
+
+  \subsubsection*{Partial charges and Delocalization}
+
+  ....
+
+  \subsubsection*{New Builders}
+
+  ....
+
+  \subsubsection*{InChI integration}
+
+  .... jni-inchi dep, atom numbering, tautomer generation
+
 \subsection*{Improved Coding Standards}
+
+  \subsubsection*{Peer review}
+
+  \subsubsection*{Modularization}
+
+  \subsubsection*{Documentation}
+
+  Initially done with DocCheck, replaced by OpenJavaDocCheck ...
+
+  \subsubsection*{Unit testing}
+
+  \subsubsection*{Code Quality}
+
+  PMD ...
+
+  \subsubsection*{Git, branching, and patches}
 
@@
 {\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
- \bibliographystyle{bmc_article}  % Style BST file
-  \bibliography{bmc_article} }     % Bibliography file (usually '*.bib' )
+\printbibliography
+}
 

### 2012-01-09 bmc_article.tex
Removed lots of comment lines
Added a basic README

@@
-%% BioMed_Central_Tex_Template_v1.05
-%%                                      %
-%  bmc_article.tex            ver: 1.05 %
-%                                       %
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                     %%
-%%  LaTeX template for BioMed Central  %%
-%%     journal article submissions     %%
-%%                                     %%
-%%         <27 January 2006>           %%
-%%                                     %%
-%%                                     %%
-%% Uses:                               %%
-%% cite.sty, url.sty, bmc_article.cls  %%
-%% ifthen.sty. multicol.sty		       %%
-%%									   %%
-%%                                     %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                                                 %%	
-%% For instructions on how to fill out this Tex template           %%
-%% document please refer to Readme.pdf and the instructions for    %%
-%% authors page on the biomed central website                      %%
-%% http://www.biomedcentral.com/info/authors/                      %%
-%%                                                                 %%
-%% Please do not use \input{...} to include other tex files.       %%
-%% Submit your LaTeX manuscript as one .tex document.              %%
-%%                                                                 %%
-%% All additional figures and files should be attached             %%
-%% separately and not embedded in the \TeX\ document itself.       %%
-%%                                                                 %%
-%% BioMed Central currently use the MikTex distribution of         %%
-%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
-%% http://www.miktex.org                                           %%
-%%                                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \NeedsTeXFormat{LaTeX2e}[1995/12/01]
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                              %%
-%% You may change the following style settings  %%
-%% Should you wish to format your article       %%
-%% in a publication style for printing out and  %%
-%% sharing with colleagues, but ensure that     %%
-%% before submitting to BMC that the style is   %%
-%% returned to the Review style setting.        %%
-%%                                              %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 %Review style settings
@@
 
-% Begin ...
+%%%%%%%%%%%%%%%%%%%%%%%%%%%% RESEARCH PAPER %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{document}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the title of your article here     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \title{The Chemistry Development Kit (CDK). 3. Atom typing, Rendering, Molecular Formula, ...}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors here                   %%
-%%                                          %%
-%% Ensure \and is entered between all but   %%
-%% the last two authors. This will be       %%
-%% replaced by a comma in the final article %%
-%%                                          %%
-%% Ensure there are no trailing spaces at   %% 
-%% the ends of the lines                    %%     	
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors' addresses here        %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \address{
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The Abstract begins here                 %%
-%%                                          %%
-%% The Section headings here are those for  %%
-%% a Research article submitted to a        %%
-%% BMC-Series journal.                      %%  
-%%                                          %%
-%% If your article is not of this type,     %%
-%% then refer to the Instructions for       %%
-%% authors on http://www.biomedcentral.com  %%
-%% and change the section headings          %%
-%% accordingly.                             %%   
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The Abstract begins here
 
@@
 and the need for access to a cheminformatics library that one can change remains.
-
 \paragraph*{Results:}
@@
 extensive unit and use case testing, and adopted peer review.
-
 \paragraph*{Conclusions:}
@@
 \ifthenelse{\boolean{publ}}{\begin{multicols}{2}}{}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The Main Body begins here                %%
-%%                                          %%
-%% The Section headings here are those for  %%
-%% a Research article submitted to a        %%
-%% BMC-Series journal.                      %%  
-%%                                          %%
-%% If your article is not of this type,     %%
-%% then refer to the instructions for       %%
-%% authors on:                              %%
-%% http://www.biomedcentral.com/info/authors%%
-%% and change the section headings          %%
-%% accordingly.                             %% 
-%%                                          %%
-%% See the Results and Discussion section   %%
-%% for details on how to create sub-sections%%
-%%                                          %%
-%% use \cite{...} to cite references        %%
-%%  \cite{koon} and                         %%
-%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
-%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\subsection*{New Functionality}
+\subsection*{New APIs}
 
@@
 
-  ....
+The chemical formula is the basic/simple chemical representation of a compound.
+It defines the number of atoms/isotopes and type of elements that constitute a chemical compound without describing how atoms are bonded.
+It is also the first step toward the identification of the chemical structure from a unidentified compound.
+
+CDK can handle several concepts related to chemical formulas
+\begin{itemize}
+\item Chemical formula =
+\item Set of chemical formulas =
+\item Range of chemical formulas =
+\item Adducts =
+\item Isotope Container =
+\item Isotopic pattern = 
+\item Rules = Filters in accepting an chemical formula
+\end{itemize}
+
+CDK can
+\begin{itemize}
+\item generate/print the elemental composition from a given IAtomContainer
+\item calculate the isotopic pattern from a given chemical formula
+\item determine the elemental composition from a given mass
+\item validate a chemical formula
+\item calculate the exact mass from a given chemical formula
+\end{itemize}
+
+Where can it be applied?
 
@@
 
-  ....
+Partial atomic charges is an electronic property of atoms defining the asymmetric distribution of electrons in chemical bonds.
+It is used as a quantitative correlation with certain compound's physical and chemical properties.
+
+CDK can
+\begin{itemize}
+\item generate all resonance structures from a given compound 
+\item calculate the topological and electronic weight value of each of the resonance structures
+\item calculate the partial atomic charge from a given structure 
+\end{itemize}
+
+Where can it be applied?
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                  The Bibliography                       %%
-%%                                                         %%              
-%%  Bmc_article.bst  will be used to                       %%
-%%  create a .BBL file for submission, which includes      %%
-%%  XML structured for BMC.                                %%
-%%                                                         %%
-%%                                                         %%
-%%  Note that the displayed Bibliography will not          %% 
-%%  necessarily be rendered by Latex exactly as specified  %%
-%%  in the online Instructions for Authors.                %% 
-%%                                                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 {\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
 %% Figures                       %%
-%%                               %%
-%% NB: this is for captions and  %%
-%% Titles. All graphics must be  %%
-%% submitted separately and NOT  %%
-%% included in the Tex document  %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%%
 %% Do not use \listoffigures as most will included as separate files
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
 %% Tables                        %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
 %% Additional Files              %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2012-01-30 bmc_article.tex
Added a draft paragraph on the new builders

@@
 
-\title{The Chemistry Development Kit (CDK). 3. Atom typing, Rendering, Molecular Formula, ...}
+\title{The Chemistry Development Kit (CDK). 3. Atom typing, Rendering, Molecular Formula, and Substructure Searching}
 
@@
 
-  ....
+Originally, the CDK was developed as a shared library between JChemPaint and Jmol. The former
+used a MVC approach with an event-passing mechanism to update the view when the model was
+changed. This can cause an cascade of change events being passed around. To address,
+interfaces were introduced allowing multiple implementations of the core interfaces.
+% TODO: check if the interfaces have previously been defined
+The IChemObjectBuilders play an important role here, allowing implementations of the
+interfaces to be instantiated without the need of explicitly referencing those implementations.
+
+However, the CDK 1.0 and 1.2 implementations of the IChemObjectBuilder had one method for
+each constructor, resulting in very large interface. Moreover, the API changed whenever
+a new class was introduced, and existing methods changed when constructors were updated.
+To simplify the API, the new IChemObjectBuilder collapsed all methods returning new
+implementations into a single method, which takes as a first parameter the class of the
+interface that is wished to be constructed. All further parameters are passes as
+parameter to the class constructor.
+
+For example, to construct a new atom from its element symbol, one would now write:
+
+\begin{verbatim}
+IChemObjectBuilder builder; // previously defined
+IAtom atom = builder.newInstance(IAtom.class, "C");
+\end{verbatim}
+
+Increasingly, the CDK library is now written against the interfaces, and when new instanced
+are needed, these builders are being used. This allows to run a certain CDK-based
+application with a specific builder, summarized in Table~\ref{tab:builders}.
+
+\begin{table}
+\begin{tabular}{l|l}
+builder & description \\
+\hline
+DefaultChemObjectBuilder & The original builder ... \\
+DataDebugChemObjectBuilder & A builder that creates classes that send debug messages ... \\
+SilentChemObjectBuilder & Classes that are created with this builder will not
+    send around change events. \\
+\end{tabular}
+\end{table}
 

### 2012-01-30 bmc_article.tex
Writing up coding standards...

@@
 
+As the library grew over the years, so did the maintenance become more complex. Increasingly,
+the main branch did not compile, and bug fixing become increasingly difficult, as fixing a bug
+in one part of the code, broke some other code which made the wrong assumption about the first
+code.
+
+To address these issues, we have adopted a number of coding standards. By no means there are
+meant to implement the best practices of source code development; instead, they attempt to find
+a balance between increasing code maintainability and being flexible enough to allow efficient
+code development. However, we appreciate the subjective nature of this statement, and some
+adopted guidelines have been heavily discussed in the community.
+The next sections describe some approaches the project have adopted that allows us to
+maintain the CDK library as it is today.
+
+  \subsubsection*{Modularization}
+
+One of the key approaches we have adopted, is to make the CDK more modular. The CDK assigns
+every class to a module, and defines dependencies between modules. For example, core modules
+are not allowed to depend on a module holding data classes implementing the CDK interfaces;
+instead, they may only depend on the interfaces themselves. This, for example, is to ensure
+that dependencies are minimized and to make it easier to exclude CDK functionality with
+third-party dependencies that are not needed.
+
+... describe all modules, maybe add graphviz plot ...
+
   \subsubsection*{Peer review}
 
-  \subsubsection*{Modularization}
+Another big changes since the CDK 1.2 series, is that the project adopted a peer-review
+step in its development. New functionality is now developed independently from the main
+CDK library until it is finished. When it is finished, the new code is send around to
+other developers to be reviewed, very much like how research papers are peer-reviewed
+by journals before publication. Independent developers review the code to check if it 
+meets our community's standards.
 

### 2012-10-24 bmc_article.tex
Added Ola as co-author
Writing some content

@@
 \and
-  You C B Here$^3$%
+  Ola Spjuth$^3$%
+  \email{Ola Spjuth - ola.spjuth@farmbio.uu.se}
+\and
+  You C B Here$^4$%
   \email{You C B Here - you.here@example.org}
@@
 \and
-  Rajarshi Guha$^99$% // FIXED ONE BUT LAST AUTHOR
+  Rajarshi Guha$^{99}$% // FIXED ONE BUT LAST AUTHOR
   \email{Rajarshi Guha - steinbeck@ebi.ac.uk}
 \and
-  Christoph Steinbeck$^100$% // FIXED LAST AUTHOR
+  Christoph Steinbeck$^{100}$% // FIXED LAST AUTHOR
   \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
@@
 Recent prominent use cases of the CDK~\cite{Steinbeck2003,Steinbeck2006} \ldots (EGON)
+AMBIT, ScaffoldHunter, OMG, ...
 
@@
 
-  The CDK has adopted a new atom typing algorithm \ldots (EGON)
+  The CDK has adopted a new atom typing framework, to split out the perception of atom types
+  from the algorithms that use atom type properties. Previous CDK releases had perception
+  integrated with algorithms, even though there was only one set of defined atom types.
+  This led to alternative but contradiction perception algorithms, which made maintenance
+  of these code bases impractical. For example, adding a new atom type required all alternatives
+  to be updated.
+  To solve this, a single perception class was added, called CDKAtomTypeMatcher. The actual
+  algorithm used is based
+  on heuristic rules, and alternative approach for perception exist. Possibly, in the future
+  alternative algorithms may be implemented, while that would again increase the maintenance
+  burden.
+
+  The new class can perceive the atom type of single atoms, but also for a full atom container.
+  The CDKAtomTypeMatcher class is written to handle various amounts of missing information,
+  and take into account as much information from the input. Perception
+
+  \subsubsection*{Stereochemistry}
+
+  CDK 1.4.XXXX [CHECK] introduces a new API for stereochemistry information: the interfaces
+  IStereoChemistryElement was introduced with
 
@@
 
-  \subsubsection*{InChI integration}
-
-  .... jni-inchi dep, atom numbering, tautomer generation
-
 \subsection*{Improved Coding Standards}
@@
 
+\subsection*{Binary distributions}
+
+\subsubsection*{Maven packages}
+
+INVITE NINA
+
+\subsubsection*{OSGi bundles}
+
+INVITE ARVID
+
 %%%%%%%%%%%%%%%%%%%%%%
 \section*{Conclusions}
-  Text for this section \ldots
+
+Over the past years since the last CDK publication in 2006, a lot has changed. The functionality
+has seen many additions, and the stability of the development model has significantly improved.
+This paper outlines how peer review and quality control contributed to a much wider adoption
+of our cheminformatics library and seen many new additions. With over 75 contributors,
+the CDK is alive and kicking.
+
+However, there are many unsolved issues. Performance is perhaps one of the most important ones.
+The disadvantage of having an API that supports many data models, is that the API is heavy and
+the interface implementations hard to optimize.
 

### 2013-09-02 article.tex
Molecular fingerprints in CDK
Text and table listing all molecular fingerprints

--- 2012-10-24 bmc_article.tex
@@
 \usepackage[utf8]{inputenc} %unicode support
+\usepackage{amssymb}
 %\usepackage[applemac]{inputenc} %applemac support if unicode package fails
@@
 \urlstyle{rm}
-\usepackage[citestyle=numeric-comp]{biblatex}                                                                     
-\setlength{\bibitemsep}{0cm}                                                                                                                
-\renewbibmacro{in:}{%                                                                                                                       
-  \ifentrytype{article}{}{%                                                                                                                 
-  \printtext{\bibstring{in}\intitlepunct}}}                                                                                                 
-\addbibresource{bmc_article.bib}
+\usepackage{cite}
 
@@
   Egon L Willighagen$^{1}$% // FIXED FIRST AUTHOR
-  \email{Egon L Willighagen - egon.willighagen@ki.se}
+  \email{Egon L Willighagen - egon.willighagen@maastrichtuniversity.nl}
 \and
-  Miguel Rojas Cherto$^2$%
-  \email{Miguel Rojas Cherto - steinbeck@ebi.ac.uk}
+  John May$^2$% // FIXED SECOND AUTHOR
+  \email{John May - COMPLETE@ebi.ac.uk}
+\and
+  Jonathan Alversson$^3$%
+  \email{Jonathan Alversson - COMPLETE}
+\and
+  Arvid Berg$^3$%
+  \email{Arvid Berg - COMPLETE}
+\and
+  Nina Jeliazkova$^4$%
+  \email{Nina Jeliakvova - COMPLETE}
+\and
+  Dazhi Jiao$^5$%
+  \email{Dazhi Jiao - COMPLETE}
+\and
+  Syed Asad Rahman$^6$%
+  \email{Syed Asad Rahman - COMPLETE}
+\and
+  Mark Rijnbeek$^2$%
+  \email{Mark Rijnbeek - COMPLETE}
+\and
+  Miguel Rojas Cherto$^7$%
+  \email{Miguel Rojas Cherto - COMPLETE}
 \and
@@
 \and
-  You C B Here$^4$%
-  \email{You C B Here - you.here@example.org}
+  Gilleain Torrance$^5$%
+  \email{Gilleain Torrance - COMPLETE}
 \and
-  C D K Project\correspondingauthor$^98$% // FIXED SECOND BUT LAST AUTHOR
-  \email{C D K Project - cdk-devel@lists.sourceforge.net}
+  Rajarshi Guha$^{8}$% // FIXED ONE BUT LAST AUTHOR
+  \email{Rajarshi Guha - COMPLETE}
 \and
-  Rajarshi Guha$^{99}$% // FIXED ONE BUT LAST AUTHOR
-  \email{Rajarshi Guha - steinbeck@ebi.ac.uk}
-\and
-  Christoph Steinbeck$^{100}$% // FIXED LAST AUTHOR
+  Christoph Steinbeck$^{2}$% // FIXED LAST AUTHOR
   \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
@@
 \address{
-    \iid(1)Division of Molecular Toxicology, Institute of Environmental Medicine, Karolinska Institutet, SE-17177 Stockholm, Sweden\\
-    \iid(2)Division of Analytical Biosciences, Leiden/Amsterdam Center for Drug Research, Leiden, The Netherlands \\
-    \iid(3)Example Ltd., Glasgow, UK \\
-    \iid(98)\url{http://cdk.sourceforge.net/} \\
-    \iid(99)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA \\
-    \iid(100)Chemoinformatics and Metabolism team, European Bioinformatics Institute, Hinxton, UK 
+    \iid(1)Department of Bioinformatics - BiGCaT, Maastricht University, Maastricht, NL-6200 MD, The Netherlands \\
+    \iid(2)Chemoinformatics and Metabolism team, European Bioinformatics Institute, Hinxton, UK, \\
+    \iid(3)Department of Pharmaceutical Biosciences, Uppsala University, 751 24 Uppsala, Sweden \\
+    \iid(4)Ideaconsult Ltd, A. Kanchev 4, Sofia 1000, Bulgaria \\
+    \iid(5)Example Ltd., Glasgow, UK \\
+    \iid(6)Example Ltd., Glasgow, UK \\
+    \iid(7)Division of Analytical Biosciences, Leiden/Amsterdam Center for Drug Research, Leiden, The Netherlands \\
+    \iid(8)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA
 }
@@
 Cheminformatics is a well-established field with many applications in chemistry,
-drug discovery, and the life sciences. Particularly in the latter field, however,
-molecular properties are not well-understood yet, and the interaction between
-molecules and larger systems remains a challenge to be modeled. Therefore, the
-development of new data mining and pattern recognition continues to be important
-and the need for access to a cheminformatics library that one can change remains.
+biology, drug discovery, and others. The Chemistry Development Kit (CDK) has
+become a widely used Open Source cheminformatics toolkit, providing various models to represent
+chemical structures, of which the chemical graph is essential. However,
+in the first five years of the project increased so much in size that
+interdependencies between components grew unmanageable large, resulting
+in unpredictable instabilities.
 \paragraph*{Results:}
-We here report the 1.4 series of the Chemistry Development Kit and outline how
-it evolved since the previous report. We demonstrate the new APIs we introduced
-in this version and discuss the extensive quality control mechanism we have
-adopted. New APIs have been introduced for substructure searching, rendering of
-molecules, handling of molecular formulas, atom type perception, and InChI
-support. At a quality control side we have introduced automated building,
-extensive unit and use case testing, and adopted peer review.
+We here report the improvements of the CDK since the 1.2 release series made
+to accomodate both the increased complexity of the library, as well as
+siginificant improvements of and additions to the functionality of the
+library. We here outline how the CDK evolved since the previous report and
+discuss the quality control approached we have adopted to ensure
+stability. Additionally, new APIs have been introduced for atom type
+perception, substructure searching, rendering of molecules, and handling of
+molecular formulas.
 \paragraph*{Conclusions:}
@@
 
-Open Source cheminformatics state \ldots (EGON)
+Open Source cheminformatics has made significants steps forward recently~\cite{OBoyle2011}.
+The Chemistry Development Kit (CDK) is one of the tools under the Blue Obelisk wings,
+and previously documented version have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
+For example, new tools expose CDK functionality, such as Cinfony~\cite{OBoyle2008},
+rcdk~\cite{}, and workflow plugins for Taverna\cite{Truszkowski2011} and KNIME\cite{Beisken2013}.
+Furthermore, projects with specific functionality have emerged building on the CDK, including
+jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
+Padel~\cite{yap2011padel}, AMBIT-SMARTS~\cite{jeliazkova2011ambitsmarts}, AMBIT-Tautomer~\cite{kochev2013ambit},
+and SMSD~\cite{}. Other tools used previous versions already, and follow
+newer releases of the CDK, such as Bioclipse~\cite{} and AMBIT~\cite{jeliazkova2011ambit}.
 
-Adoption of the CDK ... Cinfony~\cite{OBoyle2008}, and other bindings \ldots ...
+However, previous version of the CDK were written by a community with specific applications
+in mind, of which structure elucidation was one. It could therefore be rather competitive
+in some cheminformatics domains, such as fingerprinting~\cite{Ekins,Cannon}, in other parts
+it could be inadequate, such in the capturing of stereochemistry.
 
-Recent prominent use cases of the CDK~\cite{Steinbeck2003,Steinbeck2006} \ldots (EGON)
-AMBIT, ScaffoldHunter, OMG, ...
+Continued community development of the CDK in the last few years, however, has shown that
+Open Source software is here to stay. The adoption of automatic build systems and
+quality control methodologies such as unit testing, automated source code validation,
+and peer review by fellow developers have greatly improved the stability of the library,
+At the expense of slowing down the development, it has allowed for cleaning up interdependencies
+between modules of code, and has reset the scalability of the development model,
+allowing for a lot of new functionality in code APIs while maintaining the quality
+of code depending on those core APIs.
+
+This has resulted in important improved functionality, such as a separately published
+InChI functionality~\cite{SpjuthXXXX} and greatly improved ring detection functionality~\cite{MayYYYY},
+but also a core atom type perception module, covering a much wider set of elements than
+previous CDK versions and capturing many more charged and radical atom species,
+a more comprehensive fingerprinting API, many speed and stability improvements, and more.
 
@@
 
-Everyone is invited to contribute and write up the functionality they contributed;
-only requirement: the code must have been incorporated into the main CDK library by the
-time of submission of this paper.
+We here outline various new and improved APIs in the CDK library since the two previous
+publications.
 
@@
 
-  The CDK has adopted a new atom typing framework, to split out the perception of atom types
-  from the algorithms that use atom type properties. Previous CDK releases had perception
-  integrated with algorithms, even though there was only one set of defined atom types.
-  This led to alternative but contradiction perception algorithms, which made maintenance
-  of these code bases impractical. For example, adding a new atom type required all alternatives
-  to be updated.
-  To solve this, a single perception class was added, called CDKAtomTypeMatcher. The actual
-  algorithm used is based
-  on heuristic rules, and alternative approach for perception exist. Possibly, in the future
+  Atom type perception is a core cheminformations functionality: the atom types describe
+  chemical features of atoms, such as the number of neighbors, possible formal charges,
+  (approximate) hybridization, electron distribution over orbitals, etc. However, the
+  CDK commonly integrated atom type perception into algorithms, resulting in multiple
+  and diverging copies of similar atom type schemes. With the need to add many
+  new atom types for previously uncovered element, but also charged and radical atom
+  types, a new approach was needed.
+
+  This CDK version has isolated a new atom typing framework, removing the perception of
+  atom types from the algorithms, allowing the perception to be properly unit tested.
+  The new code defined the atom type using an atom type ontology based on the XXXXX.
+
+  A reference perception class, CDKAtomTypeMatcher, has been written that perceives these atom types, and
+  validates the perception automatically against the properties defined by the ontology.
+  This class handles a variety in types of missing information, as commonly resulting
+  from various (file) formats; for example, it can handle undefined hydrogen counts
+  and undefined double bond positions if hybridization information is provided instead.
+  That makes the code complex, and at times slow, but so far shown sufficiently
+  performant for many applications. Possibly, in the future
   alternative algorithms may be implemented, while that would again increase the maintenance
@@
 
-  The new class can perceive the atom type of single atoms, but also for a full atom container.
-  The CDKAtomTypeMatcher class is written to handle various amounts of missing information,
-  and take into account as much information from the input. Perception
+  The new code base defines XXXX atom types for YYYY elements, including ZZZZ radical
+  atom types. A full list of atom types is given in Additional File X.
+
+  The code is tested by FFFF unit tests, and has been tested against ChEBI and PubChem. REFS
 
@@
   CDK 1.4.XXXX [CHECK] introduces a new API for stereochemistry information: the interfaces
-  IStereoChemistryElement was introduced with
+  IStereoChemistryElement was introduced with ...
 
@@
 
-  Initiated by ...
+  Initiated by Niels Out ...
 
@@
 
+  \subsubsection*{SMILES parser}
+
+  ....
+
   \subsubsection*{SMARTS parser}
+
+  ....
+
+  \subsubsection*{Ring finding}
 
@@
 
+\subsubsection*{Molecular fingerprints}
+Also the molecular fingerprints in CDK has seen development since the 1.2
+release. Traditionally CDK fingerprints were represented using the
+\textit{BitSet} class shipping with Java. The benefits of the class include
+such things as already implemented methods for modifying bit sets using other
+bit sets. However the class keeps a vector of bits in memory. The solution was
+excellent for hashed, relatively small fingerprints, \textit{e.g.}, 1024 bits,
+\textit{i.e.}, 10 bits indexing space. However implementing a fingerprint
+designed to avoid collisions with, \textit{e.g.} a 32 bit indexing space using
+this approach would be highly memory-inefficient. To allow for multiple
+fingerprint representation implementations a bit fingerprint interface was
+introduced. Also, although fingerprints traditionally are bit vectors a count
+fingerprint was also introduced making fingerprints based on integer vectors
+supported in CDK as well. The fingerprints currently existing in CDK are listed
+in Table~\ref{tab:fingerprints}.
+
 \subsection*{Improved Coding Standards}
@@
 ... describe all modules, maybe add graphviz plot ...
-
-  \subsubsection*{Peer review}
-
-Another big changes since the CDK 1.2 series, is that the project adopted a peer-review
-step in its development. New functionality is now developed independently from the main
-CDK library until it is finished. When it is finished, the new code is send around to
-other developers to be reviewed, very much like how research papers are peer-reviewed
-by journals before publication. Independent developers review the code to check if it 
-meets our community's standards.
 
@@
 
-%%%%%%%%%%%%%%%%%%
-\section*{Methods}
-  \subsection*{Peer Review: Automated}
-    \subsubsection*{PMD}
-    \subsubsection*{JavaDoc Checks}
-  \subsection*{Peer Review: Manual}
-  \subsection*{Unit Testing}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Availability and requirements}
+
+\begin{itemize}
+\item \textbf{Project Name}: The Chemistry Development Kit
+\item \textbf{Project home page}: \url{http://cdk.sourceforge.net/}
+\item \textbf{Operating system(s)}: Windows, GNU/Linux, OS/X
+\item \textbf{Programming language}: Java
+\item \textbf{Other (optional) requirements}: JNI-InChI, Vecmath, BEAM, Guava, JGraphT, Signatures, CMLDOM, XOM, JavaCC
+\item \textbf{License}: LGPL v2.1 or later
+\item \textbf{Any restrictions to use by non-academics}: None additional
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Competing interests}
+XX and YY have companies that sell solutions based on the CDK.
 
@@
   \ifthenelse{\boolean{publ}}{\small}{}
-  Text for this section \ldots
+Niels Out and Stefan Kuhn are acknowledged for their contributions to the project.
 
 {\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
-\printbibliography
+ \bibliographystyle{bmc_article}  % Style BST file
+  \bibliography{article}
 }
@@
 %%
-%\section*{Tables}
-%  \subsection*{Table 1 - Sample table title}
+\section*{Tables}
+  \subsection*{Table 1 - The molecular fingerprints in CDK}
+    \begin{tabular}{lccrc}
+                             & Bit version  & Count version & Since CDK version & References \\
+  \cline{2-5}
+  Fingerprinter              & $\checkmark$ &               &                   & \\
+  EStateFingerprinter        & $\checkmark$ &               &                   & \\
+  ExtendedFingerprinter      & $\checkmark$ &               &                   & \\
+  GraphOnlyFingerprinter     & $\checkmark$ &               &                   & \\
+  HybridizationFingerprinter & $\checkmark$ &               &                   & \\
+  KlekotaRothFingerprinter   & $\checkmark$ &               &                   & \\
+  LingoFingerprinter         & $\checkmark$ &               &                   & \\
+  MACCSFingerprinter         & $\checkmark$ &               &                   & \\
+  PubchemFingerprinter       & $\checkmark$ &               &                   & \\
+  ShortestPathFingerprinter  & $\checkmark$ &               &                   & \\
+  SignatureFingerprinter     & $\checkmark$ & $\checkmark$  &                   & [pending] \\
+  SubstructureFingerprinter  & $\checkmark$ &               &                   & \\
+    \end{tabular}
+
 %    Here is an example of a \emph{small} table in \LaTeX\ using  

### 2013-10-06 bmc_article.tex
Additional authors

@@
 \urlstyle{rm}
-\usepackage[citestyle=numeric-comp]{biblatex}                                                                     
+\usepackage[backend=bibtex]{biblatex}                                                         
 \setlength{\bibitemsep}{0cm}                                                                                                                
@@
   \printtext{\bibstring{in}\intitlepunct}}}                                                                                                 
-\addbibresource{bmc_article.bib}
+\bibliography{bmc_article.bib}
 
@@
   Egon L Willighagen$^{1}$% // FIXED FIRST AUTHOR
-  \email{Egon L Willighagen - egon.willighagen@ki.se}
+  \email{Egon L Willighagen - COMPLETE}
 \and
-  Miguel Rojas Cherto$^2$%
-  \email{Miguel Rojas Cherto - steinbeck@ebi.ac.uk}
+  John May$^2$% // FIXED SECOND AUTHOR
+  \email{John May - COMPLETE@ebi.ac.uk}
+\and
+  Jonathan Alversson$^3$%
+  \email{Jonathan Alversson - COMPLETE}
+\and
+  Arvid Berg$^3$%
+  \email{Arvid Berg - COMPLETE}
+\and
+  Nina Jeliazkova$^4$%
+  \email{Nina Jeliakvova - COMPLETE}
+\and
+  Dazhi Jiao$^5$%
+  \email{Dazhi Jiao - COMPLETE}
+\and
+  Syed Asad Rahman$^6$%
+  \email{Syed Asad Rahman - COMPLETE}
+\and
+  Mark Rijnbeek$^2$%
+  \email{Mark Rijnbeek - COMPLETE}
+\and
+  Miguel Rojas Cherto$^7$%
+  \email{Miguel Rojas Cherto - COMPLETE}
 \and
@@
 \address{
-    \iid(1)Division of Molecular Toxicology, Institute of Environmental Medicine, Karolinska Institutet, SE-17177 Stockholm, Sweden\\
-    \iid(2)Division of Analytical Biosciences, Leiden/Amsterdam Center for Drug Research, Leiden, The Netherlands \\
-    \iid(3)Example Ltd., Glasgow, UK \\
-    \iid(98)\url{http://cdk.sourceforge.net/} \\
-    \iid(99)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA \\
-    \iid(100)Chemoinformatics and Metabolism team, European Bioinformatics Institute, Hinxton, UK 
+    \iid(1)Department of Bioinformatics - BiGCaT, \\
+    \iid(2)Chemoinformatics and Metabolism team, European Bioinformatics Institute, Hinxton, UK, \\
+    \iid(3)Department of Pharmaceutical Biosciences, Uppsala University, 751 24 Uppsala, Sweden \\
+    \iid(4)Ideaconsult Ltd, A. Kanchev 4, Sofia 1000, Bulgaria \\
+    \iid(5)Example Ltd., Glasgow, UK \\
+    \iid(6)Example Ltd., Glasgow, UK \\
+    \iid(7)Division of Analytical Biosciences, Leiden/Amsterdam Center for Drug Research, Leiden, The Netherlands \\
+    \iid(8)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA
 }
@@
 \paragraph*{Results:}
-We here report the 1.4 series of the Chemistry Development Kit and outline how
+We here report the improvements since the 1.2 series of the Chemistry Development Kit and outline how
 it evolved since the previous report. We demonstrate the new APIs we introduced
@@
 Recent prominent use cases of the CDK~\cite{Steinbeck2003,Steinbeck2006} \ldots (EGON)
-AMBIT, ScaffoldHunter, OMG, ...
+AMBIT, ScaffoldHunter, OMG, Padel, ...
+
+Plugins for Taverna, KNIME, ...
 
@@
 
-  \subsubsection*{Peer review}
-
-Another big changes since the CDK 1.2 series, is that the project adopted a peer-review
-step in its development. New functionality is now developed independently from the main
-CDK library until it is finished. When it is finished, the new code is send around to
-other developers to be reviewed, very much like how research papers are peer-reviewed
-by journals before publication. Independent developers review the code to check if it 
-meets our community's standards.
-
   \subsubsection*{Documentation}
@@
 
-%%%%%%%%%%%%%%%%%%
-\section*{Methods}
-  \subsection*{Peer Review: Automated}
-    \subsubsection*{PMD}
-    \subsubsection*{JavaDoc Checks}
-  \subsection*{Peer Review: Manual}
-  \subsection*{Unit Testing}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
   \ifthenelse{\boolean{publ}}{\small}{}
-  Text for this section \ldots
+Niels Out and Stefan Kuhn are acknowledged for their contributions to the project.
 

### 2013-10-06 bmc_article.tex
More papers

@@
 Recent prominent use cases of the CDK~\cite{Steinbeck2003,Steinbeck2006} \ldots (EGON)
-AMBIT, ScaffoldHunter, OMG, Padel, ...
+AMBIT~\cite{jeliazkova2011ambit}, jCompoundMapper~\cite{Hinselmann2011},
+ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012}, Padel~\cite{yap2011padel}, ...
 
-Plugins for Taverna, KNIME, ...
+Plugins for Taverna\cite{Truszkowski2011}, KNIME\cite{Beisken2013}, ...
+
+Some extensions make it back into the main library, like SMSD~\cite{} and fingerprints from Padel.
+Not all extensions made it back into the main library, such as those from
+AMBIT-SMARTS~\cite{jeliazkova2011ambitsmarts}, AMBIT-Tautomer~\cite{kochev2013ambit}, FooFromRydberg~\cite{}.
 

### 2013-10-06 article.tex
Revert to BibTeX

--- 2012-10-24 bmc_article.tex
@@
 Recent prominent use cases of the CDK~\cite{Steinbeck2003,Steinbeck2006} \ldots (EGON)
-AMBIT, ScaffoldHunter, OMG, ...
+AMBIT~\cite{jeliazkova2011ambit}, jCompoundMapper~\cite{Hinselmann2011},
+ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012}, Padel~\cite{yap2011padel}, ...
+
+Plugins for Taverna\cite{Truszkowski2011}, KNIME\cite{Beisken2013}, ...
+
+Some extensions make it back into the main library, like SMSD~\cite{} and fingerprints from Padel.
+Not all extensions made it back into the main library, such as those from
+AMBIT-SMARTS~\cite{jeliazkova2011ambitsmarts}, AMBIT-Tautomer~\cite{kochev2013ambit}, FooFromRydberg~\cite{}.
 

### 2013-10-06 article.tex
Added two missing sections

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Availability and requirements}
+
+\item \textbf{Project Name}: The Chemistry Development Kit
+\textbf{Project home page}: \url{http://cdk.sourceforge.net/}
+\textbf{Operating system(s)}: Windows, GNU/Linux, OS/X
+\textbf{Programming language}: Java
+\textbf{Other (optional) requirements}: JNI-InChI, Vecmath, BEAM, Guava, JGraphT, Signatures, CMLDOM, XOM, JavaCC
+\textbf{License}: LGPL v2.1 or later
+\textbf{Any restrictions to use by non-academics}: None additional
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Competing interests}
+XX and YY have companies that sell solutions based on the CDK.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Authors contributions}

### 2013-10-06 article.tex
Added a missing \begin{}
And item markers...

--- 2013-10-06 article.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Availability and requirements}
+
+\begin{itemize}
+\item \textbf{Project Name}: The Chemistry Development Kit
+\item \textbf{Project home page}: \url{http://cdk.sourceforge.net/}
+\item \textbf{Operating system(s)}: Windows, GNU/Linux, OS/X
+\item \textbf{Programming language}: Java
+\item \textbf{Other (optional) requirements}: JNI-InChI, Vecmath, BEAM, Guava, JGraphT, Signatures, CMLDOM, XOM, JavaCC
+\item \textbf{License}: LGPL v2.1 or later
+\item \textbf{Any restrictions to use by non-academics}: None additional
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Competing interests}
+XX and YY have companies that sell solutions based on the CDK.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Authors contributions}

### 2013-10-09 article.tex
Improved abstract

@@
 Cheminformatics is a well-established field with many applications in chemistry,
-drug discovery, and the life sciences. Particularly in the latter field, however,
-molecular properties are not well-understood yet, and the interaction between
-molecules and larger systems remains a challenge to be modeled. Therefore, the
+drug discovery, and the life sciences. It provides various models to represent
+chemical structures, of which the chemical graph is essential, and as such
+complements quantum chemical approaches. Large scale calculations relies on
+graph-based approaches, which are faster due to the graph simplification of
+chemical structures. As such, cheminformatics is much used in many life
+sciences fields, where the understanding of molecular properties and their
+interactions with the cell and the organism. Therefore, the
 development of new data mining and pattern recognition continues to be important
@@
 \paragraph*{Results:}
-We here report the improvements since the 1.2 series of the Chemistry Development Kit and outline how
-it evolved since the previous report. We demonstrate the new APIs we introduced
+We here report the improvements of the Chemistry Development Kit since the
+1.2 release series. This paper outlines how the library evolved since the
+previous report. We demonstrate the new APIs we introduced
 in this version and discuss the extensive quality control mechanism we have

### 2013-10-17 article.tex
Updated my affiliation detail
Wrote a bit of the introduction

@@
   Egon L Willighagen$^{1}$% // FIXED FIRST AUTHOR
-  \email{Egon L Willighagen - COMPLETE}
+  \email{Egon L Willighagen - egon.willighagen@maastrichtuniversity.nl}
 \and
@@
 \address{
-    \iid(1)Department of Bioinformatics - BiGCaT, \\
+    \iid(1)Department of Bioinformatics - BiGCaT, Maastricht University, Maastricht, NL-6200 MD, The Netherlands \\
     \iid(2)Chemoinformatics and Metabolism team, European Bioinformatics Institute, Hinxton, UK, \\
@@
 
-Open Source cheminformatics state \ldots (EGON)
+Open Source cheminformatics has made significants steps recently~\cite{OBoyle2011}.
+Development of the Chemistry Development Kit (CDK) has been previously documented~\cite{Steinbeck2003,Steinbeck2006},
+but followed this general trend in open source cheminformatics, and increased in functionality
+and adoption. For example, new tools expose CDK functionality, such as Cinfony~\cite{OBoyle2008},
+rcdk~\cite{}, and workflow plugins for Taverna\cite{Truszkowski2011}, KNIME\cite{Beisken2013}.
+Furthermore, projects with specific functionality have emerged building on the CDK, including
+jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
+Padel~\cite{yap2011padel}, AMBIT-SMARTS~\cite{jeliazkova2011ambitsmarts}, AMBIT-Tautomer~\cite{kochev2013ambit},
+SMSD~\cite{}, and FooFromRydberg~\cite{}. Other tools used previous versions already, and follow
+newer releases of the CDK, such as Bioclipse~\cite{} and AMBIT~\cite{jeliazkova2011ambit}.
 
-Adoption of the CDK ... Cinfony~\cite{OBoyle2008}, and other bindings \ldots ...
-
-Recent prominent use cases of the CDK~\cite{Steinbeck2003,Steinbeck2006} \ldots (EGON)
-AMBIT~\cite{jeliazkova2011ambit}, jCompoundMapper~\cite{Hinselmann2011},
-ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012}, Padel~\cite{yap2011padel}, ...
-
-Plugins for Taverna\cite{Truszkowski2011}, KNIME\cite{Beisken2013}, ...
-
-Some extensions make it back into the main library, like SMSD~\cite{} and fingerprints from Padel.
-Not all extensions made it back into the main library, such as those from
-AMBIT-SMARTS~\cite{jeliazkova2011ambitsmarts}, AMBIT-Tautomer~\cite{kochev2013ambit}, FooFromRydberg~\cite{}.
+Importantly, the CDK has seen independent validation ... of both functionality, quality, and performance...
+... fingerprint results ... performance paper ... Sayle MDL atom type work.
 

### 2014-06-25 article.tex
Streamlined the abstract

@@
 Cheminformatics is a well-established field with many applications in chemistry,
-drug discovery, and the life sciences. It provides various models to represent
-chemical structures, of which the chemical graph is essential, and as such
-complements quantum chemical approaches. Large scale calculations relies on
-graph-based approaches, which are faster due to the graph simplification of
-chemical structures. As such, cheminformatics is much used in many life
-sciences fields, where the understanding of molecular properties and their
-interactions with the cell and the organism. Therefore, the
-development of new data mining and pattern recognition continues to be important
-and the need for access to a cheminformatics library that one can change remains.
+biology, drug discovery, and others. The Chemistry Development Kit (CDK) has
+become a widely used Open Source cheminformatics toolkit, providing various models to represent
+chemical structures, of which the chemical graph is essential. However,
+in the first five years of the project increased so much in size that
+interdependencies between components grew unmanageable large, resulting
+in unpredictable instabilities.
 \paragraph*{Results:}
-We here report the improvements of the Chemistry Development Kit since the
-1.2 release series. This paper outlines how the library evolved since the
-previous report. We demonstrate the new APIs we introduced
-in this version and discuss the extensive quality control mechanism we have
-adopted. New APIs have been introduced for substructure searching, rendering of
-molecules, handling of molecular formulas, atom type perception, and InChI
-support. At a quality control side we have introduced automated building,
-extensive unit and use case testing, and adopted peer review.
+We here report the improvements of the CDK since the 1.2 release series made
+to accomodate both the increased complexity of the library, as well as
+siginificant improvements of and additions to the functionality of the
+library. We here outline how the CDK evolved since the previous report and
+discuss the quality control approached we have adopted to ensure
+stability. Additionally, new APIs have been introduced for atom type
+perception, substructure searching, rendering of molecules, and handling of
+molecular formulas.
 \paragraph*{Conclusions:}

### 2014-06-25 article.tex
Cleaned up the background/introduction

@@
 
-Open Source cheminformatics has made significants steps recently~\cite{OBoyle2011}.
-Development of the Chemistry Development Kit (CDK) has been previously documented~\cite{Steinbeck2003,Steinbeck2006},
-but followed this general trend in open source cheminformatics, and increased in functionality
-and adoption. For example, new tools expose CDK functionality, such as Cinfony~\cite{OBoyle2008},
-rcdk~\cite{}, and workflow plugins for Taverna\cite{Truszkowski2011}, KNIME\cite{Beisken2013}.
+Open Source cheminformatics has made significants steps forward recently~\cite{OBoyle2011}.
+The Chemistry Development Kit (CDK) is one of the tools under the Blue Obelisk wings,
+and previously documented version have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
+For example, new tools expose CDK functionality, such as Cinfony~\cite{OBoyle2008},
+rcdk~\cite{}, and workflow plugins for Taverna\cite{Truszkowski2011} and KNIME\cite{Beisken2013}.
 Furthermore, projects with specific functionality have emerged building on the CDK, including
@@
 Padel~\cite{yap2011padel}, AMBIT-SMARTS~\cite{jeliazkova2011ambitsmarts}, AMBIT-Tautomer~\cite{kochev2013ambit},
-SMSD~\cite{}, and FooFromRydberg~\cite{}. Other tools used previous versions already, and follow
+and SMSD~\cite{}. Other tools used previous versions already, and follow
 newer releases of the CDK, such as Bioclipse~\cite{} and AMBIT~\cite{jeliazkova2011ambit}.
 
-Importantly, the CDK has seen independent validation ... of both functionality, quality, and performance...
-... fingerprint results ... performance paper ... Sayle MDL atom type work.
+However, previous version of the CDK were written by a community with specific applications
+in mind, of which structure elucidation was one. It could therefore be rather competitive
+in some cheminformatics domains, such as fingerprinting~\cite{Ekins,Cannon}, in other parts
+it could be inadequate, such in the capturing of stereochemistry.
+
+Continued community development of the CDK in the last few years, however, has shown that
+Open Source software is here to stay. The adoption of automatic build systems and
+quality control methodologies such as unit testing, automated source code validation,
+and peer review by fellow developers have greatly improved the stability of the library,
+At the expense of slowing down the development, it has allowed for cleaning up interdependencies
+between modules of code, and has reset the scalability of the development model,
+allowing for a lot of new functionality in code APIs while maintaining the quality
+of code depending on those core APIs.
+
+This has resulted in important improved functionality, such as a separately published
+InChI functionality~\cite{SpjuthXXXX} and greatly improved ring detection functionality~\cite{MayYYYY},
+but also a core atom type perception module, covering a much wider set of elements than
+previous CDK versions and capturing many more charged and radical atom species,
+a more comprehensive fingerprinting API, many speed and stability improvements, and more.
 

### 2014-07-03 article.tex
work on fingerprint table

@@
 %%
-%\section*{Tables}
-%  \subsection*{Table 1 - Sample table title}
+\section*{Tables}
+  \subsection*{Table 1 - The molecular fingerprints in CDK}
+    \par
+    \mbox{
+    \begin{tabular}{lcclrc}
+                             & Bit version  & Count version & CDK version & Default Size & References \\
+  \cline{2-6}\noalign{\smallskip}
+  EStateFingerprinter        & $\checkmark$ &               & 1.2.0       & $79$         & \cite{Hall1995}\\
+  ExtendedFingerprinter      & $\checkmark$ &               & 1.0         &              & \\
+  Fingerprinter              & $\checkmark$ &               & 1.0         &              & \\
+  GraphOnlyFingerprinter     & $\checkmark$ &               & 1.0         &              & \\
+  HybridizationFingerprinter & $\checkmark$ &               & 1.4.0       &              & \\
+  KlekotaRothFingerprinter   & $\checkmark$ &               & 1.4.6       & $4\,860$     & \cite{Klekota2008}\\
+  LingoFingerprinter         & $\checkmark$ &               & 1.6.0 ?     &              & \\
+  MACCSFingerprinter         & $\checkmark$ &               & 1.2.0       &              & \\
+  PubchemFingerprinter       & $\checkmark$ &               & 1.4.0       &              & \\
+  ShortestPathFingerprinter  & $\checkmark$ &               & 1.6.0 ?     &              & \\
+  SignatureFingerprinter     & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $2^{32}$     & [pending] \\
+  SubstructureFingerprinter  & $\checkmark$ &               & 1.0         &              & \\
+    \end{tabular}}
+
 %    Here is an example of a \emph{small} table in \LaTeX\ using  

### 2014-07-03 article.tex
author spelling error
work on fingerprint table

@@
 \usepackage{cite}
+\usepackage[table]{xcolor}
+\usepackage{booktabs}
 
@@
 \setlength{\columnsep}{0.6cm}
+\definecolor{lightgray}{gray}{0.9}
 
@@
 \and
-  Jonathan Alversson$^3$%
-  \email{Jonathan Alversson - COMPLETE}
+  Jonathan Alvarsson$^3$%
+  \email{Jonathan Alvarsson - COMPLETE}
 \and
@@
   \subsection*{Table 1 - The molecular fingerprints in CDK}
+    \rowcolors{1}{}{lightgray}
     \par
@@
     \begin{tabular}{lcclrc}
+    \toprule
                              & Bit version  & Count version & CDK version & Default Size & References \\
-  \cline{2-6}\noalign{\smallskip}
   EStateFingerprinter        & $\checkmark$ &               & 1.2.0       & $79$         & \cite{Hall1995}\\
@@
   SubstructureFingerprinter  & $\checkmark$ &               & 1.0         &              & \\
+  CirkularFingerprinter      & $\checkmark$ & ?             & ?           & ?            & [pending?] \\
+  \bottomrule
     \end{tabular}}

### 2014-07-03 article.tex

@@
 library. We here outline how the CDK evolved since the previous report and
-discuss the quality control approached we have adopted to ensure
-stability. Additionally, new APIs have been introduced for atom type
-perception, substructure searching, rendering of molecules, and handling of
-molecular formulas.
+discuss the quality control approached we have adopted to ensure stability.
+Additionally, new APIs have been introduced for atom type perception,
+substructure searching, molecular fingerprints, rendering of molecules, and
+handling of molecular formulas.
 \paragraph*{Conclusions:}
@@
 that an open source cheminformatics project can act as a peer reviewed
-publishing platform for scientific computating software.
+publishing platform for scientific computing software.
 \end{abstract}
@@
 and SMSD~\cite{}. Other tools used previous versions already, and follow
-newer releases of the CDK, such as Bioclipse~\cite{} and AMBIT~\cite{jeliazkova2011ambit}.
+newer releases of the CDK, such as Bioclipse~\cite{spjuth2007bioclipse,
+spjuth2009bioclipse} and AMBIT~\cite{jeliazkova2011ambit}.
 
@@
 release. Traditionally CDK fingerprints were represented using the
-\textit{BitSet} class shipping with Java. The benefits of the class include
+\textit{BitSet} class shipping with Java. The benefits of that class include
 such things as already implemented methods for modifying bit sets using other
@@
                              & Bit version  & Count version & CDK version & Default Size & References \\
+  CirkularFingerprinter      & $\checkmark$ & ?             & ?           & ?            & [pending?] \\
   EStateFingerprinter        & $\checkmark$ &               & 1.2.0       & $79$         & \cite{Hall1995}\\
@@
   SubstructureFingerprinter  & $\checkmark$ &               & 1.0         &              & \\
-  CirkularFingerprinter      & $\checkmark$ & ?             & ?           & ?            & [pending?] \\
   \bottomrule

### 2014-07-04 article.tex

@@
 
+\newpage
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
   \subsection*{Table 1 - The molecular fingerprints in CDK}
+  Listed are the currently available molecular fingerprint in CDK with
+  information about whether they come as bit / count versions, what CDK version
+  they appeared in, their default size and where applicable also relevant
+  references.
+  \vskip 1\baselineskip
+
+    \begin{minipage}{1\textwidth}
+    \renewcommand*{\thempfootnote}{\fnsymbol{mpfootnote}}
+    \centering
     \rowcolors{1}{}{lightgray}
-    \par
-    \mbox{
-    \begin{tabular}{lcclrc}
+    \begin{tabular}{lcclcc}
     \toprule
                              & Bit version  & Count version & CDK version & Default Size    & References \\
-  CirkularFingerprinter      & $\checkmark$ & ?             & ?           & ?            & [pending?] \\
-  EStateFingerprinter        & $\checkmark$ &               & 1.2.0       & $79$         & \cite{Hall1995}\\
-  ExtendedFingerprinter      & $\checkmark$ &               & 1.0         &              & \\
-  Fingerprinter              & $\checkmark$ &               & 1.0         &              & \\
-  GraphOnlyFingerprinter     & $\checkmark$ &               & 1.0         &              & \\
-  HybridizationFingerprinter & $\checkmark$ &               & 1.4.0       &              & \\
-  KlekotaRothFingerprinter   & $\checkmark$ &               & 1.4.6       & $4\,860$     & \cite{Klekota2008}\\
-  LingoFingerprinter         & $\checkmark$ &               & 1.6.0 ?     &              & \\
-  MACCSFingerprinter         & $\checkmark$ &               & 1.2.0       &              & \\
-  PubchemFingerprinter       & $\checkmark$ &               & 1.4.0       &              & \\
-  ShortestPathFingerprinter  & $\checkmark$ &               & 1.6.0 ?     &              & \\
+  CirkularFingerprinter      & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $1\,024$ / $2^{32}$%
+\footnote[1]{For the CirkularFingerprinter the bit version is folded to 1024 whereas the count version is unfolded}
+                                                                                            & \cite{rogers2010extended, pending?} \\
+  EStateFingerprinter        & $\checkmark$ &               & 1.2.0       & $79$            & \cite{Hall1995} \\
+  ExtendedFingerprinter      & $\checkmark$ &               & 1.0         & $1\,024$        & \\
+  Fingerprinter              & $\checkmark$ &               & 1.0         & $1\,024$        & \\
+  GraphOnlyFingerprinter     & $\checkmark$ &               & 1.0         & $1\,024$        & \\
+  HybridizationFingerprinter & $\checkmark$ &               & 1.4.0       & $1\,024$        & \\
+  KlekotaRothFingerprinter   & $\checkmark$ &               & 1.4.6       & $4\,860$        & \cite{Klekota2008} \\
+  LingoFingerprinter         & $\checkmark$ &               & 1.6.0 ?     & NA%
+\footnote[2]{The Lingofingerprinter does not have a default size}               
+                                                                                            & \cite{vidal2005lingo} \\
+  MACCSFingerprinter         & $\checkmark$ &               & 1.2.0       & $166$           & \\
+  PubchemFingerprinter       & $\checkmark$ &               & 1.4.0       & $881$           & \cite{pubchemFP} \\
+  ShortestPathFingerprinter  & $\checkmark$ &               & 1.6.0 ?     & $1\,024$        & \\
   SignatureFingerprinter     & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $2^{32}$        & [pending] \\
-  SubstructureFingerprinter  & $\checkmark$ &               & 1.0         &              & \\
+  SubstructureFingerprinter  & $\checkmark$ &               & 1.0         & $307$           & \\
   \bottomrule
-    \end{tabular}}
+    \end{tabular}
+    \end{minipage}
 

### 2014-08-14 article.tex
Added CirkularFingerprinter reference

@@
 \footnote[1]{For the CirkularFingerprinter the bit version is folded to 1024 whereas the count version is unfolded}
-                                                                                            & \cite{rogers2010extended, pending?} \\
+                                                                                            & \cite{rogers2010extended, Clark2014} \\
   EStateFingerprinter        & $\checkmark$ &               & 1.2.0       & $79$            & \cite{Hall1995} \\

### 2015-12-23 article.tex
added signature fingerprints reference to fp table
Patches from Friday evening

@@
 \paragraph*{Results:}
-We here report the improvements of the CDK since the 1.2 release series made
+We here report improvements to the CDK since the 1.2 release series made
 to accomodate both the increased complexity of the library, as well as
 siginificant improvements of and additions to the functionality of the
-library. We here outline how the CDK evolved since the previous report and
-discuss the quality control approached we have adopted to ensure stability.
-Additionally, new APIs have been introduced for atom type perception,
-substructure searching, molecular fingerprints, rendering of molecules, and
-handling of molecular formulas.
+library. Second, we outline how the CDK evolved with respect to
+quality control and the approach we have adopted to ensure stability, including
+a peer review mechanism.
+Additionally, a selection of the new APIs that have been introduced will
+be discussed: atom type perception, substructure searching, molecular
+fingerprints, rendering of molecules, and handling of molecular formulas.
 \paragraph*{Conclusions:}
 With this paper we have shown the continued effort to provide a free, Open Source
-cheminformatics library, and show that such colleborative projects can survive for
+cheminformatics library, and show that such colleborative projects can exist over
 a long period. We have taken advantage from the community support, and show
@@
 The chemical formula is the basic/simple chemical representation of a compound.
-It defines the number of atoms/isotopes and type of elements that constitute a chemical compound without describing how atoms are bonded.
-It is also the first step toward the identification of the chemical structure from a unidentified compound.
+It defines the number of isotopes or elements that constitute a the composition
+of a compound without describing how atoms are bonded. With the rise of metabolomics,
+it has become increasingly relevant to have full support in cheminformatics
+libraries~\cite{RojasChertoXXX}.
+It is also the first step
+toward the identification of the chemical structure from a unidentified compound.
 
@@
   ....
-
-  \subsubsection*{Partial charges and Delocalization}
-
-Partial atomic charges is an electronic property of atoms defining the asymmetric distribution of electrons in chemical bonds.
-It is used as a quantitative correlation with certain compound's physical and chemical properties.
-
-CDK can
-\begin{itemize}
-\item generate all resonance structures from a given compound 
-\item calculate the topological and electronic weight value of each of the resonance structures
-\item calculate the partial atomic charge from a given structure 
-\end{itemize}
-
-Where can it be applied?
 
@@
 \ifthenelse{\boolean{publ}}{\small}{}
-Niels Out and Stefan Kuhn are acknowledged for their contributions to the project.
+The authors acknoledge the great number of people who have contributed smaller
+and larger contributions to the CDK library. A full list of contributors is
+found in the AUTHORS file~\cite{AUTHORS}.
 
@@
   ShortestPathFingerprinter  & $\checkmark$ &               & 1.6.0 ?     & $1\,024$        & \\
-  SignatureFingerprinter     & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $2^{32}$        & [pending] \\
+  SignatureFingerprinter     & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $2^{32}$        & \cite{signaturefingerprints} \\
   SubstructureFingerprinter  & $\checkmark$ &               & 1.0         & $307$           & \\

### 2015-12-23 article.tex
A bit more content

@@
 
+  New classes include:
+
   \subsubsection*{Rendering API}
@@
   Initiated by Niels Out ...
+
+  \subsubsection*{Structure Diagram Layout}
+
+  The structure diagram layout ... %FIXME: John, do you want this in?
 
@@
 
+% FIXME: add dependency plot
+
   \subsubsection*{Documentation}
@@
 
-  PMD ...
+The project continues to use PMD (http://pmd.sf.net/) for code quality checking,
+but deviates from the default rules. For example, we are more liberal with 
+variable name length. Moreover, a number of additional PMD tests have been
+developed specifically for the CDK, that, for example, test if a class uses
+the core interfaces instead of implementations of those interfaces. For example,
+that the code uses IAtom instead of Atom. That said, these tests do generate a
+few false positives, as the tests check the class name only, and not the
+Java package the class is in.
 
   \subsubsection*{Git, branching, and patches}
+
+Another change made over the past years is the move to a Git-based version
+control system. Advantages the projects makes advantage of is the distributed
+nature of Git repositories, and easier branching and making available of
+patches. GitHub has replaced SourceForge as the main source code hosting service
+where we can use novel approaches for commenting on code, pull requests, etc.
+These new features simplify our code review process.
 
@@
 
-INVITE NINA
+Another recent change is the choice of build system: we have moved away from
+Ant to Maven to deal with dependencies and compile the library.
+
+% FIXME: write more about this, and provide info about Maven Central
 
@@
 This paper outlines how peer review and quality control contributed to a much wider adoption
-of our cheminformatics library and seen many new additions. With over 75 contributors,
+of our cheminformatics library and seen many new additions. With 93 contributors,
 the CDK is alive and kicking.

### 2015-12-23 article.tex
More work

@@
 in mind, of which structure elucidation was one. It could therefore be rather competitive
-in some cheminformatics domains, such as fingerprinting~\cite{Ekins,Cannon}, in other parts
+in some cheminformatics domains, such as fingerprinting~\cite{Clark2014,Cannon}, in other parts
 it could be inadequate, such in the capturing of stereochemistry.
@@
 This has resulted in important improved functionality, such as a separately published
-InChI functionality~\cite{SpjuthXXXX} and greatly improved ring detection functionality~\cite{MayYYYY},
+InChI functionality~\cite{Spjuth2013} and greatly improved ring detection functionality~\cite{May2014},
 but also a core atom type perception module, covering a much wider set of elements than
@@
 
-  CDK 1.4.XXXX [CHECK] introduces a new API for stereochemistry information: the interfaces
-  IStereoChemistryElement was introduced with ...
-
-  New classes include:
+  The new CDK version introduces a new API for stereochemistry information: the interfaces
+  IStereoChemistryElement was introduced with the implementations TetrahedralChirality,
+  for four coordinate compounds like bromo-chloro-fluoro-iodomethane,
+  DoubleBondStereochemistry, for compounds like trans-2-butene, and
+  ExtendedTetrahedral, for compounds like allene.
 
@@
 it has become increasingly relevant to have full support in cheminformatics
-libraries~\cite{RojasChertoXXX}.
+libraries~\cite{RojasCherto2011}.
 It is also the first step
@@
 
-  ....
+  .... \cite{May2014}
 
@@
 nature of Git repositories, and easier branching and making available of
-patches. GitHub has replaced SourceForge as the main source code hosting service
+patches. GitHub (\url{https://github.com/cdk/cdk}) has replaced SourceForge
+as the main source code hosting service
 where we can use novel approaches for commenting on code, pull requests, etc.
@@
 \item \textbf{Project Name}: The Chemistry Development Kit
-\item \textbf{Project home page}: \url{http://cdk.sourceforge.net/}
+\item \textbf{Project home page}: \url{https://github.com/cdk/cdk}
 \item \textbf{Operating system(s)}: Windows, GNU/Linux, OS/X
 \item \textbf{Programming language}: Java
+% FIXME: add appropriate references below
 \item \textbf{Other (optional) requirements}: JNI-InChI, Vecmath, BEAM, Guava, JGraphT, Signatures, CMLDOM, XOM, JavaCC
@@
   \subsection*{Table 1 - The molecular fingerprints in CDK}
+  \label{tab:fingerprints}
   Listed are the currently available molecular fingerprint in CDK with

### 2015-12-24 article.tex
Added the dependency graph
Upgraded to a recent LaTeX template

@@
 \NeedsTeXFormat{LaTeX2e}[1995/12/01]
-\documentclass[10pt]{bmc_article}
+\documentclass[10pt]{bmcart}
+%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
+%%% additional documentclass options:
+%  [doublespacing]
+%  [linenumbers]   - put the line numbers on margins
 
@@
 \usepackage{url}  % Formatting web addresses  
-\usepackage{ifthen}  % Conditional 
-\usepackage{multicol}   %Columns
+\urlstyle{rm}
 \usepackage[utf8]{inputenc} %unicode support
 \usepackage{amssymb}
-%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
-%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
-\urlstyle{rm}
-\usepackage{cite}
-\usepackage[table]{xcolor}
-\usepackage{booktabs}
+%\usepackage{cite}
+\usepackage{graphicx}
 
@@
 
-\def\includegraphic{}
-\def\includegraphics{}
+%\def\includegraphic{}
+%\def\includegraphics{}
 
-\setlength{\topmargin}{0.0cm}
-\setlength{\textheight}{21.5cm}
-\setlength{\oddsidemargin}{0cm} 
-\setlength{\textwidth}{16.5cm}
-\setlength{\columnsep}{0.6cm}
-\definecolor{lightgray}{gray}{0.9}
-
-\newboolean{publ}
-
-%Review style settings
-\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
-
-%Publication style settings
-%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
+%%% Put your definitions there:
+\startlocaldefs
+\endlocaldefs
 
@@
 \begin{document}
-\begin{bmcformat}
+
+%%% Start of article front matter
+\begin{frontmatter}
+
+\begin{fmbox}
+\dochead{Research}
 
@@
 
-\author{
-  Egon L Willighagen$^{1}$% // FIXED FIRST AUTHOR
-  \email{Egon L Willighagen - egon.willighagen@maastrichtuniversity.nl}
-\and
-  John May$^2$% // FIXED SECOND AUTHOR
-  \email{John May - COMPLETE@ebi.ac.uk}
-\and
-  Jonathan Alvarsson$^3$%
-  \email{Jonathan Alvarsson - COMPLETE}
-\and
-  Arvid Berg$^3$%
-  \email{Arvid Berg - COMPLETE}
-\and
-  Nina Jeliazkova$^4$%
-  \email{Nina Jeliakvova - COMPLETE}
-\and
-  Dazhi Jiao$^5$%
-  \email{Dazhi Jiao - COMPLETE}
-\and
-  Syed Asad Rahman$^6$%
-  \email{Syed Asad Rahman - COMPLETE}
-\and
-  Mark Rijnbeek$^2$%
-  \email{Mark Rijnbeek - COMPLETE}
-\and
-  Miguel Rojas Cherto$^7$%
-  \email{Miguel Rojas Cherto - COMPLETE}
-\and
-  Ola Spjuth$^3$%
-  \email{Ola Spjuth - ola.spjuth@farmbio.uu.se}
-\and
-  Gilleain Torrance$^5$%
-  \email{Gilleain Torrance - COMPLETE}
-\and
-  Rajarshi Guha$^{8}$% // FIXED ONE BUT LAST AUTHOR
-  \email{Rajarshi Guha - COMPLETE}
-\and
-  Christoph Steinbeck$^{2}$% // FIXED LAST AUTHOR
-  \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
+% FIXME: finish migrating the author list
+
+\author[
+   addressref={um},                   % id's of addresses, e.g. {aff1,aff2}
+   %corref={aff1},                       % id of corresponding address, if any
+   %noteref={n1},                        % id's of article notes, if any
+   email={egon.willighagen@maastrichtuniversity.nl}   % email address
+]{\inits{ELW}\fnm{Egon L} \snm{Willighagen}}
+\author[
+   addressref={aff2},
+   email={???}
+]{\inits{JWM}\fnm{John W} \snm{May}}
+
+\address[id=um]{%                           % unique id
+  \orgname{Dept of Bioinformatics - BiGCaT, NUTRIM, Maastricht University}, % university, etc
+  %\street{Waterloo Road},                     %
+  \postcode{NL-6200 MD}                                % post or zip code
+  \city{Maastricht},                              % city
+  \cny{NL}                                    % country
+}
+\address[id=aff2]{%
+  \orgname{??},
+  %\street{D\"{u}sternbrooker Weg 20},
+  %\postcode{24105}
+  %\city{Kiel},
+  \cny{UK}
 }
 
-\address{
-    \iid(1)Department of Bioinformatics - BiGCaT, Maastricht University, Maastricht, NL-6200 MD, The Netherlands \\
-    \iid(2)Chemoinformatics and Metabolism team, European Bioinformatics Institute, Hinxton, UK, \\
-    \iid(3)Department of Pharmaceutical Biosciences, Uppsala University, 751 24 Uppsala, Sweden \\
-    \iid(4)Ideaconsult Ltd, A. Kanchev 4, Sofia 1000, Bulgaria \\
-    \iid(5)Example Ltd., Glasgow, UK \\
-    \iid(6)Example Ltd., Glasgow, UK \\
-    \iid(7)Division of Analytical Biosciences, Leiden/Amsterdam Center for Drug Research, Leiden, The Netherlands \\
-    \iid(8)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA
-}
+% \author{
+%   Egon L Willighagen$^{1}$% // FIXED FIRST AUTHOR
+%   \email{Egon L Willighagen - }
+% \and
+%   John May$^2$% // FIXED SECOND AUTHOR
+%   \email{John May - COMPLETE@ebi.ac.uk}
+% \and
+%   Jonathan Alvarsson$^3$%
+%   \email{Jonathan Alvarsson - COMPLETE}
+% \and
+%   Arvid Berg$^3$%
+%   \email{Arvid Berg - COMPLETE}
+% \and
+%   Nina Jeliazkova$^4$%
+%   \email{Nina Jeliakvova - COMPLETE}
+% \and
+%   Dazhi Jiao$^5$%
+%   \email{Dazhi Jiao - COMPLETE}
+% \and
+%   Syed Asad Rahman$^6$%
+%   \email{Syed Asad Rahman - COMPLETE}
+% \and
+%   Mark Rijnbeek$^2$%
+%   \email{Mark Rijnbeek - COMPLETE}
+% \and
+%   Miguel Rojas Cherto$^7$%
+%   \email{Miguel Rojas Cherto - COMPLETE}
+% \and
+%   Ola Spjuth$^3$%
+%   \email{Ola Spjuth - ola.spjuth@farmbio.uu.se}
+% \and
+%   Gilleain Torrance$^5$%
+%   \email{Gilleain Torrance - COMPLETE}
+% \and
+%   Rajarshi Guha$^{8}$% // FIXED ONE BUT LAST AUTHOR
+%   \email{Rajarshi Guha - COMPLETE}
+% \and
+%   Christoph Steinbeck$^{2}$% // FIXED LAST AUTHOR
+%   \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
+% }
+%      
+% \address{
+%     \iid(1) \\
+%     \iid(2)Chemoinformatics and Metabolism team, European Bioinformatics Institute, Hinxton, UK, \\
+%     \iid(3)Department of Pharmaceutical Biosciences, Uppsala University, 751 24 Uppsala, Sweden \\
+%     \iid(4)Ideaconsult Ltd, A. Kanchev 4, Sofia 1000, Bulgaria \\
+%     \iid(5)Example Ltd., Glasgow, UK \\
+%     \iid(6)Example Ltd., Glasgow, UK \\
+%     \iid(7)Division of Analytical Biosciences, Leiden/Amsterdam Center for Drug Research, Leiden, The Netherlands \\
+%     \iid(8)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA
+% }
 
-\maketitle
+%\begin{artnotes}
+%\note{Sample of title note}     % note to the article
+%\note[id=n1]{Equal contributor} % note, connected to author
+%\end{artnotes}
+
+\end{fmbox}% comment this for two column layout
 
 %% The Abstract begins here
+\begin{abstractbox}
 
@@
 % Do not use inserted blank lines (ie \\) until main body of text.
-\paragraph*{Background:}
+\parttitle{Background}
 Cheminformatics is a well-established field with many applications in chemistry,
@@
 in unpredictable instabilities.
-\paragraph*{Results:}
+\parttitle{Results}
 We here report improvements to the CDK since the 1.2 release series made
@@
 fingerprints, rendering of molecules, and handling of molecular formulas.
-\paragraph*{Conclusions:}
+\parttitle{Conclusions}
 With this paper we have shown the continued effort to provide a free, Open Source
@@
 
-\ifthenelse{\boolean{publ}}{\begin{multicols}{2}}{}
+\begin{keyword}
+\kwd{Java}
+\kwd{cheminformatics}
+\kwd{bioinformatics}
+\end{keyword}
+
+% MSC classifications codes, if any
+%\begin{keyword}[class=AMS]
+%\kwd[Primary ]{}
+%\kwd{}
+%\kwd[; secondary ]{}
+%\end{keyword}
+
+\end{abstractbox}
+%
+%\end{fmbox}% uncomment this for twcolumn layout
+
+\end{frontmatter}
 
@@
 
+%%%%%%%%%%%
+
+\begin{backmatter}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section*{Acknowledgements}
-\ifthenelse{\boolean{publ}}{\small}{}
 The authors acknoledge the great number of people who have contributed smaller
@@
 
-{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
- \bibliographystyle{bmc_article}  % Style BST file
+\bibliographystyle{bmc-mathphys}  % Style BST file
 \bibliography{article}
-}
-
-%%%%%%%%%%%
-
-\ifthenelse{\boolean{publ}}{\end{multicols}}{}
 
@@
 
-%\section*{Figures}
-%  \subsection*{Figure 1 - Sample figure title}
-%      A short description of the figure content
-%      should go here.
-%
+\section*{Figures}
+  \subsection*{Figure 1 - Dependencies between CDK modules.}
+  \label{fig:deps}
+      Visualization of the dependencies between CDK modules. For example,
+      the cdk-core depends on the cdk-interfaces module.
+
+\includegraphics[width=\textwidth]{cdkDeps.png}
+
 %  \subsection*{Figure 2 - Sample figure title}
@@
     \centering
-    \rowcolors{1}{}{lightgray}
     \begin{tabular}{lcclcc}
-    \toprule
                              & Bit version  & Count version & CDK version & Default Size    & References \\
@@
   SubstructureFingerprinter  & $\checkmark$ &               & 1.0         & $307$           & \\
-  \bottomrule
+
     \end{tabular}
@@
 
-\end{bmcformat}
+\end{backmatter}
+
 \end{document}

### 2015-12-24 article.tex
Some more work: removed the builder table; fixing refs

@@
 For example, new tools expose CDK functionality, such as Cinfony~\cite{OBoyle2008},
-rcdk~\cite{}, and workflow plugins for Taverna\cite{Truszkowski2011} and KNIME\cite{Beisken2013}.
+rcdk~\cite{Guha2007}, and workflow plugins for Taverna~\cite{Truszkowski2011} and KNIME~\cite{Beisken2013}.
 Furthermore, projects with specific functionality have emerged building on the CDK, including
@@
 Padel~\cite{yap2011padel}, AMBIT-SMARTS~\cite{jeliazkova2011ambitsmarts}, AMBIT-Tautomer~\cite{kochev2013ambit},
-and SMSD~\cite{}. Other tools used previous versions already, and follow
+and SMSD~\cite{Rahman2009,Rahman2014}. Other tools used previous versions already, and follow
 newer releases of the CDK, such as Bioclipse~\cite{spjuth2007bioclipse,
-spjuth2009bioclipse} and AMBIT~\cite{jeliazkova2011ambit}.
+spjuth2009bioclipse} and AMBIT~\cite{jeliazkova2011ambit}. While the above
+tools suggest that the CDK is primarily used to create general tools, it is
+important to note that it is also used to solve specific questions, like finding
+the maximally bridging rings in chemical structures~\cite{Marth2015}.
 
@@
 in mind, of which structure elucidation was one. It could therefore be rather competitive
-in some cheminformatics domains, such as fingerprinting~\cite{Clark2014,Cannon}, in other parts
+in some cheminformatics domains, such as fingerprinting~\cite{Clark2014,Cannon2006}, in other parts
 it could be inadequate, such in the capturing of stereochemistry.
@@
 are needed, these builders are being used. This allows to run a certain CDK-based
-application with a specific builder, summarized in Table~\ref{tab:builders}.
-
-\begin{table}
-\begin{tabular}{l|l}
-builder & description \\
-\hline
-DefaultChemObjectBuilder & The original builder ... \\
-DataDebugChemObjectBuilder & A builder that creates classes that send debug messages ... \\
-SilentChemObjectBuilder & Classes that are created with this builder will not
-    send around change events. \\
-\end{tabular}
-\end{table}
+application with a specific builder, aimed at a particular use case. The original
+implementation is available via the DefaultChemObjectBuilder, which creates
+classes that pass around change events, just like the original CDK data classes.
+However, a SilentChemObjectBuilder has been introduced that does not pass around
+change events, which makes code run about 10-20\% faster.
+The third builder is the DataDebugChemObjectBuilder which generates debug information
+for all changes to the content of the data classes. This can be useful for
+debugging and other forms of code inspection.
 

### 2015-12-24 article.tex
Migrated the authors

@@
 ]{\inits{JWM}\fnm{John W} \snm{May}}
+\author[
+   addressref={uppsala},
+   email={}
+]{\inits{JA}\fnm{Jonathan} \snm{Alvarsson}}
+\author[
+   addressref={uppsala},
+   email={}
+]{\inits{AB}\fnm{Arvid} \snm{Berg}}
+\author[
+   addressref={idea},
+   email={}
+]{\inits{NJ}\fnm{Nina} \snm{Jeliazkova}}
+\author[
+   addressref={???},
+   email={}
+]{\inits{MRJ}\fnm{Miguel} \snm{Rojas-Cherto}}
+\author[
+   addressref={uppsala},
+   email={}
+]{\inits{OS}\fnm{Ola} \snm{Spjuth}}
+\author[
+   addressref={nih},
+   email={}
+]{\inits{RG}\fnm{Rajarshi} \snm{Guha}}
+\author[
+   addressref={ebi},
+   email={}
+]{\inits{CS}\fnm{Christoph} \snm{Steinbeck}}
 
-\address[id=um]{%                           % unique id
+\address[id=um]{
   \orgname{Dept of Bioinformatics - BiGCaT, NUTRIM, Maastricht University}, % university, etc
   %\street{Waterloo Road},                     %
-  \postcode{NL-6200 MD}                                % post or zip code
+  \postcode{NL-6200 MD},
   \city{Maastricht},                              % city
-  \cny{NL}                                    % country
+  \cny{The Netherlands}                                    % country
 }
-\address[id=aff2]{%
+\address[id=aff2]{
   \orgname{??},
   %\street{D\"{u}sternbrooker Weg 20},
-  %\postcode{24105}
+  %\postcode{24105},
   %\city{Kiel},
@@
 }
-
-% \author{
-%   Egon L Willighagen$^{1}$% // FIXED FIRST AUTHOR
-%   \email{Egon L Willighagen - }
-% \and
-%   John May$^2$% // FIXED SECOND AUTHOR
-%   \email{John May - COMPLETE@ebi.ac.uk}
-% \and
-%   Jonathan Alvarsson$^3$%
-%   \email{Jonathan Alvarsson - COMPLETE}
-% \and
-%   Arvid Berg$^3$%
-%   \email{Arvid Berg - COMPLETE}
-% \and
-%   Nina Jeliazkova$^4$%
-%   \email{Nina Jeliakvova - COMPLETE}
-% \and
-%   Dazhi Jiao$^5$%
-%   \email{Dazhi Jiao - COMPLETE}
-% \and
-%   Syed Asad Rahman$^6$%
-%   \email{Syed Asad Rahman - COMPLETE}
-% \and
-%   Mark Rijnbeek$^2$%
-%   \email{Mark Rijnbeek - COMPLETE}
-% \and
-%   Miguel Rojas Cherto$^7$%
-%   \email{Miguel Rojas Cherto - COMPLETE}
-% \and
-%   Ola Spjuth$^3$%
-%   \email{Ola Spjuth - ola.spjuth@farmbio.uu.se}
-% \and
-%   Gilleain Torrance$^5$%
-%   \email{Gilleain Torrance - COMPLETE}
-% \and
-%   Rajarshi Guha$^{8}$% // FIXED ONE BUT LAST AUTHOR
-%   \email{Rajarshi Guha - COMPLETE}
-% \and
-%   Christoph Steinbeck$^{2}$% // FIXED LAST AUTHOR
-%   \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
-% }
-%      
-% \address{
-%     \iid(1) \\
-%     \iid(2)Chemoinformatics and Metabolism team, European Bioinformatics Institute, Hinxton, UK, \\
-%     \iid(3)Department of Pharmaceutical Biosciences, Uppsala University, 751 24 Uppsala, Sweden \\
-%     \iid(4)Ideaconsult Ltd, A. Kanchev 4, Sofia 1000, Bulgaria \\
-%     \iid(5)Example Ltd., Glasgow, UK \\
-%     \iid(6)Example Ltd., Glasgow, UK \\
-%     \iid(7)Division of Analytical Biosciences, Leiden/Amsterdam Center for Drug Research, Leiden, The Netherlands \\
-%     \iid(8)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA
-% }
+\address[id=uppsala]{
+  \orgname{Department of Pharmaceutical Biosciences, Uppsala University},
+  %\street{D\"{u}sternbrooker Weg 20},
+  \postcode{751 24},
+  \city{Uppsala},
+  \cny{Sweden}
+}
+\address[id=idea]{
+  \orgname{Ideaconsult Ltd},
+  \street{A. Kanchev 4},
+  \postcode{1000},
+  \city{Sofia},
+  \cny{Bulgaria}
+}
+\address[id=nih]{
+  \orgname{NIH Center for Translational Therapeutics},
+  \street{9800 Medical Center Drive},
+  \postcode{MD 20878},
+  \city{Rockville},
+  \cny{USA}
+}
+\address[id=ebi]{
+  \orgname{Chemoinformatics and Metabolism team, European Bioinformatics Institute},
+  %\street{A. Kanchev 4},
+  %\postcode{1000},
+  \city{Hinxton},
+  \cny{UK}
+}
 
@@
 
+\newpage
+
 %% Do not use \listoffigures as most will included as separate files
@@
 %% Tables                        %%
+
+\newpage
 

### 2015-12-26 article.tex
More detail about the deps figure
Overview of key modules and their dependencies

@@
 
-% FIXME: add dependency plot
+An overview of key modules with description, important changes, and dependencies
+on third-party libraries is given in Table~\ref{tab:modules} and the dependencies
+between the CDK modules are depicted in Figure~\ref{fig:deps}.
 
@@
       Visualization of the dependencies between CDK modules. For example,
-      the cdk-core depends on the cdk-interfaces module.
+      the cdk-core depends on the cdk-interfaces module. A few higher level
+      modules have been left out: cdk-builder3dtools, cdk-legacy, and
+      cdk-depict.
 
@@
 \section*{Tables}
-  \subsection*{Table 1 - The molecular fingerprints in CDK}
+
+  \subsection*{Table 1 - A selection of key CDK modules with major changes.}
+  \label{tab:modules}
+  An overview of a selection of often used CDK modules with description,
+  dependencies on third-party libraries, and the major changes since
+  version~1.2. Depedencies between modules are depicted in Figure~\ref{fig:deps}.
+  \vskip 1\baselineskip
+
+    \begin{minipage}{1\textwidth}
+    \renewcommand*{\thempfootnote}{\fnsymbol{mpfootnote}}
+    \centering
+    \begin{tabular}{llll}
+  Module                     & Description  & Major Changes & Depedencies \\
+  interfaces                 &              & & Vecmath 1.5.2 \\
+  core                       &              & & Google Guava 17.0 \\ % FIXME: how to add https://github.com/google/guava ?
+  standard                   &              & & \\
+  render                     &              & Redesigned to make it more modular and support multiple widget toolkits, like AWT and SWT. & \\
+  isomorphism                &              & & \\
+  atomtype                   &              & Unified approach where atom typing is separated from other algorithms. & \\
+  valencycheck               &              & & \\
+  ioformats                  &              & & \\
+  io                         &              & The MDL molfile reader has been rewritten and supports in atom types defind in the specification. & XPP3 1.1.4c \\
+  iordf                      &              & New. & Jena 2.7.4 \\
+  inchi                      &              & & JNI-InChI 0.8~\cite{Spjuth2013}  \\
+  libiocml                   &              & & XOM 1.2.5, CMLXOM 3.1~\cite{Murray-Rust2011} \\
+  smiles                     &              & SMILES support performance and converage is greatly improved. & Beam 0.9.1 \\ %FIXME add a citation
+  smarts                     &              & & Beam 0.9.1 \\ %FIXME add a citation
+  formula                    &              & New. & \\
+  fingerprint                &              & Many new fingerprint types (see text). & Apache Commons Math 3.1.1 \\
+  qsar                       &              & & XOM 1.2.5, JAMA 1.0.3~\cite{Hicklin2012} \\ % FIXME: find JAMA reference
+  signatures                 &              & & Signatures 1.1 \\ % FIXME: ask Gilleain for the reference
+  qsarmolecular              &              & & \\
+  log4j                      &              & & Log4j 1.2.17 \\
+    \end{tabular}
+    \end{minipage}
+
+  \subsection*{Table 1 - The molecular fingerprints in CDK.}
   \label{tab:fingerprints}

### 2015-12-26 article.tex
Easier to read layout

--- 2015-12-24 article.tex
@@
 \section*{Tables}
-  \subsection*{Table 1 - The molecular fingerprints in CDK}
+
+  \subsection*{Table 1 - A selection of key CDK modules with major changes.}
+  \label{tab:modules}
+  An overview of a selection of often used CDK modules with description,
+  dependencies on third-party libraries, and the major changes since
+  version~1.2. Depedencies between modules are depicted in Figure~\ref{fig:deps}.
+  \vskip 1\baselineskip
+
+    \begin{minipage}{1\textwidth}
+    \renewcommand*{\thempfootnote}{\fnsymbol{mpfootnote}}
+    \centering
+    \begin{tabular}{lp{3cm}p{3cm}l}
+  \textbf{Module}            & \textbf{Description}  & \textbf{Major Changes} & \textbf{Depedencies} \\ \hline
+  interfaces                 &              & & Vecmath 1.5.2 \\ \hline
+  core                       &              & & Google Guava 17.0 \\ \hline % FIXME: how to add https://github.com/google/guava ?
+  standard                   &              & & \\ \hline
+  render                     &              & Redesigned to make it more modular and support multiple widget toolkits, like AWT and SWT. & \\ \hline
+  isomorphism                &              & & \\ \hline
+  atomtype                   &              & Unified approach where atom typing is separated from other algorithms. & \\ \hline
+  valencycheck               &              & & \\ \hline
+  ioformats                  &              & & \\ \hline
+  io                         &              & The MDL molfile reader has been rewritten and supports in atom types defind in the specification. & XPP3 1.1.4c \\ \hline
+  iordf                      &              & New. & Jena 2.7.4 \\ \hline
+  inchi                      &              & & JNI-InChI 0.8~\cite{Spjuth2013}  \\ \hline
+  libiocml                   &              & & XOM 1.2.5, CMLXOM 3.1~\cite{Murray-Rust2011} \\ \hline
+  smiles                     &              & SMILES support performance and converage is greatly improved. & Beam 0.9.1 \\ \hline %FIXME add a citation
+  smarts                     &              & & Beam 0.9.1 \\ \hline %FIXME add a citation
+  formula                    &              & New. & \\ \hline
+  fingerprint                &              & Many new fingerprint types (see text). & Apache Commons Math 3.1.1 \\ \hline
+  qsar                       &              & & XOM 1.2.5, JAMA 1.0.3~\cite{Hicklin2012} \\ \hline % FIXME: find JAMA reference
+  signatures                 &              & & Signatures 1.1 \\ \hline % FIXME: ask Gilleain for the reference
+  qsarmolecular              &              & & \\ \hline
+  log4j                      &              & & Log4j 1.2.17 \\ \hline
+    \end{tabular}
+    \end{minipage}
+
+  \subsection*{Table 1 - The molecular fingerprints in CDK.}
   \label{tab:fingerprints}

### 2015-12-26 article.tex
Integrated the last and the first column in the FP table

@@
     \centering
-    \begin{tabular}{lcclcc}
-                             & Bit version  & Count version & CDK version & Default Size    & References \\
-  CirkularFingerprinter      & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $1\,024$ / $2^{32}$%
-\footnote[1]{For the CirkularFingerprinter the bit version is folded to 1024 whereas the count version is unfolded}
-                                                                                            & \cite{rogers2010extended, Clark2014} \\
-  EStateFingerprinter        & $\checkmark$ &               & 1.2.0       & $79$            & \cite{Hall1995} \\
-  ExtendedFingerprinter      & $\checkmark$ &               & 1.0         & $1\,024$        & \\
-  Fingerprinter              & $\checkmark$ &               & 1.0         & $1\,024$        & \\
-  GraphOnlyFingerprinter     & $\checkmark$ &               & 1.0         & $1\,024$        & \\
-  HybridizationFingerprinter & $\checkmark$ &               & 1.4.0       & $1\,024$        & \\
-  KlekotaRothFingerprinter   & $\checkmark$ &               & 1.4.6       & $4\,860$        & \cite{Klekota2008} \\
-  LingoFingerprinter         & $\checkmark$ &               & 1.6.0 ?     & NA%
+    \begin{tabular}{lcclc}
+                             & Bit version  & Count version & CDK version & Default Size    \\
+  CircularFingerprinter~\cite{rogers2010extended, Clark2014}      & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $1\,024$ / $2^{32}$%
+\footnote[1]{For the CirkularFingerprinter the bit version is folded to 1024 whereas the count version is unfolded} \\
+  EStateFingerprinter~\cite{Hall1995}       & $\checkmark$ &               & 1.2.0       & $79$            \\
+  ExtendedFingerprinter      & $\checkmark$ &               & 1.0         & $1\,024$        \\
+  Fingerprinter              & $\checkmark$ &               & 1.0         & $1\,024$        \\
+  GraphOnlyFingerprinter     & $\checkmark$ &               & 1.0         & $1\,024$        \\
+  HybridizationFingerprinter & $\checkmark$ &               & 1.4.0       & $1\,024$        \\
+  KlekotaRothFingerprinter~\cite{Klekota2008}   & $\checkmark$ &               & 1.4.6       & $4\,860$        \\
+  LingoFingerprinter~\cite{vidal2005lingo}         & $\checkmark$ &               & 1.6.0 ?     & NA%
 \footnote[2]{The Lingofingerprinter does not have a default size}               
-                                                                                            & \cite{vidal2005lingo} \\
-  MACCSFingerprinter         & $\checkmark$ &               & 1.2.0       & $166$           & \\
-  PubchemFingerprinter       & $\checkmark$ &               & 1.4.0       & $881$           & \cite{pubchemFP} \\
-  ShortestPathFingerprinter  & $\checkmark$ &               & 1.6.0 ?     & $1\,024$        & \\
-  SignatureFingerprinter     & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $2^{32}$        & \cite{signaturefingerprints} \\
-  SubstructureFingerprinter  & $\checkmark$ &               & 1.0         & $307$           & \\
+                                                                                             \\
+  MACCSFingerprinter         & $\checkmark$ &               & 1.2.0       & $166$           \\
+  PubchemFingerprinter~\cite{pubchemFP}       & $\checkmark$ &               & 1.4.0       & $881$            \\
+  ShortestPathFingerprinter  & $\checkmark$ &               & 1.6.0 ?     & $1\,024$        \\
+  SignatureFingerprinter~\cite{signaturefingerprints}     & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $2^{32}$         \\
+  SubstructureFingerprinter  & $\checkmark$ &               & 1.0         & $307$           \\
 

### 2016-01-04 article.tex
The new aromaticity framework should be discussed too
Paragraph on signatures and two references

@@
 
+  \subsection*{Signatures}
+  An implementation has been provided of the Signature structure descriptor for molecules 
+  \cite{Faulon2003}. These act as a linear notation - like the SMILES format - 
+  for the whole molecule as well as for connected substructures rooted at a single atom. The 
+  descriptor can also be canonicalized to provide isomorphism-independent representations 
+  \cite{Faulon2004}.
+
   \subsubsection*{Rendering API}
@@
   .... \cite{May2014}
+
+  \subsubsection*{Aromaticity}
+
+  ....
 

### 2016-01-26 article.tex
Filling the gaps

@@
   This CDK version has isolated a new atom typing framework, removing the perception of
-  atom types from the algorithms, allowing the perception to be properly unit tested.
-  The new code defined the atom type using an atom type ontology based on the XXXXX.
+  atom types from various algorithms, allowing the perception to be tested separately.
+  The new code defines the atom types using a list that specifies for each type
+  the element symbol, hybridization, formal charge, number of lone pairs, and
+  an enumeration of the bond orders. This list of properties captures the
+  information needed for the various algorithms in the CDK. For example,
+  hybridization information can be used in certain aromaticity models (see later),
+  and the lone pair information is needed for resonance structure calculation
+  needed, for example, for Gasteiger \pi-charges.
 
@@
 
-  The new code base defines XXXX atom types for YYYY elements, including ZZZZ radical
-  atom types. A full list of atom types is given in Additional File X.
-
-  The code is tested by FFFF unit tests, and has been tested against ChEBI and PubChem. REFS
-
   \subsubsection*{Stereochemistry}
@@
   \subsection*{Signatures}
+
   An implementation has been provided of the Signature structure descriptor for molecules 
@@
 
-  Initiated by Niels Out ...
+  A new rendering API has been introduced to make the rendering code independent
+  from Java widget toolkits. The previous code was tightly linked to the Swing
+  toolkit, but other tools use different widget toolkits. For example, Bioclipse
+  is based on Eclipse which uses the SWT~\cite{spjuth2007bioclipse}. %FIXME: write SWT in full
+
+  A second new design goal was introduced to balance between size restrictions
+  of some use cases, such as Java applets, and the rendering functionality. In
+  particular, some functionality, even after Modularization, needed considerable
+  parts of the CDK library, making creation if a small-sized applet unfeasible.
+  Therefore, the rendering API was modularized to allow splitting up rendering
+  functionality into modules, with varying CDK dependencies.
+
+  % DESCRIBE THE CORE APIs
 
@@
 
-%  \subsection*{Figure 2 - Sample figure title}
-%      Figure legend text.
+\subsection*{Figure 2 - Atom type information specified for a sp3-hybridized
+carbon.}
+\begin{verbatim}
+  <at:AtomType rdf:ID="C.sp3">
+    <at:categorizedAs rdf:resource="&cdkat;C.tetrahedral"/>
+    <at:hasElement rdf:resource="&elem;C"/>
+    <at:hybridization rdf:resource="&at;sp3"/>
+    <at:formalCharge>0</at:formalCharge>
+    <at:lonePairCount>0</at:lonePairCount>
+    <at:formalBondType rdf:resource="&bo;single"/>
+    <at:formalBondType rdf:resource="&bo;single"/>
+    <at:formalBondType rdf:resource="&bo;single"/>
+    <at:formalBondType rdf:resource="&bo;single"/>
+  </at:AtomType>
+\end{verbatim}
 

### 2016-07-16 article.tex
Started writing again; will try to send around a full draft around by the end of this week

@@
   and the lone pair information is needed for resonance structure calculation
-  needed, for example, for Gasteiger \pi-charges.
+  needed, for example, for Gasteiger $\pi$-charges.
 
@@
 
-  The structure diagram layout ... %FIXME: John, do you want this in?
+  The structure diagram layout has been rewritten and the new code solves a
+  number of long standing issues. Particularly, collission detection has been
+  greatly improved. Figure~\ref{fig:sdg} shows a ...
 
@@
 Another recent change is the choice of build system: we have moved away from
-Ant to Maven to deal with dependencies and compile the library.
+Ant to Maven to deal with dependencies and compile and test the library.
+To keep the original idea of CDK modules, this move required to split up the
+source code in multiple source folders. Fortunately, modern integrated
+development environments have no trouble handling this, removing the original
+argument to have everything in one source folder.
 
-% FIXME: write more about this, and provide info about Maven Central
+On the other hand, for many modules, the test code is now more closely closely
+linked to the code being tested: both reside in the same folder, though we
+adhere to the Maven custom to have src/main/java and a src/test/java folders.
+For a few modules, however, this solution introduces circular dependencies, in
+which case a separate Maven module is created for the tests.
+
+The Maven packages for the CDK are available from Maven Central, which makes it
+easy for other projects to use. The full library can be included in other
+software by depending on the cdk artifact (\url{http://mvnrepository.com/artifact/org.openscience.cdk/cdk})
+but dependencies can also be defined on individual CDK modules.
 
@@
 
-INVITE ARVID
+OSGi bundles are available for the CDK too, which are used by Bioclipse and
+KNIME.
+
+% ping Arvid
 
@@
 \end{verbatim}
+
+\subsection*{Figure 3 - The improved structure diagram generation has improved
+code to solve overlap. The original SDG code used general heuristics (left) and the
+OverlapResolver would fine tune the layout to ensure atoms would not be placed
+at the same location (middle). The new SDG algorithm is able to
+make more rigorous changes, making the final output must more pleasing (right).}
+
+\includegraphics[width=\textwidth]{sdg.png}
 

### 2016-07-16 article.tex
Writing more content

@@
 
-  ....
+  The SMILES parsing has been replaced code in the external Beam project~\cite{Beam}.
+  This BSD-licensed SMILES parser is a complete implementation of the SMILES
+  and OpenSMILES specifications by one of the authors, and is independent of
+  the CDK library. The SmilesParser API uses this library underneath, and the
+  Beam API is hidden by this class.
 
@@
 
-  ....
+  Aromaticity has seen many definitions in the past and for cheminformatics it
+  frequently is algorithmically defined. The outcome of an aromaticity calculation
+  depends on a number of atom type features and heuristics, which in literature
+  often are not well defined. Based on this information several different
+  algorithmic definitions of aromaticity can be defined. Older CDK versions had
+  various models implemented and the code was scattered throughout the library.
+  This was unified resulting in a three models, of which two are based on the
+  CDK atom typer, to provide the number of electrons donated by each atom to
+  the delocalized $\pi$ system. The difference between these two models is
+  how double bonds pointing outwards from a ring are contributing.
+
+  The current CDK version further generalizes the idea that aromaticity is a
+  model, and provides an API that allows you to select out of various options
+  providing greater interoperability with other toolkits. The new
+  Aromaticity class allows to build a custom model by selecting and combining
+  options. For example, to reproduce the functionality of the previous
+  CDKAromaticity class:
+
+\begin{verbatim}
+Aromaticity aromaticity = new Aromaticity(
+  ElectronDonation.cdk(), Cycles.cdkAromaticSet()
+);
+\end{verbatim}
+
+Here, the CDK model for counting donated electrons is used, as the ring systems
+the CDK previously included in aromaticty counting, which was limited in the 
+number of fused rings systems. However, an alternative aromaticity calculator
+that does consider all possible ring systems can now be easily created with:
+
+\begin{verbatim}
+Aromaticity aromaticity = new Aromaticity(
+  ElectronDonation.cdk(), Cycles.all()
+);
+\end{verbatim}
 
@@
 
-  Initially done with DocCheck, replaced by OpenJavaDocCheck ...
+  The quality of JavaDoc of the CDK was originally tested with DocCheck, and
+  later replaced by a custom written tool called OpenJavaDocCheck. With the move
+  to Maven (explained elsewhere) which does not have integration for this tool,
+  we adopted CheckStyle (\url{http://checkstyle.sourceforge.net/}). This tool
+  reports about missing documentation and on documentation which is not properly
+  annotated in the Java source files.
 
-  \subsubsection*{Unit testing}
+  \subsubsection*{Testing}
+
+  Years of development of the CDK library has resulted in a large suite of
+  tests of various kinds. This include unit tests, which test core APIs, and
+  functional testing, which test higher level functionality of the CDK. The
+  latter include tests if algorithm implementations calculate the expected
+  values, but also contain integrated tests, which involve more than one
+  algorithms, such as SMILES parsing.
 

### 2016-07-16 article.tex
More content, e.g. about the rendering

@@
 % FIXME: finish migrating the author list
+% FIXME: add Gilleain
 
@@
   The new CDK version introduces a new API for stereochemistry information: the interfaces
-  IStereoChemistryElement was introduced with the implementations TetrahedralChirality,
+  IStereoElement was introduced with the implementations TetrahedralChirality,
   for four coordinate compounds like bromo-chloro-fluoro-iodomethane,
@@
 
+  An atom container has a method to iterate over all stereo elements:
+
+\begin{verbatim}
+for (IStereoElement element : container.) {
+  if (element instanceof TetrahedralChirality) {
+    // ...
+  } else if (element instanceof DoubleBondStereochemistry) {
+    // ...
+  } else if (element instanceof ExtendedTetrahedral) {
+    // ...
+  }
+}
+\end{verbatim}
+
   \subsection*{Signatures}
 
-  An implementation has been provided of the Signature structure descriptor for molecules 
-  \cite{Faulon2003}. These act as a linear notation - like the SMILES format - 
+  An implementation has been provided of the Signature structure descriptor for
+  molecules~\cite{Faulon2003}. These act as a linear notation - like the SMILES format - 
   for the whole molecule as well as for connected substructures rooted at a single atom. The 
-  descriptor can also be canonicalized to provide isomorphism-independent representations 
-  \cite{Faulon2004}.
+  descriptor can also be canonicalized to provide isomorphism-independent
+  representations~\cite{Faulon2004}.
 
@@
   % DESCRIBE THE CORE APIs
+
+  Moreover, a simplified API has been introduced that addresses most of the
+  common rendering needs, with the DepictionGenerator class. To depict benzene
+  the following code can be used:
+
+\begin{verbatim}
+new DepictionGenerator()
+  .withSize(300, 300)
+  .depict(benzene)
+  .writeTo("BenzeneLocalized.png");
+\end{verbatim}
+
+  Many of the rendering options are available as parameters in the core API
+  and as methods on the DepictionGenerator class. This includes subscructure
+  coloring, exemplified with this example reaction (see Fig.~\ref{fig:depiction}):
+
+\begin{verbatim}
+TODO
+\end{verbatim}
 
@@
 
+However, perhaps the biggest factor in improved code quality is that we
+instantiated a peer review process where any functionality changing patch is
+required to be reviewed by one indepdent, senior CDK developer for  the development
+branch, and two reviewers for stable branches. This patch development system
+is supported by a number of automated validations steps as outlined below.
+
   \subsubsection*{Modularization}
@@
 
-The project continues to use PMD (http://pmd.sf.net/) for code quality checking,
+The project continues to use PMD (url{http://pmd.sf.net/}) for code quality checking,
 but deviates from the default rules. For example, we are more liberal with 
@@
 % FIXME: add appropriate references below
-\item \textbf{Other (optional) requirements}: JNI-InChI, Vecmath, BEAM, Guava, JGraphT, Signatures, CMLDOM, XOM, JavaCC
+\item \textbf{Other (optional) requirements}: JNI-InChI, Vecmath, Beam, Guava, JGraphT, Signatures, CMLDOM, XOM, JavaCC
 \item \textbf{License}: LGPL v2.1 or later
@@
 \section*{Figures}
-  \subsection*{Figure 1 - Dependencies between CDK modules.}
-  \label{fig:deps}
+  \subsection*{Figure 1 - Dependencies between CDK modules.}\label{fig:deps}
       Visualization of the dependencies between CDK modules. For example,
@@
 \subsection*{Figure 2 - Atom type information specified for a sp3-hybridized
-carbon.}
+carbon.}\label{fig:atomtype}
 \begin{verbatim}
@@
 at the same location (middle). The new SDG algorithm is able to
-make more rigorous changes, making the final output must more pleasing (right).}
+make more rigorous changes, making the final output must more pleasing
+(right).}\label{fig:sdg}
 
 \includegraphics[width=\textwidth]{sdg.png}
+
+\subsection*{Figure 4 - Integrated example showing the rendering and SMILES
+parsing functionality.}\label{fig:depiction}
+
+\includegraphics[width=\textwidth]{reaction.png}
 
@@
 
-  \subsection*{Table 1 - A selection of key CDK modules with major changes.}
-  \label{tab:modules}
+  \subsection*{Table 1 - A selection of key CDK modules with major changes.}\label{tab:modules}
   An overview of a selection of often used CDK modules with description,
@@
   libiocml                   &              & & XOM 1.2.5, CMLXOM 3.1~\cite{Murray-Rust2011} \\ \hline
-  smiles                     &              & SMILES support performance and converage is greatly improved. & Beam 0.9.1 \\ \hline %FIXME add a citation
-  smarts                     &              & & Beam 0.9.1 \\ \hline %FIXME add a citation
+  smiles                     &              & SMILES support performance and converage is greatly improved. & Beam 0.9.1~\cite{beam} \\
+  smarts                     &              & & Beam 0.9.1~\cite{beam} \\
   formula                    &              & New. & \\ \hline
   fingerprint                &              & Many new fingerprint types (see text). & Apache Commons Math 3.1.1 \\ \hline
-  qsar                       &              & & XOM 1.2.5, JAMA 1.0.3~\cite{Hicklin2012} \\ \hline % FIXME: find JAMA reference
+  qsar                       &              & & XOM 1.2.5, JAMA 1.0.3~\cite{Hicklin2012} \\ \hline
   signatures                 &              & & Signatures 1.1 \\ \hline % FIXME: ask Gilleain for the reference

### 2016-07-16 article.tex
Cleaned up the chem formula section

@@
 % FIXME: add Gilleain
+% FIXME: invite Tomas Pluskal
 
@@
   number of long standing issues. Particularly, collission detection has been
-  greatly improved. Figure~\ref{fig:sdg} shows a ...
+  greatly improved. Figure~\ref{fig:sdg} shows a difference in output between
+  the old code base, with and without overlap resolving, and with the new code.
+  The new code can be used like:
+
+\begin{verbatim}
+TODO
+\end{verbatim}
 
@@
 The chemical formula is the basic/simple chemical representation of a compound.
-It defines the number of isotopes or elements that constitute a the composition
-of a compound without describing how atoms are bonded. With the rise of metabolomics,
-it has become increasingly relevant to have full support in cheminformatics
-libraries~\cite{RojasCherto2011}.
-It is also the first step
-toward the identification of the chemical structure from a unidentified compound.
+It defines the number of isotopes or elements that constitute the composition
+of a compound without describing how atoms are bonded. With the rise of
+metabolomics it has become increasingly relevant to have full support for these
+in cheminformatics libraries~\cite{RojasCherto2011}.
 
-CDK can handle several concepts related to chemical formulas
-\begin{itemize}
-\item Chemical formula =
-\item Set of chemical formulas =
-\item Range of chemical formulas =
-\item Adducts =
-\item Isotope Container =
-\item Isotopic pattern = 
-\item Rules = Filters in accepting an chemical formula
-\end{itemize}
+The CDK interfaces can handle several concepts related to chemical formulas:
+the formula itself, sets of formulas, chemical formula ranges, adducts,
+isotope containers and patterns, and rules to filter formula sets. These new
+tools can be used for a number of tasks, including calculate the isotopic
+pattern from a given chemical formula, determining the possible elemental
+compositions for a given mass, and calculate the exact mass from a given
+chemical formula.
 
-CDK can
-\begin{itemize}
-\item generate/print the elemental composition from a given IAtomContainer
-\item calculate the isotopic pattern from a given chemical formula
-\item determine the elemental composition from a given mass
-\item validate a chemical formula
-\item calculate the exact mass from a given chemical formula
-\end{itemize}
+The following code calculates all possible chemical formula for a given
+accurate mass, within some allowed counts for each element:
 
-Where can it be applied?
+\begin{verbatim}
+Isotopes ifac = Isotopes.getInstance();
+MolecularFormulaRange range =
+  new MolecularFormulaRange();
+range.addIsotope( ifac.getMajorIsotope("C"), 8, 20);
+range.addIsotope( ifac.getMajorIsotope("H"), 0, 20);
+range.addIsotope( ifac.getMajorIsotope("O"), 0, 1);
+range.addIsotope( ifac.getMajorIsotope("N"), 0, 1);
+
+MolecularFormulaGenerator tool =
+  new MolecularFormulaGenerator(
+    SilentChemObjectBuilder.getInstance(),
+    133.0, 133.1, range
+  );
+IMolecularFormulaSet mfSet = tool.getAllFormulas();
+for (mf in mfSet) {
+  println MolecularFormulaManipulator.getString(mf) + " " +
+    MolecularFormulaManipulator.getTotalExactMass(mf)
+}
+\end{verbatim}
+
+This gives the following output:
+
+\begin{verbatim}
+C11H 133.007825032
+C9H11N 133.089149352
+C9H9O 133.065339908
+C8H7NO 133.052763844
+\end{verbatim}
 
@@
   libiocml                   &              & & XOM 1.2.5, CMLXOM 3.1~\cite{Murray-Rust2011} \\ \hline
-  smiles                     &              & SMILES support performance and converage is greatly improved. & Beam 0.9.1~\cite{beam} \\
-  smarts                     &              & & Beam 0.9.1~\cite{beam} \\
+  smiles                     &              & SMILES support performance and converage is greatly improved. & Beam 0.9.1~\cite{Beam} \\
+  smarts                     &              & & Beam 0.9.1~\cite{Beam} \\
   formula                    &              & New. & \\ \hline

### 2016-07-23 article.tex
Some spell checking

@@
 We here report improvements to the CDK since the 1.2 release series made
-to accomodate both the increased complexity of the library, as well as
-siginificant improvements of and additions to the functionality of the
+to accommodate both the increased complexity of the library, as well as
+significant improvements of and additions to the functionality of the
 library. Second, we outline how the CDK evolved with respect to
@@
 With this paper we have shown the continued effort to provide a free, Open Source
-cheminformatics library, and show that such colleborative projects can exist over
+cheminformatics library, and show that such collaborative projects can exist over
 a long period. We have taken advantage from the community support, and show
@@
 
-Open Source cheminformatics has made significants steps forward recently~\cite{OBoyle2011}.
+Open Source cheminformatics has made significant steps forward recently~\cite{OBoyle2011}.
 The Chemistry Development Kit (CDK) is one of the tools under the Blue Obelisk wings,
@@
   Many of the rendering options are available as parameters in the core API
-  and as methods on the DepictionGenerator class. This includes subscructure
+  and as methods on the DepictionGenerator class. This includes substructure
   coloring, exemplified with this example reaction (see Fig.~\ref{fig:depiction}):
@@
   The structure diagram layout has been rewritten and the new code solves a
-  number of long standing issues. Particularly, collission detection has been
+  number of long standing issues. Particularly, collision detection has been
   greatly improved. Figure~\ref{fig:sdg} shows a difference in output between
@@
 Here, the CDK model for counting donated electrons is used, as the ring systems
-the CDK previously included in aromaticty counting, which was limited in the 
+the CDK previously included in aromaticity counting, which was limited in the 
 number of fused rings systems. However, an alternative aromaticity calculator
@@
 instantiated a peer review process where any functionality changing patch is
-required to be reviewed by one indepdent, senior CDK developer for  the development
+required to be reviewed by one independent, senior CDK developer for  the development
 branch, and two reviewers for stable branches. This patch development system
@@
   dependencies on third-party libraries, and the major changes since
-  version~1.2. Depedencies between modules are depicted in Figure~\ref{fig:deps}.
+  version~1.2. Dependencies between modules are depicted in Figure~\ref{fig:deps}.
   \vskip 1\baselineskip
@@
     \begin{tabular}{lp{3cm}p{3cm}l}
-  \textbf{Module}            & \textbf{Description}  & \textbf{Major Changes} & \textbf{Depedencies} \\ \hline
+  \textbf{Module}            & \textbf{Description}  & \textbf{Major Changes} & \textbf{Dependencies} \\ \hline
   interfaces                 &              & & Vecmath 1.5.2 \\ \hline

### 2016-07-23 article.tex
Added code examples for SMILES

@@
 
-  \subsubsection*{SMILES parser}
+  \subsubsection*{SMILES parser and generator}
 
@@
   This BSD-licensed SMILES parser is a complete implementation of the SMILES
-  and OpenSMILES specifications by one of the authors, and is independent of
+  and OpenSMILES (\url{http://opensmiles.org/}) specifications by one of the
+  authors (including stereochemistry), and is independent of
   the CDK library. The SmilesParser API uses this library underneath, and the
-  Beam API is hidden by this class.
+  Beam API is hidden by this class. The most significant change here is that
+  the SMILES parser now automatically locates the positions of double bonds,
+  but this can be turned off:
+
+\begin{verbatim}
+  SmilesParser parser = new SmilesParser(
+    SilentChemObjectBuilder.getInstance()
+  );
+  parser.kekulise(false);
+  mol = parser.parseSmiles(smi);
+\end{verbatim}
+
+  The SMILES generation API has also been simplified and made more flexible.
+  Creating unique SMILES or aromatic SMILES is now done by first creating a
+  SMILES generator instance which can make generic or unique SMILES, and with
+  or without simplifying aromatic rings using the lower case element symbols
+  for the organic subset. This generator can then be repeatedly called:
+
+\begin{verbatim}
+  uniqueGenerator = SmilesGenerator.unique()
+  aromaticGenerator = SmilesGenerator.generic().aromatic()
+  smiles = uniqueGenerator.createSMILES(mol)
+  smiles2 = atomaticGenerator.createSMILES(mol)
+\end{verbatim}
 

### 2016-07-23 article.tex
Added some words on ring detection

@@
 
-  \subsubsection*{SMARTS parser}
-
-  ....
+  % \subsubsection*{SMARTS parser}
 
@@
 
-  .... \cite{May2014}
+  Ring finding is a key functionality in a cheminformatics library, and the CDK
+  knows a long history of ring finding~\cite{May2014}. Particularly, non-redundant
+  ring sets have seen particular interest, such as the smallest set of smallest
+  rings, for which the CDK implements two classical algorithms~\cite{Figueras1996,Berger2004}.
+  Recent work by one the authors has implemented a new, faster algorithm, allowing
+  searching for various types of (non-redundant) ring sets~\cite{May2014}. These
+  are available via the new Cycles API:
+
+\begin{verbatim}
+  allCycles = Cycles.all(mol)
+  relevantCycles = Cycles.relevant(mol)
+  essentialCycles = Cycles.essential(mol)
+  sssrCycles = Cycles.sssr(mol)
+\end{verbatim}
 

### 2016-07-23 article.tex
Removed some template code
Added missing periods

@@
   CircularFingerprinter~\cite{rogers2010extended, Clark2014}      & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $1\,024$ / $2^{32}$%
-\footnote[1]{For the CirkularFingerprinter the bit version is folded to 1024 whereas the count version is unfolded} \\
+\footnote[1]{For the CirkularFingerprinter the bit version is folded to 1024 whereas the count version is unfolded.} \\
   EStateFingerprinter~\cite{Hall1995}       & $\checkmark$ &               & 1.2.0       & $79$            \\
@@
   LingoFingerprinter~\cite{vidal2005lingo}         & $\checkmark$ &               & 1.6.0 ?     & NA%
-\footnote[2]{The Lingofingerprinter does not have a default size}               
+\footnote[2]{The Lingofingerprinter does not have a default size.}
                                                                                              \\
@@
 
-%    Here is an example of a \emph{small} table in \LaTeX\ using  
-%    \verb|\tabular{...}|. This is where the description of the table 
-%    should go. \par \mbox{}
-%    \par
-%    \mbox{
-%      \begin{tabular}{|c|c|c|}
-%        \hline \multicolumn{3}{|c|}{My Table}\\ \hline
-%        A1 & B2  & C3 \\ \hline
-%        A2 & ... & .. \\ \hline
-%        A3 & ..  & .  \\ \hline
-%      \end{tabular}
-%      }
-%  \subsection*{Table 2 - Sample table title}
-%    Large tables are attached as separate files but should
-%    still be described here.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Additional Files              %%
-
-%\section*{Additional Files}
-%  \subsection*{Additional file 1 --- Sample additional file title}
-%    Additional file descriptions text (including details of how to
-%    view the file, if it is in a non-standard format or the file extension).  This might
-%    refer to a multi-page table or a figure.
-%
-%  \subsection*{Additional file 2 --- Sample additional file title}
-%    Additional file descriptions text.
-
 \end{backmatter}

### 2016-07-23 article.tex
Email addresses

@@
 % FIXME: finish migrating the author list
-% FIXME: add Gilleain
-% FIXME: invite Tomas Pluskal
 
@@
    addressref={aff2},
-   email={???}
+   email={john@nextmovesoftware.com}
 ]{\inits{JWM}\fnm{John W} \snm{May}}
@@
    addressref={uppsala},
-   email={}
+   email={jonathan.alvarsson@farmbio.uu.se}
 ]{\inits{JA}\fnm{Jonathan} \snm{Alvarsson}}
@@
    addressref={uppsala},
-   email={}
+   email={berg.arvid@gmail.com}
 ]{\inits{AB}\fnm{Arvid} \snm{Berg}}
@@
    addressref={idea},
-   email={}
+   email={jeliazkova.nina@gmail.com}
 ]{\inits{NJ}\fnm{Nina} \snm{Jeliazkova}}
@@
    addressref={???},
-   email={}
-]{\inits{MRJ}\fnm{Miguel} \snm{Rojas-Cherto}}
+   email={miguelrojasch@googlemail.com}
+]{\inits{MRJ}\fnm{Miguel}~\snm{Rojas-Cherto}}
+\author[
+   addressref={???},
+   email={pluskal@wi.mit.edu}
+]{\inits{TP}\fnm{Tomáš}~\snm{Pluskal}}
 \author[
    addressref={uppsala},
-   email={}
+   email={ola.spjuth@farmbio.uu.se}
 ]{\inits{OS}\fnm{Ola} \snm{Spjuth}}
 \author[
+   addressref={???},
+   email={gilleain.torrance@gmail.com}
+]{\inits{GT}\fnm{Gilleain} \snm{Torrance}}
+\author[
    addressref={nih},
-   email={}
+   email={guhar@mail.nih.gov}
 ]{\inits{RG}\fnm{Rajarshi} \snm{Guha}}
@@
    addressref={ebi},
-   email={}
+   email={steinbeck@ebi.ac.uk}
 ]{\inits{CS}\fnm{Christoph} \snm{Steinbeck}}

### 2016-07-23 article.tex
Meta sections

@@
 \section*{Competing interests}
-XX and YY have companies that sell solutions based on the CDK.
+JWM and NJ work for a company that sell solutions based on the CDK.
 
@@
 \section*{Authors contributions}
-    Text for this section \ldots
+All authors wrote and contributed source code or documentation to the CDK
+library. Some authors have peer-reviewed source code for the library.
+ELW, JWM, RG, and CS are project leaders. All authors have contributed to the
+content of this paper and approved the final version.
 

### 2016-07-23 article.tex
Some last cleanup for today: figure numbering (latex approach doesn't work), some reshuffling)

@@
    addressref={???},
+   email={pluskal@wi.mit.edu}
+]{\inits{TP}\fnm{Tomáš}~\snm{Pluskal}}
+\author[
+   addressref={???},
    email={miguelrojasch@googlemail.com}
 ]{\inits{MRJ}\fnm{Miguel}~\snm{Rojas-Cherto}}
-\author[
-   addressref={???},
-   email={pluskal@wi.mit.edu}
-]{\inits{TP}\fnm{Tomáš}~\snm{Pluskal}}
 \author[
@@
    email={steinbeck@ebi.ac.uk}
-]{\inits{CS}\fnm{Christoph} \snm{Steinbeck}}
+]{\inits{CS}\fnm{Christoph}~\snm{Steinbeck}}
 
@@
   the element symbol, hybridization, formal charge, number of lone pairs, and
-  an enumeration of the bond orders. This list of properties captures the
+  an enumeration of the bond orders (see Figure~1). This list of properties captures the
   information needed for the various algorithms in the CDK. For example,
@@
   and as methods on the DepictionGenerator class. This includes substructure
-  coloring, exemplified with this example reaction (see Fig.~\ref{fig:depiction}):
+  coloring, exemplified with this example reaction (see Fig.~2\ref{fig:depiction}):
 
@@
   number of long standing issues. Particularly, collision detection has been
-  greatly improved. Figure~\ref{fig:sdg} shows a difference in output between
+  greatly improved. Figure~3\ref{fig:sdg} shows a difference in output between
   the old code base, with and without overlap resolving, and with the new code.
@@
 supported in CDK as well. The fingerprints currently existing in CDK are listed
-in Table~\ref{tab:fingerprints}.
+in Table~1\ref{tab:fingerprints}.
 
@@
 
-... describe all modules, maybe add graphviz plot ...
-
 An overview of key modules with description, important changes, and dependencies
-on third-party libraries is given in Table~\ref{tab:modules} and the dependencies
-between the CDK modules are depicted in Figure~\ref{fig:deps}.
+on third-party libraries is given in Table~2\ref{tab:modules} and the dependencies
+between the CDK modules are depicted in Figure~4\ref{fig:deps}.
 
@@
 
-%% Do not use \listoffigures as most will included as separate files
+\section*{Figures}
 
-\section*{Figures}
-  \subsection*{Figure 1 - Dependencies between CDK modules.}\label{fig:deps}
-      Visualization of the dependencies between CDK modules. For example,
-      the cdk-core depends on the cdk-interfaces module. A few higher level
-      modules have been left out: cdk-builder3dtools, cdk-legacy, and
-      cdk-depict.
-
-\includegraphics[width=\textwidth]{cdkDeps.png}
-
-\subsection*{Figure 2 - Atom type information specified for a sp3-hybridized
+\subsection*{Figure 1 - Atom type information specified for a sp3-hybridized
 carbon.}\label{fig:atomtype}
@@
 
+\subsection*{Figure 2 - Integrated example showing the rendering and SMILES
+parsing functionality.}\label{fig:depiction}
+
+\includegraphics[width=\textwidth]{reaction.png}
+
 \subsection*{Figure 3 - The improved structure diagram generation has improved
@@
 
-\subsection*{Figure 4 - Integrated example showing the rendering and SMILES
-parsing functionality.}\label{fig:depiction}
+  \subsection*{Figure 4 - Dependencies between CDK modules.}\label{fig:deps}
+      Visualization of the dependencies between CDK modules. For example,
+      the cdk-core depends on the cdk-interfaces module. A few higher level
+      modules have been left out: cdk-builder3dtools, cdk-legacy, and
+      cdk-depict.
 
-\includegraphics[width=\textwidth]{reaction.png}
+\includegraphics[width=\textwidth]{cdkDeps.png}
 
@@
 \section*{Tables}
-
-  \subsection*{Table 1 - A selection of key CDK modules with major changes.}\label{tab:modules}
-  An overview of a selection of often used CDK modules with description,
-  dependencies on third-party libraries, and the major changes since
-  version~1.2. Dependencies between modules are depicted in Figure~\ref{fig:deps}.
-  \vskip 1\baselineskip
-
-    \begin{minipage}{1\textwidth}
-    \renewcommand*{\thempfootnote}{\fnsymbol{mpfootnote}}
-    \centering
-    \begin{tabular}{lp{3cm}p{3cm}l}
-  \textbf{Module}            & \textbf{Description}  & \textbf{Major Changes} & \textbf{Dependencies} \\ \hline
-  interfaces                 &              & & Vecmath 1.5.2 \\ \hline
-  core                       &              & & Google Guava 17.0 \\ \hline % FIXME: how to add https://github.com/google/guava ?
-  standard                   &              & & \\ \hline
-  render                     &              & Redesigned to make it more modular and support multiple widget toolkits, like AWT and SWT. & \\ \hline
-  isomorphism                &              & & \\ \hline
-  atomtype                   &              & Unified approach where atom typing is separated from other algorithms. & \\ \hline
-  valencycheck               &              & & \\ \hline
-  ioformats                  &              & & \\ \hline
-  io                         &              & The MDL molfile reader has been rewritten and supports in atom types defind in the specification. & XPP3 1.1.4c \\ \hline
-  iordf                      &              & New. & Jena 2.7.4 \\ \hline
-  inchi                      &              & & JNI-InChI 0.8~\cite{Spjuth2013}  \\ \hline
-  libiocml                   &              & & XOM 1.2.5, CMLXOM 3.1~\cite{Murray-Rust2011} \\ \hline
-  smiles                     &              & SMILES support performance and converage is greatly improved. & Beam 0.9.1~\cite{Beam} \\
-  smarts                     &              & & Beam 0.9.1~\cite{Beam} \\
-  formula                    &              & New. & \\ \hline
-  fingerprint                &              & Many new fingerprint types (see text). & Apache Commons Math 3.1.1 \\ \hline
-  qsar                       &              & & XOM 1.2.5, JAMA 1.0.3~\cite{Hicklin2012} \\ \hline
-  signatures                 &              & & Signatures 1.1 \\ \hline % FIXME: ask Gilleain for the reference
-  qsarmolecular              &              & & \\ \hline
-  log4j                      &              & & Log4j 1.2.17 \\ \hline
-    \end{tabular}
-    \end{minipage}
 
@@
 
+      \subsection*{Table 2 - A selection of key CDK modules with major changes.}\label{tab:modules}
+  An overview of a selection of often used CDK modules with description,
+  dependencies on third-party libraries, and the major changes since
+  version~1.2. Dependencies between modules are depicted in Figure~4\ref{fig:deps}.
+  \vskip 1\baselineskip
+
+    \begin{minipage}{1\textwidth}
+    \renewcommand*{\thempfootnote}{\fnsymbol{mpfootnote}}
+    \centering
+    \begin{tabular}{lp{3cm}p{3cm}l}
+  \textbf{Module}            & \textbf{Description}  & \textbf{Major Changes} & \textbf{Dependencies} \\ \hline
+  interfaces                 &              & & Vecmath 1.5.2 \\ \hline
+  core                       &              & & Google Guava 17.0 \\ \hline % FIXME: how to add https://github.com/google/guava ?
+  standard                   &              & & \\ \hline
+  render                     &              & Redesigned to make it more modular and support multiple widget toolkits, like AWT and SWT. & \\ \hline
+  isomorphism                &              & & \\ \hline
+  atomtype                   &              & Unified approach where atom typing is separated from other algorithms. & \\ \hline
+  valencycheck               &              & & \\ \hline
+  ioformats                  &              & & \\ \hline
+  io                         &              & The MDL molfile reader has been rewritten and supports in atom types defind in the specification. & XPP3 1.1.4c \\ \hline
+  iordf                      &              & New. & Jena 2.7.4 \\ \hline
+  inchi                      &              & & JNI-InChI 0.8~\cite{Spjuth2013}  \\ \hline
+  libiocml                   &              & & XOM 1.2.5, CMLXOM 3.1~\cite{Murray-Rust2011} \\ \hline
+  smiles                     &              & SMILES support performance and converage is greatly improved. & Beam 0.9.1~\cite{Beam} \\
+  smarts                     &              & & Beam 0.9.1~\cite{Beam} \\
+  formula                    &              & New. & \\ \hline
+  fingerprint                &              & Many new fingerprint types (see text). & Apache Commons Math 3.1.1 \\ \hline
+  qsar                       &              & & XOM 1.2.5, JAMA 1.0.3~\cite{Hicklin2012} \\ \hline
+  signatures                 &              & & Signatures 1.1 \\ \hline % FIXME: ask Gilleain for the reference
+  qsarmolecular              &              & & \\ \hline
+  log4j                      &              & & Log4j 1.2.17 \\ \hline
+    \end{tabular}
+    \end{minipage}
+
 \end{backmatter}

### 2016-07-24 article.tex
Added address

@@
 \author[
-   addressref={???},
+   addressref={wi_mit},
    email={pluskal@wi.mit.edu}
@@
 }
-
+\address[id=wi_mit]{
+  \orgname{Whitehead Institute for Biomedical Research}, 
+  \street{9 Cambridge Center},
+  \postcode{MA 02142},
+  \city{Cambridge},
+  \cny{USA}
+}
 %\begin{artnotes}
@@
 
-The following code calculates all possible chemical formula for a given
-accurate mass, within some allowed counts for each element:
+The MolecularFormulaGenerator class was originally developed and evaluated 
+as part of the MZmine~2 framework~\cite{Pluskal2012}. The following code
+calculates all possible chemical formula for a given accurate mass, 
+within allowed counts for each element:
 

### 2016-08-07 article.tex
Minor tweak competing interests section to reflect that Nina and John don't work for the same company.
Flesh out stereochemistry section

@@
 
-  The new CDK version introduces a new API for stereochemistry information: the interfaces
-  IStereoElement was introduced with the implementations TetrahedralChirality,
-  for four coordinate compounds like bromo-chloro-fluoro-iodomethane,
-  DoubleBondStereochemistry, for compounds like trans-2-butene, and
-  ExtendedTetrahedral, for compounds like allene.
+  In previous versions of the API, stereochemistry was represented disparately hindering 
+  interconversion between and within file formats. In CDK 1.5.x the core representation 
+  has been standardized upon and procedures updated or added to enable duplicate checking,
+  pattern matching, and interconversion.
 
-  An atom container has a method to iterate over all stereo elements:
+  The preferred representation of stereochemistry is now for it to be stored at the molecule
+  level as a \texttt{\textbf{StereoElement}}. In abstract terms a stereo element describes local
+  geometry using a type, focus, carriers, and configuration (Figure~2\ref{fig:stereodatastructure}).
+  Currently the most common types of stereochemistry are supported: Tetrahedral, Cis-trans isomerism 
+  around a double bond, and Extended Tetrahedral. Rarer types of stereochemistry, such as: Square 
+  Planar, Trigonal Bipyramidal, Octahedral, could easily be incorporated into the chosen description 
+  given sufficient demand from the community.
 
-\begin{verbatim}
-for (IStereoElement element : container.) {
-  if (element instanceof TetrahedralChirality) {
-    // ...
-  } else if (element instanceof DoubleBondStereochemistry) {
-    // ...
-  } else if (element instanceof ExtendedTetrahedral) {
-    // ...
-  }
-}
-\end{verbatim}
+  Along with the new stereochemistry representation, algorithms were required in several areas. Generally
+  a user does not need to invoked these procedures explicitly as they are called as needed within existing
+  APIs:
+
+  \begin{itemize}
+   \item Perception from 2D coordinates
+   \item Perception from 3D coordinates
+   \item Wedge assignment
+   \item Graph (sub)isomorphism matching
+   \item SMARTS matching
+   \item Canonicalization
+  \end{itemize}
+
+  The perception from coordinates and wedge assignment algorithms are fundamental for conversion
+  between formats that store stereochemistry implicitly based on coordinates (e.g. Molfile, CML) and 
+  explicitly (e.g. SMILES, CML, InChI). Perception from 2D coordinates can optionally identify
+  perspective projects, specifically: Fischer, Haworth, and Chair projections. With the perception of 
+  perspective projections enabled, database entries currently consider distinct can be merged
+  (Figure~3\ref{fig:stereoprojections}). The perception is based on an algorithm briefly 
+  described by \cite{Karapetyan2015}.
+
+  % ChEMBL 21 ~2281 structures (~11454 TH center) drawn in perspective
+
+  Graph matching of stereochemistry with the described representation is straight forward. Given
+  the atom-atom mapping from a query structure to a target molecule, the focus and carriers of 
+  the query stereochemistry are mapped to the target. Using the permutation parity of this mapping
+  the configurations compared. SMARTS matching requires some special handling for complex cases 
+  \cite{May2014_SMARTS}. For canonicalization, a partial canonical ordering is used to assign an 
+  absolute label which can then be integrated into the ordering. More detailed descriptions of 
+  these algorithms are described in detail in Chapter 6 of \cite{May2015}.
 
@@
   number of long standing issues. Particularly, collision detection has been
-  greatly improved. Figure~3\ref{fig:sdg} shows a difference in output between
+  greatly improved. Figure~5\ref{fig:sdg} shows a difference in output between
   the old code base, with and without overlap resolving, and with the new code.
@@
 on third-party libraries is given in Table~2\ref{tab:modules} and the dependencies
-between the CDK modules are depicted in Figure~4\ref{fig:deps}.
+between the CDK modules are depicted in Figure~6\ref{fig:deps}.
 
@@
 \section*{Competing interests}
-JWM and NJ work for a company that sell solutions based on the CDK.
+JWM and NJ work for companies that sell solutions based on the CDK.
 
@@
 
-\subsection*{Figure 2 - Integrated example showing the rendering and SMILES
+\subsection*{Figure 2 - Relative storage of stereochemistry, the \textbf{type} and \textbf{focus} of stereochemsitry are fixed for a given stereocenter description but the \textbf{carriers} and \textbf{configuration} are relative.
+The multiple rows for each stereochemistry type are different internal representation that would be considered equivalent.
+In the tetrahedral types, hydrogens may be suppressed in a molecular graph so the \textbf{focus} is reused in the \textbf{carriers} list as a placeholder.}\label{fig:stereodatastructure}
+
+\includegraphics{img/stereodesc_annotated.pdf}
+
+\subsection*{Figure 3 - The raw input files of CHEMBL23970 and CHEMBL444314 are displayed (ChEMBL 21).
+Without perceiving the stereochemistry indicated by Haworth projection in CHEMBL23970, the database entries are inncorectly considered distinct.
+Down stream aggregation databases mirror this separation (PubChem CID 5280, CID 65119).}\label{fig:stereoprojections}
+
+\includegraphics{img/projections_annotated.pdf}
+
+\subsection*{Figure 4 - Integrated example showing the rendering and SMILES
 parsing functionality.}\label{fig:depiction}
@@
 
-\subsection*{Figure 3 - The improved structure diagram generation has improved
+\subsection*{Figure 5 - The improved structure diagram generation has improved
 code to solve overlap. The original SDG code used general heuristics (left) and the
@@
 
-  \subsection*{Figure 4 - Dependencies between CDK modules.}\label{fig:deps}
+  \subsection*{Figure 6 - Dependencies between CDK modules.}\label{fig:deps}
       Visualization of the dependencies between CDK modules. For example,

### 2016-08-08 article.tex

--- 2016-07-24 article.tex
@@
 
-  \subsection*{Signatures}
+  \subsubsection*{Signatures}
 
@@
   descriptor can also be canonicalized to provide isomorphism-independent
-  representations~\cite{Faulon2004}.
+  representations~\cite{Faulon2004}, and has been successfully used in QSAR modeling~\cite{signaturefingerprints,Spjuth2011DS,Moghadam2015,Alvarsson2014,Spjuth2012OS,Norinder2013}.
 
@@
 
-OSGi bundles are available for the CDK too, which are used by Bioclipse and
-KNIME.
-
-% ping Arvid
+OSGi bundles are available for the CDK too, which are used by e.g.
+Bioclipse~\cite{spjuth2007bioclipse,spjuth2009bioclipse} and KNIME~\cite{Beisken2013}.
+Due to split packages between CDK modules, CDK is bundled as a single OSGi jar, which is available from \url{http://pele.farmbio.uu.se/bioclipse/cdk/cdk-1.5.13/}.
+This constitutes an area where improvements can be done on modularization.
 
@@
 \section*{Acknowledgements}
-The authors acknoledge the great number of people who have contributed smaller
+The authors acknowledge the great number of people who have contributed smaller
 and larger contributions to the CDK library. A full list of contributors is
-found in the AUTHORS file~\cite{AUTHORS}.
+found in the AUTHORS file~\cite{AUTHORS}. OS acknowledges support from the Swedish strategic research programs eSSENCE and Swedish e-Science Research Center (SeRC).
 

### 2016-08-09 pubcounts.R
Added data, code and figure for growth in CDK paper citations

@@ -1,0 +1,8 @@
+library(ggplot2)
+library(reshape2)
+x <- read.csv('pubcounts.csv', header=TRUE)
+ggplot(x, aes(x=as.factor(year),y=cdk2))+
+  geom_bar(stat='identity', fill='beige', colour='black', position='dodge')+
+  geom_text(aes(label=cdk2,x=as.factor(year),y=cdk2),
+            position = position_dodge(width=1))+
+  xlab("Year")+ylab("Number of citations")

### 2016-08-11 article.tex
Fix NIH affiliation
Added LICSS (closes #17)

@@
 \address[id=nih]{
-  \orgname{NIH Center for Translational Therapeutics},
+  \orgname{National Center for Advancing Translational Science},
   \street{9800 Medical Center Drive},
@@
 and previously documented version have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
-For example, new tools expose CDK functionality, such as Cinfony~\cite{OBoyle2008},
-rcdk~\cite{Guha2007}, and workflow plugins for Taverna~\cite{Truszkowski2011} and KNIME~\cite{Beisken2013}.
+For example, new tools expose CDK functionality in larger platforms such as Cinfony~\cite{OBoyle2008},
+rcdk~\cite{Guha2007}, and plugins for Taverna~\cite{Truszkowski2011},
+KNIME~\cite{Beisken2013}, and LICSS for Excel~\cite{Lawson2012}.
 Furthermore, projects with specific functionality have emerged building on the CDK, including

### 2016-08-11 article.tex
Added Chris Evelo, group leader of Egon
Added ChemViz2 (closes #20)

@@
 ]{\inits{GT}\fnm{Gilleain} \snm{Torrance}}
+\author[
+   addressref={um},
+   email={chris.evelo@maastrichtuniversity.nl}
+]{\inits{CTE}\fnm{Chris T} \snm{Evelo}}
 \author[
@@
 rcdk~\cite{Guha2007}, and plugins for Taverna~\cite{Truszkowski2011},
-KNIME~\cite{Beisken2013}, and LICSS for Excel~\cite{Lawson2012}.
+KNIME~\cite{Beisken2013}, ChemViz2 for Cytoscape~\cite{ChemViz2}, and LICSS for
+Microsoft Excel~\cite{Lawson2012}.
 Furthermore, projects with specific functionality have emerged building on the CDK, including

### 2016-08-11 article.tex
Added ReactPRED (closes #16)

@@
 Padel~\cite{yap2011padel}, AMBIT-SMARTS~\cite{jeliazkova2011ambitsmarts}, AMBIT-Tautomer~\cite{kochev2013ambit},
-and SMSD~\cite{Rahman2009,Rahman2014}. Other tools used previous versions already, and follow
+ReactPRED~\cite{ReactPRED}, and SMSD~\cite{Rahman2009,Rahman2014}. Other tools used previous versions already, and follow
 newer releases of the CDK, such as Bioclipse~\cite{spjuth2007bioclipse,

### 2016-08-11 article.tex
Reformatted
Updated Table 2: added missing descriptions and the sdg module (closes #18)

@@
   \begin{itemize}
-   \item Perception from 2D coordinates
-   \item Perception from 3D coordinates
-   \item Wedge assignment
-   \item Graph (sub)isomorphism matching
-   \item SMARTS matching
-   \item Canonicalization
+   \item perception from 2D coordinates;
+   \item perception from 3D coordinates;
+   \item wedge assignment;
+   \item graph (sub)isomorphism matching;
+   \item SMARTS matching; and
+   \item canonicalization.
   \end{itemize}
@@
   \textbf{Module}            & \textbf{Description}  & \textbf{Major Changes} & \textbf{Dependencies} \\ \hline
-  interfaces                 &              & & Vecmath 1.5.2 \\ \hline
-  core                       &              & & Google Guava 17.0 \\ \hline % FIXME: how to add https://github.com/google/guava ?
-  standard                   &              & & \\ \hline
-  render                     &              & Redesigned to make it more modular and support multiple widget toolkits, like AWT and SWT. & \\ \hline
-  isomorphism                &              & & \\ \hline
-  atomtype                   &              & Unified approach where atom typing is separated from other algorithms. & \\ \hline
-  valencycheck               &              & & \\ \hline
-  ioformats                  &              & & \\ \hline
-  io                         &              & The MDL molfile reader has been rewritten and supports in atom types defind in the specification. & XPP3 1.1.4c \\ \hline
-  iordf                      &              & New. & Jena 2.7.4 \\ \hline
-  inchi                      &              & & JNI-InChI 0.8~\cite{Spjuth2013}  \\ \hline
-  libiocml                   &              & & XOM 1.2.5, CMLXOM 3.1~\cite{Murray-Rust2011} \\ \hline
-  smiles                     &              & SMILES support performance and converage is greatly improved. & Beam 0.9.1~\cite{Beam} \\
-  smarts                     &              & & Beam 0.9.1~\cite{Beam} \\
-  formula                    &              & New. & \\ \hline
-  fingerprint                &              & Many new fingerprint types (see text). & Apache Commons Math 3.1.1 \\ \hline
-  qsar                       &              & & XOM 1.2.5, JAMA 1.0.3~\cite{Hicklin2012} \\ \hline
-  signatures                 &              & & Signatures 1.1 \\ \hline % FIXME: ask Gilleain for the reference
-  qsarmolecular              &              & & \\ \hline
-  log4j                      &              & & Log4j 1.2.17 \\ \hline
+  interfaces                 & Interfaces for the data models. & & Vecmath 1.5.2 \\ \hline
+  core                       & Core functionality.             & & Google Guava 17.0 \\ \hline % FIXME: how to add https://github.com/google/guava ?
+  standard                   & Common functionality.           & & \\ \hline
+  render                     & Graphical rendering.            & Redesigned to make it more modular and support multiple widget toolkits, like AWT and SWT. & \\ \hline
+  isomorphism                & Isomorphism and substructure searching. & & \\ \hline
+  atomtype                   & Various non-core atom type schemes.     & Unified approach where atom typing is separated from other algorithms. & \\ \hline
+  ioformats                  & Definitions of (chemical) input/output formats. & & \\ \hline
+  io                         & Readers and writers for input/output formats.
+& The MDL molfile reader has been rewritten and supports in atom types defind in the specification.
+& XPP3 1.1.4c \\ \hline
+  iordf                      & Stores data models as in the Resource Description Framework serialization foramts. & New. & Jena 2.7.4 \\ \hline
+  inchi                      & IUPAC International Chemical Identifier support. & & JNI-InChI 0.8~\cite{Spjuth2013}  \\ \hline
+  libiocml                   & Writer for the Chemical Markup Language format. & & XOM 1.2.5, CMLXOM 3.1~\cite{Murray-Rust2011} \\ \hline
+  sdg                        & Structure diagram generation.  & Much improved overlap resolution. & \\ \hline
+  smiles                     & Reading and writing in the SMILES format. & SMILES support performance and converage is greatly improved. & Beam 0.9.1~\cite{Beam} \\ \hline
+  smarts                     & Substructure searching with the SMARTS format. & & Beam 0.9.1~\cite{Beam} \\ \hline
+  formula                    & Chemical formula support. & New. & \\ \hline
+  fingerprint                & Calculate fingerprints. & Many new fingerprint types (see text). & Apache Commons Math 3.1.1 \\ \hline
+  qsar and qsarmolecular                       & Molecular descriptors.  & & XOM 1.2.5, JAMA 1.0.3~\cite{Hicklin2012} \\ \hline
+  signatures                 & Calculation of molecular and atomic signatures. & & Signatures 1.1 \\ \hline % FIXME: ask Gilleain for the reference
     \end{tabular}

### 2016-08-11 article.tex
Consistent formatting of figure captions.

@@
 
-\subsection*{Figure 2 - Relative storage of stereochemistry, the \textbf{type} and \textbf{focus} of stereochemsitry are fixed for a given stereocenter description but the \textbf{carriers} and \textbf{configuration} are relative.
+\subsection*{Figure 2 - Relative storage of stereochemistry, the \textbf{type} and \textbf{focus} of stereochemsitry are fixed for a given stereocenter description but the \textbf{carriers} and \textbf{configuration} are relative.}\label{fig:stereodatastructure}
 The multiple rows for each stereochemistry type are different internal representation that would be considered equivalent.
-In the tetrahedral types, hydrogens may be suppressed in a molecular graph so the \textbf{focus} is reused in the \textbf{carriers} list as a placeholder.}\label{fig:stereodatastructure}
+In the tetrahedral types, hydrogens may be suppressed in a molecular graph so the \textbf{focus} is reused in the \textbf{carriers} list as a placeholder.
 
@@
 
-\subsection*{Figure 3 - The raw input files of CHEMBL23970 and CHEMBL444314 are displayed (ChEMBL 21).
+\subsection*{Figure 3 - The raw input files of CHEMBL23970 and CHEMBL444314 are displayed (ChEMBL 21).}\label{fig:stereoprojections}
 Without perceiving the stereochemistry indicated by Haworth projection in CHEMBL23970, the database entries are inncorectly considered distinct.
-Down stream aggregation databases mirror this separation (PubChem CID 5280, CID 65119).}\label{fig:stereoprojections}
+Down stream aggregation databases mirror this separation (PubChem CID 5280, CID 65119).
 
@@
 \subsection*{Figure 5 - The improved structure diagram generation has improved
-code to solve overlap. The original SDG code used general heuristics (left) and the
+code to solve overlap.}\label{fig:sdg}
+The original SDG code used general heuristics (left) and the
 OverlapResolver would fine tune the layout to ensure atoms would not be placed
@@
 make more rigorous changes, making the final output must more pleasing
-(right).}\label{fig:sdg}
+(right).
 

### 2016-08-11 article.tex
Updated the conclusions (closes #12)

@@
 
-Over the past years since the last CDK publication in 2006, a lot has changed. The functionality
-has seen many additions, and the stability of the development model has significantly improved.
-This paper outlines how peer review and quality control contributed to a much wider adoption
-of our cheminformatics library and seen many new additions. With 93 contributors,
-the CDK is alive and kicking.
+Since the last CDK publication in 2006, the library has been improved in many
+aspects. The functionality has seen many additions and rewrites, making the
+library more functional and a lot more faster then before. Updates on the common
+SMILES and MDL molfile formats and the improved structure diagram generation are
+very visible and benefit many of the tools using the CDK.
+Furthermore, the stability of the development model has significantly improved,
+providing greater stability of the library over time. The project adopted peer
+review and quality control.
 
-However, there are many unsolved issues. Performance is perhaps one of the most important ones.
-The disadvantage of having an API that supports many data models, is that the API is heavy and
-the interface implementations hard to optimize.
+With 93 contributors, a long list of tools based on the CDK, and hundreds
+of article citations, the CDK is alive and kicking.
 

### 2016-08-12 article.tex
Better section header name
Added two more CDK-using tools

@@
 Padel~\cite{yap2011padel}, AMBIT-SMARTS~\cite{jeliazkova2011ambitsmarts}, AMBIT-Tautomer~\cite{kochev2013ambit},
-ReactPRED~\cite{ReactPRED}, and SMSD~\cite{Rahman2009,Rahman2014}. Other tools used previous versions already, and follow
+ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014}, WhichCyp~\cite{Rostkowski2013},
+and MetaPrint2D~\cite{Carlsson2010}. Other tools used previous versions already, and follow
 newer releases of the CDK, such as Bioclipse~\cite{spjuth2007bioclipse,
@@
 
-\subsection*{New APIs}
+\subsection*{New APIs and improved implementations}
 

### 2016-08-13 formula_generator_evaluation/PFG/elements.cpp
Added the code used for chemical formula generator evaluation

@@ -1,0 +1,285 @@
+#include <iostream>
+#include <string>
+#include <map>
+#include "formula.h"
+
+using namespace std;
+
+#if defined __GNUC__
+#pragma GCC diagnostic ignored "-Wwrite-strings"
+#endif
+
+map<string, ellist> elements;
+
+void init_elementlist(){
+	
+	ellist Ac={"Actinium", "Ac", 89, 1, {227}, {227.02774700000001}, {1.0}, 3};
+	elements["Ac"]=Ac;
+	ellist Ag={"Silver", "Ag", 47, 2, {107,109}, {106.90509299999999, 108.90475600000001}, {0.51839000000000002, 0.48160999999999998}, 1};
+	elements["Ag"]=Ag;
+	ellist Al={"Aluminium", "Al", 13, 1, {27}, {26.981538440000001}, {1.0}, 3};
+	elements["Al"]=Al;
+	ellist Am={"Americium", "Am", 95, 2, {241,243}, {241.05682289999999, 243.06137269999999}, {0.0,1.0}, 3};
+	elements["Am"]=Am;
+	ellist Ar={"Argon", "Ar", 18, 2, {40,36,38}, {39.962383123000002, 35.967546280000001, 37.962732199999998}, {0.99600299999999997, 0.0033649999999999999, 0.00063199999999999997}, 0} ;
+	elements["Ar"]=Ar;
+	ellist As={"Arsenic", "As", 33, 1, {75}, {74.921596399999999}, {1.0}, 3};
+	elements["As"]=As;
+	ellist At={"Astatine", "At", 85, 2, {210,211}, {209.98713100000001,210.987481}, {0.0,1.0}, 1};
+	elements["At"]=At;
+	ellist Au={"Gold", "Au", 79, 1, {197}, {196.96655200000001}, {1.0}, 1};
+	elements["Au"]=Au;
+	ellist B={"Boron", "B", 5, 2, {10,11}, {10.012937000000001,11.0093055}, {0.19900000000000001, 0.80100000000000005}, 3};
+	elements["B"]=B;
[…]

### 2016-08-13 formula_generator_evaluation/PFG/formula.cpp

@@ -1,0 +1,300 @@
+#include <iostream>
+#include <string.h>
+#include <malloc.h>
+#include <ctime>
+#include <string.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include "formula.h"
+#include "meta.h"
+
+#if defined __GNUC__
+#pragma GCC diagnostic ignored "-Wwrite-strings"
+#endif
+
+using namespace std;
+
+double *elmass(vector<char*> compositions, int size){
+	double *_elmass;
+	_elmass = new double [size];
+	double temp;
+	for(int i = 0; i < size; i++){
+		temp = elementmass(compositions[i]);
+		_elmass[i] = temp;
+	}
+	return _elmass;
+}
+
+double delta(double measuredMass, double countedMass, string unit="ppm"){
+	// Calculate error between measured Mass and counted Mass in specified units.
+	if(strcmp(unit.c_str(),"ppm")==0)
+		return (measuredMass - countedMass) / countedMass*1000000;
+	else if(strcmp(unit.c_str(),"Da")==0)
+		return (measuredMass - countedMass);
+	else if(strcmp(unit.c_str(),"%")==0)
+		return (measuredMass - countedMass) / countedMass*100;
+	else printf("Unknown units for delta!");
+	    return false;
+}
+
+
+
+void PFG(ResultsWriter & writer, int elcount, int minimum[], int maximum[], double masses[], int *current, double pre_mass,double loMass, double hiMass)
+	//call the template metaprogramming
+{
+	int k = 0;
+	switch (elcount)
+	{
+		case 1:  formula_generator<1>::EXEC(writer, elcount, minimum, maximum, masses, current, pre_mass, loMass, hiMass, k); break;
+		case 2:  formula_generator<2>::EXEC(writer, elcount, minimum, maximum, masses, current, pre_mass, loMass, hiMass, k); break;
+		case 3:  formula_generator<3>::EXEC(writer, elcount, minimum, maximum, masses, current, pre_mass, loMass, hiMass, k); break;
+		case 4:  formula_generator<4>::EXEC(writer, elcount, minimum, maximum, masses, current, pre_mass, loMass, hiMass, k); break;
[…]

### 2016-08-13 formula_generator_evaluation/PFG/formula.h

@@ -1,0 +1,89 @@
+#ifndef __FORMULA_H   
+#define __FORMULA_H  
+#define SUBSLEN 10             
+#define EBUFLEN 128         
+#define BUFLEN 1024  
+#define ELECTRON_MASS  0.00054857990924
+#define min(a,b) a<b?a:b
+
+#include <string>
+#include <map>
+#include <iostream>
+#include <vector>
+using namespace std;
+
+typedef struct{
+	char *name;
+	char *symbol;
+	int atomicNumber;
+	int isotopecount;
+	int massnumber[10];
+	double isotope[10];
+	double abundance[10];
+	int valence;
+}ellist;
+
+typedef struct{
+	char **comp;
+	int *atomcount;
+	int elcount;
+}composition;
+
+typedef struct{
+	int *data;
+	int len;
+	double *mass;
+}result;
+
+typedef struct{
+	char *_formula;
+	double *_mass;
+	double *_mz;
+	double error;
+	double errorDa;
+	double rdbe;
+}output;
+
+
+class ResultsWriter
+{
+public:
+	ResultsWriter(string outfile,  vector<char*> compositions, vector<char*> rules, int charge, char * agentformula, int agentcharge, double currentmass, string unit);
+	~ResultsWriter();
+	void writeResults();
+	result *p_result;
+	int m_poolSize;
+private:
+	string m_outfile;
+	vector<char*> m_compositions;
+	vector<char*> m_rules;
+	int m_charge;
+	char * m_agentformula;
+	int m_agentcharge;
+	double m_currentmass;
+	string m_unit;
+	int m_elcount;
+	FILE *fpt;
+
+};
+
+//elements
+double elementmass(char *string);
+double rdbe(vector<char*> compositions, vector<int> count);
+double formulaMass(vector<char*> compositions, vector<int> count);
+
+//compound
+double mz(double currentmass,int finalcharge, int currentcharge, char *agentformula, int agentcharge);
+static double GLHC[2]={0.2, 3};
+static double GLNOPSC[4]={2, 1.2, 0.32, 0.65};
+static double GLRDBE[2]={0, 40};
+bool frules(char *string, int countC, int countH, int countO, int countN, int countP, int countS, double rdbevalue, 
+		   vector<char*> rules);
+
+//formula
+double *elmass(vector<char*> compositions, int size);
+double delta(double measuredMass, double countedMass, char units[]);
[…]

### 2016-08-13 formula_generator_evaluation/PFG/meta.h

@@ -1,0 +1,83 @@
+#include <iostream>
+#include <vector>
+#include <string>
+#include <omp.h>
+#include "formula.h"
+
+using namespace std;
+
+template<int N>
+class formula_generator{
+
+public:
+	static inline void  EXEC(ResultsWriter & writer, int elcount, int minimum[], int maximum[], double masses[], int *current, double pre_mass, double loMass, double hiMass, int k)
+	{
+		int c = min(int((hiMass-pre_mass)/masses[N-1]),maximum[N-1]);
+		if (N == elcount)
+		{
+		#pragma omp parallel for 
+			for (k = minimum[N-1]; k<=c; k++)
+			{
+				current[N-1] = k;
+				double current_mass_i= pre_mass+masses[N-1]*k;
+				int *current_i = new int [elcount];
+				for (int i=0;i<elcount;i++)
+				{
+					current_i[i]=current[i];
+				}
+
+				formula_generator<N-1>::EXEC(writer, elcount, minimum, maximum, masses, current_i, current_mass_i, loMass, hiMass, k);
+				delete []current_i;
+			}
+
+		}
+		else
+		{
+			for (int i = minimum[N-1]; i<=c; i++)
+			{
+
+				current[N-1] = i;
+
+				double current_mass_i= pre_mass+masses[N-1]*i;
+
+				formula_generator<N-1>::EXEC(writer, elcount, minimum, maximum, masses, current, current_mass_i, loMass, hiMass, k);
+
+			}
+		}
+	}
+};
+
+
+template<>
+class formula_generator<0>{
+
+public:
+	static inline void  EXEC(ResultsWriter & writer, int elcount, int minimum[], int maximum[], double masses[], int *current, double pre_mass, double loMass, double hiMass, int k)
+	{
+
+		if (pre_mass >= loMass && pre_mass <= hiMass )
+		{
+			#pragma omp critical (p_result)
+			{
+				for ( int i = 0; i < elcount; ++i)
+				{
+					if (i != elcount-1)
+					{
+						writer.p_result->data[writer.p_result->len*elcount+i] = current[i];
+					}
+					else
+					{
+						writer.p_result->data[writer.p_result->len*elcount+i] = k;
+					}
+				}
+				writer.p_result->mass[writer.p_result->len] = pre_mass;
+				writer.p_result->len++;
+				if (writer.p_result->len==writer.m_poolSize)
+				{
+					writer.writeResults();
+					writer.p_result->len = 0;
+				}
+			}
+		}
+	}
+};

### 2016-08-13 formula_generator_evaluation/measure_runtime.sh

@@ -1,0 +1,28 @@
+#!/bin/bash
+
+MASS=$1
+TOLERANCE=$2
+MIN=0
+MAX=10000
+ELEMENTS=C${MIN}-${MAX}H${MIN}-${MAX}N${MIN}-${MAX}O${MIN}-${MAX}S${MIN}-${MAX}P${MIN}-${MAX}
+
+echo Running CDK on mass ${MASS}+-${TOLERANCE}
+time -p java -cp "CDK/*:CDK" CDKFormulaGeneratorCLI $MASS $TOLERANCE $ELEMENTS > outCDK.txt
+echo
+
+echo Running PFG on mass ${MASS}+-${TOLERANCE}
+time -p PFG/PFG eval -m ${MASS} -u Da -t ${TOLERANCE} --C ${MIN}-${MAX} --H ${MIN}-${MAX} --N ${MIN}-${MAX} --O ${MIN}-${MAX} --S ${MIN}-${MAX} --P ${MIN}-${MAX} > outPFG.txt
+echo
+
+echo Running HR2 on mass ${MASS}+-${TOLERANCE}
+time -p HR2/HR2 -m ${MASS} -t ${TOLERANCE} -C ${MIN}-${MAX} -H ${MIN}-${MAX} -N ${MIN}-${MAX} -O ${MIN}-${MAX} -S ${MIN}-${MAX} -P ${MIN}-${MAX} > outHR2.txt
+echo
+
+# Clean up results
+cat outPFG.txt | cut -f 1 | grep -v formula | sort > outPFG.sorted
+cat outCDK.txt | sort > outCDK.sorted
+cat outHR2.txt | cut -f 1 | tail -n +6 | head -n -3 | sort > outHR2.sorted
+
+echo Comparing number of results
+wc -l *.sorted | grep -v total
+

### 2016-08-15 article.tex
Two more sections to be added, based on content from John
Added Kai Dührkop as coauthor
Added more details to the molecular formula section, added a table with formula generator performance evaluation

@@
 \usepackage{graphicx}
+\usepackage{multirow}
 
@@
 \author[
+   addressref={jena},
+   email={kai.duehrkop@uni-jena.de}
+]{\inits{KD}\fnm{Kai} \snm{Dührkop}}
+\author[
    addressref={idea},
@@
 }
+\address[id=jena]{
+  \orgname{Chair for Bioinformatics, Friedrich Schiller University},
+  %\street{},
+  \postcode{07743},
+  \city{Jena},
+  \cny{Germany}
+}
 \address[id=idea]{
@@
   \cny{Bulgaria}
+}
+\address[id=wi_mit]{
+  \orgname{Whitehead Institute for Biomedical Research}, 
+  \street{9 Cambridge Center},
+  \postcode{MA 02142},
+  \city{Cambridge},
+  \cny{USA}
 }
@@
 }
-\address[id=wi_mit]{
-  \orgname{Whitehead Institute for Biomedical Research}, 
-  \street{9 Cambridge Center},
-  \postcode{MA 02142},
-  \city{Cambridge},
-  \cny{USA}
-}
+
 %\begin{artnotes}
@@
 
-The chemical formula is the basic/simple chemical representation of a compound.
+  A chemical formula is the basic/simple chemical representation of a compound.
 It defines the number of isotopes or elements that constitute the composition
@@
   The CDK interfaces can handle several concepts related to chemical formulas:
-the formula itself, sets of formulas, chemical formula ranges, adducts,
-isotope containers and patterns, and rules to filter formula sets. These new
-tools can be used for a number of tasks, including calculate the isotopic
-pattern from a given chemical formula, determining the possible elemental
-compositions for a given mass, and calculate the exact mass from a given
+the formula itself, sets of formulas, chemical formula ranges, adducts, isotope
+containers and patterns, and rules to filter formula sets. These new tools can
+be used for a number of tasks, including calculating the isotopic pattern from
+a given chemical formula, determining the possible elemental compositions for a
+given mass (mass decomposition), and calculating the exact mass from a given
 chemical formula.
 
-The MolecularFormulaGenerator class was originally developed and evaluated 
-as part of the MZmine~2 framework~\cite{Pluskal2012}. The following code
-calculates all possible chemical formula for a given accurate mass, 
-within allowed counts for each element:
+  The CDK contains two algorithms for the decomposition of mass ranges into
+possible elemental formulas.  For most inputs, a Round Robin algorithm,
+originally developed for the SIRIUS metabolite identification
+tool~\cite{Bocker2009}, is used. The algorithm discretizes the real-value mass
+decomposition problem into an integer-value knapsack
+problem~\cite{Martello1990}. It first computes a dynamic programming table and
+then backtracks it to generate matching formulas~\cite{Duehrkop2013,
+Boecker2008}. Data for the Round Robin algorithm is stored in an extended
+residue table~\cite{Bocker2005}, resulting in a low memory footprint of several
+kilobytes. For certain problem instances, such as very large mass values (above
+$400\,000$ Da) or mass range span larger than $1$ Da, the Round Robin algorithm
+is not applicable and CDK falls back to an optimized full enumeration search
+method, originally developed as part of the MZmine~2 framework for mass
+spectrometry data processing~\cite{Pluskal2012, Pluskal2010}.
+
+  The following code calculates all possible chemical formulas for a given
+accurate mass, within allowed counts for each element:
 
@@
 \end{verbatim}
+
+   To evaluate the performance of the CDK molecular formula generator, we
+compared its runtimes on a range of inputs and compared them to the runtimes of
+the classic, full enumeration-based HR2 formula generator~\cite{Kind2007} and
+those of a recently developed Parallel Formula Generator (PFG)~\cite{Zhang2016}
+(Table~1\ref{tab:formula_generators}). Although for small mass values the PFG
+algorithm achieved shorter runtimes, for larger masses the Round Robin
+algorithm implemented in CDK clearly outperformed the two other methods.
 
@@
 
-  % \subsubsection*{SMARTS parser}
+  \subsubsection*{Substructure searching}
+
+  TODO: iterator and new SMARTS
 
@@
 
+  \subsubsection*{MDL molfile improvements}
+
+  TODO: S group, V3000 formats, ...
+
   \subsubsection*{New Builders}
@@
 supported in CDK as well. The fingerprints currently existing in CDK are listed
-in Table~1\ref{tab:fingerprints}.
+in Table~2\ref{tab:fingerprints}.
 
@@
 An overview of key modules with description, important changes, and dependencies
-on third-party libraries is given in Table~2\ref{tab:modules} and the dependencies
+on third-party libraries is given in Table~3\ref{tab:modules} and the dependencies
 between the CDK modules are depicted in Figure~6\ref{fig:deps}.
@@
 
-  \subsection*{Table 1 - The molecular fingerprints in CDK.}
+  \subsection*{Table 1 - Evaluation of molecular formula generators.}
+  \label{tab:formula_generators}
+  The resulting formula counts and runtimes of the HR2, PFG, and CDK chemical
+formula generators based on input mass and mass tolerance values. Formulas were
+generated using chemical elements C,H,N,O,P,S without bounds (the allowed atom
+count was set to 0--10000 for each element). All heuristic filtering rules were
+disabled for the purpose of the evaluation. The slight differences in the
+number of generated formulas were caused by different isotope masses embedded
+in each software and/or by rounding errors during calculation. The runtimes are
+average values from three independent runs performed on three different 16-core
+Intel Xeon 2.2~GHz CPU workstations equipped with 128~GB RAM, running Ubuntu
+Linux version 12.04.5 LTS and OpenJDK Java runtime version 1.7.0\_101.
+  \vskip 1\baselineskip
+
+    \begin{minipage}{1\textwidth}
+    \centering
+    \begin{tabular}{llllllll}
+	\multirow{2}{*}{ Mass (Da) } & \multirow{2}{*}{ Tolerance ($\pm$ Da) } & \multicolumn{3}{c}{ \# of generated formulas } & \multicolumn{3}{c}{ Runtime (s) } \\
+	& & HR2 & PFG & CDK & HR2 & PFG & CDK \\
+	\multirow{2}{*}{ $100$ } & $0.001$ & $1$ & $1$ & $1$ & $0.10$ & \textbf{$0.06$} & $0.49$ \\
+	& $0.01$ & $18$ & $18$ & $18$ & $0.03$ & \textbf{$0.01$} & $0.52$ \\
+	\multirow{2}{*}{ $500$ } & $0.001$ & $430$ & $430$ & $430$ & $0.51$ & \textbf{$0.09$} & $0.67$ \\
+	& $0.01$ & $4\,265$ & $4\,265$ & $4\,264$ & $0.55$ & \textbf{$0.39$} & $0.96$ \\
+	\multirow{2}{*}{ $1\,000$ } & $0.001$ & $8\,572$ & $8\,572$ & $8\,568$ & $16.5$ & $1.5$ & \textbf{$1.2$} \\
+	& $0.01$ & $85\,542$ & $85\,542$ & $85\,537$ & $16.8$ & $5.4$ & \textbf{$3.7$} \\
+	\multirow{2}{*}{ $2\,000$ } & $0.001$ & $226\,346$ & $226\,346$ & $226\,323$ & $1156.5$ & $58.9$ & \textbf{$8.6$} \\
+	& $0.01$ & $2\,263\,410$ & $2\,263\,410$ & $2\,263\,409$ & $932.5$ & $156.3$ & \textbf{$69.7$} \\
+	\multirow{2}{*}{ $3\,000$ } & $0.001$ & $1\,647\,132$ & $1\,647\,132$ & $1\,647\,118$ & $9\,667.9$ & $603.5$ & \textbf{$59.3$} \\
+	& $0.01$ & $16\,471\,338$ & $16\,471\,338$ & $16\,471\,373$ & $9\,694.5$ & $1\,303.6$ & \textbf{$505.7$} \\
+    \end{tabular}
+    \end{minipage}
+
+  \subsection*{Table 2 - The molecular fingerprints in CDK.}
   \label{tab:fingerprints}
@@
 
-      \subsection*{Table 2 - A selection of key CDK modules with major changes.}\label{tab:modules}
+      \subsection*{Table 3 - A selection of key CDK modules with major changes.}\label{tab:modules}
   An overview of a selection of often used CDK modules with description,

### 2016-08-15 article.tex
Fixed whitespace

@@
 	& & HR2 & PFG & CDK & HR2 & PFG & CDK \\
-	\multirow{2}{*}{ $100$ } & $0.001$ & $1$ & $1$ & $1$ & $0.10$ & \textbf{$0.06$} & $0.49$ \\
-	& $0.01$ & $18$ & $18$ & $18$ & $0.03$ & \textbf{$0.01$} & $0.52$ \\
-	\multirow{2}{*}{ $500$ } & $0.001$ & $430$ & $430$ & $430$ & $0.51$ & \textbf{$0.09$} & $0.67$ \\
-	& $0.01$ & $4\,265$ & $4\,265$ & $4\,264$ & $0.55$ & \textbf{$0.39$} & $0.96$ \\
-	\multirow{2}{*}{ $1\,000$ } & $0.001$ & $8\,572$ & $8\,572$ & $8\,568$ & $16.5$ & $1.5$ & \textbf{$1.2$} \\
-	& $0.01$ & $85\,542$ & $85\,542$ & $85\,537$ & $16.8$ & $5.4$ & \textbf{$3.7$} \\
-	\multirow{2}{*}{ $2\,000$ } & $0.001$ & $226\,346$ & $226\,346$ & $226\,323$ & $1156.5$ & $58.9$ & \textbf{$8.6$} \\
-	& $0.01$ & $2\,263\,410$ & $2\,263\,410$ & $2\,263\,409$ & $932.5$ & $156.3$ & \textbf{$69.7$} \\
-	\multirow{2}{*}{ $3\,000$ } & $0.001$ & $1\,647\,132$ & $1\,647\,132$ & $1\,647\,118$ & $9\,667.9$ & $603.5$ & \textbf{$59.3$} \\
-	& $0.01$ & $16\,471\,338$ & $16\,471\,338$ & $16\,471\,373$ & $9\,694.5$ & $1\,303.6$ & \textbf{$505.7$} \\
+	$100$ & $0.001$ & $1$ & $1$ & $1$ & $0.10$ & $\mathbf{0.06}$ & $0.49$ \\
+	$100$ & $0.01$ & $18$ & $18$ & $18$ & $0.03$ & $\mathbf{0.01}$ & $0.52$ \\
+	$500$ & $0.001$ & $430$ & $430$ & $430$ & $0.51$ & $\mathbf{0.09}$ & $0.67$ \\
+	$500$ & $0.01$ & $4\,265$ & $4\,265$ & $4\,264$ & $0.55$ & $\mathbf{0.39}$ & $0.96$ \\
+	$1\,000$ & $0.001$ & $8\,572$ & $8\,572$ & $8\,568$ & $16.5$ & $1.5$ & $\mathbf{1.2}$ \\
+	$1\,000$ & $0.01$ & $85\,542$ & $85\,542$ & $85\,537$ & $16.8$ & $5.4$ & $\mathbf{3.7}$ \\
+	$2\,000$ & $0.001$ & $226\,346$ & $226\,346$ & $226\,323$ & $1156.5$ & $58.9$ & $\mathbf{8.6}$ \\
+	$2\,000$ & $0.01$ & $2\,263\,410$ & $2\,263\,410$ & $2\,263\,409$ & $932.5$ & $156.3$ & $\mathbf{69.7}$ \\
+	$3\,000$ & $0.001$ & $1\,647\,132$ & $1\,647\,132$ & $1\,647\,118$ & $9\,667.9$ & $603.5$ & $\mathbf{59.3}$ \\
+	$3\,000$ & $0.01$ & $16\,471\,338$ & $16\,471\,338$ & $16\,471\,373$ & $9\,694.5$ & $1\,303.6$ & $\mathbf{505.7}$ \\
     \end{tabular}

### 2016-08-23 article.tex
Added Stefan Kuhn
Added the RDT reference (closes #26)

@@
 ]{\inits{NJ}\fnm{Nina} \snm{Jeliazkova}}
+\author[
+   addressref={leicester},
+   email={shk12@le.ac.uk}
+]{\inits{SK}\fnm{Stefan} \snm{Kuhn}}
 \author[
@@
 }
+\address[id=leicester]{
+  \orgname{Department of Informatics, University of Leicester},
+  %\street{},
+  %\postcode{LE1 7RH},
+  \city{Leicester},
+  \cny{UK}
+}
 
@@
 Padel~\cite{yap2011padel}, AMBIT-SMARTS~\cite{jeliazkova2011ambitsmarts}, AMBIT-Tautomer~\cite{kochev2013ambit},
-ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014}, WhichCyp~\cite{Rostkowski2013},
+ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016}, WhichCyp~\cite{Rostkowski2013},
 and MetaPrint2D~\cite{Carlsson2010}. Other tools used previous versions already, and follow

### 2016-08-24 article.tex
Worked on the SDG section

@@
   and as methods on the DepictionGenerator class. This includes substructure
-  coloring, exemplified with this example reaction (see Fig.~2\ref{fig:depiction}):
+  coloring, exemplified with this example reaction (see Fig.~\ref{fig:depiction}):
 
@@
 \begin{verbatim}
-TODO
+StructureDiagramGenerator sdg = new StructureDiagramGenerator();
+sdg.setMolecule(someMolecule);
+sdg.generateCoordinates();
+IAtomContainer layedOutMol = sdg.getMolecule();
 \end{verbatim}
+
+The API has hardly changed. The internals, however, have significantly. Besides
+the aforementioned overlap resolution, the engine also handles large ring
+systems much better, using a honeycomb layout approach~\cite{FIXME}, % FIXME
+and handles stereochemistry of the input. Previously, double bond stereochemistry
+was supported, but the new engine more efficiently uses this when creating 2D
+layouts. Furthermore, the engine assigns wedge bond information based on
+tetrahedral stereochemistry. These features are exemplified by the following
+code and the resulting layout depiction in Figure~\ref{fig:sdgstereo}.
+
+\begin{verbatim}
+someMolecule = fromSMILES.....
+sdg.setMolecule(someMolecule);
+sdg.generateCoordinates();
+IAtomContainer layedOutMol = sdg.getMolecule();
+\end{verbatim}
+
+% FIXME: complete code example
 
@@
 
-  \subsection*{Figure 6 - Dependencies between CDK modules.}\label{fig:deps}
+\subsection*{Figure 6 - Structure diagram generation for structures with
+double bond and tetrahedral stereochemistry.}\label{fig:sdgstereo}
+
+\includegraphics[width=\textwidth]{sdg2.png}
+
+  \subsection*{Figure 7 - Dependencies between CDK modules.}\label{fig:deps}
       Visualization of the dependencies between CDK modules. For example,

### 2016-08-24 article.tex
Started writing the MDL molfile section

@@
 
+The MDL molfile formats are still very popular, despite it being a proprietary
+standard, but taking advantage of the detailed specification still floating
+around on the world wide web and can be found by googling for
+ctfile.pdf~\cite{ctfile.pdf}. The CDK has various related readers, reflecting
+the various formats specified in this document, as well as a more historical
+reader for the pre-V2000 formats. When this reader was written, MDL molfiles
+in this pre-V2000 format were still around, but the authors are unaware of
+existing use of this format. At the same time, V3000 does not have seen a
+significant adoption either. Furthermore, several formats exist that wrap
+molfiles in collections, such as structure data files (SD files), reaction files,
+and reaction collections. Table XXX shows an overview of the various readers
+for (partially) supported formats.
+% note, I pie chart would be cool, showing the various relative amounts of the
+% various formats.
+
+Of course, the molfile is special here and this reader has been rewritten in
+this CDK version, and now supports....
+
   TODO: S group, V3000 formats, ...

### 2016-08-24 article.tex
Placeholder section for hash codes
Misnomer to call it MDL molfile, correct casing (no capital m) in my section.

@@
   The perception from coordinates and wedge assignment algorithms are fundamental for conversion
-  between formats that store stereochemistry implicitly based on coordinates (e.g. Molfile, CML) and 
+  between formats that store stereochemistry implicitly based on coordinates (e.g. molfile, CML) and 
   explicitly (e.g. SMILES, CML, InChI). Perception from 2D coordinates can optionally identify
@@
 
-  \subsubsection*{MDL molfile improvements}
-
-The MDL molfile formats are still very popular, despite it being a proprietary
+\subsubsection*{molfile improvements}
+
+The molfile formats are still very popular, despite it being a proprietary
 standard, but taking advantage of the detailed specification still floating
@@
 the various formats specified in this document, as well as a more historical
-reader for the pre-V2000 formats. When this reader was written, MDL molfiles
+reader for the pre-V2000 formats. When this reader was written, molfiles
 in this pre-V2000 format were still around, but the authors are unaware of
@@
 
+\subsubsection*{Hash Codes}
+
+%% JWM todo
+
 \subsection*{Improved Coding Standards}
@@
 library more functional and a lot more faster then before. Updates on the common
-SMILES and MDL molfile formats and the improved structure diagram generation are
+SMILES and molfile formats and the improved structure diagram generation are
 very visible and benefit many of the tools using the CDK.
@@
   ioformats                  & Definitions of (chemical) input/output formats. & & \\ \hline
-  io                         & Readers and writers for input/output formats.
-& The MDL molfile reader has been rewritten and supports in atom types defind in the specification.
-& XPP3 1.1.4c \\ \hline
+  io                         & Readers and writers for input/output formats.  & The molfile reader has been rewritten and supports in atom types defind in the specification. & XPP3 1.1.4c \\ \hline
   iordf                      & Stores data models as in the Resource Description Framework serialization foramts. & New. & Jena 2.7.4 \\ \hline

### 2016-08-24 article.tex
Update CTfile (molfile) section and include relevant citations.

--- 2016-08-24 article.tex
@@
 
-  \subsubsection*{MDL molfile improvements}
-
-  TODO: S group, V3000 formats, ...
+\subsubsection*{CTfile Format Improvements}
+
+The molfile format is still very popular, despite it being a proprietary
+format it has become a de facto standard. The format forms the core of the larger
+CTfile family which was originally developed by MDL Information Systems \cite{Dalby92}. The
+current format specification is published by BIOVIA and available on 
+request \cite{ctfilespec}.
+
+%% CTfile Format Support
+%% To tabulate, r=read, w=write
+%% MOLfile V2000=rw, V3000=rw
+%% RXNfile V2000=rw, V3000=r
+%% SDfile MOLfile V2000=rw, MOLfile V3000=r, RGfile=n/a
+%% RGfile V2000=rw
+%% RDfile n/a
+
+The CTAB block (connection table) of a molfile comes in two versions, V2000
+and V3000. The V3000 provides several enhancements including but not
+limited to: removing atom and bond count limits, enhanced stereochemistry,
+and link nodes. For backwards compatibility V2000 is often preferred resulting
+in limited usage of V3000.
+
+CDK \textbf{2.0} adds support for V3000 and has optimized and extended
+support for V2000. Currently these are considered separate formats requiring
+a user to know what version is being read beforehand. Future APIs will aim
+to simplify this and provide a unified reader.
+
+%% TODO(JWM): Sgroup Examples
+
+CTfile Sgroups capture and organise high level information about sets of atoms
+and bonds \cite{Gushurst91}. There are four types of Sgroup: Display Shortcuts, Polymers,
+Mixtures, and Data. The most familiar Sgroups from an end user perspective are structure 
+repeat units (e.g. bracketing) and abbreviations (figure: ??). CDK \textbf{2.0} adds 
+supports for representation, reading, writing, and depiction.
+
+% note, I pie chart would be cool, showing the various relative amounts of the
+% various formats.
 

### 2016-08-24 article.tex
Sgroup example image.

@@
 
-%% TODO(JWM): Sgroup Examples
-
 CTfile Sgroups capture and organise high level information about sets of atoms
@@
 Mixtures, and Data. The most familiar Sgroups from an end user perspective are structure 
-repeat units (e.g. bracketing) and abbreviations (figure: ??). CDK \textbf{2.0} adds 
+repeat units (e.g. bracketing) and abbreviations (Figure: 8\label{fig:sgroups}). CDK \textbf{2.0} adds 
 supports for representation, reading, writing, and depiction.
@@
 
-\includegraphics[width=\textwidth]{sdg2.png}
+%\includegraphics[width=\textwidth]{sdg2.png}
 
@@
 \includegraphics[width=\textwidth]{cdkDeps.png}
+
+\subsection*{Figure 8 - Examples of Sgroups now captured by the CDK and encoded in molfiles and CXSMILES.
+a) Ethyl esterification fully expanded reaction.
+b) Using Sgroup abbreviations allows display short cuts and more compact depiction.
+c) An example of a structure repeat unit in \textit{DNA 5'-phosphate} (\textbf{CHEBI:4294})}
+\label{fig:sgroups}
+
+\includegraphics[width=\textwidth]{img/sgroups.pdf}
 

### 2016-08-25 article.tex
Rerender reaction example.
sdg2 is missing so had it commented out.
Gilleain thanked for the offer and cannot contribute to the paper (personal communication)

@@
 ]{\inits{OS}\fnm{Ola} \snm{Spjuth}}
-\author[
-   addressref={???},
-   email={gilleain.torrance@gmail.com}
-]{\inits{GT}\fnm{Gilleain} \snm{Torrance}}
 \author[
@@
 \subsection*{Figure 4 - Integrated example showing the rendering and SMILES
-parsing functionality.}\label{fig:depiction}
-
-\includegraphics[width=\textwidth]{reaction.png}
+parsing functionality. Example from U.S. Patent \textbf{US 2014 231770 A1} para 287.}\label{fig:depiction}
+
+%% SMILES=[CH3:17][CH:16]([CH3:18])[CH2:15][Mg+].[cH:7]1[cH:6][cH:5][c:4]2[c:9]([cH:8]1)[cH:10][cH:11][c:2]3[c:3]2[n:12][cH:13][n:14][c:1]3Cl.[Br-]>C/C(=C/C(=O)C)/[O-].C/C(=C/C(=O)C)/[O-].C/C(=C/C(=O)C)/[O-].[Fe+3].CN1CCCC1=O.C1CCOC1.Cl>[CH3:17][CH:16]([CH3:18])[CH2:15][c:1]1[c:2]2[cH:11][cH:10][c:9]3[cH:8][cH:7][cH:6][cH:5][c:4]3[c:3]2[n:12][cH:13][n:14]1 |f:0.2,3.4.5.6,SgD::cdk&#58;ReactionConditions:16 h&#44; RT|
+%% Kumada Coupling?
+
+\includegraphics[width=\textwidth]{img/US20140231770A1_0287.pdf}
 

### 2016-08-25 article.tex
Small typo fix

@@
 jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
-Padel~\cite{yap2011padel}, AMBIT-SMARTS~\cite{jeliazkova2011ambitsmarts}, AMBIT-Tautomer~\cite{kochev2013ambit},
+PaDEL~\cite{yap2011padel}, AMBIT-SMARTS~\cite{jeliazkova2011ambitsmarts}, AMBIT-Tautomer~\cite{kochev2013ambit},
 ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016}, WhichCyp~\cite{Rostkowski2013},

### 2016-08-25 article.tex
Added two additional keywords
Finetuning things a bit

@@
 \kwd{bioinformatics}
+\kwd{metabolomics}
+\kwd{depiction}
 \end{keyword}
@@
   (Figure~3\ref{fig:stereoprojections}). The perception is based on an algorithm briefly 
-  described by \cite{Karapetyan2015}.
+  described by~\cite{Karapetyan2015}.
 
@@
   absolute label which can then be integrated into the ordering. More detailed descriptions of 
-  these algorithms are described in detail in Chapter 6 of \cite{May2015}.
+  these algorithms are described in detail in Chapter~6 of~\cite{May2015}.
 
@@
   and as methods on the DepictionGenerator class. This includes substructure
-  coloring, exemplified with this example reaction (see Fig.~\ref{fig:depiction}):
+  coloring, exemplified with this example reaction (see Figure~FIXME\ref{fig:depiction}):
 
@@
 format it has become a de facto standard. The format forms the core of the larger
-CTfile family which was originally developed by MDL Information Systems \cite{Dalby92}. The
+CTfile family which was originally developed by MDL Information Systems~\cite{Dalby92}. The
 current format specification is published by BIOVIA and available on 
-request \cite{ctfilespec}.
+request~\cite{ctfilespec}.
 
@@
 CTfile Sgroups capture and organise high level information about sets of atoms
-and bonds \cite{Gushurst91}. There are four types of Sgroup: Display Shortcuts, Polymers,
+and bonds~\cite{Gushurst91}. There are four types of Sgroup: Display Shortcuts, Polymers,
 Mixtures, and Data. The most familiar Sgroups from an end user perspective are structure 
-repeat units (e.g. bracketing) and abbreviations (Figure: 8\label{fig:sgroups}). CDK \textbf{2.0} adds 
+repeat units (e.g. bracketing) and abbreviations (Figure~8\label{fig:sgroups}). CDK \textbf{2.0} adds 
 supports for representation, reading, writing, and depiction.

### 2016-08-25 article.tex
Per recommendation: added a note that I'm selling a book about the CDK (currently not really of financial interest)
Converted a blob to a proper table

@@
 
-%% CTfile Format Support
-%% To tabulate, r=read, w=write
-%% MOLfile V2000=rw, V3000=rw
-%% RXNfile V2000=rw, V3000=r
-%% SDfile MOLfile V2000=rw, MOLfile V3000=r, RGfile=n/a
-%% RGfile V2000=rw
-%% RDfile n/a
-
 The CTAB block (connection table) of a molfile comes in two versions, V2000
@@
 a user to know what version is being read beforehand. Future APIs will aim
-to simplify this and provide a unified reader.
+to simplify this and provide a unified reader. An overview of currently
+supported CTfile formats is given in Table~2\ref{tab:ctfileFormats}.
 
@@
 supported in CDK as well. The fingerprints currently existing in CDK are listed
-in Table~2\ref{tab:fingerprints}.
+in Table~3\ref{tab:fingerprints}.
 
@@
 An overview of key modules with description, important changes, and dependencies
-on third-party libraries is given in Table~3\ref{tab:modules} and the dependencies
+on third-party libraries is given in Table~4\ref{tab:modules} and the dependencies
 between the CDK modules are depicted in Figure~6\ref{fig:deps}.
@@
 \section*{Competing interests}
-JWM and NJ work for companies that sell solutions based on the CDK.
+JWM and NJ work for companies that sell solutions based on the CDK. ELW sells
+a book describing the CDK functionality.
 
@@
 
-  \subsection*{Table 2 - The molecular fingerprints in CDK.}
+    %% CTfile Format Support
+%% To tabulate, r=read, w=write
+      \subsection*{Table 2 - CTfile Format Support..}\label{tab:ctfileFormats}
+
+    \begin{minipage}{1\textwidth}
+    \renewcommand*{\thempfootnote}{\fnsymbol{mpfootnote}}
+    \centering
+    \begin{tabular}{lll}
+  \textbf{Format}            & \textbf{V2000}  & \textbf{V3000} \\ \hline
+    MOLfile & read and write & read and write \\
+    RXNfile & read and write & read \\
+    SDfile MOLfile & read and write & read \\ % FIXME: RGfile=n/a
+    RGfile & read and write & \\
+    RDfile & & \\ % FIXME: yes, we do have support
+    \end{tabular}
+    \end{minipage}
+
+  \subsection*{Table 3 - The molecular fingerprints in CDK.}
   \label{tab:fingerprints}
@@
 
-      \subsection*{Table 3 - A selection of key CDK modules with major changes.}\label{tab:modules}
+      \subsection*{Table 4 - A selection of key CDK modules with major changes.}\label{tab:modules}
   An overview of a selection of often used CDK modules with description,

### 2016-08-25 article.tex
Some spell checking

@@
 \subsection*{Figure 3 - The raw input files of CHEMBL23970 and CHEMBL444314 are displayed (ChEMBL 21).}\label{fig:stereoprojections}
-Without perceiving the stereochemistry indicated by Haworth projection in CHEMBL23970, the database entries are inncorectly considered distinct.
+Without perceiving the stereochemistry indicated by Haworth projection in CHEMBL23970, the database entries are incorrectly considered distinct.
 Down stream aggregation databases mirror this separation (PubChem CID 5280, CID 65119).
@@
   io                         & Readers and writers for input/output formats.  & The molfile reader has been rewritten and supports in atom types defind in the specification. & XPP3 1.1.4c \\ \hline
-  iordf                      & Stores data models as in the Resource Description Framework serialization foramts. & New. & Jena 2.7.4 \\ \hline
+  iordf                      & Stores data models as in the Resource Description Framework serialization formats. & New. & Jena 2.7.4 \\ \hline
   inchi                      & IUPAC International Chemical Identifier support. & & JNI-InChI 0.8~\cite{Spjuth2013}  \\ \hline
@@
   sdg                        & Structure diagram generation.  & Much improved overlap resolution. & \\ \hline
-  smiles                     & Reading and writing in the SMILES format. & SMILES support performance and converage is greatly improved. & Beam 0.9.1~\cite{Beam} \\ \hline
+  smiles                     & Reading and writing in the SMILES format. & SMILES support performance and coverage is greatly improved. & Beam 0.9.1~\cite{Beam} \\ \hline
   smarts                     & Substructure searching with the SMARTS format. & & Beam 0.9.1~\cite{Beam} \\ \hline

### 2016-08-27 article.tex
Put the CDK version discussed in this paper in a variable for quick changing/reference.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%% RESEARCH PAPER %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\def \cdkversion {\textbf{2.0}}
 
@@
   In previous versions of the API, stereochemistry was represented disparately hindering 
-  interconversion between and within file formats. In CDK 1.5.x the core representation 
+  interconversion between and within file formats. In CDK \cdkversion{} the core representation 
   has been standardized upon and procedures updated or added to enable duplicate checking,
@@
 
-CDK \textbf{2.0} adds support for V3000 and has optimized and extended
+CDK \cdkversion{} adds support for V3000 and has optimized and extended
 support for V2000. Currently these are considered separate formats requiring
@@
 Mixtures, and Data. The most familiar Sgroups from an end user perspective are structure 
-repeat units (e.g. bracketing) and abbreviations (Figure~8\label{fig:sgroups}). CDK \textbf{2.0} adds 
+repeat units (e.g. bracketing) and abbreviations (Figure~8\label{fig:sgroups}). CDK \cdkversion{} adds 
 supports for representation, reading, writing, and depiction.

### 2016-08-28 article.tex
Updated the table to list the latest CDK version number (closes #19)

@@
                              & Bit version  & Count version & CDK version & Default Size    \\
-  CircularFingerprinter~\cite{rogers2010extended, Clark2014}      & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $1\,024$ / $2^{32}$%
+  CircularFingerprinter~\cite{rogers2010extended, Clark2014}      & $\checkmark$ & $\checkmark$  & \cdkversion{}     & $1\,024$ / $2^{32}$%
 \footnote[1]{For the CirkularFingerprinter the bit version is folded to 1024 whereas the count version is unfolded.} \\
@@
   KlekotaRothFingerprinter~\cite{Klekota2008}   & $\checkmark$ &               & 1.4.6       & $4\,860$        \\
-  LingoFingerprinter~\cite{vidal2005lingo}         & $\checkmark$ &               & 1.6.0 ?     & NA%
+  LingoFingerprinter~\cite{vidal2005lingo}         & $\checkmark$ &               & \cdkversion{}     & NA%
 \footnote[2]{The Lingofingerprinter does not have a default size.}
@@
   PubchemFingerprinter~\cite{pubchemFP}       & $\checkmark$ &               & 1.4.0       & $881$            \\
-  ShortestPathFingerprinter  & $\checkmark$ &               & 1.6.0 ?     & $1\,024$        \\
-  SignatureFingerprinter~\cite{signaturefingerprints}     & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $2^{32}$         \\
+  ShortestPathFingerprinter  & $\checkmark$ &               & \cdkversion{}     & $1\,024$        \\
+  SignatureFingerprinter~\cite{signaturefingerprints}     & $\checkmark$ & $\checkmark$  & \cdkversion{}     & $2^{32}$         \\
   SubstructureFingerprinter  & $\checkmark$ &               & 1.0         & $307$           \\

### 2016-08-28 article.tex
Added the GtoPdb

@@
 ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016}, WhichCyp~\cite{Rostkowski2013},
-and MetaPrint2D~\cite{Carlsson2010}. Other tools used previous versions already, and follow
+MetaPrint2D~\cite{Carlsson2010}, and the IUPHAR/BPS Guide to Pharmacology database~\cite{Southan2016}.
+Other tools used previous versions already, and follow
 newer releases of the CDK, such as Bioclipse~\cite{spjuth2007bioclipse,

### 2016-08-28 article.tex
Revert "Gilleain thanked for the offer and cannot contribute to the paper (personal communication)"
This reverts commit c24afea8ae9f4f97ef5d66a439379b6af9569356.
Details Gilleain (happy to have on board) and a few other minor updates (still hoping to get a reply from Miquel)

@@
 \author[
-   addressref={???},
+   addressref={miquel},
    email={miguelrojasch@googlemail.com}
-]{\inits{MRJ}\fnm{Miguel}~\snm{Rojas-Cherto}}
+]{\inits{MRJ}\fnm{Miquel}~\snm{Rojas Chertó}}
 \author[
@@
 \author[
+   addressref={gilleain},                           % ORCID: 0000-0002-8368-6954
+   email={gilleain.torrance@gmail.com}
+]{\inits{GT}\fnm{Gilleain} \snm{Torrance}}
+\author[
    addressref={um},
@@
   \orgname{Dept of Bioinformatics - BiGCaT, NUTRIM, Maastricht University}, % university, etc
-  %\street{Waterloo Road},                     %
+  %\street{Waterloo Road},
   \postcode{NL-6200 MD},
-  \city{Maastricht},                              % city
-  \cny{The Netherlands}                                    % country
+  \city{Maastricht},
+  \cny{The Netherlands}
 }
 \address[id=aff2]{
-  \orgname{??},
+  %\orgname{??},
   %\street{D\"{u}sternbrooker Weg 20},
@@
   \cny{USA}
+}
+\address[id=miquel]{
+  %\orgname{Department of Informatics, University of Leicester},
+  %\street{4 Hanway Place},
+  %\postcode{W1T 1HD},
+  %\city{London},
+  \cny{Spain}
+}
+\address[id=gilleain]{
+  %\orgname{Department of Informatics, University of Leicester},
+  \street{4 Hanway Place},
+  \postcode{W1T 1HD},
+  \city{London},
+  \cny{UK}
 }

### 2016-09-03 article.tex
Added the missing figure (closes #21)
Worked on the signatures section (closes #31)

@@
   descriptor can also be canonicalized to provide isomorphism-independent
-  representations~\cite{Faulon2004}, and has been successfully used in QSAR modeling~\cite{signaturefingerprints,Spjuth2011DS,Moghadam2015,Alvarsson2014,Spjuth2012OS,Norinder2013}.
+  representations~\cite{Faulon2004}. Signatures can be calculated for atoms with:
+
+\begin{verbatim}
+\end{verbatim}
+
+  But they can also can be calculated for full molecules:
+
+\begin{verbatim}
+\end{verbatim}
+
+  Finally, a signatures fingerprint can be calculated for molecules, to allow
+  similarity calculations. This can then be used in QSAR modeling~\cite{signaturefingerprints,
+  Spjuth2011DS,Moghadam2015,Alvarsson2014,Spjuth2012OS,Norinder2013,Alvarsson2016}.
+  The fingerprint for a molecule can be calculated with:
+
+\begin{verbatim}
+\end{verbatim}
 
@@
 \begin{verbatim}
-someMolecule = fromSMILES.....
+someMolecule = parser.parseSmiles("OC/C(C)=C/[C@](F)(Cl)Br");
 sdg.setMolecule(someMolecule);
@@
 
-%\includegraphics[width=\textwidth]{sdg2.png}
+\includegraphics[width=0.3\textwidth]{sdg2.png}
 

### 2016-09-05 article.tex
Added my own ORCID
Added info for Miquel

@@
 \author[
-   addressref={um},                   % id's of addresses, e.g. {aff1,aff2}
-   %corref={aff1},                       % id of corresponding address, if any
-   %noteref={n1},                        % id's of article notes, if any
-   email={egon.willighagen@maastrichtuniversity.nl}   % email address
+   addressref={um},                                 % ORCID: 0000-0001-7542-0286
+   %corref={aff1},
+   %noteref={n1},
+   email={egon.willighagen@maastrichtuniversity.nl}
 ]{\inits{ELW}\fnm{Egon L} \snm{Willighagen}}
@@
 \author[
-   addressref={miquel},
+   addressref={miquel},                             % ORCID: 0000-0002-4659-1446
    email={miguelrojasch@googlemail.com}
@@
 \address[id=miquel]{
-  %\orgname{Department of Informatics, University of Leicester},
+  \orgname{Química Clínica Aplicada},
   %\street{4 Hanway Place},
   %\postcode{W1T 1HD},
-  %\city{London},
+  \city{Amposta},
   \cny{Spain}

### 2016-09-06 formula_generator_evaluation/CDK/CDKFormulaGeneratorCLI.java
Updated the formula generator evaluation code

@@ -1,3 +1,5 @@
+import java.io.BufferedReader;
+import java.io.FileReader;
 import java.util.regex.Matcher;
 import java.util.regex.Pattern;
 
@@ -13,54 +15,65 @@ import org.openscience.cdk.tools.manipulator.MolecularFormulaManipulator;
 
 public class CDKFormulaGeneratorCLI {
 
-    public static void main(String args[]) {
+	public static void main(String args[]) {
 
-        if (args.length != 3) {
-            System.err
-                    .println("Use 3 arguments: <mass> <tolerance> <elements>");
-            System.err.println("For example:");
-            System.err.println("      mass: 500.0");
-            System.err.println("      tolerance: 0.005");
-            System.err.println("      elements: C0-100H0-100N0-50O0-50S0-10P0-10");
-            System.exit(1);
-        }
+		if (args.length != 3) {
+			System.err.println("Use 3 arguments: <mass_file> <tolerance> <elements>");
+			System.err.println("For example:");
+			System.err.println("      mass_file: masses.txt");
+			System.err.println("      tolerance: 0.005");
+			System.err.println("      elements: C0-100H0-100N0-50O0-50S0-10P0-10");
+			System.exit(1);
+		}
 
-        try {
+		try {
 
-            double mass = Double.parseDouble(args[0]);
-            double tolerance = Double.parseDouble(args[1]);
-            String elements = args[2];
+			double tolerance = Double.parseDouble(args[1]);
+			String elements = args[2];
 
-            IsotopeFactory ifac = Isotopes.getInstance();
-            Pattern p = Pattern.compile("(\\p{Alpha}+)(\\d+)-(\\d+)");
-            Matcher m = p.matcher(elements);
+			IsotopeFactory ifac = Isotopes.getInstance();
+			Pattern p = Pattern.compile("(\\p{Alpha}+)(\\d+)-(\\d+)");
+			Matcher m = p.matcher(elements);
 
-            MolecularFormulaRange mfRange = new MolecularFormulaRange();
+			MolecularFormulaRange mfRange = new MolecularFormulaRange();
 
-            while (m.find()) {
-                String element = m.group(1);
-                int min = Integer.parseInt(m.group(2));
-                int max = Integer.parseInt(m.group(3));
-                IIsotope i;
-		if (element.equals("D")) i = ifac.getIsotope("H", 2);
-			else i = ifac.getMajorIsotope(element);
-                mfRange.addIsotope(i, min, max);
[…]

### 2016-09-06 formula_generator_evaluation/measure_all.sh

@@ -1,0 +1,8 @@
+#!/bin/bash
+
+./measure_runtime.sh small_masses.txt 0.001
+./measure_runtime.sh small_masses.txt 0.01
+
+./measure_runtime.sh large_masses.txt 0.001
+./measure_runtime.sh large_masses.txt 0.01
+

### 2016-09-06 formula_generator_evaluation/measure_runtime.sh

@@ -1,28 +1,28 @@
 #!/bin/bash
 
-MASS=$1
+MASS_FILE=$1
 TOLERANCE=$2
 MIN=0
 MAX=10000
-ELEMENTS=C${MIN}-${MAX}H${MIN}-${MAX}N${MIN}-${MAX}O${MIN}-${MAX}S${MIN}-${MAX}P${MIN}-${MAX}
 
-echo Running CDK on mass ${MASS}+-${TOLERANCE}
-time -p java -cp "CDK/*:CDK" CDKFormulaGeneratorCLI $MASS $TOLERANCE $ELEMENTS > outCDK.txt
+rm -rf 
+echo Running CDK on file $MASS_FILE +-${TOLERANCE} Da
+time -p java -cp "CDK/*:CDK" CDKFormulaGeneratorCLI $MASS_FILE $TOLERANCE C${MIN}-${MAX}H${MIN}-${MAX}N${MIN}-${MAX}O${MIN}-${MAX}S${MIN}-${MAX}P${MIN}-${MAX} > outCDK.txt
 echo
 
-echo Running PFG on mass ${MASS}+-${TOLERANCE}
-time -p PFG/PFG eval -m ${MASS} -u Da -t ${TOLERANCE} --C ${MIN}-${MAX} --H ${MIN}-${MAX} --N ${MIN}-${MAX} --O ${MIN}-${MAX} --S ${MIN}-${MAX} --P ${MIN}-${MAX} > outPFG.txt
+echo Running PFG on file $MASS_FILE +-${TOLERANCE} Da
+time -p while read MASS ; do PFG/PFG eval -m ${MASS} -u Da -t ${TOLERANCE} --C ${MIN}-${MAX} --H ${MIN}-${MAX} --N ${MIN}-${MAX} --O ${MIN}-${MAX} --S ${MIN}-${MAX} --P ${MIN}-${MAX} ; done < $MASS_FILE > outPFG.txt
 echo
 
-echo Running HR2 on mass ${MASS}+-${TOLERANCE}
-time -p HR2/HR2 -m ${MASS} -t ${TOLERANCE} -C ${MIN}-${MAX} -H ${MIN}-${MAX} -N ${MIN}-${MAX} -O ${MIN}-${MAX} -S ${MIN}-${MAX} -P ${MIN}-${MAX} > outHR2.txt
+echo Running HR2 on file $MASS_FILE +-${TOLERANCE} Da
[…]

### 2016-09-06 article.tex
Updated the performance evaluation of molecular formula generators

@@
 $400\,000$ Da) or mass range span larger than $1$ Da, the Round Robin algorithm
-is not applicable and CDK falls back to an optimized full enumeration search
+is not suitable and CDK falls back to an optimized full enumeration search
 method, originally developed as part of the MZmine~2 framework for mass
@@
    To evaluate the performance of the CDK molecular formula generator, we
-compared its runtimes on a range of inputs and compared them to the runtimes of
-the classic, full enumeration-based HR2 formula generator~\cite{Kind2007} and
-those of a recently developed Parallel Formula Generator (PFG)~\cite{Zhang2016}
-(Table~1\ref{tab:formula_generators}). Although for small mass values the PFG
-algorithm achieved shorter runtimes, for larger masses the Round Robin
-algorithm implemented in CDK clearly outperformed the two other methods.
+compared its runtimes to those of the classic, full enumeration-based HR2
+formula generator~\cite{Kind2007} and those of a recently developed Parallel
+Formula Generator (PFG)~\cite{Zhang2016} (Table~1\ref{tab:formula_generators}).
+As inputs, we used two sets of $10\,000$ small ($< 500$~Da) and $20$ large ($>
+1\,500$ and $< 3\,500$~Da) molecular mass values downloaded from the Global
+Natural Products Social Molecular Networking database~\cite{wang2016}. The mass
+tolerance was set to 0.001 or 0.01 Da. The CDK's Round-Robin formula generator
+outperformed the other methods in all cases, despite running in a single thread
+(PFG utilizes multiple threads).
 
@@
   The resulting formula counts and runtimes of the HR2, PFG, and CDK chemical
-formula generators based on input mass and mass tolerance values. Formulas were
-generated using chemical elements C,H,N,O,P,S without bounds (the allowed atom
-count was set to 0--10000 for each element). All heuristic filtering rules were
-disabled for the purpose of the evaluation. The slight differences in the
-number of generated formulas were caused by different isotope masses embedded
-in each software and/or by rounding errors during calculation. The runtimes are
-average values from three independent runs performed on three different 16-core
-Intel Xeon 2.2~GHz CPU workstations equipped with 128~GB RAM, running Ubuntu
-Linux version 12.04.5 LTS and OpenJDK Java runtime version 1.7.0\_101.
+formula generators on two different inputs with two different mass tolerance
+settings. For the set of small masses, $10\,000$ mass values in the range of
+$0$--$500$ Da were randomly selected from the Global Natural Products Social
+Molecular Networking database~\cite{wang2016}. For the set of large masses, $20$
+mass values in the range of $1\,500$--$3\,500$ Da were randomly selected from the same
+database. Formulas were generated using chemical elements C,H,N,O,P,S without
+bounds (the allowed atom count was set to $0$--$10\,000$ for each element). All
+heuristic filtering rules were disabled for the purpose of the evaluation. The
+slight differences in the number of generated formulas were caused by different
+isotope masses embedded in each software and/or by rounding errors during
+calculation. The runtimes are average values from three independent runs
+performed on three different 16-core Intel Xeon 2.9~GHz CPU workstations
+equipped with 189~GB RAM, running Ubuntu Linux version 12.04.5 LTS and OpenJDK
+Java runtime version 1.7.0\_101.
   \vskip 1\baselineskip
@@
     \begin{tabular}{llllllll}
-	\multirow{2}{*}{ Mass (Da) } & \multirow{2}{*}{ Tolerance ($\pm$ Da) } & \multicolumn{3}{c}{ \# of generated formulas } & \multicolumn{3}{c}{ Runtime (s) } \\
+	\multirow{2}{*}{ Input } & \multirow{2}{*}{ Mass tolerance ($\pm$ Da) } & \multicolumn{3}{c}{ \# of generated formulas } & \multicolumn{3}{c}{ Runtime (s) } \\
 	& & HR2 & PFG & CDK & HR2 & PFG & CDK \\
-	$100$ & $0.001$ & $1$ & $1$ & $1$ & $0.10$ & $\mathbf{0.06}$ & $0.49$ \\
-	$100$ & $0.01$ & $18$ & $18$ & $18$ & $0.03$ & $\mathbf{0.01}$ & $0.52$ \\
-	$500$ & $0.001$ & $430$ & $430$ & $430$ & $0.51$ & $\mathbf{0.09}$ & $0.67$ \\
-	$500$ & $0.01$ & $4\,265$ & $4\,265$ & $4\,264$ & $0.55$ & $\mathbf{0.39}$ & $0.96$ \\
-	$1\,000$ & $0.001$ & $8\,572$ & $8\,572$ & $8\,568$ & $16.5$ & $1.5$ & $\mathbf{1.2}$ \\
-	$1\,000$ & $0.01$ & $85\,542$ & $85\,542$ & $85\,537$ & $16.8$ & $5.4$ & $\mathbf{3.7}$ \\
-	$2\,000$ & $0.001$ & $226\,346$ & $226\,346$ & $226\,323$ & $1156.5$ & $58.9$ & $\mathbf{8.6}$ \\
-	$2\,000$ & $0.01$ & $2\,263\,410$ & $2\,263\,410$ & $2\,263\,409$ & $932.5$ & $156.3$ & $\mathbf{69.7}$ \\
-	$3\,000$ & $0.001$ & $1\,647\,132$ & $1\,647\,132$ & $1\,647\,118$ & $9\,667.9$ & $603.5$ & $\mathbf{59.3}$ \\
-	$3\,000$ & $0.01$ & $16\,471\,338$ & $16\,471\,338$ & $16\,471\,373$ & $9\,694.5$ & $1\,303.6$ & $\mathbf{505.7}$ \\
+	$10\,000$ small masses & $0.001$ & $616\,846$ & $616\,846$ & $616\,843$ & $669$ & $168$ & $\mathbf{41}$ \\
+	$10\,000$ small masses & $0.01$ & $6\,163\,303$ & $6\,163\,302$ & $6\,163\,326$ & $689$ & $501$ & $\mathbf{212}$ \\
+	$20$ large masses & $0.001$ & $4\,912\,939$ & $4\,912\,939$ & $4\,912\,904$ & $26\,370$ & $1\,292$ & $\mathbf{177}$ \\
+	$20$ large masses & $0.01$ & $49\,128\,811$ & $49\,128\,810$ & $49\,128\,815$ & $26\,587$ & $3\,406$ & $\mathbf{1\,580}$ \\
     \end{tabular}

### 2016-09-06 formula_generator_evaluation/measure_runtime.sh
Fixed a typo

@@ -23,6 +23,6 @@ cat outPFG.txt | cut -f 1 | grep -v formula | sort > outPFG.sorted
 cat outCDK.txt | sort > outCDK.sorted
 cat outHR2.txt | cut -f 1 | sort > outHR2.sorted
 
-echo Comparing thr numbers of results
+echo Comparing the numbers of results
 wc -l *.sorted | grep -v total
 

### 2016-09-06 article.tex
Duplicated closely

--- 2016-09-05 article.tex
@@
 
-On the other hand, for many modules, the test code is now more closely closely
+On the other hand, for many modules, the test code is now more closely
 linked to the code being tested: both reside in the same folder, though we

### 2016-09-07 article.tex
Update info Miquel
Updated the title to remove the '3'. Not uncommon in series of papers, but in the context of software versions easily confused with the release version (closes #35)

@@
 
-\title{The Chemistry Development Kit (CDK). 3. Atom typing, Rendering, Molecular Formula, and Substructure Searching}
+% previous titles:
+$
+% The Chemistry Development Kit (CDK): an open-source Java library for Chemo- and Bioinformatics.
+% Recent developments of the chemistry development kit (CDK) - an open-source java library for chemo- and bioinformatics.
+
+\title{The Chemistry Development Kit (CDK): atom typing, rendering, molecular formulas, and substructure searching}
 
@@
    addressref={miquel},                             % ORCID: 0000-0002-4659-1446
-   email={miguelrojasch@googlemail.com}
-]{\inits{MRJ}\fnm{Miquel}~\snm{Rojas Chertó}}
+   email={mrojas@qca.es}
+]{\inits{MRJ}\fnm{Miquel}~\snm{Rojas-Chertó}}
 \author[
@@
   %\street{4 Hanway Place},
-  %\postcode{W1T 1HD},
+  \postcode{43870},
   \city{Amposta},

### 2016-09-10 article.tex
Write SWT in full fixed
Added funding acknowledgment
Fixed $ -> %

@@
 % previous titles:
-$
+%
 % The Chemistry Development Kit (CDK): an open-source Java library for Chemo- and Bioinformatics.
@@
 \author[
-   addressref={wi_mit},
+   addressref={wi_mit},                             % ORCID: 0000-0002-6940-3006
    email={pluskal@wi.mit.edu}
-]{\inits{TP}\fnm{Tomáš}~\snm{Pluskal}}
+]{\inits{TP}\fnm{Tomáš} \snm{Pluskal}}
 \author[
@@
    email={steinbeck@ebi.ac.uk}
-]{\inits{CS}\fnm{Christoph}~\snm{Steinbeck}}
+]{\inits{CS}\fnm{Christoph} \snm{Steinbeck}}
 
@@
   toolkit, but other tools use different widget toolkits. For example, Bioclipse
-  is based on Eclipse which uses the SWT~\cite{spjuth2007bioclipse}. %FIXME: write SWT in full
+  is based on Eclipse which uses the Standard Widget Toolkit (SWT)~\cite{spjuth2007bioclipse}.
 
@@
 and larger contributions to the CDK library. A full list of contributors is
-found in the AUTHORS file~\cite{AUTHORS}. OS acknowledges support from the Swedish strategic research programs eSSENCE and Swedish e-Science Research Center (SeRC).
+found in the AUTHORS file~\cite{AUTHORS}. OS acknowledges support from the
+Swedish strategic research programs eSSENCE and Swedish e-Science Research
+Center (SeRC). TP is a Simons Foundation Fellow of the Helen Hay Whitney
+Foundation.
 

### 2016-09-10 article.tex
Fixed a char.
Added Tomáš' ORCID.

@@
    email={pluskal@wi.mit.edu}
-]{\inits{TP}\fnm{Tomáš} \snm{Pluskal}}
+]{\inits{TP}\fnm{Tomáš}~\snm{Pluskal}}
 \author[
@@
 and larger contributions to the CDK library. A full list of contributors is
-found in the AUTHORS file~\cite{AUTHORS}. OS acknowledges support from the
-Swedish strategic research programs eSSENCE and Swedish e-Science Research
-Center (SeRC). TP is a Simons Foundation Fellow of the Helen Hay Whitney
-Foundation.
+found in the AUTHORS file~\cite{AUTHORS}.
+OS acknowledges support from the Swedish strategic research programs eSSENCE and Swedish e-Science Research Center (SeRC).
+TP is a Simons Foundation Fellow of the Helen Hay Whitney Foundation.
 

### 2016-09-10 article.tex
Substructure/SMARTS section

@@
 
-  \subsubsection*{Substructure searching}
-
-  TODO: iterator and new SMARTS
+  \subsubsection*{Substructure and SMARTS matching}
+
+  Substructure matching is fundamental to many other procedures.
+Since CDK 1.2 functionality has been added to handle the SMARTS query language.
+The SMARTS language is supported well including: Stereochemistry, Component Grouping, Atom Maps (to match reaction transformations).
+A new \textit{Pattern} API has been added to CDK \cdkversion{}, the API simplifies finding, filtering, and transform results.
+The API is immutable allowing a pattern to be initialized once and then matched against several molecules or reactions across multiple threads, during initialization the pattern is inspected as to determine what invariants will be needed (e.g.
+ring size) and only calculating what is needed.
+The internal matching algorithms provide a lazy iterator, the lazy evaluation means the next match is only calculated when it is needed.
+The API handles reactions in addition to molecules, both can be specified as either queries or targets.
+
+  \begin{verbatim}
+    // initialize SMARTS pattern API
+    Pattern pattern = 
+      SmartsPattern.findSubstructure("O=[C,N]aa[N,O;!H0]");
+
+    IAtomContainer mol = ...;
+    IReaction      rxn = ...;
+
+    // check if the query matches, molecules and reactions
+    boolean mMatch = pattern.matches(mol);
+    boolean rMatch = pattern.matches(rxn);
+
+    // lazily iterate over all unique atom matches as query atom
+    // index bijection to atoms in 'mol', 'rxn' can also be used
+    for (int[] m : pattern.matchAll(mol)
+                          .uniqueAtoms()) {
+
+    }
+  \end{verbatim}
+
+  CDK \cdkversion{} includes large improvements to algorithm efficiency, an isolated benchmark demonstrates relative improvements between recent CDK versions and to other open source toolkits (Table~5\ref{tab:smigrep}).
+The efficiency improvements are a combination of optimising data structures and key molecule processing algorithms (e.g.
+Kekulisation and Aromaticity) needed before a SMARTS match can be run.
+
+  % More on performance?
+  % http://efficientbits.blogspot.co.uk/2013_10_01_archive.html
+  % http://efficientbits.blogspot.co.uk/2013_11_01_archive.html
 
@@
 
+\subsection*{Table 5 - SMARTS matching performance.}\label{tab:smigrep}
+  Total time taken to match the SMARTS \texttt{O=[C,N]aa[N,O;!H0]} against $\sim$1.5 million SMILES from
+  ChEMBL 21.
+The reported time is for single core execution, measure with the Unix time utility (Cent OS 7, Intel Core i7-4790 CPU @ 3.60GHz).
+The times for recent versions of OpenBabel and RDKit are included for reference but this is not intended to be a systematic comparison.
+The difference in number found is due to variations in valence and aromaticity models.
+  \vskip \baselineskip
+
+\begin{tabular}{lll}
+Toolkit                   & Time   & Matched \\
+\hline
+CDK 1.4.9                 & 64m3s  & 69,782  \\
+CDK \cdkversion{}         & 1m7s   & 62,518  \\
+OpenBabel 2.3.90          & 13m55s & 61,906  \\
+RDKit 2016.3.4            & 6m18s  & 62,402  \\
+\end{tabular}
+
 \end{backmatter}

### 2016-09-10 article.tex
[optional] Not sure why AMBIT-smarts/AMBIT-tautomer are split out looks a bit biased to the other tools.

@@
 jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
-PaDEL~\cite{yap2011padel}, AMBIT-SMARTS~\cite{jeliazkova2011ambitsmarts}, AMBIT-Tautomer~\cite{kochev2013ambit},
-ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016}, WhichCyp~\cite{Rostkowski2013},
-MetaPrint2D~\cite{Carlsson2010}, and the IUPHAR/BPS Guide to Pharmacology database~\cite{Southan2016}.
+PaDEL~\cite{yap2011padel}, ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016},
+WhichCyp~\cite{Rostkowski2013}, MetaPrint2D~\cite{Carlsson2010}, and the IUPHAR/BPS Guide to Pharmacology database~\cite{Southan2016}.
 Other tools used previous versions already, and follow
 newer releases of the CDK, such as Bioclipse~\cite{spjuth2007bioclipse,
-spjuth2009bioclipse} and AMBIT~\cite{jeliazkova2011ambit}. While the above
+spjuth2009bioclipse} and AMBIT~\cite{jeliazkova2011ambit,jeliazkova2011ambitsmarts,kochev2013ambit}. While the above
 tools suggest that the CDK is primarily used to create general tools, it is

### 2016-09-10 article.tex
Update Chris' group name, need to have the EMBL in there also IIRC.
Address for John May.

@@
 \author[
-   addressref={aff2},
-   email={john@nextmovesoftware.com}
+   addressref={nm},
+   email={john.may@cantab.net}
 ]{\inits{JWM}\fnm{John W} \snm{May}}
@@
 }
-\address[id=aff2]{
-  %\orgname{??},
+\address[id=nm]{
+  \orgname{NextMove Software Ltd},
   %\street{D\"{u}sternbrooker Weg 20},
-  %\postcode{24105},
-  %\city{Kiel},
+  \postcode{CB4 0EY},
+  \city{Cambridge},
   \cny{UK}
@@
 \address[id=ebi]{
-  \orgname{Chemoinformatics and Metabolism team, European Bioinformatics Institute},
+  \orgname{Metabolomics and Molecular Informatics, EMBL-EBI},
   %\street{A. Kanchev 4},

### 2016-09-11 article.tex
Cite RDKit/Open Babel

@@
 The reported time is for single core execution, measure with the Unix time utility (Cent OS 7, Intel Core i7-4790 CPU @ 3.60GHz).
-The times for recent versions of OpenBabel and RDKit are included for reference but this is not intended to be a systematic comparison.
+The times for recent versions of Open Babel~\cite{OBoyle2011} and RDKit~\cite{rdkit} are included for reference but this is not intended to be a systematic comparison.
 The difference in number found is due to variations in valence and aromaticity models.
@@
 CDK \cdkversion{}         & 1m7s   & 62,518  \\
-OpenBabel 2.3.90          & 13m55s & 61,906  \\
+Open Babel 2.3.90         & 13m55s & 61,906  \\
 RDKit 2016.3.4            & 6m18s  & 62,402  \\

### 2016-09-12 formula_generator_evaluation/measure_runtime.sh
Fixed a minor error

@@ -5,7 +5,6 @@ TOLERANCE=$2
 MIN=0
 MAX=10000
 
-rm -rf 
 echo Running CDK on file $MASS_FILE +-${TOLERANCE} Da
 time -p java -cp "CDK/*:CDK" CDKFormulaGeneratorCLI $MASS_FILE $TOLERANCE C${MIN}-${MAX}H${MIN}-${MAX}N${MIN}-${MAX}O${MIN}-${MAX}S${MIN}-${MAX}P${MIN}-${MAX} > outCDK.txt
 echo

### 2016-09-18 article.tex
Added Lars Carlsson, whom was involved in the signature work in the CDK
Added code examples and removed the note about being slow, as it's fast enough (closes #36)

@@
 \author[
+   addressref={azg},
+   %email={kai.duehrkop@uni-jena.de}
+]{\inits{LC}\fnm{Lars}~\snm{Carlsson}}
+\author[
    addressref={jena},
@@
   \cny{Sweden}
+}
+\address[id=azg]{
+  %\orgname{Department of Informatics, University of Leicester},
+  %\street{},
+  %\postcode{LE1 7RH},
+  \city{Gotenburg},
+  \cny{SE}
 }
@@
   and undefined double bond positions if hybridization information is provided instead.
-  That makes the code complex, and at times slow, but so far shown sufficiently
-  performant for many applications. Possibly, in the future
-  alternative algorithms may be implemented, while that would again increase the maintenance
-  burden.
+  That makes the perception code flexible but also more complex. Alternative
+  algorithms to do the perception have not been explored. This reference
+  implementation can be used on a single atom:
+
+\begin{verbatim}
+for (IAtom atom : molecule.atoms()) {
+  type = matcher.findMatchingAtomType(molecule, atom);
+}
+\end{verbatim}
+
+And on a full molecule, in which case the list of types is ordered in the same
+order as the atoms in the molecule object:
+
+\begin{verbatim}
+types = matcher.findMatchingAtomTypes(molecule);
+\end{verbatim}
 

### 2016-09-18 article.tex
Combined to sentences, to fix the grammar (closes #37)
Clarified an unclear sentence (closes #38) and added missing references

@@
 
-The API has hardly changed. The internals, however, have significantly. Besides
+The API has hardly changed, but the internals have significantly. Besides
 the aforementioned overlap resolution, the engine also handles large ring
@@
 
-Originally, the CDK was developed as a shared library between JChemPaint and Jmol. The former
-used a MVC approach with an event-passing mechanism to update the view when the model was
+Originally, the CDK was developed as a shared library between
+JChemPaint~\cite{krause2000jchempaint} and Jmol~\cite{Willighagen2007jmol,Hanson2010}.
+JChemPaint used a MVC approach with an event-passing mechanism to update the view when the model was
 changed. This can cause an cascade of change events being passed around. To address,

### 2016-09-18 article.tex
Removed unfinished and wrong info (closes #29) and commented out an empty section

@@
 the aforementioned overlap resolution, the engine also handles large ring
-systems much better, using a honeycomb layout approach~\cite{FIXME}, % FIXME
+systems much better,
 and handles stereochemistry of the input. Previously, double bond stereochemistry
@@
 
-\subsubsection*{Hash Codes}
+%% \subsubsection*{Hash Codes}
 

### 2016-09-18 article.tex
Cleaned up the 'code maintenance' intro section (closes #39)

@@
 
-As the library grew over the years, so did the maintenance become more complex. Increasingly,
-the main branch did not compile, and bug fixing become increasingly difficult, as fixing a bug
-in one part of the code, broke some other code which made the wrong assumption about the first
-code.
-
-To address these issues, we have adopted a number of coding standards. By no means there are
-meant to implement the best practices of source code development; instead, they attempt to find
-a balance between increasing code maintainability and being flexible enough to allow efficient
-code development. However, we appreciate the subjective nature of this statement, and some
-adopted guidelines have been heavily discussed in the community.
-The next sections describe some approaches the project have adopted that allows us to
-maintain the CDK library as it is today.
-
-However, perhaps the biggest factor in improved code quality is that we
+As the library grew over the years, so did the complexity of the maintenance:
+Increasingly frequent, the main branch faied to compile, and bug fixing become
+more difficult with unexpected side effects. Often fixing a bug in one part of
+the code, broke some other code which made the incorrect assumptions about the
+fixed code. Such is inevitable when the number of developers increases.
+
+To address these issues, we have adopted a number of coding standards. By no
+means these are meant to implement the best practices of source code development,
+which is not practically unfeasible with limited resources; instead, they attempt
+to find a balance between increasing code maintainability and being flexible
+enough to allow efficient code development. However, we appreciate the subjective 
+nature of this statement, and some adopted guidelines have been heavily discussed
+and debated in the CDK community.
+
+Arguably, perhaps the biggest factor in improved code quality is that we
 instantiated a peer review process where any functionality changing patch is
@@
 is supported by a number of automated validations steps as outlined below.
+
+The next sections describe some approaches the project have adopted that allows
+us to maintain the CDK library as it is today.
 

### 2016-09-18 article.tex
Cite paper using the CDK for reaction fingerprints (closes #32)

@@
 important to note that it is also used to solve specific questions, like finding
-the maximally bridging rings in chemical structures~\cite{Marth2015}.
+the maximally bridging rings in chemical structures~\cite{Marth2015}, and
+defining fingerprints for organic reactions~\cite{Segler2016}.
 

### 2016-09-18 article.tex
Added address of Lars (closes #47)
Fixed a bibtex ID clash

@@
 \author[
-   addressref={azg},
-   %email={kai.duehrkop@uni-jena.de}
+   addressref={azg},                                % ORCID: 0000-0001-9491-4134
+   email={Lars.A.Carlsson@astrazeneca.com}
 ]{\inits{LC}\fnm{Lars}~\snm{Carlsson}}
@@
 \address[id=azg]{
-  %\orgname{Department of Informatics, University of Leicester},
+  \orgname{AstraZeneca, Innovative Medicines & Early Development, Quantitative Biology},
   %\street{},
   %\postcode{LE1 7RH},
-  \city{Gotenburg},
+  \city{Mölndal},
   \cny{SE}
@@
 
-Open Source cheminformatics has made significant steps forward recently~\cite{OBoyle2011}.
+Open Source cheminformatics has made significant steps forward recently~\cite{OBoyle2011b}.
 The Chemistry Development Kit (CDK) is one of the tools under the Blue Obelisk wings,

### 2016-09-18 article.tex
Properly escape the ampersand
Removed a duplicate period

@@
 \address[id=azg]{
-  \orgname{AstraZeneca, Innovative Medicines & Early Development, Quantitative Biology},
+  \orgname{AstraZeneca, Innovative Medicines \& Early Development, Quantitative Biology},
   %\street{},
@@
 %% To tabulate, r=read, w=write
-      \subsection*{Table 2 - CTfile Format Support..}\label{tab:ctfileFormats}
+      \subsection*{Table 2 - CTfile Format Support.}\label{tab:ctfileFormats}
 

### 2016-09-25 article.tex

@@
 Natural Products Social Molecular Networking database~\cite{wang2016}. The mass
-tolerance was set to 0.001 or 0.01 Da. The CDK's Round-Robin formula generator
-outperformed the other methods in all cases, despite running in a single thread
-(PFG utilizes multiple threads).
+tolerance was set to 0.001 or 0.01 Da. The CDK \cdkversion{}'s Round-Robin 
+formula generator outperformed the other methods in all cases, despite running
+in a single thread (PFG utilizes multiple threads). The performance gain of
+the Round Robin algorithm was particularly apparent when narrow mass ranges 
+were queried (e.g., $\pm 0.001$ Da), thus showing its suitability for
+applications in high-resolution mass spectrometry.
 

### 2016-09-25 article.tex
Fixed syntax of URL
'Preprint'-like mode
Comment on peer review

@@
 \NeedsTeXFormat{LaTeX2e}[1995/12/01]
-\documentclass[10pt]{bmcart}
+%\documentclass[10pt]{bmcart}
+\documentclass[doublespacing,linenumbers]{bmcart}
 %\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
@@
 
-The project continues to use PMD (url{http://pmd.sf.net/}) for code quality checking,
+The project continues to use PMD (\url{http://pmd.sf.net/}) for code quality checking,
 but deviates from the default rules. For example, we are more liberal with 
@@
 as the main source code hosting service
-where we can use novel approaches for commenting on code, pull requests, etc.
+where we can use novel approaches for commenting on code (peer review),
+pull requests, etc.
 These new features simplify our code review process.

### 2016-09-25 article.tex
Added missing Signature code examples

@@
   descriptor can also be canonicalized to provide isomorphism-independent
-  representations~\cite{Faulon2004}. Signatures can be calculated for atoms with:
-
-\begin{verbatim}
+  representations~\cite{Faulon2004}. Signatures of depth two can be calculated
+  for atoms with:
+
+\begin{verbatim}
+for (IAtom atom : molecule.atoms()) {
+  String signature = new AtomSignature(atom, 2, molecule).toCanonicalString();
+}
 \end{verbatim}
@@
 \begin{verbatim}
+String signature = new MoleculeSignature(molecule).toCanonicalString();
 \end{verbatim}
@@
   Spjuth2011DS,Moghadam2015,Alvarsson2014,Spjuth2012OS,Norinder2013,Alvarsson2016}.
-  The fingerprint for a molecule can be calculated with:
-
-\begin{verbatim}
-\end{verbatim}
 

### 2016-09-25 article.tex
No code example
Added missing figure number

@@
   and as methods on the DepictionGenerator class. This includes substructure
-  coloring, exemplified with this example reaction (see Figure~FIXME\ref{fig:depiction}):
-
-\begin{verbatim}
-TODO
-\end{verbatim}
+  coloring, exemplified with an example reaction shown in Figure~4\ref{fig:depiction}.
 
@@
 tetrahedral stereochemistry. These features are exemplified by the following
-code and the resulting layout depiction in Figure~\ref{fig:sdgstereo}.
+code and the resulting layout depiction in Figure~6\ref{fig:sdgstereo}.
 

### 2016-09-25 article.tex
\ref is for referecing, \label for defining labels
Typo fix

@@
 Mixtures, and Data. The most familiar Sgroups from an end user perspective are structure 
-repeat units (e.g. bracketing) and abbreviations (Figure~8\label{fig:sgroups}). CDK \cdkversion{} adds 
+repeat units (e.g. bracketing) and abbreviations (Figure~8\ref{fig:sgroups}).
+CDK~\cdkversion{} adds 
 supports for representation, reading, writing, and depiction.
@@
 As the library grew over the years, so did the complexity of the maintenance:
-Increasingly frequent, the main branch faied to compile, and bug fixing become
+Increasingly frequent, the main branch failed to compile, and bug fixing become
 more difficult with unexpected side effects. Often fixing a bug in one part of

### 2016-09-25 article.tex
A number of small tweaks

@@
 
-\def \cdkversion {\textbf{2.0}}
+%\def \cdkversion {\textbf{2.0}}
+\def \cdkversion {2.0}
 
@@
 cheminformatics library, and show that such collaborative projects can exist over
-a long period. We have taken advantage from the community support, and show
+a long period and result in a high-quality and fast library. We have taken
+advantage from the community support, and show
 that an open source cheminformatics project can act as a peer reviewed
@@
                           .uniqueAtoms()) {
-
+       ...
     }
@@
 CDK~\cdkversion{} adds 
-supports for representation, reading, writing, and depiction.
+supports for representation, reading, writing, and depiction of Sgroups.
 
@@
 Since the last CDK publication in 2006, the library has been improved in many
-aspects. The functionality has seen many additions and rewrites, making the
-library more functional and a lot more faster then before. Updates on the common
+aspects. The functionality has seen numerous additions and rewrites, making the
+library more functional and a lot faster than before. Updates on the common
 SMILES and molfile formats and the improved structure diagram generation are
@@
 
-With 93 contributors, a long list of tools based on the CDK, and hundreds
-of article citations, the CDK is alive and kicking.
+With more than 90 contributors, a long list of tools based on the CDK, and
+hundreds of article citations, the CDK is alive and kicking.
 

### 2016-09-26 pubcounts.R
updated R code

@@ -3,6 +3,7 @@ library(reshape2)
 x <- read.csv('pubcounts.csv', header=TRUE)
 ggplot(x, aes(x=as.factor(year),y=cdk2))+
   geom_bar(stat='identity', fill='beige', colour='black', position='dodge')+
-  geom_text(aes(label=cdk2,x=as.factor(year),y=cdk2),
+  geom_text(aes(label=cdk2,x=as.factor(year),y=cdk2, ymax=cdk2),vjust=1.25,
             position = position_dodge(width=1))+
-  xlab("Year")+ylab("Number of citations")
+  xlab("Year")+ylab("Number of citations")#+ylim(c(0, 60))
+ggsave("cdk-paper-citation-growth.png")

### 2016-09-26 article.tex
Reverted to the main style, as the others are too hard to read
Reworked abstract Fixed width spec of multirow element

@@
 \NeedsTeXFormat{LaTeX2e}[1995/12/01]
-%\documentclass[10pt]{bmcart}
-\documentclass[doublespacing,linenumbers]{bmcart}
+\documentclass[10pt]{bmcart}
 %\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
@@
 \parttitle{Background}
-Cheminformatics is a well-established field with many applications in chemistry,
-biology, drug discovery, and others. The Chemistry Development Kit (CDK) has
-become a widely used Open Source cheminformatics toolkit, providing various models to represent
-chemical structures, of which the chemical graph is essential. However,
-in the first five years of the project increased so much in size that
-interdependencies between components grew unmanageable large, resulting
-in unpredictable instabilities.
+The Chemistry Development Kit (CDK) is a 
+widely used Open Source cheminformatics toolkit, providing data
+structures to represent chemical concepts along with methods to
+manipulate such structures and perform computations on them. The
+library implements a wide variety of cheminformatics algorithms
+ranging from chemical structure canonicalization to molecular
+descriptor calculations and pharmacophore perception. Over the
+last five years, the code based has grown significantly resulting in
+many, complex interdependencies between components and poor
+performance of many algorithms. 
 \parttitle{Results}
-We here report improvements to the CDK since the 1.2 release series made
-to accommodate both the increased complexity of the library, as well as
-significant improvements of and additions to the functionality of the
-library. Second, we outline how the CDK evolved with respect to
-quality control and the approach we have adopted to ensure stability, including
-a peer review mechanism.
-Additionally, a selection of the new APIs that have been introduced will
-be discussed: atom type perception, substructure searching, molecular
+We report improvements to the CDK since the 1.2 release series,
+specifically addressing the increase functional complexity and poor
+performance. We first summarize the addition of new, and
+improvements to existing functionality that has led to significant
+improvement in performance and support for important chemistry
+features including atom type perception, substructure searching, molecular
 fingerprints, rendering of molecules, and handling of molecular formulas.
+Second, we outline how the CDK has evolved with respect to
+quality control and the approaches we have adopted to ensure stability, including
+a code review mechanism.
 \parttitle{Conclusions}
-With this paper we have shown the continued effort to provide a free, Open Source
+This paper highlights the our continued efforts to provide a community
+driven, Open Source
 cheminformatics library, and show that such collaborative projects can exist over
-a long period and result in a high-quality and fast library. We have taken
-advantage from the community support, and show
-that an open source cheminformatics project can act as a peer reviewed
+extended periods of time, resulting in a high-quality and performant
+library. By taking 
+advantage of community support and contributions, we show 
+that an Open Source cheminformatics project can act as a peer reviewed
 publishing platform for scientific computing software.
@@
     \begin{tabular}{llllllll}
-	\multirow{2}{*}{ Input } & \multirow{2}{*}{ Mass tolerance ($\pm$ Da) } & \multicolumn{3}{c}{ \# of generated formulas } & \multicolumn{3}{c}{ Runtime (s) } \\
+	\multirow{2}{1in}{ Input } & \multirow{2}{1in}{ Mass tolerance ($\pm$ Da) } & \multicolumn{3}{c}{ \# of generated formulas } & \multicolumn{3}{c}{ Runtime (s) } \\
 	& & HR2 & PFG & CDK & HR2 & PFG & CDK \\

### 2016-09-26 article.tex
Spellchecked

@@
   CircularFingerprinter~\cite{rogers2010extended, Clark2014}      & $\checkmark$ & $\checkmark$  & \cdkversion{}     & $1\,024$ / $2^{32}$%
-\footnote[1]{For the CirkularFingerprinter the bit version is folded to 1024 whereas the count version is unfolded.} \\
+\footnote[1]{For the CircularFingerprinter the bit version is folded to 1024 whereas the count version is unfolded.} \\
   EStateFingerprinter~\cite{Hall1995}       & $\checkmark$ &               & 1.2.0       & $79$            \\

### 2016-09-26 article.tex
Modifed some of the intro

@@
 
-Open Source cheminformatics has made significant steps forward recently~\cite{OBoyle2011b}.
-The Chemistry Development Kit (CDK) is one of the tools under the Blue Obelisk wings,
-and previously documented version have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
-For example, new tools expose CDK functionality in larger platforms such as Cinfony~\cite{OBoyle2008},
-rcdk~\cite{Guha2007}, and plugins for Taverna~\cite{Truszkowski2011},
-KNIME~\cite{Beisken2013}, ChemViz2 for Cytoscape~\cite{ChemViz2}, and LICSS for
-Microsoft Excel~\cite{Lawson2012}.
-Furthermore, projects with specific functionality have emerged building on the CDK, including
+The Open Source cheminformatics community has made significant steps
+forward recently~\cite{OBoyle2011b} as evidenced by the growing number
+of tools and underlying toolkits, along with the usage of these
+software components in a variety of applications.
+The Chemistry Development Kit (CDK) is one of the tools developed
+under the aegis of the Blue Obelisk 
+and previously documented versions have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
+Adoption of the CDK ranges from inclusion of CDK functionality in
+wrapper platforms such as Cinfony~\cite{OBoyle2008}, incorporation
+within the R environment (rcdk~\cite{Guha2007}), and as plugins for
+Taverna~\cite{Truszkowski2011}, 
+KNIME~\cite{Beisken2013}, Cytoscape (ChemViz2~\cite{ChemViz2}), and for
+Microsoft Excel (LICSS~\cite{Lawson2012}).
+In contrast to scenarios that have made CDK functionality available in
+larger systems, a number of projects have employed the CDK as a
+general cheminformatics toolkit. Examples include 
 jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
@@
 WhichCyp~\cite{Rostkowski2013}, MetaPrint2D~\cite{Carlsson2010}, and the IUPHAR/BPS Guide to Pharmacology database~\cite{Southan2016}.
-Other tools used previous versions already, and follow
-newer releases of the CDK, such as Bioclipse~\cite{spjuth2007bioclipse,
-spjuth2009bioclipse} and AMBIT~\cite{jeliazkova2011ambit,jeliazkova2011ambitsmarts,kochev2013ambit}. While the above
-tools suggest that the CDK is primarily used to create general tools, it is
-important to note that it is also used to solve specific questions, like finding
+A number of such tools were initially developed using older versions
+of the CDK and are updated to new releases as they are made
+available. Examples include Bioclipse~\cite{spjuth2007bioclipse,
+spjuth2009bioclipse} and
+AMBIT~\cite{jeliazkova2011ambit,jeliazkova2011ambitsmarts,kochev2013ambit}. The
+CDK has also played a role in a number of chemical studies, such as finding
 the maximally bridging rings in chemical structures~\cite{Marth2015}, and
@@
 
-However, previous version of the CDK were written by a community with specific applications
-in mind, of which structure elucidation was one. It could therefore be rather competitive
-in some cheminformatics domains, such as fingerprinting~\cite{Clark2014,Cannon2006}, in other parts
-it could be inadequate, such in the capturing of stereochemistry.
+While the CDK has purported to be a general purpose cheminformatics
+toolkit, older versions were designed by a community with specific
+applications, primary among them being structure elucidation. In
+addition, an implicit goal of previous versions was to have the CDK
+serve as an educational resource to enable students of cheminformatics
+to understand the underlying algorithms. This resulted in certain
+functionalities, such as fingerprinting~\cite{Clark2014,Cannon2006},
+receiving more attention than others, such as stereochemistry. The
+outcome was significant variance in performance and features
+throughout the toolkit.
 

### 2016-09-27 article.tex
proofing

@@
   Along with the new stereochemistry representation, algorithms were required in several areas. Generally
-  a user does not need to invoked these procedures explicitly as they are called as needed within existing
+  a user does not need to invoke these procedures explicitly as they are called as needed within existing
   APIs:
@@
   particular, some functionality, even after Modularization, needed considerable
-  parts of the CDK library, making creation if a small-sized applet unfeasible.
+  parts of the CDK library, making creation of a small-sized applet unfeasible.
   Therefore, the rendering API was modularized to allow splitting up rendering
@@
   various models implemented and the code was scattered throughout the library.
-  This was unified resulting in a three models, of which two are based on the
+  This was unified resulting in three models, of which two are based on the
   CDK atom typer, to provide the number of electrons donated by each atom to

### 2016-09-28 article.tex
Some more clean ups for the intro. Fix [REF] items before submission

@@
 toolkit, older versions were designed by a community with specific
-applications, primary among them being structure elucidation. In
+applications in mind, primary among them being structure elucidation. In
 addition, an implicit goal of previous versions was to have the CDK
@@
 
-Continued community development of the CDK in the last few years, however, has shown that
-Open Source software is here to stay. The adoption of automatic build systems and
-quality control methodologies such as unit testing, automated source code validation,
-and peer review by fellow developers have greatly improved the stability of the library,
-At the expense of slowing down the development, it has allowed for cleaning up interdependencies
-between modules of code, and has reset the scalability of the development model,
-allowing for a lot of new functionality in code APIs while maintaining the quality
-of code depending on those core APIs.
-
-This has resulted in important improved functionality, such as a separately published
-InChI functionality~\cite{Spjuth2013} and greatly improved ring detection functionality~\cite{May2014},
-but also a core atom type perception module, covering a much wider set of elements than
-previous CDK versions and capturing many more charged and radical atom species,
-a more comprehensive fingerprinting API, many speed and stability improvements, and more.
+The growth of Open Source software over the last 10 years [REF] is evidence
+of the ability of communities of developers to develop systems and
+processes that lead to high quality software systems for long term
+use. The CDK is no different. The adoption of automatic build systems
+and quality control methodologies such as unit testing, automated
+source code validation, and peer review by fellow developers have
+greatly improved the stability of the library, While it has slowed
+development somewhat, it has allowed for cleaning up interdependencies
+between modules of code, and importantly, has improved the scalability
+of the development model.  This has resulted in significant new
+functionality in core APIs while maintaining the quality of code
+depending on those core APIs.
+
+Examples of new features supported by the improved development model
+include InChI functionality~\cite{Spjuth2013}, greatly improved ring
+detection algorithms~\cite{May2014}, improvements to the core atom
+type perception module that now covers a much more comprehensive set
+of elements, charge states and radical species than previous versions,
+a more comprehensive fingerprinting API and many speed and stability
+improvements.
 

### 2016-09-29 article.tex
smore more edits

@@
 
+In this section we described the specifics of new APIs and
+improvements to pre-existing methods that are available in the latest
+CDK.  We then discuss how we have improved and formalized the
+development model for the project using unit testing, code review and
+guidelines for handling version control. Finally we report on the
+availability of binary distributions of the library, allowing users to
+include specific modules (and their dependencies) of the CDK in their
+own projects (as opposed to developers who work on the CDK library
+itself).
+
 \subsection*{New APIs and improved implementations}
@@
 We here outline various new and improved APIs in the CDK library since the two previous
-publications.
+publications (corresponding to version XXX)
 
@@
 
-  Atom type perception is a core cheminformations functionality: the atom types describe
-  chemical features of atoms, such as the number of neighbors, possible formal charges,
-  (approximate) hybridization, electron distribution over orbitals, etc. However, the
-  CDK commonly integrated atom type perception into algorithms, resulting in multiple
-  and diverging copies of similar atom type schemes. With the need to add many
-  new atom types for previously uncovered element, but also charged and radical atom
-  types, a new approach was needed.
-
-  This CDK version has isolated a new atom typing framework, removing the perception of
-  atom types from various algorithms, allowing the perception to be tested separately.
-  The new code defines the atom types using a list that specifies for each type
-  the element symbol, hybridization, formal charge, number of lone pairs, and
-  an enumeration of the bond orders (see Figure~1). This list of properties captures the
-  information needed for the various algorithms in the CDK. For example,
-  hybridization information can be used in certain aromaticity models (see later),
-  and the lone pair information is needed for resonance structure calculation
-  needed, for example, for Gasteiger $\pi$-charges.
-
-  A reference perception class, CDKAtomTypeMatcher, has been written that perceives these atom types, and
-  validates the perception automatically against the properties defined by the ontology.
-  This class handles a variety in types of missing information, as commonly resulting
-  from various (file) formats; for example, it can handle undefined hydrogen counts
-  and undefined double bond positions if hybridization information is provided instead.
-  That makes the perception code flexible but also more complex. Alternative
-  algorithms to do the perception have not been explored. This reference
-  implementation can be used on a single atom:
+  Atom type perception is a core cheminformations functionality: the
+  atom types describe chemical features of atoms, such as the number
+  of neighbors, possible formal charges, (approximate) hybridization,
+  electron distribution over orbitals and so on. However, previous
+  versions of the CDK implemented atom type perception as part of
+  different algorithms, resulting in duplicated and sometimes
+  divergent typing schemes. As a result it was cumbersome to add new
+  atom types and implement support for new charged and radical species
+  in a consistent manner.
+
+  This CDK version has a new, centralized atom typing framework,
+  removing the perception of atom types from various algorithms. This
+  allows for a consistent and extensive typing scheme, that can be
+  also be tested independently of other code.  The new code defines
+  the atom types using a list that specifies for each type the element
+  symbol, hybridization, formal charge, number of lone pairs, and an
+  enumeration of the bond orders (see Figure~\ref{fig:atomtype}). This list of
+  properties captures the information needed for the various
+  algorithms in the CDK. For example, hybridization information can be
+  used in certain aromaticity models (see later), and the lone pair
+  information is needed for resonance structure calculation needed,
+  for example, for Gasteiger $\pi$-charges.
+
+  A reference implementation, \texttt{CDKAtomTypeMatcher}, has been
+  written that perceives these atom types, and validates the
+  perception automatically against the properties defined by the
+  ontology.  This class handles a variety in types of missing
+  information, as commonly resulting from various (file) formats; for
+  example, it can handle undefined hydrogen counts and undefined
+  double bond positions if hybridization information is provided
+  instead.  That makes the perception code flexible but also more
+  complex. Alternative algorithms for atom typing have not been
+  explored. This reference implementation can be used on a single
+  atom:
 
@@
   level as a \texttt{\textbf{StereoElement}}. In abstract terms a stereo element describes local
-  geometry using a type, focus, carriers, and configuration (Figure~2\ref{fig:stereodatastructure}).
+  geometry using a type, focus, carriers, and configuration (Figure~\ref{fig:stereodatastructure}).
   Currently the most common types of stereochemistry are supported: Tetrahedral, Cis-trans isomerism 

### 2016-09-29 article.tex
reworked figures to use \figure env

@@
   perspective projections enabled, database entries currently consider distinct can be merged
-  (Figure~3\ref{fig:stereoprojections}). The perception is based on an algorithm briefly 
+  (Figure~\ref{fig:stereoprojections}).  The perception is based on an  algorithm briefly  
   described by~\cite{Karapetyan2015}.
@@
   and as methods on the DepictionGenerator class. This includes substructure
-  coloring, exemplified with an example reaction shown in Figure~4\ref{fig:depiction}.
+  coloring, exemplified with an example reaction shown in Figure~\ref{fig:depiction}.
 
@@
   number of long standing issues. Particularly, collision detection has been
-  greatly improved. Figure~5\ref{fig:sdg} shows a difference in output between
+  greatly improved. Figure~\ref{fig:sdg} shows a difference in output
+  between  
   the old code base, with and without overlap resolving, and with the new code.
@@
 Mixtures, and Data. The most familiar Sgroups from an end user perspective are structure 
-repeat units (e.g. bracketing) and abbreviations (Figure~8\ref{fig:sgroups}).
+repeat units (e.g. bracketing) and abbreviations (Figure~\ref{fig:sgroups}).
 CDK~\cdkversion{} adds 
@@
 on third-party libraries is given in Table~4\ref{tab:modules} and the dependencies
-between the CDK modules are depicted in Figure~6\ref{fig:deps}.
+between the CDK modules are depicted in Figure~\ref{fig:deps}.
 
@@
 
-\subsection*{Figure 2 - Relative storage of stereochemistry, the \textbf{type} and \textbf{focus} of stereochemsitry are fixed for a given stereocenter description but the \textbf{carriers} and \textbf{configuration} are relative.}\label{fig:stereodatastructure}
-The multiple rows for each stereochemistry type are different internal representation that would be considered equivalent.
-In the tetrahedral types, hydrogens may be suppressed in a molecular graph so the \textbf{focus} is reused in the \textbf{carriers} list as a placeholder.
-
+\begin{figure}[h!]
+  \caption{\csentence{Relative storage of stereochemistry, the \textbf{type} and
+    \textbf{focus} of stereochemsitry are fixed for a given
+    stereocenter description but the \textbf{carriers} and
+    \textbf{configuration} are relative.} 
+  The multiple rows for each
+  stereochemistry type are different internal representation that
+  would be considered equivalent. In the tetrahedral types,
+  hydrogens may be suppressed in a molecular graph so the
+  \textbf{focus} is reused in the \textbf{carriers} list as a
+  placeholder.}
+\centering
 \includegraphics{img/stereodesc_annotated.pdf}
-
-\subsection*{Figure 3 - The raw input files of CHEMBL23970 and CHEMBL444314 are displayed (ChEMBL 21).}\label{fig:stereoprojections}
-Without perceiving the stereochemistry indicated by Haworth projection in CHEMBL23970, the database entries are incorrectly considered distinct.
-Down stream aggregation databases mirror this separation (PubChem CID 5280, CID 65119).
-
+\label{fig:stereodatastructure}
+\end{figure}
+
+\begin{figure}[h!]
+  \caption{\csentence{The raw input files of CHEMBL23970 and CHEMBL444314 are displayed (ChEMBL 21).}
+    Without perceiving the stereochemistry indicated by Haworth
+    projection in CHEMBL23970, the database entries are incorrectly
+    considered distinct. Down stream aggregation databases mirror this
+    separation (PubChem CID 5280, CID 65119).}
+    \centering
     \includegraphics{img/projections_annotated.pdf}
-
-\subsection*{Figure 4 - Integrated example showing the rendering and SMILES
-parsing functionality. Example from U.S. Patent \textbf{US 2014 231770 A1} para 287.}\label{fig:depiction}
-
-%% SMILES=[CH3:17][CH:16]([CH3:18])[CH2:15][Mg+].[cH:7]1[cH:6][cH:5][c:4]2[c:9]([cH:8]1)[cH:10][cH:11][c:2]3[c:3]2[n:12][cH:13][n:14][c:1]3Cl.[Br-]>C/C(=C/C(=O)C)/[O-].C/C(=C/C(=O)C)/[O-].C/C(=C/C(=O)C)/[O-].[Fe+3].CN1CCCC1=O.C1CCOC1.Cl>[CH3:17][CH:16]([CH3:18])[CH2:15][c:1]1[c:2]2[cH:11][cH:10][c:9]3[cH:8][cH:7][cH:6][cH:5][c:4]3[c:3]2[n:12][cH:13][n:14]1 |f:0.2,3.4.5.6,SgD::cdk&#58;ReactionConditions:16 h&#44; RT|
-%% Kumada Coupling?
-
+    \label{fig:stereoprojections}
+\end{figure}
+
+\begin{figure}[h!]
+  \caption{\csentence{Integrated example showing the rendering and SMILES
+      parsing functionality. Example from U.S. Patent \textbf{US 2014 231770 A1} para 287.}}
+  \centering
   \includegraphics[width=\textwidth]{img/US20140231770A1_0287.pdf}
-
-\subsection*{Figure 5 - The improved structure diagram generation has improved
-code to solve overlap.}\label{fig:sdg}
+  \label{fig:depiction}
+\end{figure}
+
+\begin{figure}[h!]
+  \caption{\csentence{{The improved structure diagram generation has improved
+        code to solve overlap.}}
     The original SDG code used general heuristics (left) and the
@@
     make more rigorous changes, making the final output must more pleasing
-(right).
-
+    (right).}
     \includegraphics[width=\textwidth]{sdg.png}
-
-\subsection*{Figure 6 - Structure diagram generation for structures with
-double bond and tetrahedral stereochemistry.}\label{fig:sdgstereo}
-
+    \label{fig:sdg}
+\end{figure}
+
+\begin{figure}[h!]
+  \caption{\csentence{Structure diagram generation for structures with
+      double bond and tetrahedral stereochemistry.}}
+  \centering
   \includegraphics[width=0.3\textwidth]{sdg2.png}
-
-  \subsection*{Figure 7 - Dependencies between CDK modules.}\label{fig:deps}
+  \label{fig:sdgstereo}
+\end{figure}
+
+\begin{figure}[h!]
+  \caption{\csentence{Dependencies between CDK modules.}
     Visualization of the dependencies between CDK modules. For example,
@@
     modules have been left out: cdk-builder3dtools, cdk-legacy, and
-      cdk-depict.
-
+    cdk-depict.}
+  \centering
   \includegraphics[width=\textwidth]{cdkDeps.png}
-
-\subsection*{Figure 8 - Examples of Sgroups now captured by the CDK and encoded in molfiles and CXSMILES.
-a) Ethyl esterification fully expanded reaction.
-b) Using Sgroup abbreviations allows display short cuts and more compact depiction.
-c) An example of a structure repeat unit in \textit{DNA 5'-phosphate} (\textbf{CHEBI:4294})}
+  \label{fig:deps}
+\end{figure}
+
+\begin{figure}[h!]
+  \caption{\csentence{Examples of Sgroups now captured by the CDK and
+      encoded in molfiles and CXSMILES. a) Ethyl esterification fully
+      expanded reaction. b) Using Sgroup abbreviations allows display
+      short cuts and more compact depiction. c) An example of a
+      structure repeat unit in \textit{DNA 5'-phosphate}
+      (\textbf{CHEBI:4294})}}
+  \includegraphics[width=\textwidth]{img/sgroups.pdf}
   \label{fig:sgroups}
-
-\includegraphics[width=\textwidth]{img/sgroups.pdf}
+\end{figure}
 
@@
   dependencies on third-party libraries, and the major changes since
-  version~1.2. Dependencies between modules are depicted in Figure~4\ref{fig:deps}.
+  version~1.2. Dependencies between modules are depicted in Figure~\ref{fig:deps}.
   \vskip 1\baselineskip

### 2016-09-29 article.tex

--- 2016-09-28 article.tex
@@
 }
+\address[id=leicester]{
+  \orgname{Department of Informatics, University of Leicester},
+  %\street{},
+  %\postcode{LE1 7RH},
+  \city{Leicester},
+  \cny{UK}
+}
 \address[id=wi_mit]{
@@
   \city{Hinxton},
-  \cny{UK}
-}
-\address[id=leicester]{
-  \orgname{Department of Informatics, University of Leicester},
-  %\street{},
-  %\postcode{LE1 7RH},
-  \city{Leicester},
   \cny{UK}
@@
 toolkit, older versions were designed by a community with specific
-applications in mind, primary among them being structure elucidation. In
+applications, primary among them being structure elucidation. In
 addition, an implicit goal of previous versions was to have the CDK
@@
 
-The growth of Open Source software over the last 10 years [REF] is evidence
-of the ability of communities of developers to develop systems and
-processes that lead to high quality software systems for long term
-use. The CDK is no different. The adoption of automatic build systems
-and quality control methodologies such as unit testing, automated
-source code validation, and peer review by fellow developers have
-greatly improved the stability of the library, While it has slowed
-development somewhat, it has allowed for cleaning up interdependencies
-between modules of code, and importantly, has improved the scalability
-of the development model.  This has resulted in significant new
-functionality in core APIs while maintaining the quality of code
-depending on those core APIs.
-
-Examples of new features supported by the improved development model
-include InChI functionality~\cite{Spjuth2013}, greatly improved ring
-detection algorithms~\cite{May2014}, improvements to the core atom
-type perception module that now covers a much more comprehensive set
-of elements, charge states and radical species than previous versions,
-a more comprehensive fingerprinting API and many speed and stability
-improvements.
+Continued community development of the CDK in the last few years, however, has shown that
+Open Source software is here to stay. The adoption of automatic build systems and
+quality control methodologies such as unit testing, automated source code validation,
+and peer review by fellow developers have greatly improved the stability of the library,
+At the expense of slowing down the development, it has allowed for cleaning up interdependencies
+between modules of code, and has reset the scalability of the development model,
+allowing for a lot of new functionality in code APIs while maintaining the quality
+of code depending on those core APIs.
+
+This has resulted in important improved functionality, such as a separately published
+InChI functionality~\cite{Spjuth2013} and greatly improved ring detection functionality~\cite{May2014},
+but also a core atom type perception module, covering a much wider set of elements than
+previous CDK versions and capturing many more charged and radical atom species,
+a more comprehensive fingerprinting API, many speed and stability improvements, and more.
 

### 2016-09-29 article.tex
fixed some typos

--- 2016-09-29 article.tex
@@
 
-  Finally, a signatures fingerprint can be calculated for molecules, to allow
+  Finally, a signature fingerprint can be calculated for molecules, to allow
   similarity calculations. This can then be used in QSAR modeling~\cite{signaturefingerprints,
@@
   of some use cases, such as Java applets, and the rendering functionality. In
-  particular, some functionality, even after Modularization, needed considerable
+  particular, some functionality, even after modularization, needed considerable
   parts of the CDK library, making creation of a small-sized applet unfeasible.

### 2016-09-29 article.tex
fixed last figure env

@@
 tetrahedral stereochemistry. These features are exemplified by the following
-code and the resulting layout depiction in Figure~6\ref{fig:sdgstereo}.
+code and the resulting layout depiction in Figure~\ref{fig:sdgstereo}.
 
@@
 
-\subsection*{Figure 1 - Atom type information specified for a sp3-hybridized
-carbon.}\label{fig:atomtype}
+\begin{figure}[h!]
+\caption{\csentence{Atom type information specified for a sp3-hybridized
+carbon.}}
 \begin{verbatim}
@@
 \end{verbatim}
+  \label{fig:atomtype}
+\end{figure}
 

### 2016-09-29 article.tex
Better organization of the figure/table part
Make runtimes in Table 5 consistent with Table 1

@@
 
-\newpage
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\newpage
+\clearpage
 
@@
 formula generators on two different inputs with two different mass tolerance
-settings. For the set of small masses, $10\,000$ mass values in the range of
-$0$--$500$ Da were randomly selected from the Global Natural Products Social
-Molecular Networking database~\cite{wang2016}. For the set of large masses, $20$
-mass values in the range of $1\,500$--$3\,500$ Da were randomly selected from the same
+settings. For the set of small masses, 10,000 mass values in the range of
+0--500 Da were randomly selected from the Global Natural Products Social
+Molecular Networking database~\cite{wang2016}. For the set of large masses, 20
+mass values in the range of 1,500--3,500 Da were randomly selected from the same
 database. Formulas were generated using chemical elements C,H,N,O,P,S without
-bounds (the allowed atom count was set to $0$--$10\,000$ for each element). All
+bounds (the allowed atom count was set to 0--10,000 for each element). All
 heuristic filtering rules were disabled for the purpose of the evaluation. The
@@
 	& & HR2 & PFG & CDK & HR2 & PFG & CDK \\
-	$10\,000$ small masses & $0.001$ & $616\,846$ & $616\,846$ & $616\,843$ & $669$ & $168$ & $\mathbf{41}$ \\
-	$10\,000$ small masses & $0.01$ & $6\,163\,303$ & $6\,163\,302$ & $6\,163\,326$ & $689$ & $501$ & $\mathbf{212}$ \\
-	$20$ large masses & $0.001$ & $4\,912\,939$ & $4\,912\,939$ & $4\,912\,904$ & $26\,370$ & $1\,292$ & $\mathbf{177}$ \\
-	$20$ large masses & $0.01$ & $49\,128\,811$ & $49\,128\,810$ & $49\,128\,815$ & $26\,587$ & $3\,406$ & $\mathbf{1\,580}$ \\
+	10,000 small masses & 0.001 & 616,846 & 616,846 & 616,843 & 669 & 168 & \textbf{41} \\
+	10,000 small masses & 0.01 & 6,163,303 & 6,163,302 & 6,163,326 & 689 & 501 & \textbf{212} \\
+	20 large masses & 0.001 & 4,912,939 & 4,912,939 & 4,912,904 & 26,370 & 1,292 & \textbf{177} \\
+	20 large masses & 0.01 & 49,128,811 & 49,128,810 & 49,128,815 & 26,587 & 3,406 & \textbf{1,580} \\
     \end{tabular}
@@
     \end{minipage}
+
+\newpage
 
@@
 \begin{tabular}{lll}
-Toolkit                   & Time   & Matched \\
+Toolkit                   & Time (s)   & Matched \\
 \hline
-CDK 1.4.9                 & 64m3s  & 69,782  \\
-CDK \cdkversion{}         & 1m7s   & 62,518  \\
-Open Babel 2.3.90         & 13m55s & 61,906  \\
-RDKit 2016.3.4            & 6m18s  & 62,402  \\
+CDK 1.4.9                 & 3,843  & 69,782  \\
+CDK \cdkversion{}         & 67   & 62,518  \\
+Open Babel 2.3.90         & 835 & 61,906  \\
+RDKit 2016.3.4            & 378  & 62,402  \\
 \end{tabular}

### 2016-09-30 article.tex
Some more text fixes/modifications. Added todo notes

@@
 \usepackage{multirow}
+
+%% Comment out before submission 
+\usepackage[colorinlistoftodos,textsize=small]{todonotes}
 
@@
 
-  Many of the rendering options are available as parameters in the core API
-  and as methods on the DepictionGenerator class. This includes substructure
-  coloring, exemplified with an example reaction shown in Figure~\ref{fig:depiction}.
+  Many of the rendering options are available as parameters in the
+  core API and as methods on the DepictionGenerator class. This
+  includes substructure coloring, exemplified with an example reaction
+  shown in Figure~\ref{fig:depiction}.
 
@@
   The structure diagram layout has been rewritten and the new code solves a
-  number of long standing issues. Particularly, collision detection has been
+  number of long standing issues. In particular, collision detection has been
   greatly improved. Figure~\ref{fig:sdg} shows a difference in output
   between  
-  the old code base, with and without overlap resolving, and with the new code.
-  The new code can be used like:
+  the old code base, with and without overlap resolving, and with the
+  new code. The latest implementation simplifies the generation of 2D
+  coordinates as shown below:
 
@@
 
-The API has hardly changed, but the internals have significantly. Besides
-the aforementioned overlap resolution, the engine also handles large ring
-systems much better,
-and handles stereochemistry of the input. Previously, double bond stereochemistry
-was supported, but the new engine more efficiently uses this when creating 2D
-layouts. Furthermore, the engine assigns wedge bond information based on
-tetrahedral stereochemistry. These features are exemplified by the following
-code and the resulting layout depiction in Figure~\ref{fig:sdgstereo}.
+  While the API itself has not been significantly changed, the
+  internals have been revamped. In addition to improved overlap
+  resolution noted above, the engine appropriately handles large ring
+  systems and maintains input stereochemistry. While previous CDK
+versions supported 
+  double bond stereochemistry the new engine is more efficient in
+  using this information when generate 2D layouts. Furthermore, the
+  engine assigns wedge bond information based on tetrahedral
+  stereochemistry. These features are exemplified by the following
+  code and the resulting layout depiction in
+  Figure~\ref{fig:sdgstereo}.
 
@@
 
-  A chemical formula is the basic/simple chemical representation of a compound.
-It defines the number of isotopes or elements that constitute the composition
-of a compound without describing how atoms are bonded. With the rise of
-metabolomics it has become increasingly relevant to have full support for these
-in cheminformatics libraries~\cite{RojasCherto2011}.
+A chemical formula is the simplest chemical representation of a
+compound.  It defines the number of isotopes or elements that
+compose  a compound without describing how atoms
+are bonded. With the rise of metabolomics it has become increasingly
+relevant to have full support for these in cheminformatics
+libraries~\cite{RojasCherto2011}.
 
@@
 
-  The SMILES parsing has been replaced code in the external Beam project~\cite{Beam}.
+The SMILES parsing has been replaced by code from the external Beam project~\cite{Beam}.
 This BSD-licensed SMILES parser is a complete implementation of the SMILES
@@
 
-  Substructure matching is fundamental to many other procedures.
-Since CDK 1.2 functionality has been added to handle the SMARTS query language.
-The SMARTS language is supported well including: Stereochemistry, Component Grouping, Atom Maps (to match reaction transformations).
-A new \textit{Pattern} API has been added to CDK \cdkversion{}, the API simplifies finding, filtering, and transform results.
-The API is immutable allowing a pattern to be initialized once and then matched against several molecules or reactions across multiple threads, during initialization the pattern is inspected as to determine what invariants will be needed (e.g.
-ring size) and only calculating what is needed.
-The internal matching algorithms provide a lazy iterator, the lazy evaluation means the next match is only calculated when it is needed.
-The API handles reactions in addition to molecules, both can be specified as either queries or targets.
+Substructure matching is fundamental cheminformatics operation and
+plays a key role in many other functions such as fingerprint,
+descriptor generation and atom typing. Since CDK 1.2, functionality
+has been added to handle the SMARTS query language. The SMARTS
+language is supported well including advanced features such as
+stereochemistry, component grouping, and atom
+maps (to match reaction transformations). A new \textit{Pattern} API
+has been added to CDK \cdkversion{}, which simplifies finding,
+filtering, and transform results. The API is immutable allowing a
+pattern to be initialized once and then matched against several
+molecules or reactions across multiple threads. During initialization
+the pattern is inspected as to determine what invariants will be
+needed (e.g. ring size) and only calculating what is needed. The
+internal matching algorithms provide a lazy iterator, such that the
+next match is only computed when it is needed. The API handles
+reactions in addition to molecules, and both can be specified as
+either queries or targets.
 
@@
 
-  CDK \cdkversion{} includes large improvements to algorithm efficiency, an isolated benchmark demonstrates relative improvements between recent CDK versions and to other open source toolkits (Table~5\ref{tab:smigrep}).
-The efficiency improvements are a combination of optimising data structures and key molecule processing algorithms (e.g.
-Kekulisation and Aromaticity) needed before a SMARTS match can be run.
+CDK \cdkversion{} includes large improvements to algorithm
+efficiency. A simple benchmark experiment demonstrates the relative
+improvements between recent CDK versions and other open source
+toolkits (Table~5\ref{tab:smigrep}). The efficiency improvements are a
+combination of optimising data structures and key molecule processing
+algorithms (e.g. kekulisation and aromaticity) needed before a SMARTS
+match can be run.
 
@@
 
-  Ring finding is a key functionality in a cheminformatics library, and the CDK
-  knows a long history of ring finding~\cite{May2014}. Particularly, non-redundant
-  ring sets have seen particular interest, such as the smallest set of smallest
-  rings, for which the CDK implements two classical algorithms~\cite{Figueras1996,Berger2004}.
-  Recent work by one the authors has implemented a new, faster algorithm, allowing
-  searching for various types of (non-redundant) ring sets~\cite{May2014}. These
-  are available via the new Cycles API:
+Ring finding is a key functionality in a cheminformatics library, and
+the CDK knows a long history of ring finding~\cite{May2014}. In
+particular, non-redundant ring sets have seen particular interest,
+such as the smallest set of smallest rings, for which the CDK
+implements two classical algorithms~\cite{Figueras1996,Berger2004}.
+Recent work has implemented a new, faster algorithm, allowing
+searching for various types of (non-redundant) ring
+sets~\cite{May2014}. These are available via the new Cycles API:
 
@@
 
-  Aromaticity has seen many definitions in the past and for cheminformatics it
-  frequently is algorithmically defined. The outcome of an aromaticity calculation
-  depends on a number of atom type features and heuristics, which in literature
-  often are not well defined. Based on this information several different
-  algorithmic definitions of aromaticity can be defined. Older CDK versions had
-  various models implemented and the code was scattered throughout the library.
-  This was unified resulting in three models, of which two are based on the
-  CDK atom typer, to provide the number of electrons donated by each atom to
-  the delocalized $\pi$ system. The difference between these two models is
-  how double bonds pointing outwards from a ring are contributing.
-
-  The current CDK version further generalizes the idea that aromaticity is a
-  model, and provides an API that allows you to select out of various options
-  providing greater interoperability with other toolkits. The new
-  Aromaticity class allows to build a custom model by selecting and combining
-  options. For example, to reproduce the functionality of the previous
-  CDKAromaticity class:
+Aromaticity has seen many definitions in the past and for
+cheminformatics it frequently is algorithmically defined. The outcome
+of an aromaticity calculation depends on a number of atom type
+features and heuristics, which are often ambiguously defined in the
+published literature. Based on this information several different
+algorithmic definitions of aromaticity can be defined. Older CDK
+versions had various models implemented and the code was scattered
+throughout the library, resulting in an inconsistent approach to
+calling methods to compute aromaticity and a significant maintenance
+burden.  This was unified in the current version, resulting in three
+models, of which two are based on the CDK atom typer. The difference
+between these two models is how contributions from exocyclic double
+bonds are handled.
+
+The current CDK version further generalizes the idea that aromaticity
+is a model, and provides an API that allows the user to select one of
+several aromaticity models, leading to greater interoperability with
+other toolkits. The new \texttt{Aromaticity} class allows to build a
+custom model by selecting and combining options. For example, to
+reproduce the functionality of the previous \texttt{CDKAromaticity} class:
 
@@
 
-Here, the CDK model for counting donated electrons is used, as the ring systems
-the CDK previously included in aromaticity counting, which was limited in the 
-number of fused rings systems. However, an alternative aromaticity calculator
-that does consider all possible ring systems can now be easily created with:
+Here, the CDK model for counting donated electrons is used, along with
+the rings systems that were identified by the older algorithm in
+previous versions that was limited in the number of fused rings
+systems that were considered. However, an alternative aromaticity
+calculator that considers all possible ring systems can now be
+easily created with:
 
@@
 
-The molfile format is still very popular, despite it being a proprietary
+The molfile format is still very popular and despite it being a proprietary
 format it has become a de facto standard. The format forms the core of the larger
@@
 Originally, the CDK was developed as a shared library between
-JChemPaint~\cite{krause2000jchempaint} and Jmol~\cite{Willighagen2007jmol,Hanson2010}.
-JChemPaint used a MVC approach with an event-passing mechanism to update the view when the model was
-changed. This can cause an cascade of change events being passed around. To address,
-interfaces were introduced allowing multiple implementations of the core interfaces.
+JChemPaint~\cite{krause2000jchempaint} and
+Jmol~\cite{Willighagen2007jmol,Hanson2010}.  JChemPaint used a MVC
+approach with an event-passing mechanism to update the view when the
+model was changed. This can cause an cascade of change events being
+passed around. This was not always a desirable feature, especially for
+non-UI code. To address this, interfaces were introduced allowing
+multiple implementations of the core interfaces.
 % TODO: check if the interfaces have previously been defined
-The IChemObjectBuilders play an important role here, allowing implementations of the
-interfaces to be instantiated without the need of explicitly referencing those implementations.
-
-However, the CDK 1.0 and 1.2 implementations of the IChemObjectBuilder had one method for
-each constructor, resulting in very large interface. Moreover, the API changed whenever
+The \texttt{IChemObjectBuilders} play an important role here, allowing
+implementations of the interfaces to be instantiated without the need
+of explicitly referencing those implementations.\todo{The motivation
+  for IChemObjectBuilders is not well defined.}
+
+However, the CDK 1.0 and 1.2 implementations of the \texttt{IChemObjectBuilder} had one method for
+each constructor, resulting in a very large interface. Moreover, the API changed whenever
 a new class was introduced, and existing methods changed when constructors were updated.
-To simplify the API, the new IChemObjectBuilder collapsed all methods returning new
+To simplify the API, the new \texttt{IChemObjectBuilder} collapsed all methods returning new
 implementations into a single method, which takes as a first parameter the class of the
-interface that is wished to be constructed. All further parameters are passes as
-parameter to the class constructor.
-
-For example, to construct a new atom from its element symbol, one would now write:
+interface that is wished to be constructed. All further parameters are passed as
+parameters to the class constructor.
+
+For example, to construct a new atom from its element symbol, one
+would now write:\todo{might be good to display code using older version}
 
@@
 
-Increasingly, the CDK library is now written against the interfaces, and when new instanced
-are needed, these builders are being used. This allows to run a certain CDK-based
-application with a specific builder, aimed at a particular use case. The original
-implementation is available via the DefaultChemObjectBuilder, which creates
-classes that pass around change events, just like the original CDK data classes.
-However, a SilentChemObjectBuilder has been introduced that does not pass around
-change events, which makes code run about 10-20\% faster.
-The third builder is the DataDebugChemObjectBuilder which generates debug information
-for all changes to the content of the data classes. This can be useful for
-debugging and other forms of code inspection.
+The CDK library is now written against the interfaces\todo{this may
+  not mean to a non-expert Java coder}, and when new instances are
+needed, these builders are used. This allows to run a certain
+CDK-based application with a specific builder, aimed at a particular
+use case. The original implementation is available via the
+\texttt{DefaultChemObjectBuilder}, which creates classes that pass
+around change events, as in the older CDK versions.  However, a
+\texttt{SilentChemObjectBuilder} is available, that does not support
+change events, resulting in speedups of 10\%-20\%.  The third builder
+is the \texttt{DataDebugChemObjectBuilder} which generates debug
+information for all changes to the content of the data classes. This
+can be useful for debugging and other forms of code inspection.
 
 \subsubsection*{Molecular fingerprints}
-Also the molecular fingerprints in CDK has seen development since the 1.2
-release. Traditionally CDK fingerprints were represented using the
-\textit{BitSet} class shipping with Java. The benefits of that class include
-such things as already implemented methods for modifying bit sets using other
-bit sets. However the class keeps a vector of bits in memory. The solution was
-excellent for hashed, relatively small fingerprints, \textit{e.g.}, 1024 bits,
-\textit{i.e.}, 10 bits indexing space. However implementing a fingerprint
-designed to avoid collisions with, \textit{e.g.} a 32 bit indexing space using
-this approach would be highly memory-inefficient. To allow for multiple
-fingerprint representation implementations a bit fingerprint interface was
-introduced. Also, although fingerprints traditionally are bit vectors a count
-fingerprint was also introduced making fingerprints based on integer vectors
-supported in CDK as well. The fingerprints currently existing in CDK are listed
-in Table~3\ref{tab:fingerprints}.
+Molecular fingerprints are another area that seen significant
+development. Traditionally fingerprints were represented using the
+\texttt{BitSet} class from the Java library. While using this class
+allowed the use of pre-existing methods to manipulate bit strings, it
+keeps a vector of bits in memory. The solution was excellent for
+hashed, relatively small fingerprints, \textit{e.g.}, 1024 bits,
+\textit{i.e.}, 10 bits indexing space. However implementing a
+fingerprint designed to avoid collisions with, \textit{e.g.} a 32 bit
+indexing space using this approach would be highly
+memory-inefficient. To allow for multiple fingerprint representation
+implementations a bit fingerprint interface was introduced\todo{is
+  there any more information about this?}. Also,
+although fingerprints traditionally are bit vectors a count
+fingerprint was also introduced making fingerprints based on integer
+vectors supported in CDK as well. The fingerprints currently existing
+in CDK are listed in Table~3\ref{tab:fingerprints}.
 
@@
 
-As the library grew over the years, so did the complexity of the maintenance:
-Increasingly frequent, the main branch failed to compile, and bug fixing become
-more difficult with unexpected side effects. Often fixing a bug in one part of
-the code, broke some other code which made the incorrect assumptions about the
-fixed code. Such is inevitable when the number of developers increases.
-
-To address these issues, we have adopted a number of coding standards. By no
-means these are meant to implement the best practices of source code development,
-which is not practically unfeasible with limited resources; instead, they attempt
-to find a balance between increasing code maintainability and being flexible
-enough to allow efficient code development. However, we appreciate the subjective 
-nature of this statement, and some adopted guidelines have been heavily discussed
-and debated in the CDK community.
-
-Arguably, perhaps the biggest factor in improved code quality is that we
-instantiated a peer review process where any functionality changing patch is
-required to be reviewed by one independent, senior CDK developer for the development
-branch, and two reviewers for stable branches. This patch development system
-is supported by a number of automated validations steps as outlined below.
+As the library grew over the years, so did the complexity of the
+maintenance. The main branch frequently failed to compile and bug
+fixes became more onerous due to unexpected side effects.  Often
+fixing a bug in one part of the code, broke some other code which made
+the incorrect assumptions about the fixed code. With the increase in
+the CDK developer community, such issues were inevitable, in the
+absence of any formal coding and testing standards.
+
+To address these issues, we have adopted a number of coding
+standards. While not a comprehensive implementation of software
+engineering best practices, they attempt to find a balance between
+increasing code maintainability and being flexible enough to allow
+efficient code development. However, we appreciate the subjective
+nature of this statement, and some adopted guidelines have been
+heavily discussed and debated in the CDK community.
+
+Arguably, perhaps the biggest factor in improved code quality is a
+peer review process where any functionality changing patch is required
+to be reviewed by one independent, senior CDK developer for the
+development branch, and two reviewers for stable branches. This patch
+development system is supported by a number of automated validations
+steps as outlined below.
 
@@
 on third-party libraries is given in Table~4\ref{tab:modules} and the dependencies
-between the CDK modules are depicted in Figure~\ref{fig:deps}.
+between the CDK modules are depicted in
+Figure~\ref{fig:deps}.\todo{the dependency graph is not very legible}
 
@@
 
-  The quality of JavaDoc of the CDK was originally tested with DocCheck, and
+The quality of the JavaDocs was originally tested with DocCheck, and
 later replaced by a custom written tool called OpenJavaDocCheck. With the move
@@
 we adopted CheckStyle (\url{http://checkstyle.sourceforge.net/}). This tool
-  reports about missing documentation and on documentation which is not properly
+reports on missing documentation and on documentation which is not properly
 annotated in the Java source files.
@@
 
-Another change made over the past years is the move to a Git-based version
-control system. Advantages the projects makes advantage of is the distributed
-nature of Git repositories, and easier branching and making available of
-patches. GitHub (\url{https://github.com/cdk/cdk}) has replaced SourceForge
-as the main source code hosting service
-where we can use novel approaches for commenting on code (peer review),
-pull requests, etc.
-These new features simplify our code review process.
+Older versions of the CDK employed Subversion for version control. A
+few years back, the project switched to the Git version control
+system. A key advantage of this shift is the ability to have
+distributed repositories, easier branching and provision for
+patches. GitHub (\url{https://github.com/cdk/cdk}) has replaced
+SourceForge as the main source code hosting service where we can use
+novel approaches for commenting on code (peer review), pull requests,
+etc.  These new features simplify our code review process.
 
@@
 
-Another recent change is the choice of build system: we have moved away from
-Ant to Maven to deal with dependencies and compile and test the library.
-To keep the original idea of CDK modules, this move required to split up the
-source code in multiple source folders. Fortunately, modern integrated
-development environments have no trouble handling this, removing the original
-argument to have everything in one source folder.
-
-On the other hand, for many modules, the test code is now more closely
-linked to the code being tested: both reside in the same folder, though we
-adhere to the Maven custom to have src/main/java and a src/test/java folders.
-For a few modules, however, this solution introduces circular dependencies, in
-which case a separate Maven module is created for the tests.
+The build system has been converted from Ant to Maven. The shift was
+motivated by the easier dependency handling, cleaner separation of
+testing code from the main library and automated packaging. The move
+to modules necessitated splitting the original monolithic source code
+tree in to per-module source folders. While this makes the on-disk
+layout of the source code more complex, this is usually hidden by
+modern IDEs.
+
+As a result for many modules, the test code is now more closely linked
+to the code being tested: both reside in the same folder, though we
+adhere to the Maven custom to have \texttt{src/main/java} and a
+\texttt{src/test/java} folders.  For a few modules, however, this
+solution introduces circular dependencies, in which case a separate
+Maven module is created for the tests.
 
@@
 
-Since the last CDK publication in 2006, the library has been improved in many
-aspects. The functionality has seen numerous additions and rewrites, making the
-library more functional and a lot faster than before. Updates on the common
-SMILES and molfile formats and the improved structure diagram generation are
-very visible and benefit many of the tools using the CDK.
-Furthermore, the stability of the development model has significantly improved,
-providing greater stability of the library over time. The project adopted peer
-review and quality control.
+Since the last CDK publication in 2006, the library has been improved
+in many aspects including architecture, new functionality and improved
+code testing, management and deployment. These changes have led a more
+functionally rich cheminformatics library, with significant
+performance improvements. Updates on the common SMILES and molfile
+formats and the improved structure diagram generation are very visible
+and benefit many of the tools using the CDK.  Furthermore, the
+stability of the development model has significantly improved,
+providing greater stability of the library over time. The project
+adopted peer review and quality control.
 

### 2016-10-05 article.tex
BMC doesn't have that 'e'

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Acknowledgements}
+\section*{Acknowledgments}
 The authors acknowledge the great number of people who have contributed smaller

### 2016-10-05 article.tex
The term 'open source' is so common now, that it no longer needs to be capitalized

@@
 The Chemistry Development Kit (CDK) is a 
-widely used Open Source cheminformatics toolkit, providing data
+widely used open source cheminformatics toolkit, providing data
 structures to represent chemical concepts along with methods to
@@
 This paper highlights the our continued efforts to provide a community
-driven, Open Source
+driven, open source
 cheminformatics library, and show that such collaborative projects can exist over
@@
 advantage of community support and contributions, we show 
-that an Open Source cheminformatics project can act as a peer reviewed
+that an open source cheminformatics project can act as a peer reviewed
 publishing platform for scientific computing software.
@@
 
-The Open Source cheminformatics community has made significant steps
+The open source cheminformatics community has made significant steps
 forward recently~\cite{OBoyle2011b} as evidenced by the growing number
@@
 
-The growth of Open Source software over the last 10 years [REF] is evidence
+The growth of open source software over the last 10 years [REF] is evidence
 of the ability of communities of developers to develop systems and

### 2016-10-05 article.tex
Ten years by now

@@
 descriptor calculations and pharmacophore perception. Over the
-last five years, the code based has grown significantly resulting in
+last ten years, the code based has grown significantly resulting in
 many, complex interdependencies between components and poor

### 2016-10-05 article.tex
Small tweaks

@@
 and previously documented versions have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
-Adoption of the CDK ranges from inclusion of CDK functionality in
+Use of the CDK ranges from inclusion of CDK functionality in
 wrapper platforms such as Cinfony~\cite{OBoyle2008}, incorporation
@@
 the maximally bridging rings in chemical structures~\cite{Marth2015}, and
-defining fingerprints for organic reactions~\cite{Segler2016}.
+prediction of organic reactions~\cite{Segler2016}.
 
@@
 
-The growth of open source software over the last 10 years [REF] is evidence
+The growth of open source software over the last 10 years is evidence           % add REF?
 of the ability of communities of developers to develop systems and
@@
 source code validation, and peer review by fellow developers have
-greatly improved the stability of the library, While it has slowed
+greatly improved the stability of the library. While it has slowed
 development somewhat, it has allowed for cleaning up interdependencies
-between modules of code, and importantly, has improved the scalability
+between modules of functionality, and importantly, has improved the scalability
 of the development model.  This has resulted in significant new
@@
 of elements, charge states and radical species than previous versions,
-a more comprehensive fingerprinting API and many speed and stability
-improvements.
+a more comprehensive fingerprinting API, new depiction functionality,
+and many speed and stability improvements.
 

### 2016-10-05 article.tex
More small tweaks

@@
 of the development model.  This has resulted in significant new
-functionality in core APIs while maintaining the quality of code
-depending on those core APIs.
+functionality in core application programming interfaces (APIs)
+while maintaining the quality of code depending on those core APIs.
 
@@
 
-In this section we described the specifics of new APIs and
+This section describes the specifics of new APIs and
 improvements to pre-existing methods that are available in the latest
@@
 We here outline various new and improved APIs in the CDK library since the two previous
-publications (corresponding to version XXX)
+publications (corresponding to version XXX).                                    % FIXME !!!!
 
@@
 
-  Atom type perception is a core cheminformations functionality: the
+  Atom type perception is core cheminformations functionality: the
   atom types describe chemical features of atoms, such as the number

### 2016-10-05 article.tex
Functional use of whitespace increasing readability

@@
 
-  Atom type perception is core cheminformations functionality: the
+  Atom type perception is core cheminformatics functionality: the
   atom types describe chemical features of atoms, such as the number
@@
 
+\vspace{0.2cm}
 \begin{verbatim}
@@
 \end{verbatim}
+\vspace{0.2cm}
 
@@
 
+\vspace{0.2cm}
   \begin{itemize}
@@
   \end{itemize}
+\vspace{0.2cm}
 
@@
 
+\vspace{0.2cm}
   \begin{verbatim}
@@
   \end{verbatim}
+\vspace{0.2cm}
 

### 2016-10-05 article.tex
More tweaks (it's actually a pretty good read)

@@
 CDK has also played a role in a number of chemical studies, such as finding
-the maximally bridging rings in chemical structures~\cite{Marth2015}, and
-prediction of organic reactions~\cite{Segler2016}.
+the maximally bridging rings in chemical structures~\cite{Marth2015},
+prediction of organic reactions~\cite{Segler2016}, and bioactivities
+of compounds~\cite{Alvarsson2016}.
 
@@
 
-  In previous versions of the API, stereochemistry was represented disparately hindering 
-  interconversion between and within file formats. In CDK \cdkversion{} the core representation 
-  has been standardized upon and procedures updated or added to enable duplicate checking,
-  pattern matching, and interconversion.
+  Previous versions of the API presented stereochemistry disparately hindering 
+  interconversion between and within file formats. CDK \cdkversion{} standardizes
+  upon a new core representation and procedures have been updated or added to
+  enable duplicate checking, pattern matching, and interconversion.
 
@@
   \begin{itemize}
-   \item perception from 2D coordinates;
-   \item perception from 3D coordinates;
-   \item wedge assignment;
-   \item graph (sub)isomorphism matching;
-   \item SMARTS matching; and
+   \item perception from 2D coordinates,
+   \item perception from 3D coordinates,
+   \item wedge assignment,
+   \item graph (sub)isomorphism matching,
+   \item SMARTS matching, and
    \item canonicalization.
@@
 
-  \subsubsection*{Signatures}
+  \subsubsection*{Atomic and Molecular Signatures}
 
@@
   Finally, a signature fingerprint can be calculated for molecules, to allow
-  similarity calculations. This can then be used in QSAR modeling~\cite{signaturefingerprints,
-  Spjuth2011DS,Moghadam2015,Alvarsson2014,Spjuth2012OS,Norinder2013,Alvarsson2016}.
+  similarity calculations. This can then be used in QSAR modeling~\cite{Alvarsson2016,
+  signaturefingerprints,Spjuth2011DS,Moghadam2015,Alvarsson2014,Spjuth2012OS,Norinder2013}.
 
@@
   shown in Figure~\ref{fig:depiction}.
+  When missing, 2D coordinates are generated on the fly with the new structure
+  diagram layout functionality.
 

### 2016-10-06 article.tex
Typo
The previous two papers do not list which versions of the CDK they discussed
These tasks are done

@@
 \title{The Chemistry Development Kit (CDK): atom typing, rendering, molecular formulas, and substructure searching}
-
-% FIXME: finish migrating the author list
 
@@
 
-We here outline various new and improved APIs in the CDK library since the two previous
-publications (corresponding to version XXX).                                    % FIXME !!!!
+We here outline various new and improved APIs in the CDK library since the
+two previous publications in 2003 and 2006~\cite{Steinbeck2003,Steinbeck2006}.
 
@@
 \vspace{0.2cm}
-
-% FIXME: complete code example
 
@@
 \item \textbf{Programming language}: Java
-% FIXME: add appropriate references below
 \item \textbf{Other (optional) requirements}: JNI-InChI, Vecmath, Beam, Guava, JGraphT, Signatures, CMLDOM, XOM, JavaCC
@@
   qsar and qsarmolecular                       & Molecular descriptors.  & & XOM 1.2.5, JAMA 1.0.3~\cite{Hicklin2012} \\ \hline
-  signatures                 & Calculation of molecular and atomic signatures. & & Signatures 1.1 \\ \hline % FIXME: ask Gilleain for the reference
+  signatures                 & Calculation of molecular and atomic signatures. & & Signatures 1.1 \\ \hline
     \end{tabular}

### 2016-10-07 article.tex
Fixed a typo
Fixed typo. HT Stefan Kuhn.

@@
 We report improvements to the CDK since the 1.2 release series,
-specifically addressing the increase functional complexity and poor
+specifically addressing the increased functional complexity and poor
 performance. We first summarize the addition of new, and
@@
 \item \textbf{Programming language}: Java
-\item \textbf{Other (optional) requirements}: JNI-InChI, Vecmath, Beam, Guava, JGraphT, Signatures, CMLDOM, XOM, JavaCC
+\item \textbf{Other (optional) requirements}: JNI-InChI, Vecmath, Beam, Guava, JGraphT, Signatures, CMLXOM, XOM, JavaCC
 \item \textbf{License}: LGPL v2.1 or later

### 2016-10-09 article.tex
simpler file name
More consistent with the text after the code example

@@
   .depict(benzene)
-  .writeTo("BenzeneLocalized.png");
+  .writeTo("benzene.png");
 \end{verbatim}
@@
   the old code base, with and without overlap resolving, and with the
-  new code. The latest implementation simplifies the generation of 2D
-  coordinates as shown below:
+  new code. Generation of 2D coordinates is done as shown below:
 

### 2016-10-09 article.tex
More consistent with the rest of the paper
More high impact metabolomics work using the CDK

@@
 PaDEL~\cite{yap2011padel}, ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016},
-WhichCyp~\cite{Rostkowski2013}, MetaPrint2D~\cite{Carlsson2010}, and the IUPHAR/BPS Guide to Pharmacology database~\cite{Southan2016}.
+WhichCyp~\cite{Rostkowski2013}, MetaPrint2D~\cite{Carlsson2010}, MetFrag~\cite{Wolf2010},
+and the IUPHAR/BPS Guide to Pharmacology database~\cite{Southan2016}.
 A number of such tools were initially developed using older versions
@@
   code and the resulting layout depiction in
-  Figure~\ref{fig:sdgstereo}.
+  Figure~\ref{fig:sdgstereo}:
 
@@
 relevant to have full support for these in cheminformatics
-libraries~\cite{RojasCherto2011}.
+libraries~\cite{Wolf2010,RojasCherto2011,Pluskal2012,Pluskal2010}.
 

### 2016-10-09 article.tex
Small tweaks

@@
 the Round Robin algorithm was particularly apparent when narrow mass ranges 
-were queried (e.g., $\pm 0.001$ Da), thus showing its suitability for
+were queried (e.g. $\pm 0.001$ Da), thus showing its suitability for
 applications in high-resolution mass spectrometry.
@@
 or without simplifying aromatic rings using the lower case element symbols
-for the organic subset. This generator can then be repeatedly called:
+for the organic subset. This generator can then be called repeatedly:
 

### 2016-10-09 article.tex
More minor tweaks

@@
 Substructure matching is fundamental cheminformatics operation and
-plays a key role in many other functions such as fingerprint,
-descriptor generation and atom typing. Since CDK 1.2, functionality
+plays a key role in many other functions such as fingerprint and
+descriptor generation, and atom typing. Since CDK 1.2, functionality
 has been added to handle the SMARTS query language. The SMARTS
@@
 has been added to CDK \cdkversion{}, which simplifies finding,
-filtering, and transform results. The API is immutable allowing a
+filtering, and transforming search results. The API is immutable allowing a
 pattern to be initialized once and then matched against several
@@
 
-Ring finding is a key functionality in a cheminformatics library, and
-the CDK knows a long history of ring finding~\cite{May2014}. In
-particular, non-redundant ring sets have seen particular interest,
+Ring finding is another key functionality in a cheminformatics library, and
+the CDK knows a long history of ring finding~\cite{Berger2004,May2014}. Specifically,
+non-redundant ring sets have seen particular interest,
 such as the smallest set of smallest rings, for which the CDK

### 2016-10-09 article.tex
More tweaks and updated the builder section based on the TODOs given before

@@
 features and heuristics, which are often ambiguously defined in the
-published literature. Based on this information several different
+published literature. Based on the information used, several different
 algorithmic definitions of aromaticity can be defined. Older CDK
-versions had various models implemented and the code was scattered
-throughout the library, resulting in an inconsistent approach to
-calling methods to compute aromaticity and a significant maintenance
-burden.  This was unified in the current version, resulting in three
+versions had various aromaticity models implemented but the code was scattered
+throughout the library, resulting in an inconsistent API
+to compute aromaticity and a significant maintenance
+burden.  The API was unified in the current version, resulting in three
 models, of which two are based on the CDK atom typer. The difference
@@
 
-\subsubsection*{New Builders}
+\subsubsection*{New Object Builders}
 
@@
 non-UI code. To address this, interfaces were introduced allowing
-multiple implementations of the core interfaces.
-% TODO: check if the interfaces have previously been defined
-The \texttt{IChemObjectBuilders} play an important role here, allowing
+multiple implementations of the core interfaces. With much code of the CDK
+library no longer based on the original data model, a builder is needed to
+create objects of that data model, such as an implementation of the IAtom.
+The new \texttt{IChemObjectBuilders} allow implementations to be created, allowing
 implementations of the interfaces to be instantiated without the need
-of explicitly referencing those implementations.\todo{The motivation
-  for IChemObjectBuilders is not well defined.}
-
-However, the CDK 1.0 and 1.2 implementations of the \texttt{IChemObjectBuilder} had one method for
-each constructor, resulting in a very large interface. Moreover, the API changed whenever
-a new class was introduced, and existing methods changed when constructors were updated.
-To simplify the API, the new \texttt{IChemObjectBuilder} collapsed all methods returning new
-implementations into a single method, which takes as a first parameter the class of the
-interface that is wished to be constructed. All further parameters are passed as
+of explicitly referencing those implementations. This way, any algorithm
+implementation in the CDK can use any of the data model interface
+implementations.
+
+The CDK 1.0 and 1.2 implementations of the \texttt{IChemObjectBuilder} had,
+however, one method for each data object constructor, resulting in a very large
+interface. Moreover, this interface API had to be updated each time a new class
+was introduced, and when existing methods changed and constructors were updated.
+To simplify the API, the new \texttt{IChemObjectBuilder} collapses all methods
+into a single method, which takes as a first parameter the class of the
+interface that is to be constructed. All further parameters are passed as
 parameters to the class constructor.
@@
 For example, to construct a new atom from its element symbol, one
-would now write:\todo{might be good to display code using older version}
-
-\vspace{0.2cm}
-\begin{verbatim}
-IChemObjectBuilder builder; // previously defined
+would write previously:
+
+\vspace{0.2cm}
+\begin{verbatim}
+IChemObjectBuilder builder = ...;
+IAtom atom = builder.newAtom("C");
+\end{verbatim}
+\vspace{0.2cm}
+
+With the new builder, the code looks like:
+
+\vspace{0.2cm}
+\begin{verbatim}
+IChemObjectBuilder builder = ...;
 IAtom atom = builder.newInstance(IAtom.class, "C");
@@
 
-The CDK library is now written against the interfaces\todo{this may
-  not mean to a non-expert Java coder}, and when new instances are
-needed, these builders are used. This allows to run a certain
-CDK-based application with a specific builder, aimed at a particular
-use case. The original implementation is available via the
-\texttt{DefaultChemObjectBuilder}, which creates classes that pass
-around change events, as in the older CDK versions.  However, a
-\texttt{SilentChemObjectBuilder} is available, that does not support
-change events, resulting in speedups of 10\%-20\%.  The third builder
+The CDK library is now mostly written that it no longer depends on a specific
+implementation of the \texttt{IChemObjectBuilder}, allowing the user of the
+CDK to select a builder suitable to their software. Therefore, if software
+depends on event passing, then the \texttt{DefaultChemObjectBuilder} can be
+used:
+
+\vspace{0.2cm}
+\begin{verbatim}
+IChemObjectBuilder builder = DefaultChemObjectBuilder.getInstance();
+\end{verbatim}
+\vspace{0.2cm}
+
+However, if events are not needed, a \texttt{SilentChemObjectBuilder} is
+available, resulting in a typical speedup of 10\%-20\%.  The third builder
 is the \texttt{DataDebugChemObjectBuilder} which generates debug
@@
 
-\subsubsection*{Molecular fingerprints}
+\subsubsection*{Molecular Fingerprints}
 Molecular fingerprints are another area that seen significant

### 2016-10-09 article.tex
More details about the fingerprints

@@
 \subsubsection*{Molecular Fingerprints}
-Molecular fingerprints are another area that seen significant
-development. Traditionally fingerprints were represented using the
-\texttt{BitSet} class from the Java library. While using this class
+Molecular fingerprints have also seen significant development in this CDK version.
+Previously, fingerprints were represented using the \texttt{BitSet} class from
+the Java library. While using this class
 allowed the use of pre-existing methods to manipulate bit strings, it
@@
 hashed, relatively small fingerprints, \textit{e.g.}, 1024 bits,
-\textit{i.e.}, 10 bits indexing space. However implementing a
+\textit{i.e.} with a 10 bits indexing space. However, implementing a
 fingerprint designed to avoid collisions with, \textit{e.g.} a 32 bit
@@
 memory-inefficient. To allow for multiple fingerprint representation
-implementations a bit fingerprint interface was introduced\todo{is
-  there any more information about this?}. Also,
-although fingerprints traditionally are bit vectors a count
+implementations a bit fingerprint interface was introduced:
+\texttt{IBitFingerprint}.
+
+\vspace{0.2cm}
+\begin{verbatim}
+IFingerprinter fingerprinter = new Fingerprinter();
+IBitFingerprint fingerprint = fingerprinter.getBitFingerprint(mol);
+\end{verbatim}
+\vspace{0.2cm}
+
+Also, although fingerprints traditionally are bit vectors a count
 fingerprint was also introduced making fingerprints based on integer
-vectors supported in CDK as well. The fingerprints currently existing
-in CDK are listed in Table~3\ref{tab:fingerprints}.
+vectors supported in CDK as well. This is useful for fingerprints that are based
+on substructures, such as the \texttt{SignatureFingerprinter} which uses the
+aforementioned atomic signatures. The counts in the fingerprint then represent
+how often this substructure is found in the molecule it represents. For
+example, for signature depth 1:
+
+\vspace{0.2cm}
+\begin{verbatim}
+SignatureFingerprinter fingerprinter = new SignatureFingerprinter(2);
+ICountFingerprint bitFP = fingerprinter.getCountFingerprint(mol);
+\end{verbatim}
+\vspace{0.2cm}
+
+The fingerprints currently provided by the CDK are listed in Table~3\ref{tab:fingerprints}.
 

### 2016-10-09 article.tex
More tweaks

@@
 how often this substructure is found in the molecule it represents. For
-example, for signature depth 1:
+example, for a signature of depth 1:
 
@@
 
-As the library grew over the years, so did the complexity of the
+As the CDK library grew over the years, so did the complexity of the
 maintenance. The main branch frequently failed to compile and bug
@@
 fixing a bug in one part of the code, broke some other code which made
-the incorrect assumptions about the fixed code. With the increase in
-the CDK developer community, such issues were inevitable, in the
+the incorrect assumptions about the fixed code. With the increased size of
+the CDK developer community, such issues were inevitable in the
 absence of any formal coding and testing standards.
@@
 increasing code maintainability and being flexible enough to allow
-efficient code development. However, we appreciate the subjective
+efficient code development. We appreciate the subjective
 nature of this statement, and some adopted guidelines have been
@@
 to be reviewed by one independent, senior CDK developer for the
-development branch, and two reviewers for stable branches. This patch
+development branch, and by two reviewers for stable branches. This patch
 development system is supported by a number of automated validations
 steps as outlined below.
-
 The next sections describe some approaches the project have adopted that allows
@@
 
-One of the key approaches we have adopted, is to make the CDK more modular. The CDK assigns
+One of the central approaches we have adopted, is to make the CDK more modular. The CDK assigns
 every class to a module, and defines dependencies between modules. For example, core modules
-are not allowed to depend on a module holding data classes implementing the CDK interfaces;
-instead, they may only depend on the interfaces themselves. This, for example, is to ensure
-that dependencies are minimized and to make it easier to exclude CDK functionality with
-third-party dependencies that are not needed.
-
+are not allowed to depend on modules with data classes implementing the CDK interfaces;
+instead, they may only depend on the interfaces themselves. This ensures that
+dependencies are minimized. Furthermore, it also allows cherrypicking CDK
+functionality, reducing the number of third-party library dependencies that are
+needed.
 An overview of key modules with description, important changes, and dependencies
@@
 between the CDK modules are depicted in
-Figure~\ref{fig:deps}.\todo{the dependency graph is not very legible}
+Figure~\ref{fig:deps}.
+
+% \todo{the dependency graph is not very legible}
+% EW: not sure how to solve that
 

### 2016-10-09 article.tex
Almost there...

@@
 later replaced by a custom written tool called OpenJavaDocCheck. With the move
-to Maven (explained elsewhere) which does not have integration for this tool,
+to Maven (explained later), which does not have integration for this tool,
 we adopted CheckStyle (\url{http://checkstyle.sourceforge.net/}). This tool
@@
 values, but also contain integrated tests, which involve more than one
-algorithms, such as SMILES parsing.
+algorithm, such as SMILES parsing. The suite consists of more than 23 thousand
+tests.
 
@@
 developed specifically for the CDK, that, for example, test if a class uses
-the core interfaces instead of implementations of those interfaces. For example,
-that the code uses IAtom instead of Atom. That said, these tests do generate a
+the core interfaces instead of implementations of those interfaces. That is,
+that the code uses \texttt{IAtom} instead of \texttt{Atom}. However, these tests do generate a
 few false positives, as the tests check the class name only, and not the
@@
 OSGi bundles are available for the CDK too, which are used by e.g.
-Bioclipse~\cite{spjuth2007bioclipse,spjuth2009bioclipse} and KNIME~\cite{Beisken2013}.
-Due to split packages between CDK modules, CDK is bundled as a single OSGi jar, which is available from \url{http://pele.farmbio.uu.se/bioclipse/cdk/cdk-1.5.13/}.
-This constitutes an area where improvements can be done on modularization.
+Bioclipse~\cite{spjuth2007bioclipse,spjuth2009bioclipse} and
+KNIME~\cite{Beisken2013}. However, because CDK Java packages are occassionally
+split between CDK modules, the CDK currently needs to be bundled as a single
+OSGi jar. The bundle is available from \url{http://pele.farmbio.uu.se/bioclipse/cdk/cdk-1.5.13/}.
+This Java package and bundle incompatibilities are currently being explored and
+constitutes an area where improvements can be done on modularization.
 

### 2016-10-09 article.tex
Last glance of the conclusion section

@@
 
-Since the last CDK publication in 2006, the library has been improved
-in many aspects including architecture, new functionality and improved
-code testing, management and deployment. These changes have led a more
+Since the second CDK publication, in 2006, the library has been improved
+in many aspects including architecture, new functionality, improved
+code testing, management, peer review, and deployment. These changes have led a more
 functionally rich cheminformatics library, with significant
@@
 stability of the development model has significantly improved,
-providing greater stability of the library over time. The project
-adopted peer review and quality control.
-
+providing greater stability of the library over time.
 With more than 90 contributors, a long list of tools based on the CDK, and

### 2016-10-09 article.tex
Done

@@
   Listed are the currently available molecular fingerprint in CDK with
-  information about whether they come as bit / count versions, what CDK version
-  they appeared in, their default size and where applicable also relevant
-  references.
+  information about whether they come as a bit and/or count version, what CDK version
+  they were introduced in, their default size, and relevant
+  references, where applicable.
   \vskip 1\baselineskip
@@
   LingoFingerprinter~\cite{vidal2005lingo}         & $\checkmark$ &               & \cdkversion{}     & NA%
-\footnote[2]{The Lingofingerprinter does not have a default size.}
+\footnote[2]{The LingoFingerprinter does not have a default size.}
                                                                                              \\

### 2016-10-10 article.tex
Per request from Kai
Added a citation to CSI:FingerID

@@
 ]{\inits{LC}\fnm{Lars}~\snm{Carlsson}}
-\author[
-   addressref={jena},
-   email={kai.duehrkop@uni-jena.de}
-]{\inits{KD}\fnm{Kai} \snm{Dührkop}}
+% \author[
+%    addressref={jena},
+%    email={kai.duehrkop@uni-jena.de}
+% ]{\inits{KD}\fnm{Kai} \snm{Dührkop}}
 \author[
@@
 }
-\address[id=jena]{
-  \orgname{Chair for Bioinformatics, Friedrich Schiller University},
-  %\street{},
-  \postcode{07743},
-  \city{Jena},
-  \cny{Germany}
-}
+% \address[id=jena]{
+%   \orgname{Chair for Bioinformatics, Friedrich Schiller University},
+%   %\street{},
+%   \postcode{07743},
+%   \city{Jena},
+%   \cny{Germany}
+% }
 \address[id=idea]{
@@
 relevant to have full support for these in cheminformatics
-libraries~\cite{Wolf2010,RojasCherto2011,Pluskal2012,Pluskal2010}.
+libraries~\cite{Wolf2010,RojasCherto2011,Pluskal2012,Pluskal2010,Duhrkop2015}.
 
@@
 TP is a Simons Foundation Fellow of the Helen Hay Whitney Foundation.
+We also thank K. Dührkop for his contributions during the writing of this paper.
 

### 2016-10-11 article.tex
Typos

@@
 descriptor calculations and pharmacophore perception. Over the
-last ten years, the code based has grown significantly resulting in
+last ten years, the code base has grown significantly resulting in
 many, complex interdependencies between components and poor

### 2016-10-11 article.tex
between -> among

@@
 last ten years, the code base has grown significantly resulting in
-many, complex interdependencies between components and poor
+many, complex interdependencies among components and poor
 performance of many algorithms. 

### 2016-10-11 article.tex
Removed a comma

@@
 last ten years, the code base has grown significantly resulting in
-many, complex interdependencies among components and poor
+many complex interdependencies among components and poor
 performance of many algorithms. 

### 2016-10-11 article.tex

@@
 %\def \cdkversion {\textbf{2.0}}
-\def \cdkversion {2.0}
+\def \cdkversion {v2.0}
 
@@
    email={john.may@cantab.net}
-]{\inits{JWM}\fnm{John W} \snm{May}}
+]{\inits{JWM}\fnm{John W} \snm{Mayfield}}
 \author[
@@
 \parttitle{Results}
-We report improvements to the CDK since the 1.2 release series,
+We report improvements to the CDK \cdkversion{} since the v1.2 release series,
 specifically addressing the increased functional complexity and poor
@@
 
-  Previous versions of the API presented stereochemistry disparately hindering 
-  interconversion between and within file formats. CDK \cdkversion{} standardizes
+  Previous versions of the API represented stereochemistry in different ways, this hindered
+  interconversion between and within file 
+  .formats. CDK \cdkversion{} standardizes
   upon a new core representation and procedures have been updated or added to
@@
 
+  Rendering is split up into several generation steps, previous versions split
+  up bond from atom rendering, heteroatom symbols were simply drawn over lines
+  representing bonds using a white rectangle to mask. A new \texttt{StandardGenerator}
+  has been introduced that does bond and atom rendering at the same time, it incorporates
+  many ideas described by Alex Clark~\cite{Clark10,Clark13}. The depictions generated are of much
+  higher quality and suitable for publication.
+
   % DESCRIBE THE CORE APIs
 
-  Moreover, a simplified API has been introduced that addresses most of the
-  common rendering needs, with the DepictionGenerator class. To depict benzene
-  the following code can be used:
-
-\vspace{0.2cm}
-\begin{verbatim}
+  Moreover, a simplified high-level API has been introduced that addresses most of the
+  common rendering needs, with the DepictionGenerator class. To depict a molecule
+  loaded into a variable `benzene' the following code can be used:
+
+\vspace{0.2cm}
+\begin{verbatim}
+IAtomContainer benzene = ...;
 new DepictionGenerator()
@@
 
-  The structure diagram layout has been rewritten and the new code solves a
-  number of long standing issues. In particular, collision detection has been
+  The structure diagram layout has been improved and the new code solves a
+  number of long standing issues. In particular, collision avoidance has been
   greatly improved. Figure~\ref{fig:sdg} shows a difference in output
-  between  
-  the old code base, with and without overlap resolving, and with the
-  new code. Generation of 2D coordinates is done as shown below:
-
-\vspace{0.2cm}
-\begin{verbatim}
+  between the old code base, with and without overlap resolving, and with the
+  new refinement based implementation\cite{Helson07}. Generation of 2D coordinates is done as shown below:
+
+\vspace{0.2cm}
+\begin{verbatim}
+IAtomContainer mol = ...;
 StructureDiagramGenerator sdg = new StructureDiagramGenerator();
-sdg.setMolecule(someMolecule);
-sdg.generateCoordinates();
-IAtomContainer layedOutMol = sdg.getMolecule();
+sdg.generateCoordinates(mol);
 \end{verbatim}
@@
   resolution noted above, the engine appropriately handles large ring
-  systems and maintains input stereochemistry. While previous CDK
-versions supported 
+  systems, maintains input stereochemistry, and makes use of a large
+  template library. Templates are useful for laying out substructure
+  While previous CDK versions partially supported 
   double bond stereochemistry the new engine is more efficient in
@@
 \begin{verbatim}
-someMolecule = parser.parseSmiles("OC/C(C)=C/[C@](F)(Cl)Br");
-sdg.setMolecule(someMolecule);
-sdg.generateCoordinates();
-IAtomContainer layedOutMol = sdg.getMolecule();
+String smi = "C[C@@H](\N=C(/S)\Nc1ccccn1)C2CCCCC2 CHEMBL305150";
+IAtomContainer mol = parser.parseSmiles(smi);
+sdg.generateCoordinates(mol);
 \end{verbatim}
@@
 the CDK library. The SmilesParser API uses this library underneath, and the
-Beam API is hidden by this class. The most significant change here is that
-the SMILES parser now automatically locates the positions of double bonds,
-but this can be turned off:
-
-\vspace{0.2cm}
-\begin{verbatim}
-  SmilesParser parser = new SmilesParser(
-    SilentChemObjectBuilder.getInstance()
-  );
-  parser.kekulise(false);
-  mol = parser.parseSmiles(smi);
-\end{verbatim}
-\vspace{0.2cm}
-
-The SMILES generation API has also been simplified and made more flexible.
-Creating unique SMILES or aromatic SMILES is now done by first creating a
-SMILES generator instance which can make generic or unique SMILES, and with
-or without simplifying aromatic rings using the lower case element symbols
-for the organic subset. This generator can then be called repeatedly:
-
-\vspace{0.2cm}
-\begin{verbatim}
-  uniqueGenerator = SmilesGenerator.unique()
-  aromaticGenerator = SmilesGenerator.generic().aromatic()
-  smiles = uniqueGenerator.createSMILES(mol)
-  smiles2 = atomaticGenerator.createSMILES(mol)
-\end{verbatim}
+Beam API is hidden by this class. Basic usage is as follows:
+
+\vspace{0.2cm}
+\begin{verbatim}
+  IChemObjectBuilder bldr 
+     = SilentChemObjectBuilder.getInstance()
+  SmilesParser       smipar = new SmilesParser(bldr);
+  IAtomContainer     mol    = smipar.parseSmiles('[nH]1cccc1');
+\end{verbatim}
+\vspace{0.2cm}
+
+The most significant functional change here is that the SMILES parser automatically 
+locates the positions of double bonds in de-localised aromatic systems (Kekulisation).
+If this invariant can not be meet the SMILES is rejected as invalid. It is 
+possible to override this check but this is strongly discouraged as rejected
+molecules do not have a fixed formula or tautomer.
+
+The SMILES generation API has also been simplified and made more flexible able 
+to produce several different flavours. The \texttt{SmiFlavor} flags are used
+to control the type of SMILES generated. Historically the terms: generic, 
+isomeric, unique, absolute have been used in other toolkits and are also 
+supported.
+
+\vspace{0.2cm}
+\begin{verbatim}
+  SmilesGenerator smigen = new SmilesGenerator(SmiFlavor.Canonical |
+                                               SmiFlavor.StereoTetrahedral)
+\end{verbatim}
+\vspace{0.2cm}
+
+Support for ChemAxon Extended SMILES (CXSMILES)\cite{CXSMILES} layers has been added
+to the CDK \cdkversion{}. CXSMILES provides a powerful means of including auxiliary 
+information in a SMILES string such as 2D/3D coordinates, atom values, generic labels, 
+repeat units, and positional variation. CXSMILES is achieving by placing additional
+information between pipe characters (`\texttt{|}') in the SMILES title field. Information
+is annotated based on the order of the atoms in the SMILES string. An example CXSMILES
+for a generic structure is shown below.
+
+\vspace{0.2cm}
+\small
+\begin{verbatim}
+c1(:*:c2c(:*:c1*)C(N(N2)*)=O)* |$;Y;;;X;;R10;;;;Z;;R11$| US 2007/0129374 (I)
+<- SMILES -------------------> <- CXSMILES ------------> <- Title --------->
+\end{verbatim}
+\normalsize
 \vspace{0.2cm}
@@
 plays a key role in many other functions such as fingerprint and
-descriptor generation, and atom typing. Since CDK 1.2, functionality
+descriptor generation, and atom typing. Since CDK v1.2, functionality
 has been added to handle the SMARTS query language. The SMARTS
-language is supported well including advanced features such as
+language is supported well including features such as
 stereochemistry, component grouping, and atom
@@
 the pattern is inspected as to determine what invariants will be
-needed (e.g. ring size) and only calculating what is needed. The
+needed (e.g. ring size) and only required invariants are calculated. The
 internal matching algorithms provide a lazy iterator, such that the
@@
 
+For SMARTS matching and SMILES generation a model based
+on Daylight \cite{DaylightCIS} can be used and offers significant
+speed improvements to the one based on CDK Atom Types. 
+This model has recently been documented as part of the
+OpenSMILES specification (\url{http://opensmiles.org/}):
+
+\vspace{0.2cm}
+\begin{verbatim}
+Aromaticity aromaticity = new Aromaticity(
+  ElectronDonation.daylight(), Cycles.all()
+);
+\end{verbatim}
+\vspace{0.2cm}
+
 \subsubsection*{CTfile Format Improvements}
@@
 CTfile Sgroups capture and organise high level information about sets of atoms
-and bonds~\cite{Gushurst91}. There are four types of Sgroup: Display Shortcuts, Polymers,
+and bonds~\cite{Gushurst91}. There are four types of Sgroup: Display Short-cuts, Polymers,
 Mixtures, and Data. The most familiar Sgroups from an end user perspective are structure 
 repeat units (e.g. bracketing) and abbreviations (Figure~\ref{fig:sgroups}).
-CDK~\cdkversion{} adds 
-supports for representation, reading, writing, and depiction of Sgroups.
-
-% note, I pie chart would be cool, showing the various relative amounts of the
+CDK~\cdkversion{} adds supports for representation, reading, writing, and depiction of Sgroups.
+
+% note, a pie chart would be cool, showing the various relative amounts of the
 % various formats.
@@
 depends on event passing, then the \texttt{DefaultChemObjectBuilder} can be
-used:
-
-\vspace{0.2cm}
-\begin{verbatim}
-IChemObjectBuilder builder = DefaultChemObjectBuilder.getInstance();
-\end{verbatim}
-\vspace{0.2cm}
-
-However, if events are not needed, a \texttt{SilentChemObjectBuilder} is
-available, resulting in a typical speedup of 10\%-20\%.  The third builder
-is the \texttt{DataDebugChemObjectBuilder} which generates debug
+used, in most cases this isn't needed and the \texttt{SilentChemObjectBuilder} 
+is preferred resulting in a typical speed up of 10\%-20\%:
+
+\vspace{0.2cm}
+\begin{verbatim}
+IChemObjectBuilder builder = SilentChemObjectBuilder.getInstance();
+\end{verbatim}
+\vspace{0.2cm}
+
+The third builder is the \texttt{DataDebugChemObjectBuilder} which generates debug
 information for all changes to the content of the data classes. This
@@
 hashed, relatively small fingerprints, \textit{e.g.}, 1024 bits,
-\textit{i.e.} with a 10 bits indexing space. However, implementing a
-fingerprint designed to avoid collisions with, \textit{e.g.} a 32 bit
-indexing space using this approach would be highly
-memory-inefficient. To allow for multiple fingerprint representation
-implementations a bit fingerprint interface was introduced:
-\texttt{IBitFingerprint}.
-
-\vspace{0.2cm}
-\begin{verbatim}
-IFingerprinter fingerprinter = new Fingerprinter();
-IBitFingerprint fingerprint = fingerprinter.getBitFingerprint(mol);
+\textit{i.e.} with a $2^{10}$ indexing space (128 B). However, implementing a
+fingerprint designed to avoid collisions with a $2^{32}$ bit
+indexing space using this approach would be memory-inefficient (512 MiB).
+To allow for multiple fingerprint representations, a bit
+fingerprint interface was introduced: \texttt{IBitFingerprint}.
+
+\vspace{0.2cm}
+\begin{verbatim}
+IFingerprinter  ecfp = new CircularFingerprinter();
+IBitFingerprint fp   = ecfp.getBitFingerprint(mol);
 \end{verbatim}
@@
 fingerprint was also introduced making fingerprints based on integer
-vectors supported in CDK as well. This is useful for fingerprints that are based
-on substructures, such as the \texttt{SignatureFingerprinter} which uses the
-aforementioned atomic signatures. The counts in the fingerprint then represent
-how often this substructure is found in the molecule it represents. For
-example, for a signature of depth 1:
-
-\vspace{0.2cm}
-\begin{verbatim}
-SignatureFingerprinter fingerprinter = new SignatureFingerprinter(2);
-ICountFingerprint bitFP = fingerprinter.getCountFingerprint(mol);
+vectors supported in CDK as well. The counts in the fingerprint then 
+represent how often this substructure is found in the molecule it
+represents.
+
+\vspace{0.2cm}
+\begin{verbatim}
+IFingerprinter    ecfp = new CircularFingerprinter(CLASS_ECFP4);
+ICountFingerprint fp   = ecfp.getCountFingerprint(mol);
 \end{verbatim}
@@
 The fingerprints currently provided by the CDK are listed in Table~3\ref{tab:fingerprints}.
-
-%% \subsubsection*{Hash Codes}
-
-%% JWM todo
 
@@
 easy for other projects to use. The full library can be included in other
-software by depending on the cdk artifact (\url{http://mvnrepository.com/artifact/org.openscience.cdk/cdk})
+software by depending on the cdk artifact (\url{http://search.maven.org/#search|ga|1|org.openscience})
 but dependencies can also be defined on individual CDK modules.
@@
 and larger contributions to the CDK library. A full list of contributors is
-found in the AUTHORS file~\cite{AUTHORS}.
+found in the Maven parent POM~\cite{AUTHORS}.
 OS acknowledges support from the Swedish strategic research programs eSSENCE and Swedish e-Science Research Center (SeRC).
@@
   \centering
-  \includegraphics[width=0.3\textwidth]{sdg2.png}
+  \includegraphics[width=0.3\textwidth]{sdg2.pdf}
   \label{fig:sdgstereo}
@@
   smarts                     & Substructure searching with the SMARTS format. & & Beam 0.9.1~\cite{Beam} \\ \hline
+  hash                       & Molecular hash codes~\cite{Ihlenfeldt94}. & & \\ \hline
   formula                    & Chemical formula support. & New. & \\ \hline

### 2016-10-11 article.tex
Fix reviewer comments.

@@
   Previous versions of the API represented stereochemistry in different ways, this hindered
-  interconversion between and within file 
-  .formats. CDK \cdkversion{} standardizes
+  interconversion between and within file formats CDK \cdkversion{} standardizes
   upon a new core representation and procedures have been updated or added to
@@
 locates the positions of double bonds in de-localised aromatic systems (Kekulisation).
-If this invariant can not be meet the SMILES is rejected as invalid. It is 
+If this invariant cannot be met the SMILES is rejected as invalid. It is 
 possible to override this check but this is strongly discouraged as rejected
@@
 Support for ChemAxon Extended SMILES (CXSMILES)\cite{CXSMILES} layers has been added
-to the CDK \cdkversion{}. CXSMILES provides a powerful means of including auxiliary 
+to CDK \cdkversion{}. CXSMILES provides a powerful means of including auxiliary 
 information in a SMILES string such as 2D/3D coordinates, atom values, generic labels, 

### 2016-10-12 article.tex
Lost two replies to inquires on cdk-user: added two further papers

@@
 general cheminformatics toolkit. Examples include 
-jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
-PaDEL~\cite{yap2011padel}, ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016},
+jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive,Klein2013}, OMG~\cite{Peironcely2012},
+PaDEL~\cite{yap2011padel}, ChemDes~\cite{Dong2015},
+ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016},
 WhichCyp~\cite{Rostkowski2013}, MetaPrint2D~\cite{Carlsson2010}, MetFrag~\cite{Wolf2010},

### 2016-10-12 article.tex
Another one I missed

--- 2016-10-11 article.tex
@@
 general cheminformatics toolkit. Examples include 
-jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
-PaDEL~\cite{yap2011padel}, ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016},
+jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive,Klein2013}, OMG~\cite{Peironcely2012},
+PaDEL~\cite{yap2011padel}, ChemDes~\cite{Dong2015},
+ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016},
 WhichCyp~\cite{Rostkowski2013}, MetaPrint2D~\cite{Carlsson2010}, MetFrag~\cite{Wolf2010},
-and the IUPHAR/BPS Guide to Pharmacology database~\cite{Southan2016}.
+and the IUPHAR/BPS Guide to Pharmacology~\cite{Southan2016} and
+QSAR DataBank~\cite{Ruusmann2015} databases.
 A number of such tools were initially developed using older versions

### 2016-10-12 article.tex
Fixed formatting in Table 1

@@
     \begin{tabular}{llllllll}
-	\multirow{2}{1in}{ Input } & \multirow{2}{1in}{ Mass tolerance ($\pm$ Da) } & \multicolumn{3}{c}{ \# of generated formulas } & \multicolumn{3}{c}{ Runtime (s) } \\
+	\multirow{2}{1in}{Input} & \multirow{2}{1in}{Mass tolerance ($\pm$~Da)} & \multicolumn{3}{c}{\# of generated formulas} & \multicolumn{3}{c}{Runtime (s)} \\
 	& & HR2 & PFG & CDK & HR2 & PFG & CDK \\

### 2016-10-13 article.tex
Added v for consistency

@@
 
-The CDK 1.0 and 1.2 implementations of the \texttt{IChemObjectBuilder} had,
+The CDK v1.0 and v1.2 implementations of the \texttt{IChemObjectBuilder} had,
 however, one method for each data object constructor, resulting in a very large
@@
 \footnote[1]{For the CircularFingerprinter the bit version is folded to 1024 whereas the count version is unfolded.} \\
-  EStateFingerprinter~\cite{Hall1995}       & $\checkmark$ &               & 1.2.0       & $79$            \\
-  ExtendedFingerprinter      & $\checkmark$ &               & 1.0         & $1\,024$        \\
-  Fingerprinter              & $\checkmark$ &               & 1.0         & $1\,024$        \\
-  GraphOnlyFingerprinter     & $\checkmark$ &               & 1.0         & $1\,024$        \\
-  HybridizationFingerprinter & $\checkmark$ &               & 1.4.0       & $1\,024$        \\
-  KlekotaRothFingerprinter~\cite{Klekota2008}   & $\checkmark$ &               & 1.4.6       & $4\,860$        \\
+  EStateFingerprinter~\cite{Hall1995}       & $\checkmark$ &               & v1.2.0       & $79$            \\
+  ExtendedFingerprinter      & $\checkmark$ &               & v1.0         & $1\,024$        \\
+  Fingerprinter              & $\checkmark$ &               & v1.0         & $1\,024$        \\
+  GraphOnlyFingerprinter     & $\checkmark$ &               & v1.0         & $1\,024$        \\
+  HybridizationFingerprinter & $\checkmark$ &               & v1.4.0       & $1\,024$        \\
+  KlekotaRothFingerprinter~\cite{Klekota2008}   & $\checkmark$ &               & v1.4.6       & $4\,860$        \\
   LingoFingerprinter~\cite{vidal2005lingo}         & $\checkmark$ &               & \cdkversion{}     & NA%
@@
                                                                                              \\
-  MACCSFingerprinter         & $\checkmark$ &               & 1.2.0       & $166$           \\
-  PubchemFingerprinter~\cite{pubchemFP}       & $\checkmark$ &               & 1.4.0       & $881$            \\
+  MACCSFingerprinter         & $\checkmark$ &               & v1.2.0       & $166$           \\
+  PubchemFingerprinter~\cite{pubchemFP}       & $\checkmark$ &               & v1.4.0       & $881$            \\
   ShortestPathFingerprinter  & $\checkmark$ &               & \cdkversion{}     & $1\,024$        \\
   SignatureFingerprinter~\cite{signaturefingerprints}     & $\checkmark$ & $\checkmark$  & \cdkversion{}     & $2^{32}$         \\
-  SubstructureFingerprinter  & $\checkmark$ &               & 1.0         & $307$           \\
+  SubstructureFingerprinter  & $\checkmark$ &               & v1.0         & $307$           \\
 
@@
 \hline
-CDK 1.4.9                 & 3,843  & 69,782  \\
+CDK v1.4.9                 & 3,843  & 69,782  \\
 CDK \cdkversion{}         & 67   & 62,518  \\

### 2016-10-17 article.tex
Typo fix.

@@
 \parttitle{Conclusions}
-This paper highlights the our continued efforts to provide a community
+This paper highlights our continued efforts to provide a community
 driven, open source

### 2016-10-17 article.tex
Updated the title to make clear it's not the first CDK paper.
Added a line to the introduction to indicate what the CDK is used for.
Suggested by John.
Two more tweaks (HT Christoph)

@@
 
-\title{The Chemistry Development Kit (CDK): atom typing, rendering, molecular formulas, and substructure searching}
+\title{Updated Chemistry Development Kit (CDK): atom typing, depiction, molecular formulas, and substructure searching}
 
@@
 ranging from chemical structure canonicalization to molecular
-descriptor calculations and pharmacophore perception. Over the
+descriptor calculations and pharmacophore perception. It is used
+in drug discovery, metabolomics, and toxicology. Over the
 last ten years, the code base has grown significantly resulting in
@@
 driven, open source
-cheminformatics library, and show that such collaborative projects can exist over
+cheminformatics library, and shows that such collaborative projects can thrive over
 extended periods of time, resulting in a high-quality and performant

### 2016-10-18 article.tex
Better title (HT to Tomáš)
Restored The.
More no-break spaces.

@@
 
-\title{Updated Chemistry Development Kit (CDK): atom typing, depiction, molecular formulas, and substructure searching}
+\title{The Chemistry Development Kit (CDK) v2.0: atom typing, depiction, molecular formulas, and substructure searching}
 
@@
    email={chris.evelo@maastrichtuniversity.nl}
-]{\inits{CTE}\fnm{Chris T} \snm{Evelo}}
+]{\inits{CTE}\fnm{Chris~T}~\snm{Evelo}}
 \author[

### 2017-01-03 article.tex
OK, no names are broken up anymore
Updating the material to send in as the first submission was sent back
Pkg no longer used
No figures in the submission
For Software the Implementation and the Results is the same thing
JChemInf uses the Vancouver style
Address change
Whitehead Institute changed its official mailing address
Update John Mayfield Email
Hopefully somewhat less confusing - could also add a footnote. Previous email is still valid.

@@
 \NeedsTeXFormat{LaTeX2e}[1995/12/01]
-\documentclass[10pt]{bmcart}
+\documentclass[doublespacing]{bmcart}
 %\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
@@
 %% Comment out before submission 
-\usepackage[colorinlistoftodos,textsize=small]{todonotes}
+%%\usepackage[colorinlistoftodos,textsize=small]{todonotes}
 
@@
    addressref={nm},
-   email={john.may@cantab.net}
+   email={john.mayfield@cantab.net}
 ]{\inits{JWM}\fnm{John W} \snm{Mayfield}}
@@
    email={jeliazkova.nina@gmail.com}
-]{\inits{NJ}\fnm{Nina} \snm{Jeliazkova}}
+]{\inits{NJ}\fnm{Nina}~\snm{Jeliazkova}}
 \author[
@@
 %%
-\section*{Results and Discussion}
+\section*{Implementation and Results}
 
@@
 
-\bibliographystyle{bmc-mathphys}  % Style BST file
+\bibliographystyle{vancouver}  % Style BST file
 \bibliography{article}
@@
 \centering
-\includegraphics{img/stereodesc_annotated.pdf}
+%\includegraphics{img/stereodesc_annotated.pdf}
 \label{fig:stereodatastructure}
@@
     \centering
-    \includegraphics{img/projections_annotated.pdf}
+%    \includegraphics{img/projections_annotated.pdf}
     \label{fig:stereoprojections}
@@
   \centering
-  \includegraphics[width=\textwidth]{img/US20140231770A1_0287.pdf}
+%  \includegraphics[width=\textwidth]{img/US20140231770A1_0287.pdf}
   \label{fig:depiction}
@@
     (right).}
-    \includegraphics[width=\textwidth]{sdg.png}
+%    \includegraphics[width=\textwidth]{sdg.png}
     \label{fig:sdg}
@@
   \centering
-  \includegraphics[width=0.3\textwidth]{sdg2.pdf}
+%  \includegraphics[width=0.3\textwidth]{sdg2.pdf}
   \label{fig:sdgstereo}
@@
   \centering
-  \includegraphics[width=\textwidth]{cdkDeps.png}
+%  \includegraphics[width=\textwidth]{cdkDeps.png}
   \label{fig:deps}
@@
       (\textbf{CHEBI:4294})}}
-  \includegraphics[width=\textwidth]{img/sgroups.pdf}
+%  \includegraphics[width=\textwidth]{img/sgroups.pdf}
   \label{fig:sgroups}

### 2017-03-25 article.tex
Add ORCID
Small language corrections (rev #4)

@@
 \author[
-   addressref={nm},
+   addressref={nm},                                 % ORCID: 0000-0001-7730-2646
    email={john.mayfield@cantab.net}
@@
   \orgname{Whitehead Institute for Biomedical Research}, 
-  \street{9 Cambridge Center},
+  \street{455 Main Street},
   \postcode{MA 02142},
@@
   perception automatically against the properties defined by the
-  ontology.  This class handles a variety in types of missing
+  ontology.  This class handles a variety of types of missing
   information, as commonly resulting from various (file) formats; for
@@
   explicitly (e.g. SMILES, CML, InChI). Perception from 2D coordinates can optionally identify
-  perspective projects, specifically: Fischer, Haworth, and Chair projections. With the perception of 
+  perspective projections, specifically: Fischer, Haworth, and Chair projections. With the perception of 
   perspective projections enabled, database entries currently consider distinct can be merged
@@
   the query stereochemistry are mapped to the target. Using the permutation parity of this mapping
-  the configurations compared. SMARTS matching requires some special handling for complex cases 
+  the configurations were compared. SMARTS matching requires some special handling for complex cases 
   \cite{May2014_SMARTS}. For canonicalization, a partial canonical ordering is used to assign an 

### 2017-03-25 article.tex
Another small type (rev #2)

@@
   perspective projections, specifically: Fischer, Haworth, and Chair projections. With the perception of 
-  perspective projections enabled, database entries currently consider distinct can be merged
+  perspective projections enabled, database entries currently considered distinct can be merged
   (Figure~\ref{fig:stereoprojections}).  The perception is based on an  algorithm briefly  

### 2017-03-26 article.tex
Removed an outcommented author
Updated the main website

@@
 ]{\inits{LC}\fnm{Lars}~\snm{Carlsson}}
-% \author[
-%    addressref={jena},
-%    email={kai.duehrkop@uni-jena.de}
-% ]{\inits{KD}\fnm{Kai} \snm{Dührkop}}
 \author[
@@
 }
-% \address[id=jena]{
-%   \orgname{Chair for Bioinformatics, Friedrich Schiller University},
-%   %\street{},
-%   \postcode{07743},
-%   \city{Jena},
-%   \cny{Germany}
-% }
 \address[id=idea]{
@@
 problem~\cite{Martello1990}. It first computes a dynamic programming table and
-then backtracks it to generate matching formulas~\cite{Duehrkop2013,
+then backtracks within it to generate matching formulas~\cite{Duehrkop2013,
 Boecker2008}. Data for the Round Robin algorithm is stored in an extended
@@
 distributed repositories, easier branching and provision for
-patches. GitHub (\url{https://github.com/cdk/cdk}) has replaced
+patches. GitHub (\url{https://cdk.github.io/}) has replaced
 SourceForge as the main source code hosting service where we can use
@@
 \item \textbf{Project Name}: The Chemistry Development Kit
-\item \textbf{Project home page}: \url{https://github.com/cdk/cdk}
+\item \textbf{Project home page}: \url{https://cdk.github.io/}
 \item \textbf{Operating system(s)}: Windows, GNU/Linux, OS/X

### 2017-03-26 article.tex
Added more info about available documentation
Added information about automatic building and testing

@@
 reports on missing documentation and on documentation which is not properly
-annotated in the Java source files.
+annotated in the Java source files. The new website lists a few resources to
+help starting CDK users, including a book~\cite{GroovyCDK} and the
+Chemistry Toolkit Rosetta Wiki (\url{http://ctr.wikia.com/wiki/Chemistry_Toolkit_Rosetta_Wiki}).
 
@@
 Java package the class is in.
+
+\subsubsection*{Continuous integration}
+
+The CDK has had an automated build system for many years now. Originally,
+Nightly integrated various tools (building, testing, JavaDoc, etc)~\cite{Guha2006}.
+After the move to Maven, running various steps could be done with Maven,
+and Jenkins was used to execute the steps (one instance is still running
+at \url{https://jenkins.bigcat.unimaas.nl/job/cdk/}. The online Travis-CI
+service is used to build all branches, including pull requests, to ensure
+everything properly compiles: \url{https://travis-ci.org/cdk/cdk}.
 

### 2017-03-26 article.tex
Added a section on user fora

@@
 
+\subsubsection*{Support}
+
+Besides the aforementioned sources of documentation, the project has additional
+sources of support. First, the issue tracker welcomes questions and other types
+of support requests, available at \url{https://github.com/cdk/cdk/issues}.
+The mailing list is another place where support can be requested, while the
+archives document many past user questions. The list and archives can be
+accessed from \url{https://sourceforge.net/p/cdk/mailman/cdk-user/}.
+
 \subsection*{Binary distributions}

### 2017-03-26 article.tex
Shorted the sentences (rev #2)

@@
 
-  Previous versions of the API represented stereochemistry in different ways, this hindered
-  interconversion between and within file formats CDK \cdkversion{} standardizes
+  Previous versions of the API represented stereochemistry in different ways. This hindered
+  interconversion between and within file formats. CDK \cdkversion{} standardizes
   upon a new core representation and procedures have been updated or added to
@@
 
-  Rendering is split up into several generation steps, previous versions split
-  up bond from atom rendering, heteroatom symbols were simply drawn over lines
+  Rendering is split up into several generation steps: previous versions split
+  up bond from atom rendering. Heteroatom symbols were simply drawn over lines
   representing bonds using a white rectangle to mask. A new \texttt{StandardGenerator}
-  has been introduced that does bond and atom rendering at the same time, it incorporates
+  has been introduced that does bond and atom rendering at the same time. It incorporates
   many ideas described by Alex Clark~\cite{Clark10,Clark13}. The depictions generated are of much

### 2017-03-26 article.tex
Update article.tex
Updated my address

@@
 \author[
-   addressref={ebi},
-   email={steinbeck@ebi.ac.uk}
+   addressref={jena},
+   email={christoph.steinbeck@uni-jena.de}
 ]{\inits{CS}\fnm{Christoph}~\snm{Steinbeck}}
@@
 }
-\address[id=ebi]{
-  \orgname{Metabolomics and Molecular Informatics, EMBL-EBI},
-  %\street{A. Kanchev 4},
-  %\postcode{1000},
-  \city{Hinxton},
-  \cny{UK}
+\address[id=jena]{
+  \orgname{Institute for Inorganic and Analytical Chemistry, Friedrich-Schiller-University},
+  \street{Lessingstr. 8},
+  \postcode{07743},
+  \city{Jena},
+  \cny{Germany}
 }

### 2017-03-27 article.tex
Add section on version identifiers.

@@
 
+\subsubsection*{Stability and Version Identifier}
+
+Prior to CDK \cdkversion{}, the parity of the version identifier's second digit 
+indicated stability. Even numbers (v1.2.x, v1.4.x) indicating API stability and
+odd numbers (v1.3.x, v1.5.x) indicating potential API instability. Versions 
+v1.4.x and v1.5.x were developed in parallel, where possible patches were
+applied to both. As the APIs diverged the amount of effort to port patches
+from the development but more robust v1.5.x to v1.4.x 
+became unmanageable for the core development team. This even-odd version scheme
+was adopted from old Linux kernal versioning that was subsequently
+abandoned in 2004 for time-based releases~\cite{LinuxKernalVersionNumbering}.
+
+At the time of writing the development branch is more than 3000 commits ahead
+of v1.4.x. As the the v1.5.x API has become stable it became time to release 
+v1.6.x. Due to significant API changes in 2011\footnote{https://github.com/cdk/cdk/commit/2fc6b61972af834c1fea7fcb64287363ecbcb188}
+it was felt a larger digit increment was needed. This provided the 
+opportunity to change to a more manageable and intuitive version identifier.
+
+From CDK \cdkversion{} a new sequence based version scheme will be used. The
+version identifier indicates change significance as follows:
+
+\begin{verbatim}
+{MAJOR}.{MINOR}.{PATCH}
+
+MAJOR: API change/removed
+MINOR: New feature (API added)
+PATCH: Bug fix
+\end{verbatim}
+
+Due to limited developer resources we envision that releases will primarily 
+increment the minor version with the occasional patch release. As per Maven 
+convention, development versions are suffixed with \texttt{-SNAPSHOT}. 
+There are no API changes from v1.5.x and v2.0.
+
 \subsubsection*{Modularization}

### 2017-04-01 benchmark/cdk-1.4.19/src/main/java/org/openscience/cdk/Cansmi.java
Systematic benchmarking framework and preliminary results.

@@ -1,0 +1,39 @@
+/*
+ * Copyright (c) 2017. NextMove Software Ltd.
+ */
+
+package org.openscience.cdk;
+
+import org.openscience.cdk.exception.CDKException;
+import org.openscience.cdk.interfaces.IAtomContainer;
+import org.openscience.cdk.smiles.SmilesGenerator;
+import org.openscience.cdk.tools.manipulator.AtomContainerManipulator;
+
+import java.io.BufferedWriter;
+import java.io.IOException;
+
+public class Cansmi extends CliExecutable {
+
+  SmilesGenerator smigen = new SmilesGenerator();
+
+  public Cansmi()
+  {
+    super("cansmi");
+  }
+
+  @Override
+  boolean processMolecule(BufferedWriter out, IAtomContainer mol) throws CDKException, IOException
+  {
+    AtomContainerManipulator.removeHydrogens(mol);
+    out.write(smigen.createSMILES(mol));
+
+    Object title = mol.getProperty(CDKConstants.TITLE);
+    if (title != null) {
+      out.write(' ');
+      out.write(title.toString());
+    }
+    out.write('\n');
+    return true;
+  }
+
+}

### 2017-04-01 benchmark/cdk-1.4.19/src/main/java/org/openscience/cdk/Convert.java

@@ -1,0 +1,115 @@
+/*
+ * Copyright (c) 2017. NextMove Software Ltd.
+ */
+
+package org.openscience.cdk;
+
+import joptsimple.OptionSet;
+import joptsimple.OptionSpec;
+import org.openscience.cdk.exception.CDKException;
+import org.openscience.cdk.graph.ConnectivityChecker;
+import org.openscience.cdk.interfaces.IAtomContainer;
+import org.openscience.cdk.interfaces.IAtomContainerSet;
+import org.openscience.cdk.interfaces.IMolecule;
+import org.openscience.cdk.io.SDFWriter;
+import org.openscience.cdk.layout.StructureDiagramGenerator;
+import org.openscience.cdk.smiles.SmilesGenerator;
+
+import java.io.BufferedWriter;
+import java.io.IOException;
+import java.io.OutputStream;
+import java.util.Arrays;
+
+public class Convert extends CliExecutable {
+
+  FileFormat fmt = null;
+
+  protected final OptionSpec<String> ofmtSpec;
+  protected final OptionSpec<?> gen2d;
+
+  private SmilesGenerator smigen = null;
+  private SDFWriter       sdfw   = null;
+
+  private StructureDiagramGenerator sdg = null;
+
+  public Convert()
+  {
+    super("convert");
+    ofmtSpec = optpar.accepts("ofmt", "Output format")
+                     .withRequiredArg()
+                     .ofType(String.class);
+    gen2d = optpar.accepts("gen2d", "Generate 2D coordinates");
+  }
+
+  @Override
+  void configure(OptionSet optset, OutputStream out) throws IOException
+  {
+    if (optset.has(ofmtSpec)) {
+      fmt = getFmt(optset.valueOf(ofmtSpec));
+      if (fmt == FileFormat.UNK) {
+        System.err.println("[FATAl] Unknown output format: " + optset.valuesOf(ofmtSpec));
+        System.exit(1);
+      }
+    } else if (optset.has(outputSpec)){
+      String ext = getExt(optset.valueOf(outputSpec).getName());
+      fmt = getFmt(ext);
+      if (fmt == FileFormat.UNK) {
+        System.err.println("[FATAl] Unknown output format: " + ext);
+        System.exit(1);
+      }
+    }
+    if (fmt == FileFormat.UNK || fmt == null) {
+      System.err.println("[FATAl] No format specified");
+      System.exit(1);
+    }
[…]

### 2017-04-01 benchmark/cdk-1.4.19/src/main/java/org/openscience/cdk/FpGen.java

@@ -1,0 +1,81 @@
+/*
+ * Copyright (c) 2017. NextMove Software Ltd.
+ */
+
+package org.openscience.cdk;
+
+import joptsimple.OptionSet;
+import joptsimple.OptionSpec;
+import org.openscience.cdk.exception.CDKException;
+import org.openscience.cdk.fingerprint.Fingerprinter;
+import org.openscience.cdk.fingerprint.IFingerprinter;
+import org.openscience.cdk.fingerprint.MACCSFingerprinter;
+import org.openscience.cdk.interfaces.IAtomContainer;
+
+import java.io.BufferedWriter;
+import java.io.File;
+import java.io.IOException;
+import java.io.OutputStream;
+import java.nio.charset.StandardCharsets;
+import java.util.BitSet;
+import java.util.Locale;
+
+public final class FpGen extends CliExecutable {
+
+  private final OptionSpec<String> typeSpec = optpar.accepts("type", "Fingerprint type, [circ, maccs, path]")
+                                                    .withRequiredArg()
+                                                    .ofType(String.class)
+                                                    .defaultsTo("circ");
+
+  IFingerprinter fpr;
+  int num_bits;
+
+  public FpGen()
+  {
+    super("fpgen");
+  }
+
+  @Override
+  void configure(OptionSet optset, OutputStream out) throws IOException
+  {
+    final String fptype;
+
+    final String fpkey = typeSpec.value(optset).toLowerCase(Locale.ROOT);
+    switch (fpkey) {
+      case "maccs":
+      case "maccs166":
+        fptype = "maccs166";
+        fpr = new MACCSFingerprinter();
+        break;
+      case "path":
+      case "daylight":
+      case "dy":
+        fptype = "path/length=6";
+        fpr = new Fingerprinter(1024, 6);
+        break;
+      default:
+        System.err.println("FATAL - Unknown fp type: " + fpkey);
+        return;
+    }
+
+    num_bits = fpr.getSize();
+
+    out.write("#FPS1\n".getBytes(StandardCharsets.UTF_8));
+    writeFpsHeader(out, "num_bits", Integer.toString(num_bits));
+    writeFpsHeader(out, "toolkit", "CDK/2.0");
+    writeFpsHeader(out, "type", fptype);
+    writeFpsHeader(out, "date", getDateString());
+    for (File file : optset.valuesOf(inputSpec))
[…]

### 2017-04-01 benchmark/cdk-1.4.19/src/main/java/org/openscience/cdk/Main.java

@@ -1,0 +1,46 @@
+/*
+ * Copyright (c) 2017. NextMove Software Ltd.
+ */
+
+package org.openscience.cdk;
+
+import java.util.Arrays;
+import java.util.HashMap;
+import java.util.Locale;
+import java.util.Map;
+
+public class Main {
+
+  private static final Map<String, CliExecutable> map = new HashMap<>();
+
+  static {
+    add(new Cansmi());
+    add(new FpGen());
+    add(new HeavyAtomCount());
+    add(new Rings());
+    add(new Convert());
+  }
+
+  private static void add(CliExecutable exec) {
+    map.put(exec.getName(), exec);
+  }
+
+  public static void main(String[] args)
+  {
+    if (args.length == 0) {
+      System.err.println("Usage: java -jar cdk-1.4-demo.jar " + map.keySet() + " {input} -out {output}");
+      return;
+    }
+
+    String        cmd   = args[0];
+    CliExecutable bmark = map.get(cmd.toLowerCase(Locale.ROOT));
+    if (bmark == null) {
+      System.err.println("No such command: " + cmd);
+      System.err.println("Usage: ./cdk " + map.keySet() + " {input} -out {output}");
+      return;
+    }
+
+    bmark.process(Arrays.copyOfRange(args, 1, args.length));
+  }
+
+}

### 2017-04-01 benchmark/cdk-1.4.19/src/main/java/org/openscience/cdk/Rings.java

@@ -1,0 +1,92 @@
+/*
+ * Copyright (c) 2017. NextMove Software Ltd.
+ */
+
+package org.openscience.cdk;
+
+import joptsimple.OptionSet;
+import joptsimple.OptionSpec;
+import org.openscience.cdk.exception.CDKException;
+import org.openscience.cdk.graph.SpanningTree;
+import org.openscience.cdk.interfaces.IAtomContainer;
+import org.openscience.cdk.ringsearch.AllRingsFinder;
+import org.openscience.cdk.ringsearch.SSSRFinder;
+
+import java.io.BufferedWriter;
+import java.io.IOException;
+import java.io.OutputStream;
+
+public class Rings extends CliExecutable {
+
+  final OptionSpec<?> markSpec;
+  final OptionSpec<?> sssrSpec;
+  final OptionSpec<?> allSpec;
+
+  private enum RingSet {
+    MARK,
+    SSSR,
+    ALL;
+  }
+
+  RingSet rset = null;
+
+  public Rings()
+  {
+    super("rings");
+    markSpec = optpar.accepts("mark", "Mark ring membership");
+    sssrSpec = optpar.accepts("sssr", "Smallest Set of Smallest Ring");
+    allSpec  = optpar.accepts("all",  "All rings (size≤12)");
+  }
+
+  @Override
+  void configure(OptionSet optset, OutputStream out) throws IOException
+  {
+    if (optset.has(markSpec))
+      rset = RingSet.MARK;
+    if (optset.has(sssrSpec)) {
+      if (rset != null) {
+        System.err.println("[ERROR] please specify one of -mark, -sssr, or -all");
+        System.exit(1);
+      }
+      rset = RingSet.SSSR;
+    }
+    if (optset.has(allSpec)) {
+      if (rset != null) {
+        System.err.println("[ERROR] please specify one of -mark, -sssr, or -all");
+        System.exit(1);
+      }
+      rset = RingSet.ALL;
+    }
+    if (rset == null) {
+      System.err.println("[ERROR] please specify one of -mark, -sssr, or -all");
+      System.exit(1);
+    }
+  }
+
+  @Override
+  boolean processMolecule(BufferedWriter out, IAtomContainer mol) throws CDKException, IOException
+  {
+    switch (rset) {
+      case MARK:
+        SpanningTree spanningtree = new SpanningTree(mol);
+        out.write(Integer.toString(spanningtree.getBondsAcyclicCount()));
+        break;
+      case SSSR:
[…]

### 2017-04-01 benchmark/cdk-2.0/src/main/java/org/openscience/cdk/Convert.java

@@ -1,0 +1,97 @@
+/*
+ * Copyright (c) 2017. NextMove Software Ltd.
+ */
+
+package org.openscience.cdk;
+
+import joptsimple.OptionSet;
+import joptsimple.OptionSpec;
+import org.openscience.cdk.exception.CDKException;
+import org.openscience.cdk.interfaces.IAtomContainer;
+import org.openscience.cdk.io.SDFWriter;
+import org.openscience.cdk.layout.StructureDiagramGenerator;
+import org.openscience.cdk.smiles.SmiFlavor;
+import org.openscience.cdk.smiles.SmilesGenerator;
+
+import java.io.BufferedWriter;
+import java.io.IOException;
+import java.io.OutputStream;
+
+public class Convert extends CliExecutable {
+
+  FileFormat fmt = null;
+
+  protected final OptionSpec<String> ofmtSpec;
+  protected final OptionSpec<?> gen2d;
+
+  private SmilesGenerator smigen = null;
+  private SDFWriter       sdfw   = null;
+
+  private StructureDiagramGenerator sdg = null;
+
+  public Convert()
+  {
+    super("convert");
+    ofmtSpec = optpar.accepts("ofmt", "Output format")
+                     .withRequiredArg()
+                     .ofType(String.class);
+    gen2d = optpar.accepts("gen2d", "Generate 2D coordinates");
+  }
+
+  @Override
+  void configure(OptionSet optset, OutputStream out) throws IOException
+  {
+    if (optset.has(ofmtSpec)) {
+      fmt = getFmt(optset.valueOf(ofmtSpec));
+      if (fmt == FileFormat.UNK) {
+        System.err.println("[FATAl] Unknown output format: " + optset.valuesOf(ofmtSpec));
+        System.exit(1);
+      }
+    } else if (optset.has(outputSpec)){
+      String ext = getExt(optset.valueOf(outputSpec).getName());
+      fmt = getFmt(ext);
+      if (fmt == FileFormat.UNK) {
+        System.err.println("[FATAl] Unknown output format: " + ext);
+        System.exit(1);
+      }
+    }
+    if (fmt == FileFormat.UNK || fmt == null) {
+      System.err.println("[FATAl] No format specified");
+      System.exit(1);
+    }
+    if (optset.has(gen2d)) {
+      sdg = new StructureDiagramGenerator();
+    }
+  }
+
+  @Override
+  boolean processMolecule(BufferedWriter out, IAtomContainer mol) throws CDKException, IOException
+  {
+    // Gen 2D
[…]

### 2017-04-01 benchmark/cdk-2.0/src/main/java/org/openscience/cdk/HeavyAtomCount.java

@@ -1,0 +1,49 @@
+/*
+ * Copyright (c) 2017. NextMove Software Ltd.
+ */
+
+package org.openscience.cdk;
+
+import joptsimple.OptionSet;
+import joptsimple.OptionSpec;
+import org.openscience.cdk.exception.CDKException;
+import org.openscience.cdk.graph.Cycles;
+import org.openscience.cdk.interfaces.IAtom;
+import org.openscience.cdk.interfaces.IAtomContainer;
+import org.openscience.cdk.io.MDLV2000Reader;
+import org.openscience.cdk.silent.AtomContainer;
+
+import java.io.BufferedWriter;
+import java.io.File;
+import java.io.IOException;
+import java.io.InputStream;
+import java.io.OutputStream;
+import java.nio.charset.StandardCharsets;
+import java.util.Collections;
+import java.util.List;
+
+public class HeavyAtomCount extends CliExecutable {
+
+  public HeavyAtomCount()
+  {
+    super("countheavy");
+  }
+
+  @Override
+  boolean processMolecule(BufferedWriter out, IAtomContainer mol) throws CDKException, IOException
+  {
+    int numHeavy = 0;
+    for (IAtom atom : mol.atoms()) {
+      if (atom.getAtomicNumber() != null && atom.getAtomicNumber() != 1)
+        numHeavy++;
+    }
+    out.write(Integer.toString(numHeavy));
+    String title = mol.getProperty(CDKConstants.TITLE);
+    if (title != null) {
+      out.write(' ');
+      out.write(title);
+    }
+    out.write('\n');
+    return true;
+  }
+}

### 2017-04-01 benchmark/cdk-2.0/src/main/java/org/openscience/cdk/Rings.java

@@ -1,0 +1,101 @@
+/*
+ * Copyright (c) 2017. NextMove Software Ltd.
+ */
+
+package org.openscience.cdk;
+
+import joptsimple.OptionSet;
+import joptsimple.OptionSpec;
+import org.openscience.cdk.exception.CDKException;
+import org.openscience.cdk.graph.Cycles;
+import org.openscience.cdk.interfaces.IAtomContainer;
+import org.openscience.cdk.interfaces.IBond;
+import org.openscience.cdk.silent.SilentChemObjectBuilder;
+import org.openscience.cdk.smiles.SmilesParser;
+
+import java.io.BufferedReader;
+import java.io.BufferedWriter;
+import java.io.File;
+import java.io.IOException;
+import java.io.InputStream;
+import java.io.InputStreamReader;
+import java.io.OutputStream;
+import java.io.OutputStreamWriter;
+import java.util.Collections;
+import java.util.List;
+
+public class Rings extends CliExecutable {
+
+  final OptionSpec<?> markSpec;
+  final OptionSpec<?> sssrSpec;
+  final OptionSpec<?> allSpec;
+
+  private enum RingSet {
+    MARK,
+    SSSR,
+    ALL;
+  }
+
+  RingSet rset = null;
+
+  public Rings()
+  {
+    super("rings");
+    markSpec = optpar.accepts("mark", "Mark ring membership");
+    sssrSpec = optpar.accepts("sssr", "Smallest Set of Smallest Ring");
+    allSpec  = optpar.accepts("all",  "All rings (size≤12)");
+  }
+
+  @Override
+  void configure(OptionSet optset, OutputStream out) throws IOException
+  {
+    if (optset.has(markSpec))
+      rset = RingSet.MARK;
+    if (optset.has(sssrSpec)) {
+      if (rset != null) {
+        System.err.println("[ERROR] please specify one of -mark, -sssr, or -all");
+        System.exit(1);
+      }
+      rset = RingSet.SSSR;
+    }
+    if (optset.has(allSpec)) {
+      if (rset != null) {
+        System.err.println("[ERROR] please specify one of -mark, -sssr, or -all");
+        System.exit(1);
+      }
+      rset = RingSet.ALL;
+    }
+    if (rset == null) {
+      System.err.println("[ERROR] please specify one of -mark, -sssr, or -all");
+      System.exit(1);
+    }
+  }
+
+  @Override
[…]

### 2017-04-02 article.tex
Systematic benchmarking framework and preliminary results.
cdkversion everywhere

@@
 
-\title{The Chemistry Development Kit (CDK) v2.0: atom typing, depiction, molecular formulas, and substructure searching}
+\title{The Chemistry Development Kit (CDK) \cdkversion{}: atom typing, depiction, molecular formulas, and substructure searching}
 
@@
 convention, development versions are suffixed with \texttt{-SNAPSHOT}. 
-There are no API changes from v1.5.x and v2.0.
+There are no API changes from v1.5.x and \cdkversion{}.
 
@@
 
+\subsection*{Table ? }
+\begin{tabular}{llllllll}
+Benchmark  & Data Set & Input Format & \multicolumn{2}{c}{CDK 1.4.19} & \multicolumn{2}{c}{CDK \cdkversion{}} & Improvement \\
+           &          &              & Skipped & Elapsed Time         & Skipped  & Elapsed Time \\
+\hline
+\multirow{4}{*}{countheavy} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 37.14s  & 9    & 3.32s & x11.18 \\
+                            &                            & 
+    SDfile & 0              & 15.97s  & 25   & 10.45s & x1.52 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES &                &         & 9    & 14.75s \\
+                            &                              & 
+    SDfile &                &         & 0    & 1m12.08s \\
+\hline
+\multirow{4}{*}{rings --mark} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 37.53s  & 9    & 4.42s & x8.49 \\
+                            &                            & 
+    SDfile & 0              & 18.64s  & 25   & 12.26s & x1.52 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES &                &         & 9    & 34.48s \\
+                            &                              & 
+    SDfile &                &         & 0    & 1m24.59s \\
+\hline
+\multirow{4}{*}{rings --sssr} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 48.83s  & 9    & 5.17s & x9.44 \\
+                            &                            & 
+    SDfile & 0              & 28.24s  & 25   & 12.54s & x2.25 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES &                &         & 9    & 32.91s \\
+                            &                              & 
+    SDfile &                &         & 0    & 1m28.02s \\
+\hline
+\multirow{4}{*}{rings --all} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2126           & 1m5.74s  & 26    & 12.96s & x5.07 \\
+                            &                            & 
+    SDfile & 16             & 49.91s  & 40   & 4.53s & x11.01 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES &                &         & 9    & 30.03s \\
+                            &                              & 
+    SDfile &                &         & 0    & 1m28.65s \\
+\hline
+\multirow{4}{*}{cansmi} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 58.43s  & 9    & 6.00s & x9.74 \\
+                            &                            & 
+    SDfile & 0              & 37.28s  & 26   & 13.82s & x2.69 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES &                &         & 9    & 46.43s \\
+                            &                              & 
+    SDfile &                &         & 0    & 1m48.43s \\
+\hline
+\multirow{4}{*}{convert -ofmt=smi} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 54.96s  & 9    & 4.85s \\
+                            &                            & 
+    SDfile & 35             & 47.05s  & 29   & 11.70s \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES &                &         & 9    & 35.42s \\
+                            &                              & 
+    SDfile &                &         & 0    & 1m38.08s \\
+\hline
+\multirow{4}{*}{convert -ofmt=sdf} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 53.61s  & 9    & 19.47s & x2.82 \\
+                            &                            & 
+    SDfile & 13             & 38.31s  & 29   & 35.08s & x1.09 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES &                &         & 9    & 5m44.59s \\
+                            &                              & 
+    SDfile &                &         & 0    & 7m0.96s \\
+\hline
+\multirow{4}{*}{convert --gen2d -ofmt=sdf} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 36m49.59s  & 9    & 1m9.82s & x31.64 \\
+                            &                            & 
+    SDfile & 13             & 25m39.99s  & 29   & 1m29.09s & x17.28 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES &                &         & 9    & 18m55.64s & \\
+                            &                              & 
+    SDfile &                &         & 0    & 20m55.64s \\
+\hline
+\multirow{4}{*}{genfp --type=maccs} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2150           & 100m1.61s  & 9    & 26.53s & x226 \\
+                            &                            & 
+    SDfile & ?              & ??.??s     & 25   & 35.46s \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES &                &         & 9    & 13m48.73s \\
+                            &                              & 
+    SDfile &                &         & 0    & 14m4.49s \\
+\hline
+\multirow{4}{*}{genfp --type=path} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2150           & ??.??s     & 9    & 3.32s \\
+                            &                            & 
+    SDfile & ?              & ??.??s     & 25   & 10.45s \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES &                &         & 9    & 50.12s \\
+                            &                              & 
+    SDfile &                &         & 0    & 57.71s \\
+\hline
+\multirow{4}{*}{genfp --type=ecfp4} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2150           & ??.??s     & 26    & 50.12s \\
+                            &                            & 
+    SDfile & ?              & ??.??s     & 40   & 57.71s \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES &                &         & 9    & 22m50.2s \\
+                            &                              & 
+    SDfile &                &         & 0    & 23m58.27s \\\hline
+
+\end{tabular}
+
 \end{backmatter}

### 2017-04-02 benchmark/cdk-1.4.19/src/main/java/org/openscience/cdk/FpGen.java
Exit on unrecognised fp type.

@@ -55,6 +55,7 @@ public final class FpGen extends CliExecutable {
         break;
       default:
         System.err.println("FATAL - Unknown fp type: " + fpkey);
+        System.exit(1);
         return;
     }
 

### 2017-04-02 benchmark/cdk-1.4.19/src/main/java/org/openscience/cdk/Main.java
Missing import

@@ -6,6 +6,7 @@ package org.openscience.cdk;
 
 import java.util.Arrays;
 import java.util.HashMap;
+import java.util.LinkedHashMap;
 import java.util.Locale;
 import java.util.Map;
 

### 2017-04-05 benchmark/util/tabprep.R
Should report elapsed time rather than user (CPU). Include raw results and scripts to help build tables. Include num mols per min, useful to put speed/improvement in gaugeable "human" terms.

@@ -1,0 +1,49 @@
+cdk_1.4 <- read.table("~/workspace/github/cdk/cdk-paper-3/benchmark/results/cdk-1.4.19-results", sep='\t')
+cdk_2.0 <- read.table("~/workspace/github/cdk/cdk-paper-3/benchmark/results/cdk-2.0-results", sep='\t')
+
+colnames(cdk_1.4) <- c('benchmark', 'ifmt', 'dataset', 'skipped', 'tElap', 't_s')
+colnames(cdk_2.0) <- c('benchmark', 'ifmt', 'dataset', 'skipped', 'tElap', 't_s')
+
+cdk_1.4$benchmark <- as.character(cdk_1.4$benchmark)
+cdk_2.0$benchmark <- as.character(cdk_2.0$benchmark)
+
+cdk_1.4$version = "CDK 1.4.19"
+cdk_2.0$version = "CDK 2.0"
+
+d <- data.frame();
+for (i in 1:nrow(cdk_2.0)) {
+  row_2.0 <- cdk_2.0[i,];
+  row_1.4 <- subset(cdk_1.4,
+         benchmark == row_2.0$benchmark &
+         ifmt      == row_2.0$ifmt &
+         dataset   == row_2.0$dataset)
+  if (nrow(row_1.4) == 1) {
+    row <- cbind(row_2.0[-7], row_1.4$skipped, row_1.4$tElap, row_1.4$t_s);
+    names(row) <- c('benchmark', 'ifmt', 'dataset', 'v2.0_skipped', 'v2.0_tElap', 'v2.0_t_s', 'v1.4_skipped', 'v1.4_tElap', 'v1.4_t_s')
+    d <- rbind(d, row);
+  } else {
+    row <- cbind(row_2.0[-7], 0, 0, 0);
+    names(row) <- c('benchmark', 'ifmt', 'dataset', 'v2.0_skipped', 'v2.0_tElap', 'v2.0_t_s', 'v1.4_skipped', 'v1.4_tElap', 'v1.4_t_s')
+    d <- rbind(d, row);
+  }
+
+}
+
+d$v2.0_mpm <- c()
+d$v1.4_mpm <- c()
+
+chembl <- which(d$dataset == 'chembl_22_1');
+chebi <- which(d$dataset == 'chebi_149');
+d$v1.4_mpm[chembl] <- round(60*(1678393/d$v1.4_t_s[chembl]))
[…]

### 2017-04-06 article.tex
More benchmarks finished overnight.
Should report elapsed time rather than user (CPU). Include raw results and scripts to help build tables. Include num mols per min, useful to put speed/improvement in gaugeable "human" terms.

--- 2017-03-27 article.tex
@@
 
+\subsection*{Table ? }
+\begin{tabular}{lllrclrcll}
+Benchmark  & Data Set & Input Format & \multicolumn{3}{c}{CDK 1.4.19} & \multicolumn{3}{c}{CDK 2.0} & Improvement \\
+           &          &              & Skipped & Time & Per Min & Skipped  & Time  & Per Min \\
+\hline
+\multirow{4}{*}{countheavy} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 22.51s   & 113k & 9    & 0.97s & 2.6m & x23.21 \\
+                            &                            & 
+    SDfile & 0              & 7.21s   & 355k & 25   & 3.75s & 686k & x1.92 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 0              & 8m39.3s  & 193k & 9    & 12.17s & 8.3m & x42.67 \\
+                            &                              & 
+    SDfile & 0              & 3m17.29s & 510k & 0    & 1m07.23 & 1.5m & x2.93 \\
+\hline
+\multirow{4}{*}{rings -mark} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 22.68   & 112k & 9    & 1.42s & 1.8m & x15.97 \\
+                            &                            & 
+    SDfile & 0              & 9.36   & & 273k  & 4.37s & 568k & x2.14 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 0              & 8m41.57    & 193k & 9    & 30.58s & 3.3m & x17.06 \\
+                            &                              & 
+    SDfile & 0              & 4m9.82 & 403k & 0    & 1.20.60s & 1.2m & x3.10 \\
+\hline
+\multirow{4}{*}{rings -sssr} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 27.40s    & 93k & 9    & 1.48s & 1.7m & x18.51 \\
+                            &                            & 
+    SDfile & 0              & 11.85s    & 216k & 25   & 4.08s & 628k & x2.9 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 0              & 12m4.62s & 138k & 9    & 28.07s & 3.6m & x25.81 \\
+                            &                              & 
+    SDfile & 0              & 7m09.58s  & 234k & 0    & 1m24.16s & 1.2m & x5.10 \\
+\hline
+\multirow{4}{*}{rings -all} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2126           & 45.28s  & 56k & 26    & 1.27s & 2m & x35.65 \\
+                            &                            & 
+    SDfile & 16             & 36.56  & 70k & 40   & 4.13s & 620k & x8.85 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 88             & 12m40.20s & 132k & 9    & 26.48s & 3.8m & x28.71 \\
+                            &                              & 
+    SDfile & 90             & 8m05.64s  & 207k & 0    & 1m18.99s & 1.3m & x6.15 \\
+\hline
+\multirow{4}{*}{cansmi} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 36.58s  & 70k & 9    & 1.83s & 1.4m & x19.99 \\
+                            &                            & 
+    SDfile & 0              & 21.15s  & 121k & 26   & 4.77s & 537k & x4.43 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 14             & 14m33.86s & 115k & 9    & 42.41s & 2.4m & x20.61 \\
+                            &                              & 
+    SDfile & 0              & 8m59.82s & 186k & 0    & 1m44.86s & 960k & x5.15 \\
+\hline
+\multirow{4}{*}{convert -ofmt=smi} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 35.63s  & 71k & 9    & 1.37s & 1.9m & x26.01 \\
+                            &                            & 
+    SDfile & 35             & 20.91s  & 122k & 29   & 4.21s & 608k & x4.97 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 14             & 14m26.02s & 116k & 9    & 28.65s & 3.5m & x30.23 \\
+                            &                              & 
+    SDfile & 0              & 8m59.38s & 186k & 0    & 1m27.35s & 1.2m & x6.17 \\
+\hline
+\multirow{4}{*}{convert -ofmt=sdf} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 32.42s  & 79k & 9    & 10.33s & 248k & x3.14 \\
+                            &                            & 
+    SDfile & 13             & 17.00s  & 150k & 29   & 14.49s & 176k & x1.17 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 0              & 14m25.82s & 116k & 9    & 5m30.66s & 304k & x2.62 \\
+                            &                              & 
+    SDfile & 1              & 8m51.33s & 189k & 0    & 6m50.92s & 245k & x1.29 \\
+\hline
+\multirow{4}{*}{convert -gen2d -ofmt=sdf} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 24m28.02s  & 1.7k & 9    & 35.7s & 72k & x41.12 \\
+                            &                            & 
+    SDfile & 13             & 35m12.03s  & 1.2k & 29   & 42.08s & 61k & x50.19 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 0              & 3h27m07s  & 8.1k & 9    & 18m02.59s & 93k & x11.48 \\
+                            &                              & 
+    SDfile & 1              & 5h58m30s  & 4.6k & 0    & 19m54.12s & 84k & x18.01 \\
+\hline
+\multirow{4}{*}{fpgen -type=maccs} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2150           & 1h37m35s  & 438 & 9    & 19.8s & 129k & x295.71 \\
+                            &                            & 
+    SDfile & 48             & 1h44m17s & 409 & 25   & 22.33s & 114k & x280.21 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 214            & 20h24m57s & 1.4k  & 9    & 13m27.11s & 124k & x91.06 \\
+                            &                              & 
+    SDfile & 225            & 24h41m46s & 1.1k & 0    & 13m41.83s & 122k & x108.18 \\
+\hline
+\multirow{4}{*}{fpgen -type=path} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 1m14.50s   & 34k & 26    & 35.27s & 73k & x2.11 \\
+                            &                            & 
+    SDfile & 0              & 1m52.17s   & 23k & 40    & 38.79s & 66k & x2.89 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 0              & 35m02.80s       & 48k & 9    & 22m7.95 & 76k & x1.58 \\
+                            &                              & 
+    SDfile & 0              & 38m51.73s       & 43k & 0    & 23m08.95s & 73k & x1.68 \\
+\hline
+\multirow{4}{*}{fpgen -type=ecfp4} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & -              &  N/A        && 9    & 4.47s & 573k & \\
+                            &                            & 
+    SDfile & -              &  N/A        && 25   & 7.04s & 364k & \\
+                            & \multirow{2}{*}{ChEMBL 22.1} &
+    SMILES & -              &  N/A        && 9    & 2m45.68s & 608k & \\
+                            &                              & 
+    SDfile & -              &  N/A        && 0    & 3m59.23s & 421k & \\
+\hline
+\end{tabular}
+
 \end{backmatter}

### 2017-04-09 article.tex
Include optimised path fingerprint times and table caption.

--- 2017-03-27 article.tex
@@
 
+\subsection*{Table ? - Summary of systematic benchmark comparing v1.4.19 to \cdkversion{}.}\label{tab:benchmark}
+  The total elapsed real time was measured with the GNU time utility. The 
+  throughput is reported in molecules per minute (k=thousand, m=million) as a 
+  relatable metric.
+  This throughput was calculated by taking the total elapsed
+  time and dividing it by the number of molecule in the dataset (42704 for 
+  ChEBI 149, and 1678393 for ChEMBL 22.1). Nine records skipped for SMILES input for
+  \cdkversion{} is because the parser now validates on read and both ChEBI/
+  ChEMBL contain invalid SMILES (9 each, they are different records). The SDfile skips
+  were molecules in ChEBI with query bonds. The ChEBI SMILES input contains 
+  2107 blank (but valid) inputs, this accounts for the majority skipped 
+  in v1.4.19. Additional skipped records were due to failures somewhere else,
+  ChEBI contains some exotic molecules causing a few skipped records where
+  ChEMBL has none.
+
+\begin{tabular}{lllrclrcll}
+Benchmark  & Data Set & Input Format & \multicolumn{3}{c}{CDK 1.4.19} & \multicolumn{3}{c}{CDK \cdkversion{}} & Improvement \\
+           &          &              & Skip & Time & Per Min & Skip  & Time  & Per Min \\
+\hline
+\multirow{4}{*}{countheavy} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 22.51s   & 113k & 9    & 0.97s & 2.6m & 23.21 \\
+                            &                            & 
+    SDfile & 0              & 7.21s   & 355k & 25   & 3.75s & 686k & 1.92 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 0              & 8m39.3s  & 193k & 9    & 12.17s & 8.3m & 42.67 \\
+                            &                              & 
+    SDfile & 0              & 3m17.29s & 510k & 0    & 1m07.23 & 1.5m & 2.93 \\
+\hline
+\multirow{4}{*}{rings -mark} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 22.68   & 112k & 9    & 1.42s & 1.8m & 15.97 \\
+                            &                            & 
+    SDfile & 0              & 9.36   & 273k  & 0    & 4.37s & 568k & 2.14 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 0              & 8m41.57    & 193k & 9    & 30.58s & 3.3m & 17.06 \\
+                            &                              & 
+    SDfile & 0              & 4m9.82 & 403k & 0    & 1.20.60s & 1.2m & 3.10 \\
+\hline
+\multirow{4}{*}{rings -sssr} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 27.40s    & 93k & 9    & 1.48s & 1.7m & 18.51 \\
+                            &                            & 
+    SDfile & 0              & 11.85s    & 216k & 25   & 4.08s & 628k & 2.9 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 0              & 12m4.62s & 138k & 9    & 28.07s & 3.6m & 25.81 \\
+                            &                              & 
+    SDfile & 0              & 7m09.58s  & 234k & 0    & 1m24.16s & 1.2m & 5.10 \\
+\hline
+\multirow{4}{*}{rings -all} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2126           & 45.28s  & 56k & 26    & 1.27s & 2m & 35.65 \\
+                            &                            & 
+    SDfile & 16             & 36.56  & 70k & 40   & 4.13s & 620k & 8.85 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 88             & 12m40.20s & 132k & 9    & 26.48s & 3.8m & 28.71 \\
+                            &                              & 
+    SDfile & 90             & 8m05.64s  & 207k & 0    & 1m18.99s & 1.3m & 6.15 \\
+\hline
+\multirow{4}{*}{cansmi} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 36.58s  & 70k & 9    & 1.83s & 1.4m & 19.99 \\
+                            &                            & 
+    SDfile & 0              & 21.15s  & 121k & 26   & 4.77s & 537k & 4.43 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 14             & 14m33.86s & 115k & 9    & 42.41s & 2.4m & 20.61 \\
+                            &                              & 
+    SDfile & 0              & 8m59.82s & 186k & 0    & 1m44.86s & 960k & 5.15 \\
+\hline
+\multirow{4}{*}{convert -ofmt=smi} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 35.63s  & 71k & 9    & 1.37s & 1.9m & 26.01 \\
+                            &                            & 
+    SDfile & 35             & 20.91s  & 122k & 29   & 4.21s & 608k & 4.97 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 14             & 14m26.02s & 116k & 9    & 28.65s & 3.5m & 30.23 \\
+                            &                              & 
+    SDfile & 0              & 8m59.38s & 186k & 0    & 1m27.35s & 1.2m & 6.17 \\
+\hline
+\multirow{4}{*}{convert -ofmt=sdf} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 32.42s  & 79k & 9    & 10.33s & 248k & 3.14 \\
+                            &                            & 
+    SDfile & 13             & 17.00s  & 150k & 29   & 14.49s & 176k & 1.17 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 0              & 14m25.82s & 116k & 9    & 5m30.66s & 304k & 2.62 \\
+                            &                              & 
+    SDfile & 1              & 8m51.33s & 189k & 0    & 6m50.92s & 245k & 1.29 \\
+\hline
+\multirow{4}{*}{convert -gen2d -ofmt=sdf} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 24m28.02s  & 1.7k & 9    & 35.7s & 72k & 41.12 \\
+                            &                            & 
+    SDfile & 13             & 35m12.03s  & 1.2k & 29   & 42.08s & 61k & 50.19 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 0              & 3h27m07s  & 8.1k & 9    & 18m02.59s & 93k & 11.48 \\
+                            &                              & 
+    SDfile & 1              & 5h58m30s  & 4.6k & 0    & 19m54.12s & 84k & 18.01 \\
+\hline
+\multirow{4}{*}{fpgen -type=maccs} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2150           & 1h37m35s  & 438 & 9    & 19.8s & 129k & 295.71 \\
+                            &                            & 
+    SDfile & 48             & 1h44m17s & 409 & 25   & 22.33s & 114k & 280.21 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 214            & 20h24m57s & 1.4k  & 9    & 13m27.11s & 124k & 91.06 \\
+                            &                              & 
+    SDfile & 225            & 24h41m46s & 1.1k & 0    & 13m41.83s & 122k & 108.18 \\
+\hline
+\multirow{4}{*}{fpgen -type=path} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & 2112           & 1m14.50s   & 34k & 26    & 8.22s & 312k & 9.06 \\
+                            &                            & 
+    SDfile & 0              & 1m52.17s   & 23k & 40    & 11.67s & 220k & 9.61 \\
+                            & \multirow{2}{*}{ChEMBL 22.1} & 
+    SMILES & 0              & 35m02.80s       & 48k & 9    & 5m35.25s & 300k & 6.27 \\
+                            &                              & 
+    SDfile & 0              & 38m51.73s       & 43k & 0    & 6m7.04s & 274k & 6.35 \\
+\hline
+\multirow{4}{*}{fpgen -type=ecfp4} & \multirow{2}{*}{ChEBI 149}   & 
+    SMILES & -              &  N/A        && 9    & 4.47s & 573k & \\
+                            &                            & 
+    SDfile & -              &  N/A        && 25   & 7.04s & 364k & \\
+                            & \multirow{2}{*}{ChEMBL 22.1} &
+    SMILES & -              &  N/A        && 9    & 2m45.68s & 608k & \\
+                            &                              & 
+    SDfile & -              &  N/A        && 0    & 3m59.23s & 421k & \\
+\hline
+\end{tabular}
+
 \end{backmatter}

### 2017-04-20 benchmark/cdk-1.4.19/src/main/java/org/openscience/cdk/Rings.java
Tweak ring benchmark

@@ -69,7 +69,8 @@ public class Rings extends CliExecutable {
     switch (rset) {
       case MARK:
         SpanningTree spanningtree = new SpanningTree(mol);
-        out.write(Integer.toString(spanningtree.getBondsAcyclicCount()));
+        // spanningtree.getBondsCyclicCount() wrong answer?
+        out.write(Integer.toString(spanningtree.getCyclicFragmentsContainer().getBondCount()));
         break;
       case SSSR:
         SSSRFinder finder = new SSSRFinder(mol);

### 2017-04-20 benchmark/cdk-2.0/src/main/java/org/openscience/cdk/Rings.java

@@ -8,8 +8,10 @@ import joptsimple.OptionSet;
 import joptsimple.OptionSpec;
 import org.openscience.cdk.exception.CDKException;
 import org.openscience.cdk.graph.Cycles;
+import org.openscience.cdk.graph.GraphUtil;
 import org.openscience.cdk.interfaces.IAtomContainer;
 import org.openscience.cdk.interfaces.IBond;
+import org.openscience.cdk.ringsearch.RingSearch;
 import org.openscience.cdk.silent.SilentChemObjectBuilder;
 import org.openscience.cdk.smiles.SmilesParser;
 
@@ -76,11 +78,17 @@ public class Rings extends CliExecutable {
   {
     switch (rset) {
       case MARK:
-        Cycles.markRingAtomsAndBonds(mol);
+        int[][] adjlist = GraphUtil.toAdjList(mol);
+        RingSearch ringSearch = new RingSearch(mol, adjlist);
         int count = 0;
-        for (IBond bond : mol.bonds())
-          if (bond.isInRing())
-            count++;
+        for (int beg = 0; beg < adjlist.length; beg++) {
+          if (ringSearch.cyclic(beg)) {
+            for (int end : adjlist[beg]) {
+              if (end < beg && ringSearch.cyclic(beg, end))
+                count++;
+            }
+          }
+        }
         out.write(Integer.toString(count));
         break;
       case SSSR:

### 2017-04-20 article.tex
Update tables and include a "delta" table that provides more comparable times.

@@
 
+\begin{tabular}{lllrclrclc}
+Benchmark  & \multicolumn{2}{c}{Data Set} & \multicolumn{3}{c}{CDK 1.4.19} & \multicolumn{3}{c}{CDK \cdkversion{}} & Improvement \\
+           &          &              & Skip & Time & Per Min & Skip  & Time  & Per Min \\
+\hline
+\multirow{4}{*}{countheavy} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 22.51s & 108.2K & 9 & 0.97s & 2.5M & 23.21\\
+ &  & sdf & 0 & 7.21s & 355.4K & 25 & 3.75s & 683.3K & 1.92\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 8m39.3s & 193.9K & 9 & 12.17s & 8.3M & 42.67\\
+ &  & sdf & 0 & 3m17.29s & 510.4K & 0 & 1m7.23s & 1.5M & 2.93\\
+\hline
+\multirow{4}{*}{rings-mark} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 22.91s & 106.3K & 9 & 1.03s & 2.4M & 22.24\\
+ &  & sdf & 0 & 8.71s & 294.2K & 25 & 3.93s & 652K & 2.22\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 8m45.78s & 191.5K & 9 & 17.98s & 5.6M & 29.24\\
+ &  & sdf & 0 & 4m12.01s & 399.6K & 0 & 1m14.95s & 1.3M & 3.36\\
+\hline
+\multirow{4}{*}{rings-sssr} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 27.4s & 88.9K & 9 & 1.48s & 1.6M & 18.51\\
+ &  & sdf & 0 & 11.84s & 216.4K & 25 & 4.08s & 628K & 2.9\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 12m4.62s & 139K & 9 & 28.07s & 3.6M & 25.81\\
+ &  & sdf & 0 & 7m9.58s & 234.4K & 0 & 1m24.16s & 1.2M & 5.1\\
+\hline
+\multirow{4}{*}{rings-all} & \multirow{2}{*}{ChEBI 149} & smi & 2126 & 45.28s & 53.8K & 26 & 1.27s & 1.9M & 35.65\\
+ &  & sdf & 16 & 36.56s & 70.1K & 40 & 4.13s & 620.4K & 8.85\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 88 & 12m40.2s & 132.5K & 9 & 26.48s & 3.8M & 28.71\\
+ &  & sdf & 90 & 8m5.64s & 207.4K & 0 & 1m18.99s & 1.3M & 6.15\\
+\hline
+\multirow{4}{*}{cansmi} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 36.58s & 66.6K & 9 & 1.83s & 1.3M & 19.99\\
+ &  & sdf & 35 & 21.15s & 121.1K & 26 & 4.77s & 537.2K & 4.43\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 14 & 14m33.86s & 115.2K & 9 & 42.41s & 2.4M & 20.61\\
+ &  & sdf & 0 & 8m59.82s & 186.6K & 0 & 1m44.86s & 960.4K & 5.15\\
+\hline
+\multirow{4}{*}{convert-ofmtsmi} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 35.63s & 68.4K & 16 & 1.37s & 1.8M & 26.01\\
+ &  & sdf & 35 & 20.91s & 122.5K & 29 & 4.21s & 608.6K & 4.97\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 14 & 14m26.02s & 116.3K & 34 & 28.65s & 3.5M & 30.23\\
+ &  & sdf & 0 & 8m59.38s & 186.7K & 4 & 1m27.35s & 1.2M & 6.17\\
+\hline
+\multirow{4}{*}{convert-ofmtsdf} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 32.42s & 75.1K & 9 & 10.33s & 235.8K & 3.14\\
+ &  & sdf & 13 & 17s & 150.7K & 29 & 14.49s & 176.8K & 1.17\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 14m25.82s & 116.3K & 9 & 5m30.66s & 304.6K & 2.62\\
+ &  & sdf & 1 & 8m51.33s & 189.5K & 0 & 6m50.92s & 245.1K & 1.29\\
+\hline
+\multirow{4}{*}{convert--gen2d-ofmtsdf} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 24m28.02s & 1.7K & 9 & 35.7s & 68.2K & 41.12\\
+ &  & sdf & 13 & 35m12.03s & 1.2K & 29 & 42.08s & 60.9K & 50.19\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 3h27m7s & 8.1K & 9 & 18m2.59s & 93K & 11.48\\
+ &  & sdf & 1 & 5h58m30s & 4.7K & 0 & 19m54.12s & 84.3K & 18.01\\
+\hline
+\multirow{4}{*}{fpgen-type=path} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 1m14.5s & 32.7K & 26 & 8.22s & 296.3K & 9.06\\
+ &  & sdf & 0 & 1m52.17s & 22.8K & 40 & 11.67s & 219.6K & 9.61\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 35m2.8s & 47.9K & 9 & 5m35.25s & 300.4K & 6.27\\
+ &  & sdf & 0 & 38m51.73s & 43.2K & 0 & 6m7.04s & 274.4K & 6.35\\
+\hline
+\multirow{4}{*}{fpgen-type=maccs} & \multirow{2}{*}{ChEBI 149} & smi & 2150 & 1h37m35s & 416 & 9 & 19.8s & 123K & 295.71\\
+ &  & sdf & 48 & 1h44m17s & 409 & 25 & 22.33s & 114.7K & 280.21\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 214 & 20h24m57s & 1.4K & 9 & 13m27.11s & 124.8K & 91.06\\
+ &  & sdf & 225 & 24h41m46s & 1.1K & 0 & 13m41.83s & 122.5K & 108.18\\
+\hline
+\multirow{4}{*}{fpgen-type=circ} & \multirow{2}{*}{ChEBI 149} & smi & 0 & NA & - & 9 & 4.47s & 544.9K & 0\\
+ &  & sdf & 0 & NA & - & 25 & 7.04s & 364K & 0\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & NA & - & 9 & 2m45.68s & 607.8K & 0\\
+ &  & sdf & 0 & NA & - & 0 & 3m59.23s & 420.9K & 0\\
+
+\end{tabular}
+
 \begin{tabular}{lllrclrcll}
-Benchmark  & Data Set & Input Format & \multicolumn{3}{c}{CDK 1.4.19} & \multicolumn{3}{c}{CDK \cdkversion{}} & Improvement \\
+Benchmark  & \multicolumn{2}{c}{Data Set} & \multicolumn{3}{c}{CDK 1.4.19} & \multicolumn{3}{c}{CDK \cdkversion{}} & Improvement \\
            &          &              & Skip & Time & Per Min & Skip  & Time  & Per Min \\
 \hline
-\multirow{4}{*}{countheavy} & \multirow{2}{*}{ChEBI 149}   & 
-    SMILES & 2112           & 22.51s   & 113k & 9    & 0.97s & 2.6m & 23.21 \\
-                            &                            & 
-    SDfile & 0              & 7.21s   & 355k & 25   & 3.75s & 686k & 1.92 \\
-                            & \multirow{2}{*}{ChEMBL 22.1} & 
-    SMILES & 0              & 8m39.3s  & 193k & 9    & 12.17s & 8.3m & 42.67 \\
-                            &                              & 
-    SDfile & 0              & 3m17.29s & 510k & 0    & 1m07.23 & 1.5m & 2.93 \\
-\hline
-\multirow{4}{*}{rings -mark} & \multirow{2}{*}{ChEBI 149}   & 
-    SMILES & 2112           & 22.68   & 112k & 9    & 1.42s & 1.8m & 15.97 \\
-                            &                            & 
-    SDfile & 0              & 9.36   & 273k  & 0    & 4.37s & 568k & 2.14 \\
-                            & \multirow{2}{*}{ChEMBL 22.1} & 
-    SMILES & 0              & 8m41.57    & 193k & 9    & 30.58s & 3.3m & 17.06 \\
-                            &                              & 
-    SDfile & 0              & 4m9.82 & 403k & 0    & 1.20.60s & 1.2m & 3.10 \\
-\hline
-\multirow{4}{*}{rings -sssr} & \multirow{2}{*}{ChEBI 149}   & 
-    SMILES & 2112           & 27.40s    & 93k & 9    & 1.48s & 1.7m & 18.51 \\
-                            &                            & 
-    SDfile & 0              & 11.85s    & 216k & 25   & 4.08s & 628k & 2.9 \\
-                            & \multirow{2}{*}{ChEMBL 22.1} & 
-    SMILES & 0              & 12m4.62s & 138k & 9    & 28.07s & 3.6m & 25.81 \\
-                            &                              & 
-    SDfile & 0              & 7m09.58s  & 234k & 0    & 1m24.16s & 1.2m & 5.10 \\
-\hline
-\multirow{4}{*}{rings -all} & \multirow{2}{*}{ChEBI 149}   & 
-    SMILES & 2126           & 45.28s  & 56k & 26    & 1.27s & 2m & 35.65 \\
-                            &                            & 
-    SDfile & 16             & 36.56  & 70k & 40   & 4.13s & 620k & 8.85 \\
-                            & \multirow{2}{*}{ChEMBL 22.1} & 
-    SMILES & 88             & 12m40.20s & 132k & 9    & 26.48s & 3.8m & 28.71 \\
-                            &                              & 
-    SDfile & 90             & 8m05.64s  & 207k & 0    & 1m18.99s & 1.3m & 6.15 \\
-\hline
-\multirow{4}{*}{cansmi} & \multirow{2}{*}{ChEBI 149}   & 
-    SMILES & 2112           & 36.58s  & 70k & 9    & 1.83s & 1.4m & 19.99 \\
-                            &                            & 
-    SDfile & 0              & 21.15s  & 121k & 26   & 4.77s & 537k & 4.43 \\
-                            & \multirow{2}{*}{ChEMBL 22.1} & 
-    SMILES & 14             & 14m33.86s & 115k & 9    & 42.41s & 2.4m & 20.61 \\
-                            &                              & 
-    SDfile & 0              & 8m59.82s & 186k & 0    & 1m44.86s & 960k & 5.15 \\
-\hline
-\multirow{4}{*}{convert -ofmt=smi} & \multirow{2}{*}{ChEBI 149}   & 
-    SMILES & 2112           & 35.63s  & 71k & 9    & 1.37s & 1.9m & 26.01 \\
-                            &                            & 
-    SDfile & 35             & 20.91s  & 122k & 29   & 4.21s & 608k & 4.97 \\
-                            & \multirow{2}{*}{ChEMBL 22.1} & 
-    SMILES & 14             & 14m26.02s & 116k & 9    & 28.65s & 3.5m & 30.23 \\
-                            &                              & 
-    SDfile & 0              & 8m59.38s & 186k & 0    & 1m27.35s & 1.2m & 6.17 \\
-\hline
-\multirow{4}{*}{convert -ofmt=sdf} & \multirow{2}{*}{ChEBI 149}   & 
-    SMILES & 2112           & 32.42s  & 79k & 9    & 10.33s & 248k & 3.14 \\
-                            &                            & 
-    SDfile & 13             & 17.00s  & 150k & 29   & 14.49s & 176k & 1.17 \\
-                            & \multirow{2}{*}{ChEMBL 22.1} & 
-    SMILES & 0              & 14m25.82s & 116k & 9    & 5m30.66s & 304k & 2.62 \\
-                            &                              & 
-    SDfile & 1              & 8m51.33s & 189k & 0    & 6m50.92s & 245k & 1.29 \\
-\hline
-\multirow{4}{*}{convert -gen2d -ofmt=sdf} & \multirow{2}{*}{ChEBI 149}   & 
-    SMILES & 2112           & 24m28.02s  & 1.7k & 9    & 35.7s & 72k & 41.12 \\
-                            &                            & 
-    SDfile & 13             & 35m12.03s  & 1.2k & 29   & 42.08s & 61k & 50.19 \\
-                            & \multirow{2}{*}{ChEMBL 22.1} & 
-    SMILES & 0              & 3h27m07s  & 8.1k & 9    & 18m02.59s & 93k & 11.48 \\
-                            &                              & 
-    SDfile & 1              & 5h58m30s  & 4.6k & 0    & 19m54.12s & 84k & 18.01 \\
-\hline
-\multirow{4}{*}{fpgen -type=maccs} & \multirow{2}{*}{ChEBI 149}   & 
-    SMILES & 2150           & 1h37m35s  & 438 & 9    & 19.8s & 129k & 295.71 \\
-                            &                            & 
-    SDfile & 48             & 1h44m17s & 409 & 25   & 22.33s & 114k & 280.21 \\
-                            & \multirow{2}{*}{ChEMBL 22.1} & 
-    SMILES & 214            & 20h24m57s & 1.4k  & 9    & 13m27.11s & 124k & 91.06 \\
-                            &                              & 
-    SDfile & 225            & 24h41m46s & 1.1k & 0    & 13m41.83s & 122k & 108.18 \\
-\hline
-\multirow{4}{*}{fpgen -type=path} & \multirow{2}{*}{ChEBI 149}   & 
-    SMILES & 2112           & 1m14.50s   & 34k & 26    & 8.22s & 312k & 9.06 \\
-                            &                            & 
-    SDfile & 0              & 1m52.17s   & 23k & 40    & 11.67s & 220k & 9.61 \\
-                            & \multirow{2}{*}{ChEMBL 22.1} & 
-    SMILES & 0              & 35m02.80s       & 48k & 9    & 5m35.25s & 300k & 6.27 \\
-                            &                              & 
-    SDfile & 0              & 38m51.73s       & 43k & 0    & 6m7.04s & 274k & 6.35 \\
-\hline
-\multirow{4}{*}{fpgen -type=ecfp4} & \multirow{2}{*}{ChEBI 149}   & 
-    SMILES & -              &  N/A        && 9    & 4.47s & 573k & \\
-                            &                            & 
-    SDfile & -              &  N/A        && 25   & 7.04s & 364k & \\
-                            & \multirow{2}{*}{ChEMBL 22.1} &
-    SMILES & -              &  N/A        && 9    & 2m45.68s & 608k & \\
-                            &                              & 
-    SDfile & -              &  N/A        && 0    & 3m59.23s & 421k & \\
+\multirow{4}{*}{countheavy}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 0s & - & 0 & 0s & - & \\
+ &  & sdf & 0 & 0s & - & 0 & 0s & - & \\
+ & \multirow{2}{*}{ChEMBL 22.1}  & smi & 0 & 0s & - & 0 & 0s & - & \\
+ &  & sdf & 0 & 0s & - & 0 & 0s & - & \\
+\hline
+\multirow{4}{*}{rings-mark}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 0.4s & 6.1M & 0 & 0.06s & 40.6M & 6.7\\
+ &  & sdf & 0 & 1.5s & 1.7M & 0 & 0.18s & 14.2M & 8.3\\
+ & \multirow{2}{*}{ChEMBL 22.1}  & smi & 0 & 6.48s & 15.5M & 0 & 5.81s & 17.3M & 1.1\\
+ &  & sdf & 0 & 54.72s & 1.8M & 0 & 7.72s & 13M & 7.1\\
+\hline
+\multirow{4}{*}{rings-sssr}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 4.89s & 498.1K & 0 & 0.51s & 4.8M & 9.6\\
+ &  & sdf & 0 & 4.63s & 553.4K & 0 & 0.33s & 7.8M & 14\\
+ & \multirow{2}{*}{ChEMBL 22.1}  & smi & 0 & 3m25.32s & 490.5K & 0 & 15.9s & 6.3M & 12.9\\
+ &  & sdf & 0 & 3m52.29s & 433.5K & 0 & 16.93s & 5.9M & 13.7\\
+\hline
+\multirow{4}{*}{rings-all}  & \multirow{2}{*}{ChEBI 149}  & smi & 14 & 22.77s & 107K & 17 & 0.3s & 8.1M & 75.9\\
+ &  & sdf & 16 & 29.35s & 87.3K & 15 & 0.38s & 6.7M & 77.2\\
+ & \multirow{2}{*}{ChEMBL 22.1}  & smi & 88 & 4m0.9s & 418K & 0 & 14.31s & 7M & 16.8\\
+ &  & sdf & 90 & 4m48.35s & 349.2K & 0 & 11.76s & 8.6M & 24.5\\
+\hline
+\multirow{4}{*}{cansmi}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 14.07s & 173.1K & 0 & 0.86s & 2.8M & 16.4\\
+ &  & sdf & 35 & 13.94s & 183.8K & 1 & 1.02s & 2.5M & 13.7\\
+ & \multirow{2}{*}{ChEMBL 22.1}  & smi & 14 & 5m54.56s & 284K & 0 & 30.24s & 3.3M & 11.7\\
+ &  & sdf & 0 & 5m42.53s & 294K & 0 & 37.63s & 2.7M & 9.1\\
+\hline
+\multirow{4}{*}{convert-ofmtsmi}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 13.12s & 185.7K & 7 & 0.4s & 6.1M & 32.8\\
+ &  & sdf & 35 & 13.7s & 187K & 4 & 0.46s & 5.6M & 29.8\\
+ & \multirow{2}{*}{ChEMBL 22.1}  & smi & 14 & 5m46.72s & 290.4K & 25 & 16.48s & 6.1M & 21\\
+ &  & sdf & 0 & 5m42.09s & 294.4K & 4 & 20.12s & 5M & 17\\
+\hline
+\multirow{4}{*}{convert-ofmtsdf}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 9.91s & 245.8K & 0 & 9.36s & 260.2K & 1.1\\
+ &  & sdf & 13 & 9.79s & 261.7K & 4 & 10.74s & 238.6K & 0.9\\
+ & \multirow{2}{*}{ChEMBL 22.1}  & smi & 0 & 5m46.52s & 290.6K & 0 & 5m18.49s & 316.2K & 1.1\\
+ &  & sdf & 1 & 5m34.04s & 301.5K & 0 & 5m43.69s & 293K & 1\\
+\hline
+\multirow{4}{*}{convert--gen2d-ofmtsdf}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 24m5.51s & 1.7K & 0 & 34.73s & 70.1K & 41.6\\
+ &  & sdf & 13 & 35m4.82s & 1.2K & 4 & 38.33s & 66.8K & 54.9\\
+ & \multirow{2}{*}{ChEMBL 22.1}  & smi & 0 & 3h18m28s & 8.5K & 0 & 17m50.42s & 94.1K & 11.1\\
+ &  & sdf & 1 & 5h55m13s & 4.7K & 0 & 18m46.89s & 89.4K & 18.9\\
+\hline
+\multirow{4}{*}{fpgen-type=path}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 51.99s & 46.9K & 17 & 7.25s & 336K & 7.2\\
+ &  & sdf & 0 & 1m44.96s & 24.4K & 15 & 7.92s & 323.5K & 13.3\\
+ & \multirow{2}{*}{ChEMBL 22.1}  & smi & 0 & 26m23.5s & 63.6K & 0 & 5m23.08s & 311.7K & 4.9\\
+ &  & sdf & 0 & 35m34.44s & 47.2K & 0 & 4m59.81s & 335.9K & 7.1\\
+\hline
+\multirow{4}{*}{fpgen-type=maccs}  & \multirow{2}{*}{ChEBI 149}  & smi & 38 & 1h37m12s & 418 & 0 & 18.83s & 129.4K & 309.7\\
+ &  & sdf & 48 & 1h44m10s & 410 & 0 & 18.58s & 137.9K & 336.4\\
+ & \multirow{2}{*}{ChEMBL 22.1}  & smi & 214 & 20h16m18s & 1.4K & 0 & 13m14.94s & 126.7K & 91.8\\
+ &  & sdf & 225 & 24h38m29s & 1.1K & 0 & 12m34.6s & 133.5K & 117.6\\
+\hline
+\multirow{4}{*}{fpgen-type=circ}  & \multirow{2}{*}{ChEBI 149}  & smi & -2112 &  & - & 0 & 3.5s & 695.9K & \\
+ &  & sdf & 0 &  & - & 0 & 3.29s & 778.8K & \\
+ & \multirow{2}{*}{ChEMBL 22.1}  & smi & 0 &  & - & 0 & 2m33.51s & 656K & \\
+ &  & sdf & 0 &  & - & 0 & 2m52s & 585.5K & \\
 \hline

### 2017-04-20 benchmark/util/tabprep.R

@@ -1,3 +1,55 @@
+chebi_149_sdf_size <- 42704
+chebi_149_smi_size <- chebi_149_sdf_size-2107
+chembl_22_sdf_size <- 1678393
+chembl_22_smi_size <- chembl_22_sdf_size
+
+toseconds <- function(str) {
+  parts <- unlist(strsplit(as.character(str), ":"))
+  if (length(parts) == 3) {
+    return (3600*as.numeric(parts[1]) + 60*as.numeric(parts[2]) + as.numeric(parts[3]));
+  } else if (length(parts) == 2) {
+    return (60*as.numeric(parts[1]) + as.numeric(parts[2]));
+  } else if (length(parts) == 1) {
+    return (as.numeric(parts[1]));
+  }
+}
+
+tohms <- function(sec) {
+  res <- ""
+  if (is.na(sec))
+    return (NA);
+  decimal <- T
+  if (sec > 3600) {
+    h <- floor(sec / 3600)
+    sec <- sec - (h*3600);
+    res <- paste(res, h, "h", sep="")
+    decimal <- F
+  }
+  if (sec > 60) {
+    m <- floor(sec / 60)
+    sec <- sec - (m*60);
+    res <- paste(res, m, "m", sep="")
+  }
+  if (decimal) {
+    res <- paste(res, round(sec, 2), "s", sep="")
+  } else {
+    res <- paste(res, round(sec), "s", sep="")
+  }
+  return (res)
+}
+
+tohuman <- function(num) {
+  if (is.na(num) || is.infinite(num))
+    return ("-");
+  if (num > 1e6) {
+    return (paste(round(num/1e6,1), "M", sep=""))
+  }
+  if (num > 1e3) {
+    return (paste(round(num/1e3,1), "K", sep=""))
+  }
+  return(num);
+}
+
 cdk_1.4 <- read.table("~/workspace/github/cdk/cdk-paper-3/benchmark/results/cdk-1.4.19-results", sep='\t')
 cdk_2.0 <- read.table("~/workspace/github/cdk/cdk-paper-3/benchmark/results/cdk-2.0-results", sep='\t')
 
@@ -26,24 +78,102 @@ for (i in 1:nrow(cdk_2.0)) {
[…]

### 2017-04-23 article.tex

@@
 
+\subsection*{Systematic Benchmark}
+
+A systematic benchmark was performed to evaluate and quantify 
+performance improvements from v1.4.19 to \cdkversion{}. The benchmark is 
+divided into several cheminformatics tasks for common use cases. Each task 
+was evaluated on input from ChEBI 149 \cite{ChEBI13} and ChEMBL 22.1 
+\cite{ChEMBL14} as both SMILES and SDF.
+
+The benchmark was run on Java SE 8, Cent OS 7, Intel Core i7-4790 CPU @ 
+3.60GHz with 16 GB of RAM. The code to run the benchmark is available in 
+the supplementary material allowing numbers to be recorded on the
+reader's system.
+
+The results of benchmark are summarised in Table 6 and 7. The total elapsed
+times are reported in Table 6, Table 7 subtracts the first tasks results 
+(Count Heavy Atoms) to provide a comparable measure without the overhead of
+input read time. The throughput as molecules per minute is reported but is less
+accurate for v. fast running tasks.
+
+\subsubsection*{Count Heavy Atoms}
+
+This task highlights improvements in raw read performance. Each record is read
+in to a resident memory connection table and the number of heavy (non-hydrogen)
+atoms counted by iterating over the atoms sequentially.
+
+The improvement on this task is most noticeable for SMILES input, previously
+it would take more than 8 minutes to read ChEMBL 22.1 but this is reduced
+to only 12 seconds. On top of this improvement SMILES input is now validated 
+and assigned a Kekul$\mathrm{\acute{e}}$ structure. This identifies 9 invalid
+entries in ChEBI and another 9 in ChEMBL. Most of these rejected SMILES
+are due to the wrong encoding of Cis/Trans double bond stereochemistry at ring 
+closures. The ChEBI 149 SMILES input has 2107 empty records that v1.4.19 skip,
+\cdkversion{} simply reads these as empty molecules. Input from SDF also 
+improved from $\sim$3 minutes to $\sim$1 minute for ChEMBL. The SDF input in
+\cdkversion{} now includes perception of stereochemistry and reading CTfile 
+Sgroups. There are 9 entries from ChEBI's SDF that are rejected because they 
+contain CTfile query features (e.g. any bond order).
+
+\subsubsection*{Rings}
+
+Ring perception is a fundamental step in many other algorithms. 
+The rings task is divided as three subtasks: mark, sssr, and all.
+
+\textbf{-mark}, The first subtask measures the performance in marking 
+ring membership and reporting the number of ring bonds in each record.
+This requires a linear algorithm based on a depth first search. 
+The original code used a weighted spanning tree to compute the membership in 
+linearithmic time. The run times are similar for these datasets (Table
+7), larger differences are seen for more complex cage molecules such fullerenes
+\cite{May2014}.
+
+\textbf{-sssr}, The second subtask computes the Smallest Set of Smallest Rings 
+(SSSR) and reports the size of the SSSR (circuit rank) for each record. Although 
+circuit rank can be computed more efficiently 
+with a linear traversal (counting DFS back-edges) or with Euler's 
+polyhedron formula we are testing the time to enumerate the SSSR set. In general
+SSSR is considered unfavourable due to the 
+non-uniqueness of the set and need for Gaussian matrix elimination (cubic 
+runtime). With some bookkeeping the time spent in the matrix elimination has been
+reduced~\cite{May2014}. For ChEMBL we see the time to generate the SSSR is now
+$\sim$16 seconds when it previously took around $\sim$3.5 minutes (Table 7).
+
+\textbf{-all}, The third subtask counts the number of all rings up to or equal
+to size 12. This includes rings that encompass other smaller rings, for 
+example, 1H-indole has rings of size 5, 6, and 9. In general this problem is 
+exponential and so an adjustable threshold or timeout is used to avoid problematic 
+molecules. CDK v1.4.19 used a timeout based threshold (default 5 seconds) 
+whilst \cdkversion{} uses a 
+counter based on properties of algorithm~\cite{May2014}. In \cdkversion{} there
+were 15/17\footnote{2 records from SDF use query bond features and are skipped
+when read} records skipped from ChEBI that have complex cage-like ring systems 
+(e.g. CHEBI:33611), no records in ChEMBL reached the threshold. By comparison in
+v1.4.19 there were 14/16 records skipped from ChEBI and 88/90 in ChEMBL due to
+reaching the time out. The number fluctuates here as a timeout based threshold 
+depends on external factors such as CPU load.
+
+The speed-up in \cdkversion{} is slightly better than the SSSR task. ChEMBL 
+previously spent 4-5 minutes and now takes only $\sim$14 seconds (Table 7).
+In \cdkversion{} finding all rings ($\leq$ 12 bonds) runs faster than the 
+non-unique SSSR computation.
+
+\subsection*{Canonical SMILES}
+
+This task measures the generation of a Unique SMILES string. These can be
+used to compare dataset intersection and exact lookup. From SMILES input
+\cdkversion{} the total elapsed time is $\sim$20 times faster for both ChEBI
+and ChEMBL. For ChEMBL it now takes just under 43 seconds to read, reorder, 
+and write the SMILES compared to more than 14 minutes previously.
+
+\subsection*{Convert}
+
+This tasks tests the non-canonical conversion between SDF and SMILES input.
+
+\textbf{-ofmt smi}: SMILES is a very compact means of storing connection tables, v1.4.19 could 
+only write canonical SMILES, \cdkversion{} allows different SMILES flavours 
+to be generated including a non-canonical variant. This task outputs
+CXSMILES that includes additional fields such as repeat groups (used by some 
+ChEBI entries). As
+expected the v1.4.19 execution time is the same as for the Canonical SMILES 
+task but \cdkversion{} can generate the non-canonical SMILES faster taking less
+than 30 seconds for ChEMBL.
+
+Assigning double-bond configurations in SMILES is 
+non-trivial and \cdkversion{} has some safety checks, since the SMILES output is 
+Keklu$\mathrm{\acute{e}}$ but input was aromatic, when the bond orders are 
+assigned an extra double-bond may be accidental encoded in the SMILES output, 
+this is sometimes acceptable but will currently report an error. A few errors (4
+in ChEBI, 3 in ChEMBl) occur from Sgroup processing that have now been addressed.
+
+\textbf{-ofmt sdf}: For writing SDF output there is minimal improvement from v1.14.19, when 
+discounting improvements in read performance the SDF generation for ChEBI 
+actually runs slightly slower than v1.4.19 (Table 7). This can be partially 
+explained by the more comprehensive SDF generation that now writes Sgroups 
+as well as computing values for atom parity and valence columns.
+
+\textbf{-gen2d -ofmt sdf}: When writing SDF the only portable way to store stereochemistry is with the 
+inclusion of coordinates, this is specified with the \texttt{-gen2d} option. 
+The overhaul in layout generation discussed early provides better layouts but 
+also included performance tweaks, in CDK v1.4.19 generating coordinates and 
+writing an SDF for ChEMBL took almost 3.5 hours but now only takes just over
+18 minutes.
+
+\subsection*{Fingerprint Generation}
+
+This task tests the generation of fingerprints for similarity and substructure
+screening. Three different types of fingerprints were tested, a Daylight-like 
+Hashed Path Fingerprint, MACCS-like 166 Keys, and Pipeline Pilot-like Hashed 
+Circular Fingerprint (ECFP4). The task generates a hexadecimal FPS file that
+can be used with chemfp~\cite{Dalke2013}.
+
+\textbf{-type path}: Path based fingerprints encode paths of length 1...n. n was
+choose to be 6 for this benchmark. Path based fingerprints can be used for both
+substructure and similarity screening. The algorithm was
+tweaked for \cdkversion{} to hash the paths without pre-computing all paths
+upfront and without needing to generate character strings before hashing. An
+abort threshold was also placed to catch pathological molecules (shown as 
+skipped in ChEBI) that previously took a long time - this will be optimized and 
+removed in future. The time to encode ChEMBL previously took 26-35 minutes now 
+only takes $\sim$5 minutes.
+
+\textbf{-type maccs}: The CDK MACCS fingerprint uses 166 keys to encode
+features of a structure and can be used for similarity searching. This encoding
+uses different aspects of the library including ring perception and the new
+aromaticity perception but the speed is primarily dependant on SMARTS matching 
+performance. Generating the fingerprint previously took almost 1 day for ChEMBL 
+and 2 hours for ChEBI. This has been reduced to less than 
+15 minutes for ChEMBL and less than 30 seconds for ChEBI.
+
+\textbf{-type circ}: Circular fingerprints can only be used for similarity and
+could not be generated in v1.4.19. However, the fingerprints are known to perform
+well for retrieval performance~\cite{OBoyle2016}. The times are included here 
+to show they are faster to calculate than path or MACCS-like keys and therefore
+recommended. CDK includes two implementations based on signatures or extended 
+connectivity~\cite{Clark2014}.
+
+\subsubsection*{Benchmark Summary}
+
+In all tasks, the total elapsed time is better in \cdkversion{} compared to 
+v1.4.19. On many tasks the improvement is more than ten times faster.
+Not only is the execution time improved but improvements in robustness and 
+correctness means \cdkversion{} is often doing much more work than the 
+equivalent procedures in v1.4.19.
+
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-\subsection*{Table ? - Summary of systematic benchmark comparing v1.4.19 to \cdkversion{}.}\label{tab:benchmark}
-  The total elapsed real time was measured with the GNU time utility. The 
-  throughput is reported in molecules per minute (k=thousand, m=million) as a 
+\subsection*{Table 6 - Summary of systematic benchmark comparing v1.4.19 to \cdkversion{}.}\label{tab:benchmark}
+  The total elapsed \textit{real} time was measured with the unix time utility. The 
+  throughput is reported in molecules per minute (K=thousand, M=million) as a 
   relatable metric.
@@
   time and dividing it by the number of molecule in the dataset (42704 for 
-  ChEBI 149, and 1678393 for ChEMBL 22.1). Nine records skipped for SMILES input for
-  \cdkversion{} is because the parser now validates on read and both ChEBI/
-  ChEMBL contain invalid SMILES (9 each, they are different records). The SDfile skips
-  were molecules in ChEBI with query bonds. The ChEBI SMILES input contains 
+  ChEBI 149, and 1678393 for ChEMBL 22.1). The ChEBI SMILES input contains 
   2107 blank (but valid) inputs, this accounts for the majority skipped 
-  in v1.4.19. Additional skipped records were due to failures somewhere else,
-  ChEBI contains some exotic molecules causing a few skipped records where
-  ChEMBL has none.
+  in v1.4.19. The throughput calculation was adjust to account for this.
 
@@
 \hline
-\multirow{4}{*}{rings-mark} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 22.91s & 106.3K & 9 & 1.03s & 2.4M & 22.24\\
+\multirow{4}{*}{\shortstack[l]{rings \\ -mark}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 22.91s & 106.3K & 9 & 1.03s & 2.4M & 22.24\\
  &  & sdf & 0 & 8.71s & 294.2K & 25 & 3.93s & 652K & 2.22\\
@@
 \hline
-\multirow{4}{*}{rings-sssr} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 27.4s & 88.9K & 9 & 1.48s & 1.6M & 18.51\\
+\multirow{4}{*}{\shortstack[l]{rings \\ -sssr}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 27.4s & 88.9K & 9 & 1.48s & 1.6M & 18.51\\
  &  & sdf & 0 & 11.84s & 216.4K & 25 & 4.08s & 628K & 2.9\\
@@
 \hline
-\multirow{4}{*}{rings-all} & \multirow{2}{*}{ChEBI 149} & smi & 2126 & 45.28s & 53.8K & 26 & 1.27s & 1.9M & 35.65\\
+\multirow{4}{*}{\shortstack[l]{rings \\ -all}} & \multirow{2}{*}{ChEBI 149} & smi & 2126 & 45.28s & 53.8K & 26 & 1.27s & 1.9M & 35.65\\
  &  & sdf & 16 & 36.56s & 70.1K & 40 & 4.13s & 620.4K & 8.85\\
@@
 \hline
-\multirow{4}{*}{convert-ofmtsmi} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 35.63s & 68.4K & 16 & 1.37s & 1.8M & 26.01\\
+\multirow{4}{*}{\shortstack[l]{convert \\ -ofmt smi}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 35.63s & 68.4K & 16 & 1.37s & 1.8M & 26.01\\
  &  & sdf & 35 & 20.91s & 122.5K & 29 & 4.21s & 608.6K & 4.97\\
@@
 \hline
-\multirow{4}{*}{convert-ofmtsdf} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 32.42s & 75.1K & 9 & 10.33s & 235.8K & 3.14\\
+\multirow{4}{*}{\shortstack[l]{convert \\ -ofmt sdf}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 32.42s & 75.1K & 9 & 10.33s & 235.8K & 3.14\\
  &  & sdf & 13 & 17s & 150.7K & 29 & 14.49s & 176.8K & 1.17\\
@@
 \hline
-\multirow{4}{*}{convert--gen2d-ofmtsdf} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 24m28.02s & 1.7K & 9 & 35.7s & 68.2K & 41.12\\
+\multirow{4}{*}{\shortstack[l]{convert \\ -gen2d \\ -ofmt smi}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 24m28.02s & 1.7K & 9 & 35.7s & 68.2K & 41.12\\
  &  & sdf & 13 & 35m12.03s & 1.2K & 29 & 42.08s & 60.9K & 50.19\\
@@
 \hline
-\multirow{4}{*}{fpgen-type=path} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 1m14.5s & 32.7K & 26 & 8.22s & 296.3K & 9.06\\
+\multirow{4}{*}{\shortstack[l]{fpgen \\ -type path}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 1m14.5s & 32.7K & 26 & 8.22s & 296.3K & 9.06\\
  &  & sdf & 0 & 1m52.17s & 22.8K & 40 & 11.67s & 219.6K & 9.61\\
@@
 \hline
-\multirow{4}{*}{fpgen-type=maccs} & \multirow{2}{*}{ChEBI 149} & smi & 2150 & 1h37m35s & 416 & 9 & 19.8s & 123K & 295.71\\
+\multirow{4}{*}{\shortstack[l]{fpgen \\ -type maccs}} & \multirow{2}{*}{ChEBI 149} & smi & 2150 & 1h37m35s & 416 & 9 & 19.8s & 123K & 295.71\\
  &  & sdf & 48 & 1h44m17s & 409 & 25 & 22.33s & 114.7K & 280.21\\
@@
 \hline
-\multirow{4}{*}{fpgen-type=circ} & \multirow{2}{*}{ChEBI 149} & smi & 0 & NA & - & 9 & 4.47s & 544.9K & 0\\
+\multirow{4}{*}{\shortstack[l]{fpgen \\ -type circ}} & \multirow{2}{*}{ChEBI 149} & smi & 0 & NA & - & 9 & 4.47s & 544.9K & 0\\
  &  & sdf & 0 & NA & - & 25 & 7.04s & 364K & 0\\
@@
 \end{tabular}
+
+\subsection*{Table 7 - Summary of systematic benchmark comparing v1.4.19 to \cdkversion{} without read times.}
+\label{tab:benchmark-delta}
+The number of records skipped and time to run the countheavy benchmark has been subtracted from Table 6.
+The remaining results provides a relative comparison without accounting for the overhead of reading the input.
 
@@
 \hline
-\multirow{4}{*}{rings-mark}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 0.4s & 6.1M & 0 & 0.06s & 40.6M & 6.7\\
+\multirow{4}{*}{\shortstack[l]{rings \\ -mark}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 0.4s & 6.1M & 0 & 0.06s & 40.6M & 6.7\\
  &  & sdf & 0 & 1.5s & 1.7M & 0 & 0.18s & 14.2M & 8.3\\
@@
 \hline
-\multirow{4}{*}{rings-sssr}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 4.89s & 498.1K & 0 & 0.51s & 4.8M & 9.6\\
+\multirow{4}{*}{\shortstack[l]{rings \\ -sssr}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 4.89s & 498.1K & 0 & 0.51s & 4.8M & 9.6\\
  &  & sdf & 0 & 4.63s & 553.4K & 0 & 0.33s & 7.8M & 14\\
@@
 \hline
-\multirow{4}{*}{rings-all}  & \multirow{2}{*}{ChEBI 149}  & smi & 14 & 22.77s & 107K & 17 & 0.3s & 8.1M & 75.9\\
+\multirow{4}{*}{\shortstack[l]{rings \\ -all}}  & \multirow{2}{*}{ChEBI 149}  & smi & 14 & 22.77s & 107K & 17 & 0.3s & 8.1M & 75.9\\
  &  & sdf & 16 & 29.35s & 87.3K & 15 & 0.38s & 6.7M & 77.2\\
@@
 \hline
-\multirow{4}{*}{convert-ofmtsmi}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 13.12s & 185.7K & 7 & 0.4s & 6.1M & 32.8\\
+\multirow{4}{*}{\shortstack[l]{convert \\ -ofmt smi}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 13.12s & 185.7K & 7 & 0.4s & 6.1M & 32.8\\
  &  & sdf & 35 & 13.7s & 187K & 4 & 0.46s & 5.6M & 29.8\\
@@
 \hline
-\multirow{4}{*}{convert-ofmtsdf}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 9.91s & 245.8K & 0 & 9.36s & 260.2K & 1.1\\
+\multirow{4}{*}{\shortstack[l]{convert \\ -ofmt sdf}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 9.91s & 245.8K & 0 & 9.36s & 260.2K & 1.1\\
  &  & sdf & 13 & 9.79s & 261.7K & 4 & 10.74s & 238.6K & 0.9\\
@@
 \hline
-\multirow{4}{*}{convert--gen2d-ofmtsdf}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 24m5.51s & 1.7K & 0 & 34.73s & 70.1K & 41.6\\
+\multirow{4}{*}{\shortstack[l]{convert \\ -gen2d \\ -ofmt smi}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 24m5.51s & 1.7K & 0 & 34.73s & 70.1K & 41.6\\
  &  & sdf & 13 & 35m4.82s & 1.2K & 4 & 38.33s & 66.8K & 54.9\\
@@
 \hline
-\multirow{4}{*}{fpgen-type=path}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 51.99s & 46.9K & 17 & 7.25s & 336K & 7.2\\
+\multirow{4}{*}{\shortstack[l]{fpgen \\ -type path}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 51.99s & 46.9K & 17 & 7.25s & 336K & 7.2\\
  &  & sdf & 0 & 1m44.96s & 24.4K & 15 & 7.92s & 323.5K & 13.3\\
@@
 \hline
-\multirow{4}{*}{fpgen-type=maccs}  & \multirow{2}{*}{ChEBI 149}  & smi & 38 & 1h37m12s & 418 & 0 & 18.83s & 129.4K & 309.7\\
+\multirow{4}{*}{\shortstack[l]{fpgen \\ -type maccs}}  & \multirow{2}{*}{ChEBI 149}  & smi & 38 & 1h37m12s & 418 & 0 & 18.83s & 129.4K & 309.7\\
  &  & sdf & 48 & 1h44m10s & 410 & 0 & 18.58s & 137.9K & 336.4\\
@@
 \hline
-\multirow{4}{*}{fpgen-type=circ}  & \multirow{2}{*}{ChEBI 149}  & smi & -2112 &  & - & 0 & 3.5s & 695.9K & \\
+\multirow{4}{*}{\shortstack[l]{fpgen \\ -type circ}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 &  & - & 0 & 3.5s & 695.9K & \\
  &  & sdf & 0 &  & - & 0 & 3.29s & 778.8K & \\

### 2017-04-26 article.tex
Benchmark fpgen --type=path and convert -ofmt=smi rerun for 2.0 now patches have been applied, simplifies explanation of why there were skipped entires.
Add footnote on he MDL parity value, not sure if there's a better way to fit in this digression.

@@
   The perception from coordinates and wedge assignment algorithms are fundamental for conversion
-  between formats that store stereochemistry implicitly based on coordinates (e.g. molfile, CML) and 
-  explicitly (e.g. SMILES, CML, InChI). Perception from 2D coordinates can optionally identify
+  between formats that store stereochemistry implicitly based on coordinates (e.g. molfile\footnote{Molfiles 
+  can also store tetrahedral stereochemistry as a parity value, this is read if no coordinates are specified.
+  In general there is no guarantee the parity value is read and the only portable way to store
+  stereochemistry in a molfile is with coordinates.}, CML) and 
+  explicitly (e.g. SMILES, CML, InChI). Perception from 2D coordinates 
+  can optionally identify
   perspective projections, specifically: Fischer, Haworth, and Chair projections. With the perception of 
@@
 assigned an extra double-bond may be accidental encoded in the SMILES output, 
-this is sometimes acceptable but will currently report an error. A few errors (4
-in ChEBI, 3 in ChEMBl) occur from Sgroup processing that have now been addressed.
+this is sometimes acceptable but will currently report an error.
 
@@
 tweaked for \cdkversion{} to hash the paths without pre-computing all paths
-upfront and without needing to generate character strings before hashing. An
-abort threshold was also placed to catch pathological molecules (shown as 
-skipped in ChEBI) that previously took a long time - this will be optimized and 
-removed in future. The time to encode ChEMBL previously took 26-35 minutes now 
-only takes $\sim$5 minutes.
+upfront and without needing to generate character strings before hashing. The 
+time to encode ChEMBL previously took 26-35 minutes now only takes $\sim$5 
+minutes.
 
@@
 \hline
-\multirow{4}{*}{\shortstack[l]{convert \\ -ofmt smi}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 35.63s & 68.4K & 16 & 1.37s & 1.8M & 26.01\\
- &  & sdf & 35 & 20.91s & 122.5K & 29 & 4.21s & 608.6K & 4.97\\
- & \multirow{2}{*}{ChEMBL 22.1} & smi & 14 & 14m26.02s & 116.3K & 34 & 28.65s & 3.5M & 30.23\\
- &  & sdf & 0 & 8m59.38s & 186.7K & 4 & 1m27.35s & 1.2M & 6.17\\
+\multirow{4}{*}{\shortstack[l]{convert \\ -ofmt smi}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 35.63s & 68.4K & 16 & 1.6s & 1.5M & 22.27 \\
+ &  & sdf & 35 & 20.91s & 122.5K & 25 & 4.47s & 573.2K & 4.68 \\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 14 & 14m26.02s & 116.3K & 34 & 28.14s & 3.6M & 30.78 \\
+ &  & sdf & 0 & 8m59.38s & 186.7K & 1 & 1m19.86s & 1.3M & 6.75 \\
 \hline
@@
 \hline
-\multirow{4}{*}{\shortstack[l]{fpgen \\ -type path}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 1m14.5s & 32.7K & 26 & 8.22s & 296.3K & 9.06\\
- &  & sdf & 0 & 1m52.17s & 22.8K & 40 & 11.67s & 219.6K & 9.61\\
- & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 35m2.8s & 47.9K & 9 & 5m35.25s & 300.4K & 6.27\\
- &  & sdf & 0 & 38m51.73s & 43.2K & 0 & 6m7.04s & 274.4K & 6.35\\
+\multirow{4}{*}{\shortstack[l]{fpgen \\ -type path}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 1m14.5s & 32.7K & 9 & 8.67s & 280.9K & 8.59 \\
+ &  & sdf & 0 & 1m52.17s & 22.8K & 25 & 12.15s & 210.9K & 9.23 \\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 35m2.8s & 47.9K & 9 & 5m31.52s & 303.8K & 6.34 \\
+ &  & sdf & 0 & 38m51.73s & 43.2K & 0 & 6m19.63s & 265.3K & 6.14 \\
 \hline
@@
 \hline
-\multirow{4}{*}{\shortstack[l]{convert \\ -ofmt smi}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 13.12s & 185.7K & 7 & 0.4s & 6.1M & 32.8\\
- &  & sdf & 35 & 13.7s & 187K & 4 & 0.46s & 5.6M & 29.8\\
- & \multirow{2}{*}{ChEMBL 22.1}  & smi & 14 & 5m46.72s & 290.4K & 25 & 16.48s & 6.1M & 21\\
- &  & sdf & 0 & 5m42.09s & 294.4K & 4 & 20.12s & 5M & 17\\
+\multirow{4}{*}{\shortstack[l]{convert \\ -ofmt smi}}  & \multirow{2}{*}{ChEBI 149} & smi & 0 & 13.12s & 185.7K & 7 & 0.63s & 3.9M & 20.8 \\
+ &  & sdf & 35 & 13.7s & 187K & 0 & 0.72s & 3.6M & 19 \\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 14 & 5m46.72s & 290.4K & 25 & 15.97s & 6.3M & 21.7 \\
+ &  & sdf & 0 & 5m42.09s & 294.4K & 1 & 12.63s & 8M & 27.1 \\
 \hline
@@
 \hline
-\multirow{4}{*}{\shortstack[l]{fpgen \\ -type path}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 51.99s & 46.9K & 17 & 7.25s & 336K & 7.2\\
- &  & sdf & 0 & 1m44.96s & 24.4K & 15 & 7.92s & 323.5K & 13.3\\
- & \multirow{2}{*}{ChEMBL 22.1}  & smi & 0 & 26m23.5s & 63.6K & 0 & 5m23.08s & 311.7K & 4.9\\
- &  & sdf & 0 & 35m34.44s & 47.2K & 0 & 4m59.81s & 335.9K & 7.1\\
+\multirow{4}{*}{\shortstack[l]{fpgen \\ -type path}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 51.99s & 46.9K & 0 & 7.7s & 316.3K & 6.8 \\
+ &  & sdf & 0 & 1m44.96s & 24.4K & 0 & 8.4s & 305K & 12.5 \\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 26m23.5s & 63.6K & 0 & 5m19.35s & 315.3K & 5 \\
+ &  & sdf & 0 & 35m34.44s & 47.2K & 0 & 5m12.4s & 322.4K & 6.8 \\
 \hline

### 2017-04-26 article.tex
Remove the SMARTS benchmark - we can summarise the same point with the more comprehensive benchmark.

@@
 CDK \cdkversion{} includes large improvements to algorithm
-efficiency. A simple benchmark experiment demonstrates the relative
-improvements between recent CDK versions and other open source
-toolkits (Table~5\ref{tab:smigrep}). The efficiency improvements are a
-combination of optimising data structures and key molecule processing
+efficiency. This is emphasised in the systematic benchmark
+of MACCS-like 166 key generation (Table~5). The efficiency improvements
+are a combination of optimising data structures and key molecule processing
 algorithms (e.g. kekulisation and aromaticity) needed before a SMARTS
@@
 
-The results of benchmark are summarised in Table 6 and 7. The total elapsed
-times are reported in Table 6, Table 7 subtracts the first tasks results 
+The results of benchmark are summarised in Table 5 and 6. The total elapsed
+times are reported in Table 5, Table 6 subtracts the first tasks results 
 (Count Heavy Atoms) to provide a comparable measure without the overhead of
@@
 reduced~\cite{May2014}. For ChEMBL we see the time to generate the SSSR is now
-$\sim$16 seconds when it previously took around $\sim$3.5 minutes (Table 7).
+$\sim$16 seconds when it previously took around $\sim$3.5 minutes (Table 6).
 
@@
 The speed-up in \cdkversion{} is slightly better than the SSSR task. ChEMBL 
-previously spent 4-5 minutes and now takes only $\sim$14 seconds (Table 7).
+previously spent 4-5 minutes and now takes only $\sim$14 seconds (Table 6).
 In \cdkversion{} finding all rings ($\leq$ 12 bonds) runs faster than the 
@@
 discounting improvements in read performance the SDF generation for ChEBI 
-actually runs slightly slower than v1.4.19 (Table 7). This can be partially 
+actually runs slightly slower than v1.4.19 (Table 6). This can be partially 
 explained by the more comprehensive SDF generation that now writes Sgroups 
@@
 
-\subsection*{Table 5 - SMARTS matching performance.}\label{tab:smigrep}
-  Total time taken to match the SMARTS \texttt{O=[C,N]aa[N,O;!H0]} against $\sim$1.5 million SMILES from
-  ChEMBL 21.
-The reported time is for single core execution, measure with the Unix time utility (Cent OS 7, Intel Core i7-4790 CPU @ 3.60GHz).
-The times for recent versions of Open Babel~\cite{OBoyle2011} and RDKit~\cite{rdkit} are included for reference but this is not intended to be a systematic comparison.
-The difference in number found is due to variations in valence and aromaticity models.
-  \vskip \baselineskip
-
-\begin{tabular}{lll}
-Toolkit                   & Time (s)   & Matched \\
-\hline
-CDK v1.4.9                 & 3,843  & 69,782  \\
-CDK \cdkversion{}         & 67   & 62,518  \\
-Open Babel 2.3.90         & 835 & 61,906  \\
-RDKit 2016.3.4            & 378  & 62,402  \\
-\end{tabular}
-
-\subsection*{Table 6 - Summary of systematic benchmark comparing v1.4.19 to \cdkversion{}.}\label{tab:benchmark}
+\subsection*{Table 5 - Summary of systematic benchmark comparing v1.4.19 to \cdkversion{}.}\label{tab:benchmark}
   The total elapsed \textit{real} time was measured with the unix time utility. The 
@@
 
-\subsection*{Table 7 - Summary of systematic benchmark comparing v1.4.19 to \cdkversion{} without read times.}
+\subsection*{Table 6 - Summary of systematic benchmark comparing v1.4.19 to \cdkversion{} without read times.}
 \label{tab:benchmark-delta}

### 2017-04-27 benchmark/cdk-1.4.19/src/main/java/org/openscience/cdk/FpGen.java
Generate fingerprints of length 7 which is more typical for Daylight-like path based FPs. Results is the same but the times are slightly slower due to the increased cost.

@@ -50,8 +50,8 @@ public final class FpGen extends CliExecutable {
       case "path":
       case "daylight":
       case "dy":
-        fptype = "path/length=6";
-        fpr = new Fingerprinter(1024, 6);
+        fptype = "path/length=7";
+        fpr = new Fingerprinter(1024, 7);
         break;
       default:
         System.err.println("FATAL - Unknown fp type: " + fpkey);

### 2017-04-27 article.tex
Generate fingerprints of length 7 which is more typical for Daylight-like path based FPs. Results is the same but the times are slightly slower due to the increased cost.
Timing updates in text.

@@
 
-\textbf{-type path}: Path based fingerprints encode paths of length 1...n. n was
-choose to be 6 for this benchmark. Path based fingerprints can be used for both
-substructure and similarity screening. The algorithm was
-tweaked for \cdkversion{} to hash the paths without pre-computing all paths
-upfront and without needing to generate character strings before hashing. The 
-time to encode ChEMBL previously took 26-35 minutes now only takes $\sim$5 
-minutes.
+\textbf{-type path}: Path based fingerprints encode paths of length 0 to 7. Path
+based fingerprints can be used for both substructure and similarity screening. 
+The algorithm was tweaked for \cdkversion{} to hash the paths without 
+pre-computing all paths upfront and without needing to generate character 
+strings before hashing. The time to encode ChEMBL previously took 42-47 minutes 
+now only takes 6-7 minutes.
 
@@
 \hline
-\multirow{4}{*}{\shortstack[l]{fpgen \\ -type path}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 1m14.5s & 32.7K & 9 & 8.67s & 280.9K & 8.59 \\
- &  & sdf & 0 & 1m52.17s & 22.8K & 25 & 12.15s & 210.9K & 9.23 \\
- & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 35m2.8s & 47.9K & 9 & 5m31.52s & 303.8K & 6.34 \\
- &  & sdf & 0 & 38m51.73s & 43.2K & 0 & 6m19.63s & 265.3K & 6.14 \\
+\multirow{4}{*}{\shortstack[l]{fpgen \\ -type path}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 1m38s & 24.9K & 9 & 10.4s & 234.2K & 9.42 \\
+ &  & sdf & 0 & 2m11.03s & 19.6K & 25 & 14.22s & 180.2K & 9.21 \\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 42m56.15s & 39.1K & 9 & 6m38.67s & 252.6K & 6.46 \\
+ &  & sdf & 0 & 47m5.58s & 35.6K & 0 & 7m46.85s & 215.7K & 6.05 \\
 \hline
@@
 \hline
-\multirow{4}{*}{\shortstack[l]{fpgen \\ -type path}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 51.99s & 46.9K & 0 & 7.7s & 316.3K & 6.8 \\
- &  & sdf & 0 & 1m44.96s & 24.4K & 0 & 8.4s & 305K & 12.5 \\
- & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 26m23.5s & 63.6K & 0 & 5m19.35s & 315.3K & 5 \\
- &  & sdf & 0 & 35m34.44s & 47.2K & 0 & 5m12.4s & 322.4K & 6.8 \\
+\multirow{4}{*}{\shortstack[l]{fpgen \\ -type path}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 1m15.49s & 32.3K & 0 & 9.43s & 258.3K & 8 \\
+ &  & sdf & 0 & 2m3.82s & 20.7K & 0 & 10.47s & 244.7K & 11.8 \\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 34m16.85s & 49K & 0 & 6m26.5s & 260.6K & 5.3 \\
+ &  & sdf & 0 & 43m48.29s & 38.3K & 0 & 6m39.62s & 252K & 6.6 \\
 \hline

### 2017-04-27 article.tex
More citations for performance improvements.

@@
 
-The most significant functional change here is that the SMILES parser automatically 
-locates the positions of double bonds in de-localised aromatic systems (Kekulisation).
-If this invariant cannot be met the SMILES is rejected as invalid. It is 
-possible to override this check but this is strongly discouraged as rejected
-molecules do not have a fixed formula or tautomer.
+The most significant functional change here is that the SMILES parser 
+automatically locates the positions of double bonds in de-localised aromatic 
+systems (Kekulisation). If this invariant cannot be met the SMILES is rejected 
+as invalid. It is possible to override this check but this is strongly 
+discouraged as rejected molecules do not have a fixed formula or tautomer~\cite{May2015}.
 
@@
 algorithms (e.g. kekulisation and aromaticity) needed before a SMARTS
-match can be run.
+match can be run~\cite{MayBlog2013a,MayBlog2013b,May2015}.
 

### 2017-04-27 article.tex
Make stereochemistry citation more prominent and provide the basic description of the aromaticity.

@@
   perspective projections enabled, database entries currently considered distinct can be merged
-  (Figure~\ref{fig:stereoprojections}).  The perception is based on an  algorithm briefly  
-  described by~\cite{Karapetyan2015}.
+  (Figure~\ref{fig:stereoprojections}).  The perception from projections is based on an  algorithm briefly  
+  described by~\cite{Karapetyan2015}. Other aspects of stereochemistry handling
+  are detailed thoroughly in Chapter 6 of ~\cite{May2015}.
 
@@
 
-  % More on performance?
-  % http://efficientbits.blogspot.co.uk/2013_10_01_archive.html
-  % http://efficientbits.blogspot.co.uk/2013_11_01_archive.html
-
 \subsubsection*{Ring finding}
@@
 \vspace{0.2cm}
+
+The aromaticity algorithm is straight forward, the potential electron donation
+is calculated for each atom as -1 (not aromatic), 0, 1, 2. The set of cycles
+provided in the constructor is then generated and each is checked for H{\"u}ckel's 
+rule ($4n+2$).
 

### 2017-04-27 article.tex

@@
   perspective projections enabled, database entries currently considered distinct can be merged
-  (Figure~\ref{fig:stereoprojections}).  The perception from projections is based on an  algorithm briefly  
-  described by~\cite{Karapetyan2015}. Other aspects of stereochemistry handling
-  are detailed thoroughly in Chapter 6 of ~\cite{May2015}.
+  (Figure~\ref{fig:stereoprojections}).
 
@@
 
-  Graph matching of stereochemistry with the described representation is straight forward. Given
+  Pattern matching of stereochemistry with the described representation is straight forward. Given
   the atom-atom mapping from a query structure to a target molecule, the focus and carriers of 
@@
   \cite{May2014_SMARTS}. For canonicalization, a partial canonical ordering is used to assign an 
-  absolute label which can then be integrated into the ordering. More detailed descriptions of 
-  these algorithms are described in detail in Chapter~6 of~\cite{May2015}.
+  absolute label which can then be integrated into the ordering. The algorithms used
+  for stereochemistry are thoroughly detailed in Chapter~6 of~\cite{May2015}. The perception
+  from projections is based on an algorithm briefly described by~\cite{Karapetyan2015}.
 

### 2017-04-29 article.tex
Typo fix.

@@
 became unmanageable for the core development team. This even-odd version scheme
-was adopted from old Linux kernal versioning that was subsequently
+was adopted from old Linux kernel versioning that was subsequently
 abandoned in 2004 for time-based releases~\cite{LinuxKernalVersionNumbering}.

### 2017-04-29 article.tex
Fixed a few typos

@@
 Bioclipse~\cite{spjuth2007bioclipse,spjuth2009bioclipse} and
-KNIME~\cite{Beisken2013}. However, because CDK Java packages are occassionally
+KNIME~\cite{Beisken2013}. However, because CDK Java packages are occasionally
 split between CDK modules, the CDK currently needs to be bundled as a single
@@
 uses different aspects of the library including ring perception and the new
-aromaticity perception but the speed is primarily dependant on SMARTS matching 
+aromaticity perception but the speed is primarily dependent on SMARTS matching 
 performance. Generating the fingerprint previously took almost 1 day for ChEMBL 
@@
   \caption{\csentence{Relative storage of stereochemistry, the \textbf{type} and
-    \textbf{focus} of stereochemsitry are fixed for a given
+    \textbf{focus} of stereochemistry are fixed for a given
     stereocenter description but the \textbf{carriers} and
@@
   ioformats                  & Definitions of (chemical) input/output formats. & & \\ \hline
-  io                         & Readers and writers for input/output formats.  & The molfile reader has been rewritten and supports in atom types defind in the specification. & XPP3 1.1.4c \\ \hline
+  io                         & Readers and writers for input/output formats.  & The molfile reader has been rewritten and supports atom types defined in the specification. & XPP3 1.1.4c \\ \hline
   iordf                      & Stores data models as in the Resource Description Framework serialization formats. & New. & Jena 2.7.4 \\ \hline

### 2017-04-30 article.tex
Small tweaks

@@
 in drug discovery, metabolomics, and toxicology. Over the
-last ten years, the code base has grown significantly resulting in
+last ten years, the code base has grown significantly, however, resulting in
 many complex interdependencies among components and poor
@@
 
-  Along with the new stereochemistry representation, algorithms were required in several areas. Generally
+  Along with the new stereochemistry representation, algorithms were required in several areas. Generally,
   a user does not need to invoke these procedures explicitly as they are called as needed within existing
@@
 
-The SMILES parsing has been replaced by code from the external Beam project~\cite{Beam}.
+The SMILES parsing has been replaced by code from the external Beam project~\cite{Beam,Weininger1988}.
 This BSD-licensed SMILES parser is a complete implementation of the SMILES
@@
 \begin{verbatim}
-  SmilesGenerator smigen = new SmilesGenerator(SmiFlavor.Canonical |
+  SmilesGenerator smigen
+    = new SmilesGenerator(SmiFlavor.Canonical |
                           SmiFlavor.StereoTetrahedral)
@@
 The molfile format is still very popular and despite it being a proprietary
-format it has become a de facto standard. The format forms the core of the larger
+format, it has become a de facto standard. The format forms the core of the larger
 CTfile family which was originally developed by MDL Information Systems~\cite{Dalby92}. The
@@
 used, in most cases this isn't needed and the \texttt{SilentChemObjectBuilder} 
-is preferred resulting in a typical speed up of 10\%-20\%:
+is preferred resulting in a typical speed up of 10\% to 20\%:
 
@@
 
-The benchmark was run on Java SE 8, Cent OS 7, Intel Core i7-4790 CPU @ 
+The benchmark was run on Java SE 8, CentOS 7, Intel Core i7-4790 CPU @ 
 3.60GHz with 16 GB of RAM. The code to run the benchmark is available in 

### 2017-05-01 article.tex
Moved the reference earlier.
Rerun the numbers on latest 2.0 code base.

@@
 
-The SMILES parsing has been replaced by code from the external Beam project~\cite{Beam,Weininger1988}.
+The SMILES~\cite{Weininger1988} parsing has been replaced by code from the external Beam project~\cite{Beam}.
 This BSD-licensed SMILES parser is a complete implementation of the SMILES
@@
 \hline
-\multirow{4}{*}{countheavy} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 22.51s & 108.2K & 9 & 0.97s & 2.5M & 23.21\\
- &  & sdf & 0 & 7.21s & 355.4K & 25 & 3.75s & 683.3K & 1.92\\
- & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 8m39.3s & 193.9K & 9 & 12.17s & 8.3M & 42.67\\
- &  & sdf & 0 & 3m17.29s & 510.4K & 0 & 1m7.23s & 1.5M & 2.93\\
-\hline
-\multirow{4}{*}{\shortstack[l]{rings \\ -mark}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 22.91s & 106.3K & 9 & 1.03s & 2.4M & 22.24\\
- &  & sdf & 0 & 8.71s & 294.2K & 25 & 3.93s & 652K & 2.22\\
- & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 8m45.78s & 191.5K & 9 & 17.98s & 5.6M & 29.24\\
- &  & sdf & 0 & 4m12.01s & 399.6K & 0 & 1m14.95s & 1.3M & 3.36\\
-\hline
-\multirow{4}{*}{\shortstack[l]{rings \\ -sssr}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 27.4s & 88.9K & 9 & 1.48s & 1.6M & 18.51\\
- &  & sdf & 0 & 11.84s & 216.4K & 25 & 4.08s & 628K & 2.9\\
- & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 12m4.62s & 139K & 9 & 28.07s & 3.6M & 25.81\\
- &  & sdf & 0 & 7m9.58s & 234.4K & 0 & 1m24.16s & 1.2M & 5.1\\
-\hline
-\multirow{4}{*}{\shortstack[l]{rings \\ -all}} & \multirow{2}{*}{ChEBI 149} & smi & 2126 & 45.28s & 53.8K & 26 & 1.27s & 1.9M & 35.65\\
- &  & sdf & 16 & 36.56s & 70.1K & 40 & 4.13s & 620.4K & 8.85\\
- & \multirow{2}{*}{ChEMBL 22.1} & smi & 88 & 12m40.2s & 132.5K & 9 & 26.48s & 3.8M & 28.71\\
- &  & sdf & 90 & 8m5.64s & 207.4K & 0 & 1m18.99s & 1.3M & 6.15\\
-\hline
-\multirow{4}{*}{cansmi} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 36.58s & 66.6K & 9 & 1.83s & 1.3M & 19.99\\
- &  & sdf & 35 & 21.15s & 121.1K & 26 & 4.77s & 537.2K & 4.43\\
- & \multirow{2}{*}{ChEMBL 22.1} & smi & 14 & 14m33.86s & 115.2K & 9 & 42.41s & 2.4M & 20.61\\
- &  & sdf & 0 & 8m59.82s & 186.6K & 0 & 1m44.86s & 960.4K & 5.15\\
-\hline
-\multirow{4}{*}{\shortstack[l]{convert \\ -ofmt smi}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 35.63s & 68.4K & 16 & 1.6s & 1.5M & 22.27 \\
- &  & sdf & 35 & 20.91s & 122.5K & 25 & 4.47s & 573.2K & 4.68 \\
- & \multirow{2}{*}{ChEMBL 22.1} & smi & 14 & 14m26.02s & 116.3K & 34 & 28.14s & 3.6M & 30.78 \\
- &  & sdf & 0 & 8m59.38s & 186.7K & 1 & 1m19.86s & 1.3M & 6.75 \\
-\hline
-\multirow{4}{*}{\shortstack[l]{convert \\ -ofmt sdf}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 32.42s & 75.1K & 9 & 10.33s & 235.8K & 3.14\\
- &  & sdf & 13 & 17s & 150.7K & 29 & 14.49s & 176.8K & 1.17\\
- & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 14m25.82s & 116.3K & 9 & 5m30.66s & 304.6K & 2.62\\
- &  & sdf & 1 & 8m51.33s & 189.5K & 0 & 6m50.92s & 245.1K & 1.29\\
-\hline
-\multirow{4}{*}{\shortstack[l]{convert \\ -gen2d \\ -ofmt smi}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 24m28.02s & 1.7K & 9 & 35.7s & 68.2K & 41.12\\
- &  & sdf & 13 & 35m12.03s & 1.2K & 29 & 42.08s & 60.9K & 50.19\\
- & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 3h27m7s & 8.1K & 9 & 18m2.59s & 93K & 11.48\\
- &  & sdf & 1 & 5h58m30s & 4.7K & 0 & 19m54.12s & 84.3K & 18.01\\
-\hline
-\multirow{4}{*}{\shortstack[l]{fpgen \\ -type path}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 1m38s & 24.9K & 9 & 10.4s & 234.2K & 9.42 \\
- &  & sdf & 0 & 2m11.03s & 19.6K & 25 & 14.22s & 180.2K & 9.21 \\
- & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 42m56.15s & 39.1K & 9 & 6m38.67s & 252.6K & 6.46 \\
- &  & sdf & 0 & 47m5.58s & 35.6K & 0 & 7m46.85s & 215.7K & 6.05 \\
-\hline
-\multirow{4}{*}{\shortstack[l]{fpgen \\ -type maccs}} & \multirow{2}{*}{ChEBI 149} & smi & 2150 & 1h37m35s & 416 & 9 & 19.8s & 123K & 295.71\\
- &  & sdf & 48 & 1h44m17s & 409 & 25 & 22.33s & 114.7K & 280.21\\
- & \multirow{2}{*}{ChEMBL 22.1} & smi & 214 & 20h24m57s & 1.4K & 9 & 13m27.11s & 124.8K & 91.06\\
- &  & sdf & 225 & 24h41m46s & 1.1K & 0 & 13m41.83s & 122.5K & 108.18\\
-\hline
-\multirow{4}{*}{\shortstack[l]{fpgen \\ -type circ}} & \multirow{2}{*}{ChEBI 149} & smi & 0 & NA & - & 9 & 4.47s & 544.9K & 0\\
- &  & sdf & 0 & NA & - & 25 & 7.04s & 364K & 0\\
- & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & NA & - & 9 & 2m45.68s & 607.8K & 0\\
- &  & sdf & 0 & NA & - & 0 & 3m59.23s & 420.9K & 0\\
-
+\multirow{4}{*}{countheavy} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 22.51s & 108.2K & 9 & 0.85s & 2.9M & 26.48\\
+ &  & sdf & 0 & 7.21s & 355.4K & 25 & 3s & 854.1K & 2.4\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 8m39.3s & 193.9K & 9 & 10.74s & 9.4M & 48.35\\
+ &  & sdf & 0 & 3m17.29s & 510.4K & 0 & 53.27s & 1.9M & 3.7\\
+\hline
+\multirow{4}{*}{\shortstack[l]{rings \\-mark}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 22.91s & 106.3K & 9 & 1.06s & 2.3M & 21.61\\
+ &  & sdf & 0 & 8.71s & 294.2K & 25 & 3.11s & 823.9K & 2.8\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 8m45.78s & 191.5K & 9 & 17.09s & 5.9M & 30.77\\
+ &  & sdf & 0 & 4m12.01s & 399.6K & 0 & 1m6.54s & 1.5M & 3.79\\
+\hline
+\multirow{4}{*}{\shortstack[l]{rings \\-sssr}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 27.4s & 88.9K & 9 & 1.43s & 1.7M & 19.16\\
+ &  & sdf & 0 & 11.84s & 216.4K & 25 & 3.78s & 677.8K & 3.13\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 12m4.62s & 139K & 9 & 27.16s & 3.7M & 26.68\\
+ &  & sdf & 0 & 7m9.58s & 234.4K & 0 & 1m8.17s & 1.5M & 6.3\\
+\hline
+\multirow{4}{*}{\shortstack[l]{rings \\-all}} & \multirow{2}{*}{ChEBI 149} & smi & 2126 & 45.28s & 53.8K & 26 & 1.26s & 1.9M & 35.94\\
+ &  & sdf & 16 & 36.56s & 70.1K & 40 & 3.51s & 730K & 10.42\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 88 & 12m40.2s & 132.5K & 9 & 24.97s & 4M & 30.44\\
+ &  & sdf & 90 & 8m5.64s & 207.4K & 0 & 1m5.68s & 1.5M & 7.39\\
+\hline
+\multirow{4}{*}{cansmi} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 36.58s & 66.6K & 9 & 1.91s & 1.3M & 19.15\\
+ &  & sdf & 35 & 21.15s & 121.1K & 26 & 4.37s & 586.3K & 4.84\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 14 & 14m33.86s & 115.2K & 9 & 40.84s & 2.5M & 21.4\\
+ &  & sdf & 0 & 8m59.82s & 186.6K & 0 & 1m29.33s & 1.1M & 6.04\\
+\hline
+\multirow{4}{*}{\shortstack[l]{convert \\-ofmt smi}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 35.63s & 68.4K & 16 & 1.47s & 1.7M & 24.24\\
+ &  & sdf & 35 & 20.91s & 122.5K & 25 & 4.55s & 563.1K & 4.6\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 14 & 14m26.02s & 116.3K & 37 & 26.2s & 3.8M & 33.05\\
+ &  & sdf & 0 & 8m59.38s & 186.7K & 1 & 1m12.49s & 1.4M & 7.44\\
+\hline
+\multirow{4}{*}{\shortstack[l]{convert \\-ofmt sdf}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 32.42s & 75.1K & 9 & 10.39s & 234.4K & 3.12\\
+ &  & sdf & 13 & 17s & 150.7K & 25 & 13.96s & 183.5K & 1.22\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 14m25.82s & 116.3K & 9 & 5m26.29s & 308.6K & 2.65\\
+ &  & sdf & 1 & 8m51.33s & 189.5K & 0 & 6m34.5s & 255.3K & 1.35\\
+\hline
+\multirow{4}{*}{\shortstack[l]{convert \\-gen2d \\-ofmt sdf}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 24m28.02s & 1.7K & 9 & 35.86s & 67.9K & 40.94\\
+ &  & sdf & 13 & 35m12.03s & 1.2K & 25 & 42.43s & 60.4K & 49.78\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 3h27m7s & 8.1K & 9 & 17m44.64s & 94.6K & 11.67\\
+ &  & sdf & 1 & 5h58m30s & 4.7K & 0 & 19m42.77s & 85.1K & 18.19\\
+\hline
+\multirow{4}{*}{\shortstack[l]{fpgen \\-type path}} & \multirow{2}{*}{ChEBI 149} & smi & 2112 & 1m38s & 24.9K & 9 & 10.28s & 236.9K & 9.53\\
+ &  & sdf & 0 & 2m11.03s & 19.6K & 25 & 13.03s & 196.6K & 10.06\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 42m56.15s & 39.1K & 9 & 6m34.67s & 255.2K & 6.53\\
+ &  & sdf & 0 & 47m5.58s & 35.6K & 0 & 7m52.32s & 213.2K & 5.98\\
+\hline
+\multirow{4}{*}{\shortstack[l]{fpgen \\-type maccs}} & \multirow{2}{*}{ChEBI 149} & smi & 2150 & 1h37m35s & 416 & 9 & 19.51s & 124.8K & 300.1\\
+ &  & sdf & 48 & 1h44m17s & 409 & 25 & 21.25s & 120.6K & 294.45\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 214 & 20h24m57s & 1.4K & 9 & 13m31.21s & 124.1K & 90.6\\
+ &  & sdf & 225 & 24h41m46s & 1.1K & 0 & 13m26.41s & 124.9K & 110.25\\
+\hline
+\multirow{4}{*}{\shortstack[l]{fpgen \\-type circ}} & \multirow{2}{*}{ChEBI 149} & smi & 0 &  & - & 9 & 4.37s & 557.4K & 0\\
+ &  & sdf & 0 &  & - & 25 & 6.81s & 376.2K & 0\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 &  & - & 9 & 2m43.45s & 616.1K & 0\\
+ &  & sdf & 0 &  & - & 0 & 3m42.01s & 453.6K & 0\\
 \end{tabular}
@@
 \hline
-\multirow{4}{*}{\shortstack[l]{rings \\ -mark}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 0.4s & 6.1M & 0 & 0.06s & 40.6M & 6.7\\
- &  & sdf & 0 & 1.5s & 1.7M & 0 & 0.18s & 14.2M & 8.3\\
- & \multirow{2}{*}{ChEMBL 22.1}  & smi & 0 & 6.48s & 15.5M & 0 & 5.81s & 17.3M & 1.1\\
- &  & sdf & 0 & 54.72s & 1.8M & 0 & 7.72s & 13M & 7.1\\
-\hline
-\multirow{4}{*}{\shortstack[l]{rings \\ -sssr}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 4.89s & 498.1K & 0 & 0.51s & 4.8M & 9.6\\
- &  & sdf & 0 & 4.63s & 553.4K & 0 & 0.33s & 7.8M & 14\\
- & \multirow{2}{*}{ChEMBL 22.1}  & smi & 0 & 3m25.32s & 490.5K & 0 & 15.9s & 6.3M & 12.9\\
- &  & sdf & 0 & 3m52.29s & 433.5K & 0 & 16.93s & 5.9M & 13.7\\
-\hline
-\multirow{4}{*}{\shortstack[l]{rings \\ -all}}  & \multirow{2}{*}{ChEBI 149}  & smi & 14 & 22.77s & 107K & 17 & 0.3s & 8.1M & 75.9\\
- &  & sdf & 16 & 29.35s & 87.3K & 15 & 0.38s & 6.7M & 77.2\\
- & \multirow{2}{*}{ChEMBL 22.1}  & smi & 88 & 4m0.9s & 418K & 0 & 14.31s & 7M & 16.8\\
- &  & sdf & 90 & 4m48.35s & 349.2K & 0 & 11.76s & 8.6M & 24.5\\
-\hline
-\multirow{4}{*}{cansmi}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 14.07s & 173.1K & 0 & 0.86s & 2.8M & 16.4\\
- &  & sdf & 35 & 13.94s & 183.8K & 1 & 1.02s & 2.5M & 13.7\\
- & \multirow{2}{*}{ChEMBL 22.1}  & smi & 14 & 5m54.56s & 284K & 0 & 30.24s & 3.3M & 11.7\\
- &  & sdf & 0 & 5m42.53s & 294K & 0 & 37.63s & 2.7M & 9.1\\
-\hline
-\multirow{4}{*}{\shortstack[l]{convert \\ -ofmt smi}}  & \multirow{2}{*}{ChEBI 149} & smi & 0 & 13.12s & 185.7K & 7 & 0.63s & 3.9M & 20.8 \\
- &  & sdf & 35 & 13.7s & 187K & 0 & 0.72s & 3.6M & 19 \\
- & \multirow{2}{*}{ChEMBL 22.1} & smi & 14 & 5m46.72s & 290.4K & 25 & 15.97s & 6.3M & 21.7 \\
- &  & sdf & 0 & 5m42.09s & 294.4K & 1 & 12.63s & 8M & 27.1 \\
-\hline
-\multirow{4}{*}{\shortstack[l]{convert \\ -ofmt sdf}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 9.91s & 245.8K & 0 & 9.36s & 260.2K & 1.1\\
- &  & sdf & 13 & 9.79s & 261.7K & 4 & 10.74s & 238.6K & 0.9\\
- & \multirow{2}{*}{ChEMBL 22.1}  & smi & 0 & 5m46.52s & 290.6K & 0 & 5m18.49s & 316.2K & 1.1\\
- &  & sdf & 1 & 5m34.04s & 301.5K & 0 & 5m43.69s & 293K & 1\\
-\hline
-\multirow{4}{*}{\shortstack[l]{convert \\ -gen2d \\ -ofmt smi}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 24m5.51s & 1.7K & 0 & 34.73s & 70.1K & 41.6\\
- &  & sdf & 13 & 35m4.82s & 1.2K & 4 & 38.33s & 66.8K & 54.9\\
- & \multirow{2}{*}{ChEMBL 22.1}  & smi & 0 & 3h18m28s & 8.5K & 0 & 17m50.42s & 94.1K & 11.1\\
- &  & sdf & 1 & 5h55m13s & 4.7K & 0 & 18m46.89s & 89.4K & 18.9\\
-\hline
-\multirow{4}{*}{\shortstack[l]{fpgen \\ -type path}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 & 1m15.49s & 32.3K & 0 & 9.43s & 258.3K & 8 \\
- &  & sdf & 0 & 2m3.82s & 20.7K & 0 & 10.47s & 244.7K & 11.8 \\
- & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 34m16.85s & 49K & 0 & 6m26.5s & 260.6K & 5.3 \\
- &  & sdf & 0 & 43m48.29s & 38.3K & 0 & 6m39.62s & 252K & 6.6 \\
-\hline
-\multirow{4}{*}{\shortstack[l]{fpgen \\ -type maccs}}  & \multirow{2}{*}{ChEBI 149}  & smi & 38 & 1h37m12s & 418 & 0 & 18.83s & 129.4K & 309.7\\
- &  & sdf & 48 & 1h44m10s & 410 & 0 & 18.58s & 137.9K & 336.4\\
- & \multirow{2}{*}{ChEMBL 22.1}  & smi & 214 & 20h16m18s & 1.4K & 0 & 13m14.94s & 126.7K & 91.8\\
- &  & sdf & 225 & 24h38m29s & 1.1K & 0 & 12m34.6s & 133.5K & 117.6\\
-\hline
-\multirow{4}{*}{\shortstack[l]{fpgen \\ -type circ}}  & \multirow{2}{*}{ChEBI 149}  & smi & 0 &  & - & 0 & 3.5s & 695.9K & \\
- &  & sdf & 0 &  & - & 0 & 3.29s & 778.8K & \\
- & \multirow{2}{*}{ChEMBL 22.1}  & smi & 0 &  & - & 0 & 2m33.51s & 656K & \\
- &  & sdf & 0 &  & - & 0 & 2m52s & 585.5K & \\
-\hline
+\multirow{4}{*}{\shortstack[l]{rings \\-mark}} & \multirow{2}{*}{ChEBI 149} & smi & 0 & 0.4s & 6.1M & 0 & 0.21s & 11.6M & 1.9\\
+ &  & sdf & 0 & 1.5s & 1.7M & 0 & 0.11s & 23.3M & 13.6\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 6.48s & 15.5M & 0 & 6.35s & 15.9M & 1\\
+ &  & sdf & 0 & 54.72s & 1.8M & 0 & 13.27s & 7.6M & 4.1\\
+\hline
+\multirow{4}{*}{\shortstack[l]{rings \\-sssr}} & \multirow{2}{*}{ChEBI 149} & smi & 0 & 4.89s & 498.1K & 0 & 0.58s & 4.2M & 8.4\\
+ &  & sdf & 0 & 4.63s & 553.4K & 0 & 0.78s & 3.3M & 5.9\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 3m25.32s & 490.5K & 0 & 16.42s & 6.1M & 12.5\\
+ &  & sdf & 0 & 3m52.29s & 433.5K & 0 & 14.9s & 6.8M & 15.6\\
+\hline
+\multirow{4}{*}{\shortstack[l]{rings \\-all}} & \multirow{2}{*}{ChEBI 149} & smi & 14 & 22.77s & 107K & 17 & 0.41s & 5.9M & 55.5\\
+ &  & sdf & 16 & 29.35s & 87.3K & 15 & 0.51s & 5M & 57.5\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 88 & 4m0.9s & 418K & 0 & 14.23s & 7.1M & 16.9\\
+ &  & sdf & 90 & 4m48.35s & 349.2K & 0 & 12.41s & 8.1M & 23.2\\
+\hline
+\multirow{4}{*}{cansmi} & \multirow{2}{*}{ChEBI 149} & smi & 0 & 14.07s & 173.1K & 0 & 1.06s & 2.3M & 13.3\\
+ &  & sdf & 35 & 13.94s & 183.8K & 1 & 1.37s & 1.9M & 10.2\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 14 & 5m54.56s & 284K & 0 & 30.1s & 3.3M & 11.8\\
+ &  & sdf & 0 & 5m42.53s & 294K & 0 & 36.06s & 2.8M & 9.5\\
+\hline
+\multirow{4}{*}{\shortstack[l]{convert \\-ofmt smi}} & \multirow{2}{*}{ChEBI 149} & smi & 0 & 13.12s & 185.7K & 7 & 0.62s & 3.9M & 21.2\\
+ &  & sdf & 35 & 13.7s & 187K & 0 & 1.55s & 1.7M & 8.8\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 14 & 5m46.72s & 290.4K & 28 & 15.46s & 6.5M & 22.4\\
+ &  & sdf & 0 & 5m42.09s & 294.4K & 1 & 19.22s & 5.2M & 17.8\\
+\hline
+\multirow{4}{*}{\shortstack[l]{convert \\-ofmt sdf}} & \multirow{2}{*}{ChEBI 149} & smi & 0 & 9.91s & 245.8K & 0 & 9.54s & 255.3K & 1\\
+ &  & sdf & 13 & 9.79s & 261.7K & 0 & 10.96s & 233.8K & 0.9\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 5m46.52s & 290.6K & 0 & 5m15.55s & 319.1K & 1.1\\
+ &  & sdf & 1 & 5m34.04s & 301.5K & 0 & 5m41.23s & 295.1K & 1\\
+\hline
+\multirow{4}{*}{\shortstack[l]{convert \\-gen2d \\-ofmt sdf}} & \multirow{2}{*}{ChEBI 149} & smi & 0 & 24m5.51s & 1.7K & 0 & 35.01s & 69.6K & 41.3\\
+ &  & sdf & 13 & 35m4.82s & 1.2K & 0 & 39.43s & 65K & 53.4\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 3h18m28s & 8.5K & 0 & 17m33.9s & 95.6K & 11.3\\
+ &  & sdf & 1 & 5h55m13s & 4.7K & 0 & 18m49.5s & 89.2K & 18.9\\
+\hline
+\multirow{4}{*}{\shortstack[l]{fpgen \\-type path}} & \multirow{2}{*}{ChEBI 149} & smi & 0 & 1m15.49s & 32.3K & 0 & 9.43s & 258.3K & 8\\
+ &  & sdf & 0 & 2m3.82s & 20.7K & 0 & 10.03s & 255.5K & 12.3\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 & 34m16.85s & 49K & 0 & 6m23.93s & 262.3K & 5.4\\
+ &  & sdf & 0 & 43m48.29s & 38.3K & 0 & 6m59.05s & 240.3K & 6.3\\
+\hline
+\multirow{4}{*}{\shortstack[l]{fpgen \\-type maccs}} & \multirow{2}{*}{ChEBI 149} & smi & 38 & 1h37m12s & 418 & 0 & 18.66s & 130.5K & 312.6\\
+ &  & sdf & 48 & 1h44m10s & 410 & 0 & 18.25s & 140.4K & 342.5\\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 214 & 20h16m18s & 1.4K & 0 & 13m20.47s & 125.8K & 91.2\\
+ &  & sdf & 225 & 24h38m29s & 1.1K & 0 & 12m33.14s & 133.7K & 117.8\\
+\hline
+\multirow{4}{*}{\shortstack[l]{fpgen \\-type circ}} & \multirow{2}{*}{ChEBI 149} & smi & 0 &  & - & 0 & 3.52s & 692K & \\
+ &  & sdf & 0 &  & - & 0 & 3.81s & 672.5K & \\
+ & \multirow{2}{*}{ChEMBL 22.1} & smi & 0 &  & - & 0 & 2m32.71s & 659.4K & \\
+ &  & sdf & 0 &  & - & 0 & 2m48.74s & 596.8K & \\
 \end{tabular}

### 2017-05-01 article.tex
Update numbers and wording in benchmark section.

@@
 it would take more than 8 minutes to read ChEMBL 22.1 but this is reduced
-to only 12 seconds. On top of this improvement SMILES input is now validated 
+to less than 11 seconds. On top of this improvement SMILES input is now validated 
 and assigned a Kekul$\mathrm{\acute{e}}$ structure. This identifies 9 invalid
@@
 linearithmic time. The run times are similar for these datasets (Table
-7), larger differences are seen for more complex cage molecules such fullerenes
-\cite{May2014}.
+7), larger differences are only seen for more complex cage molecules such 
+fullerenes~\cite{May2014}.
 
@@
 v1.4.19 there were 14/16 records skipped from ChEBI and 88/90 in ChEMBL due to
-reaching the time out. The number fluctuates here as a timeout based threshold 
-depends on external factors such as CPU load.
+reaching the time out.
 
 The speed-up in \cdkversion{} is slightly better than the SSSR task. ChEMBL 
-previously spent 4-5 minutes and now takes only $\sim$14 seconds (Table 6).
+previously spent 4-5 minutes and now takes only $\sim$12-14 seconds (Table 6).
 In \cdkversion{} finding all rings ($\leq$ 12 bonds) runs faster than the 
@@
 \cdkversion{} the total elapsed time is $\sim$20 times faster for both ChEBI
-and ChEMBL. For ChEMBL it now takes just under 43 seconds to read, reorder, 
+and ChEMBL. For ChEMBL it now takes just under 41 seconds to read, reorder, 
 and write the SMILES compared to more than 14 minutes previously.
@@
 task but \cdkversion{} can generate the non-canonical SMILES faster taking less
-than 30 seconds for ChEMBL.
+than 30 seconds for SMILES from ChEMBL.
 
@@
 
-\textbf{-gen2d -ofmt sdf}: When writing SDF the only portable way to store stereochemistry is with the 
-inclusion of coordinates, this is specified with the \texttt{-gen2d} option. 
-The overhaul in layout generation discussed early provides better layouts but 
-also included performance tweaks, in CDK v1.4.19 generating coordinates and 
-writing an SDF for ChEMBL took almost 3.5 hours but now only takes just over
-18 minutes.
+\textbf{-gen2d -ofmt sdf}: When writing SDF the only portable way to store 
+stereochemistry is with the inclusion of coordinates, this is specified with 
+the \texttt{-gen2d} option. The overhaul in layout generation discussed early 
+provides better layouts but also included performance tweaks, in CDK v1.4.19 
+generating coordinates and writing an SDF for ChEMBL took almost 3.5 hours but
+now only takes $\sim$18 minutes.
 
@@
 strings before hashing. The time to encode ChEMBL previously took 42-47 minutes 
-now only takes 6-7 minutes.
+now only takes 6-8 minutes.
 
@@
 aromaticity perception but the speed is primarily dependent on SMARTS matching 
-performance. Generating the fingerprint previously took almost 1 day for ChEMBL 
-and 2 hours for ChEBI. This has been reduced to less than 
-15 minutes for ChEMBL and less than 30 seconds for ChEBI.
+performance. Generating the fingerprint previously took $\sim$1 day for ChEMBL 
+and $\sim$1.75 hours for ChEBI. This has been reduced to less than 
+13.5 minutes for ChEMBL and $\sim$20 seconds for ChEBI.
 

### 2017-05-01 article.tex
Use v1.4.19 everywhere
A bit more textual variance

@@
   A reference implementation, \texttt{CDKAtomTypeMatcher}, has been
-  written that perceives these atom types, and validates the
+  written in such a way that perceives these atom types, and validates the
   perception automatically against the properties defined by the
@@
 
-The CDK library is now mostly written that it no longer depends on a specific
+The CDK library is now mostly refactored and no longer depends on a specific
 implementation of the \texttt{IChemObjectBuilder}, allowing the user of the
@@
 \begin{tabular}{lllrclrclc}
-Benchmark  & \multicolumn{2}{c}{Data Set} & \multicolumn{3}{c}{CDK 1.4.19} & \multicolumn{3}{c}{CDK \cdkversion{}} & Improvement \\
+Benchmark  & \multicolumn{2}{c}{Data Set} & \multicolumn{3}{c}{CDK v1.4.19} & \multicolumn{3}{c}{CDK \cdkversion{}} & Improvement \\
            &          &              & Skip & Time & Per Min & Skip  & Time  & Per Min \\
@@
 \begin{tabular}{lllrclrcll}
-Benchmark  & \multicolumn{2}{c}{Data Set} & \multicolumn{3}{c}{CDK 1.4.19} & \multicolumn{3}{c}{CDK \cdkversion{}} & Improvement \\
+Benchmark  & \multicolumn{2}{c}{Data Set} & \multicolumn{3}{c}{CDK v1.4.19} & \multicolumn{3}{c}{CDK \cdkversion{}} & Improvement \\
            &          &              & Skip & Time & Per Min & Skip  & Time  & Per Min \\

### 2017-05-01 article.tex
Consistency

@@
 to understand the underlying algorithms. This resulted in certain
-functionalities, such as fingerprinting~\cite{Clark2014,Cannon2006},
+functionalities, such as molecular fingerprinting~\cite{Clark2014,Cannon2006},
 receiving more attention than others, such as stereochemistry. The

### 2017-05-02 article.tex
proofing

@@
   systems, maintains input stereochemistry, and makes use of a large
-  template library. Templates are useful for laying out substructure
+  template library. Templates are useful for laying out substructure.
   While previous CDK versions partially supported 
   double bond stereochemistry the new engine is more efficient in
-  using this information when generate 2D layouts. Furthermore, the
+  using this information when generating 2D layouts. Furthermore, the
   engine assigns wedge bond information based on tetrahedral
-  stereochemistry. These features are exemplified by the following
+  ste\-reo\-chem\-is\-try. These features are exemplified by the following
   code and the resulting layout depiction in
@@
 approach with an event-passing mechanism to update the view when the
-model was changed. This can cause an cascade of change events being
+model was changed. This can cause a cascade of change events being
 passed around. This was not always a desirable feature, especially for

### 2017-05-07 article.tex
Fix wording.
Added the BRENDA database (still using CDK fingerprints, but they're looking for a PHP alternative; closes #83)

--- 2017-05-01 article.tex
@@
 WhichCyp~\cite{Rostkowski2013}, MetaPrint2D~\cite{Carlsson2010}, MetFrag~\cite{Wolf2010},
-and the IUPHAR/BPS Guide to Pharmacology~\cite{Southan2016} and
+and the IUPHAR/BPS Guide to Pharmacology~\cite{Southan2016}, BRENDA~\cite{Placzek2017} and
 QSAR DataBank~\cite{Ruusmann2015} databases.
@@
 molecules or reactions across multiple threads. During initialization
-the pattern is inspected as to determine what invariants will be
+the pattern is inspected so as to determine what invariants will be
 needed (e.g. ring size) and only required invariants are calculated. The

### 2017-05-07 article.tex
Clarified what the Blue Obelisk is (to improve the impact of the paper outside the journal; closes #79)

@@
 The Chemistry Development Kit (CDK) is one of the tools developed
-under the aegis of the Blue Obelisk 
-and previously documented versions have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
+under the aegis of the Blue Obelisk, a movement promoting Open Data,
+Open Source, and Open Standards in chemistry~\cite{OBoyle2011b,Guha2006}.
+Previously documented CDK versions have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
 Use of the CDK ranges from inclusion of CDK functionality in

### 2017-05-07 article.tex
Restructured to distinguish new from improved functionality (closes #80)

@@
 specifically addressing the increased functional complexity and poor
-performance. We first summarize the addition of new, and
-improvements to existing functionality that has led to significant
-improvement in performance and support for important chemistry
-features including atom type perception, substructure searching, molecular
-fingerprints, rendering of molecules, and handling of molecular formulas.
+performance. We first summarize the addition of new functionality, such
+atom typing and molecular formula handling, and improvement to existing
+functionality that has led to significantly better performance for
+substructure searching, molecular fingerprints, and rendering of molecules.
 Second, we outline how the CDK has evolved with respect to

### 2017-05-07 article.tex
Repeat what the CDK is, which was written in the abstract, but not in the main text (closes #81)

@@
 Open Source, and Open Standards in chemistry~\cite{OBoyle2011b,Guha2006}.
+The CDK providing data structures to represent chemical concepts along with
+methods to manipulate such structures and perform computations on them.
 Previously documented CDK versions have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
